### 2016-07-25 drafts/writeup_07_25_16.tex
starting writeup

\documentclass{article}

\usepackage[numbers, compress]{natbib}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{amsthm,amsmath,amssymb}
\usepackage{macros}
\usepackage{subcaption}
\usepackage[textfont=small, labelfont=small]{caption}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}

\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage[color=yellow]{todonotes}
\usepackage{booktabs}
\usepackage[inline]{enumitem}
\usepackage{verbatim}

\usepackage[margin=1in]{geometry}

\usepackage{setspace}

\title{Discrete and continuous latent states of neural activity in \textit{Caenorhabditis Elegans}}

\author{Scott W. Linderman 
  and
  Liam Paninski \\
  Columbia University
}

\begin{document}

\singlespacing
\maketitle

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  immobilized C. elegans \cite{kato2015global}. Moreover, since some
  neurons are known to reliably indicate the onset or offset of
  particular behaviors, like ventral and dorsal turns, behavioral
  state can be decoded from the simultaneous population recordings.
  These datasets provide unique visibility into the relationship
  between neural activity and behavior.  While it seems clear that
  activity is inherently lower dimensional than the number of neurons
  due to strong correlations between cells, the nature of the latent brain state
  remains unclear. For example, is brain state better thought of as
  discrete or continuous, or perhaps a combination of the two? Does it
  obey linear or nonlinear dynamics?  We propose a generative approach
  to probing these questions. We model the neural activity as a
  \emph{switching linear dynamical system} (SLDS), with both discrete
  and continuous latent states, and conditionally linear dynamics.  We
  then analyze the posterior distribution over states implied by the
  neural recordings and find that the discrete states correspond to
  stereotypical motor sequences. In contrast to previous work, these
  states are exposed in an entirely unsupervised manner.
\end{abstract}

% Body
\onehalfspacing

\section{Model}

Assume the instantaneous neural activity at time~$t$ for a population
of~$N$ neurons is represented as a vector,~$\by_t \in \reals^N$. In
calcium imaging settings, the entries in this vector may be
instaneous~$\Delta F/F$ measurements, or another signal that captures
neural activity. In this experiment, we use the smoothed time
derivative of~$\Delta F/F$. Over the course of an experiment, we
measure a sequence of vectors, which we combine into a matrix denoted
by~$\by_{1:T}$.

Our model is based on the following assumptions: (i) the instantaneous
neural activity,~$\by_t$, reflects an underlying, low-dimensional
latent state; (ii) this state has a discrete component,~$z_t \in \{1, \ldots, K\}$,
and a continuous component,~$\bx_t \in \reals^D$; (iii) the continuous
latent state has linear dynamics governed by the corresponding discrete
latent state; and (iv) the observed neural activity is a linear function
of the underlying states with additive Gaussian noise.

\TODO{Discuss the motivation for these modeling assumptions.}

These assumptions
are combined in a switching linear dynamical system, which we formalize 
with the following generative model:
\begin{align}
  p(\by_{1:T}, \bx_{1:T}, \bz_{1:T} \given \bTheta) &= 
  \prod_{t=1}^T
  p(z_t \given z_{t-1}, \bTheta) \, 
  p(\bx_t \given z_{t-1}, \bx_{t-1}, \bTheta) \, 
  p(\by_t \given z_t, \bx_t, \bTheta).
\end{align}
Our beliefs about the dynamics of these latent states are encoded in the 
form of the conditional distributions for~$z_t$ and~$\bx_t$. First,
we assume the discrete states follow a Markov process,
\begin{align}
  p(z_t \given z_{t-1}, \bTheta) &\sim \distDiscrete(\bpi^{(z_{t-1})}).
\end{align}
Next, we imbue the continuous latent states with linear Gaussian dynamics,
\begin{align}
  p(\bx_t \given \bx_{t-1}, z_{t-1}, \bTheta) 
  &\sim \distNormal(\bA^{(z_{t-1})} \bx_{t-1} + \bb^{(z_{t-1})}, \bQ^{(z_{t-1})}).
\end{align}
Finally, we impose the assumption of linear observations via the conditional
distribution,
\begin{align}
  p(\by_t \given \bx_{t}, z_{t}, \bTheta) 
  &\sim \distNormal(\bC^{(z_{t})} \bx_{t} + \bd^{(z_{t})}, \bR^{(z_t)}).
\end{align}
Thus, the parameters of the model are,
\begin{align}
  \bTheta &= \left\{ \bA^{(k)}, \bb^{(k)}, \bQ^{(k)}, \bC^{(k)}, \bd^{(k)}, \bR^{(k)}, \bpi^{(k)} \right\}_{k=1}^K .
\end{align}

\TODO{Discuss interpretation of the parameters. Set the stage for visualizing~$\bA^{(k)}$ 
in later sections.}

\section{Inference}
\TODO{Describe block Gibbs updates for~$\bz_{1:T} \given \bx_{1:T},
  \by_{1:T}$ and~$\bx_{1:T} \given \bz_{1:T}, \by_{1:T}$.  Given the
  latent states, the remaining parameters are sampled as in Bayesian
  linear regression. }

\section{Results}

\bibliographystyle{unsrt}
\bibliography{writeup}

\end{document}

### 2016-07-26 drafts/writeup_07_25_16.tex
working on short writeup

@@
 
-\usepackage[margin=1in]{geometry}
+\usepackage[left=1.5in, right=1.5in, top=1.25in, bottom=1.25in]{geometry}
 
@@
 
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{slds} 
+\caption{Simulated data from a switching linear dynamical system.
+The population stochastically jumps between discrete states,~$z_t \in \{a,b,c,d\}$,
+each of which is color coded for visualization.
+These discrete states govern the linear dynamics of the continuous latent 
+state,~$\bx_t \in \reals^D$ (here,~$D=2$). For example, state~$a$ corresponds to a simple
+random walk, whereas state~$d$ has fast, oscillatory dynamics. Finally,
+the observed signals,~$\by_t \in \reals^N$ (here,~$N=8$), are obtained via a linear 
+transformation of the underlying, continuous state,~$\bx_t$. The correlations
+and dynamics in the observations are inherited from the dynamics of the
+latent states.}
+\vspace{-0.5cm}
+\label{fig:slds_ex}
+\end{figure}
+
 Assume the instantaneous neural activity at time~$t$ for a population
@@
 \begin{align}
+  \label{eq:joint}
   p(\by_{1:T}, \bx_{1:T}, \bz_{1:T} \given \bTheta) &= 
+  p(\bTheta)
   \prod_{t=1}^T
@@
 \end{align}
+
+% Figure illustrating latent dynamics
+\begin{figure}[t]
+\centering%
+\begin{subfigure}{.49\textwidth}
+\includegraphics[width=\textwidth]{dynamics_1}
+\end{subfigure} 
+\begin{subfigure}{.49\textwidth}
+\includegraphics[width=\textwidth]{dynamics_4}
+\end{subfigure} 
+\caption{The dynamics corresponding to discrete latent state~$k$ may
+  be visualized as a vector field where the arrows point to the
+  expected next state,~$\bbE[\bx_{t+1}]=\bA^{(k)}\bx_t + \bb^{(k)}$.
+Here, we show the dynamics for the first and fourth discrete states 
+in the synthetic example from Figure~\ref{fig:slds_ex}. The first 
+discrete state is a random walk with a slight decay toward the origin;
+the fourth discrete state corresponds to fast, oscillatory dynamics.}
+\label{fig:slds_dynamics_ex}
+\vspace{-0.5cm}
+\end{figure}
+
 Finally, we impose the assumption of linear observations via the conditional
@@
 
-\section{Inference}
-\TODO{Describe block Gibbs updates for~$\bz_{1:T} \given \bx_{1:T},
-  \by_{1:T}$ and~$\bx_{1:T} \given \bz_{1:T}, \by_{1:T}$.  Given the
-  latent states, the remaining parameters are sampled as in Bayesian
-  linear regression. }
-
 \section{Results}
@@
 
+\appendix
+
+\section{Bayesian Inference for Switching Linear Dynamical Systems}
+Our goal is to estimate the posterior probability of a sequence 
+of latent states and a set of parameters given the observed data.
+From Bayes' rule, we have,
+\begin{align}
+  p(\bz_{1:T}, \bx_{1:T}, \bTheta \given \by_{1:T}) 
+  &= 
+  \frac{p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta)}{p(\by_{1:T})}.
+\end{align}
+The numerator is the joint probability given by Eq.~\eqref{eq:joint}, and
+the denominator,~$p(\by_{1:T})$, which is also known as the
+\emph{marginal likelihood}, is given by an integral over possible
+latent states and parameters,
+\begin{align}
+  p(\by_{1:T}) &= \int p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta) 
+  \, \mathrm{d}\bx_{1:T} \, \mathrm{d}\bz_{1:T} \, \mathrm{d}\bTheta.
+\end{align}
+Unfortunately, this integral is not efficiently computable for complex
+models like the SLDS, forcing us to seek approximate inference methods
+instead. Markov chain Monte Carlo (MCMC) methods \todo{cite} offer one
+such approach.
+
+To construct our MCMC algorithm, we iteratively sample one set of 
+latent states or parameters from its conditional distribution, holding 
+the rest fixed, in a technique known as Gibbs sampling \todo{cite}. 
+There are five main sets of parameters to sample, detailed below.
+\begin{enumerate}
+  \item \textit{Gibbs sampling the discrete latent states,~$\bz_{1:T}$:}
+
+    Given the continuous latent states,~$\bx_{1:T}$, and the
+    parameters,~$\bTheta$, the conditional distribution over discrete
+    latent states is the same as in a standard hidden Markov model. A
+    joint sample from~$p(\bz_{1:T} \given \bx_{1:T}, \by_{1:T},
+    \bTheta)$ can be generated using the forward filtering backward
+    sampling (FFBS) algorithm.
+
+  \item \textit{Gibbs sampling the continuous latent states,~$\bx_{1:T}$:}
+
+    Given the discrete latent states,~$\bz_{1:T}$, the observations,~$\by_{1:T}$,
+    and the parameters,~$\bTheta$, the conditional distribution of the 
+    continuous latent states is linear and Gaussian. As with the discrete
+    latent states, a joint sample of~$p(\bx_{1:T} \given \bz_{1:T}, \by_{1:T}, \bTheta)$
+    can be generated using an FFBS algorithm.
+
+  \item \textit{Gibbs sampling the dynamics parameters,~$\{\bA^{(k)}, \bb^{(k)}, \bQ^{(k)}\}_{k=1}^K$:}
+
+    For fixed latent state sequences, the dynamics model reduces to a simple 
+    multivariate regression problem. We have,
+    \begin{multline}
+      p(\bA^{(k)}, \bb^{(k)}, \bQ^{(k)}  \given \bz_{1:T}, \bx_{1:T}, \by_{1:T}, \bTheta)
+      \\ 
+      \propto
+      p(\bA^{(k)}, \bb^{(k)}, \bQ^{(k)})
+      \prod_{t=1}^T \left[ \,
+        \distNormal(\bx_{t} \given \bA^{(k)} \bx_{t-1} + \bb^{(k)},\, \bQ^{(k)} \right]^{\bbI[z_{t-1}=k]}.
+    \end{multline}
+    If the prior distribution is the form of a matrix normal inverse Wishart (MNIW) prior, 
+    then this conditional distribution will be as well.
+
+  \item \textit{Gibbs sampling the observation parameters,~$\{\bC^{(k)}, \bd^{(k)}, \bR^{(k)}\}_{k=1}^K$:}
+
+    As with the dynamics parameters, 
+    for fixed latent state sequences, the observation model is also a 
+    multivariate regression problem. We have,
+    \begin{multline}
+      p(\bC^{(k)}, \bd^{(k)}, \bR^{(k)}  \given \bz_{1:T}, \bx_{1:T}, \by_{1:T}, \bTheta)
+      \\ 
+      \propto
+      p(\bC^{(k)}, \bd^{(k)}, \bR^{(k)})
+      \prod_{t=1}^T \left[ \,
+        \distNormal(\by_{t} \given \bC^{(k)} \bx_{t} + \bd^{(k)},\, \bR^{(k)} \right]^{\bbI[z_{t}=k]}.
+    \end{multline}
+    This, too, is conjugate when the prior distribution assumes the
+    form of a matrix normal inverse Wishart (MNIW) distribution.
+
+  \item \textit{Gibbs sampling the Markov parameters,~$\{\bpi^{(k)}\}_{k=1}^K$:}
+
+    Finally, we must sample the Markov transition matrix. We separate
+    this into its~$K$ rows, each of which specifies a probability
+    distribution,~$p(z_t \given z_{t-1}=k) = \bpi^{(k)}$.  For a fixed
+    discrete latent state sequence, the conditional distribution of~$\bpi^{(k)}$ 
+    is,
+    \begin{align}
+      p(\bpi^{(k)} \given \bz_{1:T}) &\propto
+      p(\bpi^{(k)}) \prod_{t=1}^T \left[ \pi_{z_t}^{(k)} \right]^{\bbI[z_{t-1}=k]}.
+    \end{align}
+    If the prior distribution is~$p(\bpi^{(k)}) = \distDirichlet(\bpi^{(k)} \given \alpha)$, then this
+    conditional distribution is a Dirichlet as well,
+    \begin{align}
+      p(\bpi^{(k)} \given \bz_{1:T}) 
+      &= \distDirichlet(\bpi^{(k)} \given \widetilde{\balpha}^{(k)}) \\
+      \widetilde{\alpha}_{k'}^{(k)} &= \alpha + \sum_{t=1}^T \bbI[z_t = k'] \, \bbI[z_{t-1}=k].
+    \end{align}
+\end{enumerate}
+
+Each one of these five steps leaves the desired posterior distribution as the 
+unique stationary distribution of the Markov chain. Thus, by iterating these 
+steps, the sampled states and parameters will eventually be distributed according
+to their posterior probability given the observed data. Critically, the rate 
+at which the Markov chain converges to its stationary distribution is determined
+in part by the correlation between the sampled latent states at one iteration and
+those at the next. If the chain only makes minor updates to the latent state sequence,
+it will likely take a long time to converge to the desired posterior distribution.  
+By performing joint, ``block'' updates of~$\bx_{1:T}$ and~$\bz_{1:T}$ in steps 
+1 and 2, we find that the latent state sequences are able to be explored more 
+efficiently.
+
 \end{document}

### 2016-07-29 drafts/writeup_07_25_16.tex
refactoring, adding notebooks

@@
 \includegraphics[width=5.5in]{slds} 
-\caption{Simulated data from a switching linear dynamical system.
-The population stochastically jumps between discrete states,~$z_t \in \{a,b,c,d\}$,
-each of which is color coded for visualization.
-These discrete states govern the linear dynamics of the continuous latent 
-state,~$\bx_t \in \reals^D$ (here,~$D=2$). For example, state~$a$ corresponds to a simple
-random walk, whereas state~$d$ has fast, oscillatory dynamics. Finally,
-the observed signals,~$\by_t \in \reals^N$ (here,~$N=8$), are obtained via a linear 
-transformation of the underlying, continuous state,~$\bx_t$. The correlations
-and dynamics in the observations are inherited from the dynamics of the
-latent states.}
+\caption{Simulated data from a switching linear dynamical system.  The
+  population stochastically switches between discrete states,~$z_t \in
+  \{1,\ldots,K\}$ (here,~$K=5$), each of which is color coded for
+  visualization.  These discrete states govern the linear dynamics of
+  the continuous latent state,~$\bx_t \in \reals^D$ (here,~$D=2$). For
+  example, state~$z_t=1$ corresponds to a simple random walk, whereas
+  state~$z_t=5$ has fast, oscillatory dynamics. Finally, the observed
+  signals,~$\by_t \in \reals^N$ (here,~$N=8$), are obtained via a
+  linear transformation of the underlying, continuous
+  state,~$\bx_t$. The correlations and dynamics in the observations
+  are inherited from the dynamics of the latent states.}
 \vspace{-0.5cm}

### 2016-11-17 drafts/2016_11_16/writeup.tex
adding updated writeup

--- 2016-07-29 drafts/writeup_07_25_16.tex
@@
 
-\usepackage[numbers, compress]{natbib}
+\usepackage{natbib}
 
@@
 
-\author{Scott W. Linderman 
+\author{Scott W. Linderman,
+  David M. Blei,
   and
@@
   simultaneous measurements of the majority of head ganglia neurons in
-  immobilized C. elegans \cite{kato2015global}. Moreover, since some
+  immobilized \celegans~\citep{kato2015global}. Moreover, since some
   neurons are known to reliably indicate the onset or offset of
@@
 
-\section{Model}
+\section{Introduction}
+% The world as it was:
+The nematode \celegans~is unique among the model organism of
+neuroscience. For decades, we have known the connectome of its 302
+neurons, yet a comprehensive understanding of hows its neural activity
+reflects sensory processes and produces behavior still eludes
+us. While this connectome, or wiring diagram, has provided invaluable
+information about the structure of these neural circuits, the critical
+missing piece has been large scale functional measurements of these
+circuits in action. With the advent of optical recording technologies,
+we now possess a mounting set of tools for measuring the activity of large
+fractions of these neurons simultaneously. These technologies provide
+exciting opportunities to probe the link between neural activity,
+sensory processing, and, ultimately, behavior.
+
+Recently, \citet{kato2015global} have harnessed these methodological
+advances to study the coordinated activity of hundreds of neurons
+in head-fixed \celegans. Across multiple organisms, they have found that
+neural activity reliably traces out smooth trajectories in the subspace
+spanned by the first three principal components.
+% TODO: Figure showing these trajectories
+Upon closer evaluation, they have found that different components of these
+trajectories correspond to different elements of behavior, like
+forward and reverse crawling, dorsal and ventral turns, etc. These
+results raise a number of interesting questions: can we learn a
+(potentially nonlinear) dynamical system that approximates these
+low-dimensional dynamics; is there a more appropriate subspace, or
+collection of subspaces, for capturing these dynamics; can we gain
+statistical power by partial, noisy recordings across separate worms
+and trials; and, can we segment these trajectories in an unsupervised
+manner to glean further insight into the relationship between
+low-dimensional state and behavior?
+
+% Then one day:
+We have developed a probabilistic framework for investigating these questions.
+The findings of \citet{kato2015global} --- namely, that neural activity
+is low dimensional with behavior-specific dynamics --- suggest that
+these multineuronal recordings may be aptly modeled in terms of
+a dynamic, low-dimensional latent states. Moreover, they suggest
+that neural activity may be parsed into a sequence of discrete
+actions, like turning or crawling forward. These two properties are
+naturally combined by \emph{switching linear dynamical systems} (SLDS),
+a class of probabilistic time series characterized by co-evolving
+discrete and continuous latent states.  We have extended this
+class of models with a hierarchical Bayesian framework that captures
+many of the nuances of these whole-brain recordings. In fitting these
+models to recordings from multiple \celegans, we characterize the
+globally nonlinear dynamics that govern neural activity, and we
+parse these recordings into an interpretable sequence of behavioral segments.
+
+% Raising the stakes:
+% Moment of truth:
+% World as it is now:
+The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
+we introduce the class of swithing linear dynamical systems and provide
+some insight into the latent states and model parameters. Section~\ref{sec:hslds}
+develops a hierarchical extension of the SLDS to share statistical
+strength by combining recordings from multiple organisms with partially
+overlapping sets of observed neurons.  Then, Section~\ref{sec:results}
+presents results of applying these models to a set of recordings from
+five separate \celegans. We conclude with directions for further research.
+
+\section{Switching Linear Dynamical Systems}
+\label{sec:slds}
 
@@
 \centering%
-\includegraphics[width=5.5in]{slds} 
+\includegraphics[width=5.5in]{slds.png} 
 \caption{Simulated data from a switching linear dynamical system.  The
   population stochastically switches between discrete states,~$z_t \in
-  \{1,\ldots,K\}$ (here,~$K=5$), each of which is color coded for
+  \{1,\ldots,K\}$ (here,~$K=4$), each of which is color coded for
   visualization.  These discrete states govern the linear dynamics of
   the continuous latent state,~$\bx_t \in \reals^D$ (here,~$D=2$). For
-  example, state~$z_t=1$ corresponds to a simple random walk, whereas
-  state~$z_t=5$ has fast, oscillatory dynamics. Finally, the observed
-  signals,~$\by_t \in \reals^N$ (here,~$N=8$), are obtained via a
-  linear transformation of the underlying, continuous
-  state,~$\bx_t$. The correlations and dynamics in the observations
-  are inherited from the dynamics of the latent states.}
-\vspace{-0.5cm}
+  example, these states correspond to oscillatory dynamics with
+  different frequencies. Finally, the observed signals,~$\by_t \in
+  \reals^N$ (here,~$N=8$), are obtained via a linear transformation of
+  the underlying, continuous state,~$\bx_t$, plus Gaussian noise. The
+  correlations and dynamics in the observations are inherited from the
+  dynamics of the latent states.}
 \label{fig:slds_ex}
@@
 
-Our model is based on the following assumptions: (i) the instantaneous
+The SLDS model is based on the following assumptions: (i) the instantaneous
 neural activity,~$\by_t$, reflects an underlying, low-dimensional
@@
 latent state; and (iv) the observed neural activity is a linear function
-of the underlying states with additive Gaussian noise.
+of the underlying states with additive Gaussian noise.  Assumptions (i)
+and (iv) are justified by the low-dimensional trajectories that
+\citet{kato2015global} revealed with PCA (a linear dimensionality
+reduction method). Assumptions (ii) and (iii) follow from the
+trajectories naturally segment into discrete units, each with relatively
+simple dynamics.
 
-\TODO{Discuss the motivation for these modeling assumptions.}
-
-These assumptions
-are combined in a switching linear dynamical system, which we formalize 
-with the following generative model:
+These assumptions are formalized with the following generative model:
 \begin{align}
@@
 \end{align}
-Our beliefs about the dynamics of these latent states are encoded in the 
+Beliefs about the dynamics of these latent states are encoded in the 
 form of the conditional distributions for~$z_t$ and~$\bx_t$. First,
@@
 \begin{align}
-  p(z_t \given z_{t-1}, \bTheta) &\sim \distDiscrete(\bpi^{(z_{t-1})}).
+  p(z_t \given z_{t-1}, \bTheta) &\sim \bpi^{(z_{t-1})}.
 \end{align}
-Next, we imbue the continuous latent states with linear Gaussian dynamics,
+Next, the continuous latent state is imbued with linear Gaussian dynamics,
 \begin{align}
@@
 \centering%
-\begin{subfigure}{.49\textwidth}
-\includegraphics[width=\textwidth]{dynamics_1}
+\begin{subfigure}{.24\textwidth}
+\includegraphics[width=\textwidth]{rotation_1}
 \end{subfigure} 
-\begin{subfigure}{.49\textwidth}
-\includegraphics[width=\textwidth]{dynamics_4}
+\begin{subfigure}{.24\textwidth}
+\includegraphics[width=\textwidth]{rotation_2}
+\end{subfigure} 
+\begin{subfigure}{.24\textwidth}
+\includegraphics[width=\textwidth]{rotation_3}
+\end{subfigure} 
+\begin{subfigure}{.24\textwidth}
+\includegraphics[width=\textwidth]{rotation_4}
 \end{subfigure} 
@@
 \label{fig:slds_dynamics_ex}
-\vspace{-0.5cm}
 \end{figure}
@@
 
-\TODO{Discuss interpretation of the parameters. Set the stage for visualizing~$\bA^{(k)}$ 
-in later sections.}
+Figure~\ref{fig:slds_ex} illustrates a sample from this generative model.
+The observations are a sequence of vectors, in this case they are ~$N=3$ dimensional
+vectors,~$\by_t$.  These states are a noisy, linear projection of an underlying
+continuous latent state,~$\bx_t$, which is~$D=2$ dimensional in this case.
+Here, the continuous state switches between~$K=4$ discrete regimes (four
+different colors), each corresponding to different frequencies of oscillation.
+The discrete state, encoded by~$z_t \in \{1, 2, 3, 4\}$ (blue, red, yellow, green),
+follows a simple Markov process with transition matrix~${\bP = \{\pi^{(k)}\}_{k=1}^K}$
+(not shown).
+
+How should we interpret the these parameters? Linear dynamical
+systems can essentially capture simple dynamics, like rotations, exponential
+growth, and exponential decay. The stability of the system is determined
+by the eigenvalues of the dynamics matrix,~$\bA^{(k)}$ --- if the eigenvalues
+all have modulus less than one, the system is asymptotically stable, and
+converges to a fixed point at~$(\bI - \bA^{(k)})^{-1} \bb^{(k)}$. If
+the largest eigenvalue is exactly one, the system oscillates in perpetuity.
+If any eigenvalue exceeds one in magnitude, the system diverges exponentially
+quickly.  A variety of linear dynamical systems are illustrated in
+Figure~\ref{fig:slds_dynamics_ex}. All of these systems are stable, and their
+fixed points are shown as dots.
+
+% Nonlinearity of compositional system
+While the capacity of linear dynamical systems is somewhat limited,
+the composition of many linear dynamical modes, as in an SLDS, yields
+a highly nonlinear system. This is particularly important for modeling
+neural data, which is believed to have nonlinear and often behavior-dependent
+dynamics.
+
+Given observations,~$\by_{1:T}$, our goal is to infer the discrete
+latent states,~$bz_{1:T}$, and the continuous latent states,~$\bx_{1:T}$,
+as well as to learn the parameters,~$\bTheta$.  We take a Bayesian
+approach, specifying conjugate prior distributions for the parameters,
+and using Markov chain Monte Carlo (MCMC) to estimate the posterior
+distribution of the parameters and states given the observed data.
+Details of this MCMC algorithm are provided in Appendix~\ref{app:mcmc}.
+
+\section{Hierarchical Switching Linear Dynamical Systems}
+\label{sec:hslds}
+
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{identified_neurons.pdf} 
+\caption{Identified neurons in the five worms from \citet{kato2015global}.
+  Across all five worms, these 60 labeled neurons were identified in at
+  least one worm. In addition to these labeled neurons,
+  each worm contains roughly 70 unlabeled neurons that inform our
+  estimate of that worm's latent states, but which we cannot share
+  across worms. }
+\label{fig:identified_neurons}
+\end{figure}
+
+The neural recordings of \citet{kato2015global} present additional
+opportunities to extend the SLDS. In these experiments, recordings
+have been made in five separate worms. Each recording contains roughly
+100 of the worm's 302 neurons, but the subset of observed neurons
+varies from worm to worm. Moreover, while each of these 302 neurons
+has a unique name, only about 30 out of the 100 neurons could be
+reliably identified in each recording. For example, in the recording
+of worm A, we may observe 100 neurons of which 30 have been identified
+as, say, {AAV}, {RIM}, etc., while the remaining 70 are unidentified.
+Figure~\ref{fig:identified_neurons} shows the 60 labeled neurons which
+were identified in at least one of the worms. For example, ABL is
+identified in all five worms, whereas SIADL is only found in the
+second worm.  In addition to these identified neurons, we also have
+roughly 70 unlabeled neurons for each worm. While these unlabeled
+neurons do not provide information that can be shared across worms,
+they do provide information about that worm's latent states.  Our goal
+is to combine these recordings to extract a shared model of the neural
+dynamics of \celegans.
+
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{hslds.png} 
+\caption{A schematic of the hierarchical SLDS for combining information
+  across multiple worms in order to learn a shared dynamical system
+  for \celegans.  We have a set of shared set of model parameters,~$\bTheta$,
+  represented by the large node at the top of the graphical model. Each
+  worm,~$w$ has its own set of discrete latent states (colored nodes)
+  and continuous latent states (nodes with colored outlines) and observed neurons.
+  For example, neuron 1 may appear in all neurons whereas neuron 2
+  may only be seen in worm 2. Thus, each worm provides information
+  about the mapping from latent states to observations for only a
+  subset of neurons. In terms of the model, this corresponds to a
+  subset of rows in the matrix~$\bC$ and the vector~$\bd$. Finally,
+  we allow each worm to have local parameters that capture, for example,
+  the amount of fluorescent protein expressed in a particular neuron
+  for a given worm. 
+}
+\label{fig:hslds}
+\end{figure}
+
+We propose the following hierarchical SLDS (hSLDS) to combine data from
+multiple worms. Since these worms are genetically identical, we hypothesize
+that their dynamics should be the same. That is, they share the same set
+of~$K$ discrete states, as well as the
+parameters~$\{\bA^{(k)}, \bb^{(k)}, \bQ^{(k)}, \bpi^{(k)}\}_{k=1}^K$.
+We assume the following mapping from latent states to the neural activity
+of worm~$w$, time~$t$, and neuron~$n$:
+\begin{align}
+  y_{w,t,n} &= \gamma_{w,n} \bc_n^\trans \bx_{w,t} + d_{w,n} + \epsilon_{w,t,n}, \\
+  \epsilon_{w,t,n} &\sim \distNormal(0, r_{w,n}^2).
+\end{align}
+Thus, the mapping from latent states to the activity of neuron~$n$ is
+partially shared across worms, in that all worms share the
+vector~$\bc_n$.  These constitute the rows of an emission
+matrix,~$\bC$. Note that, for simplicity, we have required all
+discrete states,~$z_t$ to share the same emission matrix. However, we
+have introduced a set of worm-specific parameters,~$\{\gamma_{w,n},
+d_{w,n}, r_{w,n}\}$. The gain,~$\gamma_{w,n}$, captures differences in
+amplitude of the neural signal that may arise from variation in
+fluorescent protein expression from worm to worm.  Likewise, the
+bias,~$d_{w,n}$, accounts for different baseline levels of
+activity. Finally, the noise variance,~$r_{w,n}^2$, enables neuron~$n$
+to have different marginal variance from worm to worm, which may arise
+due to a number of factors, like protein expression levels and the
+clarity of a particular recording.  Figure~\ref{fig:hslds} provides a
+graphical representation of this model.
 
 \section{Results}
+\label{sec:results}
 
-\bibliographystyle{unsrt}
+\bibliographystyle{plainnat}
 \bibliography{writeup}
@@
 \section{Bayesian Inference for Switching Linear Dynamical Systems}
+\label{app:mcmc}
 Our goal is to estimate the posterior probability of a sequence 

### 2016-12-01 drafts/2016_11_16/writeup.tex
adding hslds code; draft of writeup with results

@@
   and
-  Liam Paninski \\
+  Liam Paninski
+  \\
   Columbia University
@@
 reflects sensory processes and produces behavior still eludes
-us. While this connectome, or wiring diagram, has provided invaluable
+us \citep{white1986structure}. While this connectome, or wiring diagram, has provided invaluable
 information about the structure of these neural circuits, the critical
@@
 is low dimensional with behavior-specific dynamics --- suggest that
-these multineuronal recordings may be aptly modeled in terms of
+these multi-neuronal recordings may be aptly modeled in terms of
 a dynamic, low-dimensional latent states. Moreover, they suggest
@@
 The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
-we introduce the class of swithing linear dynamical systems and provide
+we introduce the class of switching linear dynamical systems and provide
 some insight into the latent states and model parameters. Section~\ref{sec:hslds}
@@
 
+The switching linear dynamical system (SLDS) is a powerful model for
+stochastic time series with nonlinear dynamics \citep{ackerson1970state, chang1978state,
+  hamilton1990analysis, bar1993estimation, ghahramani1996switching,
+  murphy1998switching, fox2009nonparametric}, and as such, it is
+particularly well suited to the analysis of whole-brain activity.
 Assume the instantaneous neural activity at time~$t$ for a population
@@
 calcium imaging settings, the entries in this vector may be
-instaneous~$\Delta F/F$ measurements, or another signal that captures
+instantaneous~$\Delta F/F$ measurements, or another signal that captures
 neural activity. In this experiment, we use the smoothed time
@@
 \begin{align}
-  y_{w,t,n} &= \gamma_{w,n} \bc_n^\trans \bx_{w,t} + d_{w,n} + \epsilon_{w,t,n}, \\
+  y_{w,t,n} &= \gamma_{w,n} (\bc_n^\trans \bx_{w,t} + d_{n}) + \epsilon_{w,t,n}, \\
   \epsilon_{w,t,n} &\sim \distNormal(0, r_{w,n}^2).
@@
 discrete states,~$z_t$ to share the same emission matrix. However, we
-have introduced a set of worm-specific parameters,~$\{\gamma_{w,n},
-d_{w,n}, r_{w,n}\}$. The gain,~$\gamma_{w,n}$, captures differences in
-amplitude of the neural signal that may arise from variation in
-fluorescent protein expression from worm to worm.  Likewise, the
-bias,~$d_{w,n}$, accounts for different baseline levels of
-activity. Finally, the noise variance,~$r_{w,n}^2$, enables neuron~$n$
-to have different marginal variance from worm to worm, which may arise
-due to a number of factors, like protein expression levels and the
-clarity of a particular recording.  Figure~\ref{fig:hslds} provides a
-graphical representation of this model.
+have introduced a set of worm-specific, scalar
+parameters,~$\{\gamma_{w,n}$ and~$r_{w,n}\}$. The
+gain,~$\gamma_{w,n}$, captures differences in amplitude of the neural
+signal that may arise from variation in fluorescent protein expression
+from worm to worm.  The noise variance,~$r_{w,n}^2$, enables
+neuron~$n$ to have different marginal variance from worm to worm,
+which may arise due to a number of factors, like protein expression
+levels and the clarity of a particular recording.  Whereas the general
+SLDS allows for covariance in the noise from neuron to neuron, for
+simplicity we have assumed that this matrix is worm-specific and
+diagonal. Figure~\ref{fig:hslds} provides a graphical representation
+of this model.
+
+% Talk about the prior on gain... we used a truncated normal prior,
+% maybe the fact that we needed such a constrained prior means that
+% this gain parameter is a bad idea after all...
 
@@
 \label{sec:results}
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{shared_activity_worm1.pdf} 
+\caption{Fluorescence traces of 15 neurons in the second worm. These
+  neurons were identified in all 5 worms. Axis limits are the same for
+  all neurons. }
+\label{fig:shared_activity}
+\end{figure}
+
+We fit the hSLDS to recordings of five \textit{C. Elegans} collected
+by \cite{kato2015global}. Each recording is 18 minutes long with a
+sampling frequency of 3Hz, yield 3240 time steps per worm.  The
+recordings contain~$N_w=109$, $107$, $131$, $126$, and $129$ neurons,
+respectively. Of these, a subset have been manually identified.
+Figure~\ref{fig:identified_neurons} shows the neurons that were
+identified in each worm. We will capitalize on those neurons that are
+identified in multiple worms in order to learn a shared subspace of
+neural activity and a shared set of dynamical states for all
+\textit{C. Elegans}. Figure~\ref{fig:shared_activity} shows the neural
+activity of 15 identified neurons in the second worm. We immediately
+recognize repeating patterns of activity; these will correspond to
+characteristic ``loops'' in the low dimensional portrait of the
+population activity.
+
+\paragraph{Neural Activity is Low Dimensional}
+
+% PCA trajectories of neural activity
+\begin{figure}[t]
+  \centering%
+  \begin{subfigure}[b]{.19\linewidth}
+    \includegraphics[width=\textwidth]{pca_trajectory_worm0.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{.19\linewidth}
+    \includegraphics[width=\textwidth]{pca_trajectory_worm1.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{.19\linewidth}
+    \includegraphics[width=\textwidth]{pca_trajectory_worm2.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{.19\linewidth}
+    \includegraphics[width=\textwidth]{pca_trajectory_worm3.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{.19\linewidth}
+    \includegraphics[width=\textwidth]{pca_trajectory_worm4.pdf}
+  \end{subfigure} 
+  \caption{Principal components analysis of each individual worm.
+    In all cases, the neural activity is relatively low dimensional,
+    with large percentages of the variance captured by the first three
+    principal components (PC's). Above, the projection of roughly 100-dimensional
+    neural activity onto the subspace spanned by the first three PC's.
+    The hierarchical SLDS will find a common subspace for all worms, and,
+    furthermore, segment the low dimensional trajectories into a sequence
+    of discrete pieces.}
+\label{fig:pca_trajectory}
+\end{figure}
+
+A fundamental assumption of the hSLDS is that neural activity is
+low dimensional. To test this assumption, we first run principal
+components analysis (PCA) on each worm's
+population activity (a~$T \times N_w$ matrix) separately. We find
+that the top three principal components (PC's)capture a large fraction
+of the variance for all worms, as shown in the bottom panels of
+Figure~\ref{fig:pca_trajectory}. The top panels show the projection
+of neural activity onto the first three PC's. While the PCA
+projections show qualitatively similar patterns, like intersecting
+and orthogonal loops, the subspaces are not aligned across worms
+and the trajectories are not particularly smooth or interpretable.
+The hSLDS addresses these issues by learning a shared subspace
+across all worms, and segmenting the low dimensional
+trajectories into a sequence of states with simple, linear dynamics.
+
+\paragraph{The hSLDS reveals a shared subspace and canonical dynamics across worms}
+
+% Inferred dynamical system -- set of discrete latent states
+\begin{figure}[t]
+  \centering%
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{xs_3d_worm0.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{xs_3d_worm1.pdf}
+  \end{subfigure}
+  \\
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{xs_3d_worm2.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{xs_3d_worm3.pdf}
+  \end{subfigure}  
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{xs_3d_worm4.pdf}
+  \end{subfigure}
+  \caption{Latent states discovered by the hierarchical SLDS.
+    Since the emission matrix,~$\bC$, is shared across worms, the
+    coordinates of the latent states are shared as well. This
+    reveals a common pattern of low dimensional activity characterized
+    by interlocking and opposing loops. Colors denote the discrete
+    latent state inferred for each part of the trajectory. For example,
+    one loop begins with the latent state shooting out from the origin
+    along the orange and purple states, and then returning via the blue
+    state. Another begins with the green state, which swings upward and
+    is then followed by a return via the red state. In worms 3 and 5,
+    the upward swing is slightly tilted, as captured by the pink 
+    rather than the green state. While harder to see in this view, there
+    is a third loop in yellow that lies in the~$(x_1, x_2)$ plane.
+    See supplementary videos for further detail.
+  }
+\label{fig:hslds_trajectory}
+\end{figure}
+
+The hSLDS shares information across worms in order to learn a
+canonical subspace of neural activity and a common set of
+low-dimensional dynamics. Figure~\ref{fig:hslds_trajectory} shows the
+inferred continuous latent states,~$\bx_{1:T}$, of the hSLDS for each
+of the five worms. There are three points to note, First, in contrast
+to the PCA trajectories in Figure~\ref{fig:pca_trajectory}, the
+similarity in hSLDS trajectories is readily apparent. This is due to
+the fact that all worms share the same subspace, which is defined by the
+global emission matrix~$\bC$. Second, the hSLDS trajectories
+are considerably smoother than their PCA counterparts. This is a
+direct result of the set of linear dynamics that comprise the hSLDS.
+Assuming that the latent state,~$\bx_{t+1}$, is linearly related to the
+preceding state,~$\bx_t$, induces an inductive bias toward smooth
+trajectories in latent space. This reduces the discontinuities and noise
+in the PCA trajectories and reveals the characteristic patterns of
+latent dynamics. Third, the hSLDS trajectories have been segmented into
+snippets, each corresponding to a different discrete latent state,~$z_t$, with
+its own set of linear dynamics. These discrete states illustrated
+as different colors in the hSLDS trajectories. These provide a
+handle into understanding the latent dynamics of \textit{C. Elegans}
+population activity, and a useful latent variable for relating
+activity to behavior.
+
+\paragraph{Neural dynamics switch between a set of linear regimes}
+% 3D dynamics
+\begin{figure}[th!]
+  \centering%
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_0.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_1.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_2.pdf}
+  \end{subfigure}
+  \\
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_3.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_4.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_5.pdf}
+  \end{subfigure}
+  \\
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_6.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_7.pdf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\linewidth}
+    \includegraphics[width=\textwidth]{dynamics_3d_8.pdf}
+  \end{subfigure}  
+%  \\
+%  \begin{subfigure}[b]{0.32\linewidth}
+%    \includegraphics[width=\textwidth]{dynamics_3d_9.pdf}
+%  \end{subfigure}
+%  \begin{subfigure}[b]{0.32\linewidth}
+%    \includegraphics[width=\textwidth]{dynamics_3d_10.pdf}
+%  \end{subfigure}
+%  \begin{subfigure}[b]{0.32\linewidth}
+%    \includegraphics[width=\textwidth]{dynamics_3d_11.pdf}
+%  \end{subfigure}  
+  \caption{The discrete dynamical states that compose this switching
+    linear dynamical system. The vector fields point in the direction of
+    ~$\bA^{(k)} \bx + \bb^{(k)}$~($k=1 \ldots 9$). While the dynamics are
+    defined for all~$\bx$,
+    they are only plotted at locations where the corresponding state is used.
+    States are ordered from most common (1)
+    to least common (9); see supplementary material for remainder of
+    states. The most common states (1 and 2) simply decay to the
+    origin. State 3 is a tight loop in the~$(x_1, x_2)$ plane. States
+    4 and 8 are larger loops in the~$(x_2, x_3)$ plane. States 5, 6,
+    and 9 are sections of a third loop opposite to State 3. Finally,
+    State 7 seems to bridge between loops.}
+\label{fig:dynamics}
+\end{figure}
+
+As we have stated, the hSLDS segments the low dimensional trajectories
+into a sequence of discrete states, each governed by simple linear dynamics.
+These dynamics are shown in Figure~\ref{fig:dynamics}. The states are ordered
+from most common (State 1) to least common (State 9). The model was fit with~$K=15$
+states, but only 11 were used. The remaining two are omitted from this figure
+since they are used very rarely. The vector fields shown here point in the
+direction of~$\bA^{(k)} \bx_t + \bb^{(k)}$ for~$k=1, \ldots, 9$. While these
+dynamics are defined for all~$\bx_t \in \reals^3$, for clarity, they are only plotted
+at the locations in latent space where the corresponding state is deployed.
+
+The most common states (1 and 2) correspond to decays toward the origin.
+This is to be expected, since the neurons are quiescent most of the time
+(c.f. Figure~\ref{fig:shared_activity}). The remaining states capture
+different sections of the intersecting loops seen in Figure~\ref{fig:hslds_trajectory}.
+State 3 is a short loop to the left in the~$(x_1, x_2)$ plane. States 4
+and 8 correspond to the large upward swings that trace out loops in the~$(x_2, x_3)$
+plane. States 5, 6, and 9 correspond to loops that swing out to the right
+of the origin, again in the~$(x_1, x_2)$ plane. Finally, state 7 seems to
+connect the~$(x_2, x_3)$ loop to the~$(x_1, x_2)$ loop, bypassing the origin.
+
+This discrete segmentation of the latent state trajectories serves multiple
+purposes. First, by stitching together discrete states with unique
+linear dynamics, the hSLDS models a globally \emph{nonlinear} dynamical
+system.  Second, these discrete latent states provide a valuable covariate
+for understanding switches in the neural activity. As we will see below,
+when we condition on the entry into these discrete states, characteristic
+patterns of population activity are revealed. Third, these states provide
+an interpretable summary of the complex, global brain activity. Next, we
+will show that these discrete states correspond to meaningful behaviors.
+
+\paragraph{Discrete states correspond to distinct behaviors}
+
+% Discrete state transitions mimic those of hand-labeled behavioral segments
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{discrete_state_cps_worm1.pdf} 
+\caption{The inferred discrete states align with the manual segmentation
+  of \citet{kato2015global}. \textit{Top:} samples from the posterior
+  distribution of discrete states,~$z_{1:T}$. Each row is a sample, and
+  each color denotes a distinct value from 1 to~$K=15$. \textit{Middle:}
+  from these samples we estimate the probability of a change-point
+  (i.e.~$z_t \neq z_{t-1}$). \textit{Bottom:} The inferred change-points
+  align with the manual segmentation of \citet{kato2015global}. Note,
+  however, that the colors do not perfectly align since there is not a
+  one-to-one correspondence between the inferred and manual states. For
+  example, the blue and red states of~\citet{kato2015global}, which
+  correspond to ``REVSUS'' and ``SLOW,'' respectively, are combined into
+  inferred State 1.
+}
+\label{fig:discrete_state_comparison}
+\end{figure}
+
+\citet{kato2015global} manually labeled segments of time corresponding to
+distinct behaviors, like forward or reverse crawling and dorsal or ventral
+turns. This process is both laborious and subjective. The hSLDS provides
+a principled and automated alternative to manual labeling, by automatically
+segmenting activity into snippets governed by linear dynamics. We find that
+the manual and hSLDS segmentations are closely aligned.
+
+Figure~\ref{fig:discrete_state_comparison} illustrates both the uncertainty
+of the discrete labeling as well as the alignment with the ``ground truth''
+labels. The top panel shows samples from the posterior distribution over
+latent state sequences. Each row corresponds to a sample of~$z_{1:T}$, and
+as before, the colors denote different discrete states. The variability
+from sample to sample highlights the uncertainty that is inherent in assigning
+a discrete label to a point in time --- where one state ends and the next
+begins is not entirely clear from the noisy data.
+
+From this sequence of discrete state samples we derive an estimate of the
+\emph{change-point probability}, i.e. the probability that~$z_t \neq z_{t-1}$.
+This is shown in the middle panel. In some instants it is eminently clear
+that the discrete state has switched; this is due to large changes in the
+dynamical trajectory that favor one state over another. More often, however,
+this probability is lower and distributed over a window of time. Thus, our
+Bayesian approach provides a principled means of assessing this uncertainty.
+
+% Neural reconstruction
+\begin{figure}[t]
+\centering%
+\includegraphics[width=5.5in]{neural_activity_worm_1.pdf} 
+\caption{Observed (black) and smoothed (blue) activity of six neurons
+  from worm 2. The smoothed activity is given by~$\gamma_{w,n}
+  (\bc_n^\trans \bbE[\bx_{w,t}] + d_n)$. Shading denotes ~$\pm 1$
+  standard deviation under the posterior given by~$\sqrt{\gamma_{w,n}^2 \bc_n^\trans
+    \text{Var}[\bx_{w,t}] \bc_n + \sigma^2_{w,n}}$. Inferred discrete states
+  are shown above for reference. }
+\label{fig:neural_reconstruction}
+\end{figure}
+
+Finally, we see that these change-points align with the manual segmentation
+of~\citet{kato2015global}, which is shown in the bottom panel. Note that the
+manual and inferred state labels are not in one-to-one correspondence. This
+is to be expected given both the invariance to permutation (though both are
+sorted by usage), but more importantly because some manual states are grouped
+into one inferred state. For example, the red and blue manual states, which
+are labeled ``REVSUS'' and ``SLOW'', respectively, are combined in the blue
+inferred state. Likewise, the manual sequence of purple followed by green,
+which corresponds to a dorsal turn followed by a forward crawl, is combined
+in the yellow inferred state.
+Other states, like the yellow manual state, which corresponds
+to a ventral turn, are separated into two inferred states, here red
+and green. While the exact divisions between states are somewhat arbitrary,
+the change points in these trajectories are clearly revealed by the hSLDS.
+
+\paragraph{The hSLDS captures salient patterns of neural activity}
+
+% Reconstruct the neural activity
+As a generative model, the hSLDS provides a distribution over observed
+neural activity. This provides a valuable sanity check for our results ---
+does the model capture the variability of the data?
+Figure~\ref{fig:neural_reconstruction} shows that indeed it does. We
+plot the observed traces of six neurons in black and the predicted traces
+under the posterior of the hSLDS in blue. With a Bayesian approach, we
+also obtain an estimate of the uncertainty of these predictions, shown
+here as shaded blue regions around the mean. We see that for most neurons,
+the predicted and observed trajectories are well aligned, indicating that
+the hSLDS with~$D=3$ dimensional latent states has sufficient capacity
+to capture the variability of the data; however, some spikes in the activity
+of neuron AVBL are smoothed over in this reconstruction.
+
+% Where do neurons lie in latent space?
+% - which neurons are most interesting?
+% - any listed in their paper
+\paragraph{The hSLDS discovers a low-dimensional embedding of neurons}
+\begin{figure}[t]
+\centering%
+\includegraphics[width=4.in]{neuron_embedding.pdf} 
+\caption{Three-dimensional embedding of neurons given by the
+  rows of the emission matrix,~$\bc_n$. For reference, the
+  true, one-dimensional location of the soma is shown above
+  (jittered for clarity). While most neurons are clustered
+  near the origin (reflecting the fact that most neurons
+  have low amplitude activity), the more active neurons (shown
+  above) stand out.}
+\label{fig:neuron_embedding}
+\end{figure}
+
+In addition to yielding an interpretable, low-dimensional portrait
+of neural activity, the hSLDS also provides an embedding of neurons
+in latent space via the rows of the emission matrix,~$\bC$. Nearby
+neurons in latent space will have highly correlated activity, and neurons
+that are opposed will be anti-correlated. Figure~\ref{fig:neuron_embedding}
+shows this embedding for the 60 neurons that are identified in at least
+one of the five worms. For reference, the true one-dimensional location
+of the neuron's soma along the head to tail axis is shown above.
+Most of the neurons are clustered around the origin, reflecting the fact
+that most neurons have little activity. However, two pairs of neurons
+stand out: (AVAL, AVAR) and (RIML, RIMR). These pairs of neurons have larger
+amplitude activity as shown above. These pairs both drive backward locomotion
+(reversal), and are correlated via a gap junction.
+
+% State entry triggered neural responses
+\paragraph{Discrete state onset reveal stereotyped population activity}
+
+\begin{figure}[t]
+  \centering%
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_0.pdf}
+  \end{subfigure} \\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_1.pdf}
+  \end{subfigure} \\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_2.pdf}
+  \end{subfigure}\\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_3.pdf}
+  \end{subfigure}
+  \caption{Neural activity conditioned upon entry into each discrete
+    state. The~$x$-axis is time since entering the discrete state
+    (from 0 to 5 seconds), and the $y$-axis is the average neural
+    response~$\pm 2$ standard deviations.  Light gray traces show a
+    sample of individual neural activity traces.  Each row corresponds
+    to the discrete state with the corresponding color. Here shown are
+    States 1-4.}
+  \label{fig:state_trigged_activity_1}
+\end{figure}
+
+\begin{figure}[t]
+  \centering%
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_4.pdf}
+  \end{subfigure} \\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_5.pdf}
+  \end{subfigure} \\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_6.pdf}
+  \end{subfigure}\\
+  \begin{subfigure}[b]{5.5in}
+    \includegraphics[width=\textwidth]{neural_responses_subset_7.pdf}
+  \end{subfigure}
+  \caption{Same as figure above but for States 5-8.}
+  \label{fig:state_trigged_activity_2}
+\end{figure}
+
+The discrete latent states partition neural activity into interpretable
+and behaviorally relevant segments. As such, they are natural covariates
+for neural activity. Figures~\ref{fig:state_trigged_activity_1}
+and~\ref{fig:state_trigged_activity_2} illustrate the distribution of
+neural responses upon entry into each of the top 8 discrete states.
+The~$x$ axis in these plots denotes time from 0 to 5
+seconds after entry into the corresponding state. The colored line
+denotes the mean response of the corresponding neuron (same $y$ limits
+in all panels), and the shaded regions denote~$\pm 2$ standard deviations.
+The light gray traces show a sample of randomly chosen neural responses.
+That AVA and RIM neurons are associated with reverse crawling suggests
+that states 2, 3, and 8 are related to forward locomotion (since the neural
+activity is decreasing) and states 5, 6, and 7 are related to the
+reverse locomotion (since the activity is increasing).
+
+% Residual analysis -- how well does the model fit?
+% - baselines: probabilistic PCA, classical SLDS
+% - note that these are less constrained!
+
+\section{Discussion}
+The elucidation of global brain dynamics and their relation to behavior
+is perhaps the foremost goal of systems neuroscience. \celegans, with
+its well-studied connectome and corresponding array of optical imaging tools,
+provides many exciting opportunities to make progress toward this goal.
+To do so, however, requires statistical methods capable of revealing latent
+structure in complex neural data, and combining data across many worms.
+The hierarchical SLDS satisfies many of these demands, decomposing
+high-dimensional, nonlinear dynamics into a sequence of discrete latent
+states with low-dimensional, linear dynamics. Moreover, by sharing across
+worms while allowing for individual variability, the hSLDS reveals a
+common set of latent dynamics.
+
+% Remaining questions
+Many questions remain. First, as a generative model, the hSLDS falls
+somewhat short with its relatively limited model for discrete dynamics.
+It is clear from these portraits of neural activity that discrete states
+are deployed in very localized regions of latent space. Our recent work on
+recurrent switching linear dynamical systems (rSLDS) leverages this type
+of structure \citep{linderman2016recurrent}. In future work, we intend to
+combine the hSLDS and rSLDS and apply them to this data.
+
+Second, our analyses of state-triggered neural activity have only scratched
+the surface. We intend to look more closely at the population responses
+in these distinct latent states and search for further insights into how
+this low-dimensional population activity drives behavior.
+
+Third, we have not leveraged the availability of the \celegans connectome
+in this work. In theory, this connectome should provide strong information
+about the correlational structure of population activity. It remains an
+open question how best to combine this anatomical data with state space
+models like these.
+
+Finally, while we have utilized the neurons that are found in multiple
+worms to learn a shared subspace of activity and a common set of discrete
+states, we could further improve upon this work by inferring neuron
+identities for the hundreds of unlabeled neurons.  Their rough locations
+and the correlations across worms provides a substantial amount of information
+to tackle this problem.
 
@@
 models like the SLDS, forcing us to seek approximate inference methods
-instead. Markov chain Monte Carlo (MCMC) methods \todo{cite} offer one
+instead. Markov chain Monte Carlo (MCMC) methods \citep{gilks2005markov, robert2013monte} offer one
 such approach.
@@
 latent states or parameters from its conditional distribution, holding 
-the rest fixed, in a technique known as Gibbs sampling \todo{cite}. 
+the rest fixed, in a technique known as Gibbs sampling. 
 There are five main sets of parameters to sample, detailed below.

### 2016-12-01 experiments/fit_hslds.py

@@ -1,0 +1,866 @@
+import os, pickle, copy
+
+import numpy as np
+np.random.seed(0)
+
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
+import matplotlib.gridspec as gridspec
+import matplotlib.animation as manimation
+
+from hips.plotting.colormaps import gradient_cmap
+from hips.plotting.sausage import sausage_plot
+from hips.plotting.layout import create_axis_at_location
+
+# Come up with a set of colors
+import seaborn as sns
+sns.set_style("white")
+# sns.set_context("paper")
+
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+
+colors = sns.xkcd_palette(color_names)
+cmap = gradient_cmap(colors)
+
+import importlib
+from sklearn.decomposition import PCA
+
+from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
+from pybasicbayes.util.text import progprint_xrange
+from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
+from pyslds.util import get_empirical_ar_params
+
+
+
+import zimmer.io
+importlib.reload(zimmer.io)
+from zimmer.io import WormData, N_worms, find_shared_neurons, load_key
+
+import zimmer.states
+importlib.reload(zimmer.states)
+
+import zimmer.models
+importlib.reload(zimmer.models)
+from zimmer.models import MultiEmissionWeakLimitStickyHDPHMMSLDS
+
+import zimmer.emissions
+importlib.reload(zimmer.emissions)
+from zimmer.emissions import HierarchicalDiagonalRegression
+from zimmer.emissions import HierarchicalDiagonalRegressionFixedScale
+from zimmer.emissions import HierarchicalDiagonalRegressionTruncatedScale
+
+import zimmer.plotting
+importlib.reload(zimmer.plotting)
[…]

### 2016-12-01 zimmer/emissions.py

@@ -1,0 +1,227 @@
+"""
+More sophisticated emission models than just simple regressions.
+"""
+import numpy as np
+
+from pybasicbayes.distributions import DiagonalRegression
+from pybasicbayes.util.general import objarray
+from pybasicbayes.util.stats import sample_gaussian, sample_invgamma
+
+class HierarchicalDiagonalRegression(DiagonalRegression):
+    """
+    y_i^g ~ N(a^g (c \dot x_i), eta^g)
+
+    where
+    x_i, y_i:   covariates and observations
+    c:          shared regression weights
+    a^g:        group specific scaling
+    eta^g       group specific variance
+    """
+    def __init__(self, D_out, D_in, N_groups,
+                 mu_0=None, Sigma_0=None, alpha_0=3.0, beta_0=2.0,
+                 A=None, sigmasq=None, niter=1):
+
+        # Initialize the scaling factors for each group
+        self._D_out = D_out
+        self._D_in = D_in
+        self.N_groups = N_groups
+        self._A = np.zeros((D_out, D_in))
+        self.scale = np.ones((N_groups, D_out))
+        self.sigmasq_flat = np.ones((N_groups, D_out))
+
+
+        self.affine = False  # We do not yet support affine
+
+        mu_0 = np.zeros(D_in) if mu_0 is None else mu_0
+        Sigma_0 = np.eye(D_in) if Sigma_0 is None else Sigma_0
+        assert mu_0.shape == (D_in,)
+        assert Sigma_0.shape == (D_in, D_in)
+        self.h_0 = np.linalg.solve(Sigma_0, mu_0)
+        self.J_0 = np.linalg.inv(Sigma_0)
+        self.alpha_0 = alpha_0
+        self.beta_0 = beta_0
+
+        self.niter = niter
+        self.resample(data=[], groups=[])  # initialize from prior
+
+    @property
+    def A(self):
+        return self.scale[:, :, None] * self._A[None, :, :]
+
+    @property
+    def sigma(self):
+        return np.array([np.diag(s) for s in self.sigmasq_flat])
+
+    def log_likelihood(self, xy, group=None, mask=None):
+        assert group is not None
+
+        if isinstance(xy, tuple):
+            x,y = xy
+        else:
+            x,y = xy[:,:self.D_in], xy[:,self.D_in:]
+            assert y.shape[1] == self.D_out
+
+        if mask is None:
+            mask = np.ones_like(y)
+        else:
[…]

### 2016-12-01 zimmer/io.py

@@ -24,6 +24,12 @@ def load_labels():
             "sevenStateColoring.mat"))
     return zimmer_state_labels
 
+def load_key():
+    data = load_labels()
+    key = data["sevenStateColoring"]["key"][0,0][0]
+    key = [str(k)[2:-2] for k in key]
+    return key
+
 class WormData(object):
     """
     Wrapper for basic worm dataset

### 2016-12-01 zimmer/models.py

@@ -1,0 +1,116 @@
+import numpy as np
+
+from pyhsmm.models import HMMPython, HMM, WeakLimitHDPHMM, WeakLimitStickyHDPHMM
+
+from zimmer.states import MultiEmissionSLDSStatesEigen, MultiEmissionSLDSStatesPython
+from pyslds.models import _SLDSMixin
+
+class _MultiEmissionSLDS(_SLDSMixin):
+    def __init__(self,dynamics_distns,emission_distns,init_dynamics_distns,**kwargs):
+        self.init_dynamics_distns = init_dynamics_distns
+        self.dynamics_distns = dynamics_distns
+        self.emission_distns = emission_distns
+
+        super(_SLDSMixin, self).__init__(
+            obs_distns=self.dynamics_distns, **kwargs)
+
+    def generate(self, T=100, keep=True, **kwargs):
+        s = self._states_class(model=self, T=T, initialize_from_prior=True, **kwargs)
+        s.generate_states()
+        data = self._generate_obs(s)
+        if keep:
+            self.states_list.append(s)
+        return data, s.stateseq
+
+
+class _MultiEmissionSLDSGibbsMixin(object):
+    def resample_parameters(self):
+        self.resample_lds_parameters()
+        self.resample_hmm_parameters()
+
+    def resample_lds_parameters(self):
+        self.resample_init_dynamics_distns()
+        self.resample_dynamics_distns()
+        self.resample_emission_distns()
+
+    def resample_hmm_parameters(self):
+        super(_MultiEmissionSLDSGibbsMixin, self).resample_parameters()
+
+    def resample_init_dynamics_distns(self):
+        for state, d in enumerate(self.init_dynamics_distns):
+            d.resample(
+                [s.gaussian_states[0] for s in self.states_list
+                 if s.stateseq[0] == state])
+        self._clear_caches()
+
+    def resample_dynamics_distns(self):
+        zs = [s.stateseq[:-1] for s in self.states_list]
+        xs = [np.hstack((s.gaussian_states[:-1], s.inputs[:-1]))
+              for s in self.states_list]
+        ys = [s.gaussian_states[1:] for s in self.states_list]
+
+        for state, d in enumerate(self.dynamics_distns):
+            d.resample(
+                [(x[z == state], y[z == state])
+                 for x, y, z in zip(xs, ys, zs)])
+        self._clear_caches()
+
+    def resample_emission_distns(self):
[…]

### 2016-12-01 zimmer/plotting.py

@@ -67,6 +67,41 @@ def plot_2d_continuous_states(ax, x, z, colors,
                 x[cp_start:cp_stop + 1, inds[1]],
                  '-', color=colors[z[cp_start]])
 
+def plot_3d_continuous_states(x, z, colors,
+                              ax=None,
+                              figsize=(2.5,2.5),
+                              inds=(0,1,2),
+                              title=None,
+                              results_dir=".", filename=None,
+                              **kwargs):
+
+    if ax is None:
+        fig = plt.figure(figsize=figsize)
+        ax = fig.add_subplot(111, projection="3d")
+
+    cps = states_to_changepoints(z)
+
+    # Color denotes our inferred latent discrete state
+    for cp_start, cp_stop in zip(cps[:-1], cps[1:]):
+        ax.plot(x[cp_start:cp_stop + 1, inds[0]],
+                x[cp_start:cp_stop + 1, inds[1]],
+                x[cp_start:cp_stop + 1, inds[2]],
+                '-', marker='.', markersize=3,
+                color=colors[z[cp_start]],
+                **kwargs)
+
+    ax.set_xlabel("$x_1$", labelpad=-10)
+    ax.set_ylabel("$x_2$", labelpad=-10)
+    ax.set_zlabel("$x_3$", labelpad=-10)
+    ax.set_xticklabels([])
+    ax.set_yticklabels([])
+    ax.set_zticklabels([])
+
+    if title is not None:
+        ax.set_title(title)
+
+    if filename is not None:
+        plt.savefig(os.path.join(results_dir, filename))
 
 def plot_vector_field(ax, dds, zs, xs, sigma_xs, kk, inds=(0, 1), P=3, P_in=0,
                       title="",
@@ -105,10 +140,13 @@ def plot_vector_field(ax, dds, zs, xs, sigma_xs, kk, inds=(0, 1), P=3, P_in=0,
 
     # Figure out where the dynamics take each point
     A = dds[kk].A[:, :P]
-    if P_in > 0:
-        b = dds[kk].A[:, P:]
-    else:
+    if P_in == 1:
+        b = dds[kk].A[:, P:].T
+    elif P_in == 0:
         b = 0
+    else:
+        raise Exception
+
     d_xyz = xyz.dot(A.T) + b - xyz
 
[…]

### 2016-12-01 zimmer/states.py

@@ -1,0 +1,262 @@
+# Build specific models for joint observations
+import numpy as np
+
+from pybasicbayes.distributions import DiagonalRegression
+
+from pyhsmm.internals.hmm_states import HMMStatesPython, HMMStatesEigen
+from pyslds.states import _SLDSStatesGibbs
+
+from pinkybrain.distributions import _PGLogisticRegressionBase
+
+
+class _MultiEmissionSLDSStates(_SLDSStatesGibbs):
+    """
+    Incorporating two separate Gaussian observation streams.
+    We just have to combine the two observations and compute
+    the effective information form potential to give to the
+    latent state sampler.
+
+    As long as both the observations are linear and Gaussian,
+    this should just be a matter of adding potentials.
+
+    The major concern is: how do we weight the likelihoods so
+    that the two streams are treated as equally important? If
+    the behavioral video stream is 2000 pixels and the neural
+    signal is 2 dimensional, then unless the precision is 1000
+    times larger for the neural signal, the model is only
+    incentivized to capture the behavioral video.
+
+    NOTES:
+
+        1. We assume the model emission_distns list is a set
+           of distributions for each observation modality.
+           These are shared by all discrete latent states.
+
+        2. We need to handle missing data here too. Does each
+           modality have its own mask?
+
+        3. We can either take in the data as a hstack'ed array
+           or as a list of arrays. I'm partial to the latter
+           since it is more transparent; however, it also
+           breaks with the pySLDS convention.
+
+    """
+
+    def __init__(self, model, data=None, T=None, inputs=None, mask=None, group=None, fixed_stateseq=None, **kwargs):
+        # TODO: Fix up this initialization code.
+        # TODO: It is way too complicated.
+
+        self.model = model
+        self.fixed_stateseq = fixed_stateseq
+
+        if data is None:
+            assert isinstance(T, int)
+            self.T = T
+            self.data = data
+        else:
+            if isinstance(data, list):
+                self.data = data
+            elif isinstance(data, np.ndarray) and len(self.emission_distns) == 1:
+                self.data = [data]
+            else:
+                raise Exception("data must be a list, one array "
+                                "for each observation modality")
+
[…]

### 2017-11-09 drafts/2017_11_06/zimmer.tex
tons of updates

% \documentclass{article}

% \usepackage{natbib}

% \usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
% \usepackage{amsfonts}       % blackboard math symbols
% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
% \usepackage{microtype}      % microtypography

% \usepackage{amsthm,amsmath,amssymb}
% \usepackage{macros}
% \usepackage{subcaption}
% \usepackage[textfont=small, labelfont=small]{caption}
% \usepackage{graphicx}
% \DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}

% \usepackage{algorithm}
% \usepackage{algorithmic}

% \usepackage[color=yellow]{todonotes}
% \usepackage{booktabs}
% \usepackage[inline]{enumitem}
% \usepackage{verbatim}

% \usepackage[left=1.5in, right=1.5in, top=1.25in, bottom=1.25in]{geometry}

% \usepackage{setspace}

\documentclass{article}

\input{preamble/preamble.tex}
\input{preamble/preamble_math.tex}
\input{preamble/preamble_acronyms.tex}

\newcommand{\celegans}{\textit{C. elegans}}

\title{Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}

\author{Scott W. Linderman,
  David M. Blei,
  and
  Liam Paninski
  \\
  Columbia University
}

\begin{document}

%\singlespacing
\maketitle

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving \celegans.  The dynamics of
  neural activity shed light on how \celegans~processes sensory
  information and generates motor activity.  To understand these
  dynamics, we leverage \emph{recurrent switching linear dynamical
    systems}~\citep{linderman2017recurrent}---probabilistic models
  that decompose complex time-series into segments with simple, linear
  dynamics. We incorporate these models into a robust, hierarchical
  framework for combining information across whole-brain recordings of
  multiple worms.  We use this framework to reveal latent states of
  population neural activity, along with the discrete modes of
  behavior that drive dynamics in this latent state space.  We find
  characeristic transition patterns between these discrete modes, and
  we see that transition probabilities are determined, in part, by the
  current brain state.  This probabilistic framework is a useful aid
  for neural identification, currently a laborious, manual task.
  Moreover, we find clusters of neurons that are similarly tuned in
  latent state space, and we show how the different dynamical modes
  activate these clusters.  Finally, we find a significant overlap
  between our inferred modes and the manually-labeled states
  of~\citet{kato2015global}, which were shown to correspond to
  different worm behaviors. Our framework automatically discovers
  these behaviorally-meaningful regimes directly from neural activity.
\end{abstract}

\section{Introduction}
% The world as it was:
The nematode \celegans~is a uniquely accessible model organism of
neuroscience. With recent advances in optical recording
technologies~\citep{schrodel2013brain, prevedel2014simultaneous,
  nguyen2016whole}, we now possess a mounting set of tools for
measuring the activity of large fractions of these neurons
simultaneously. These technologies provide exciting opportunities to
probe neural dynamics and their link to sensory
processing and behavior.

Recently, \citet{kato2015global} harnessed these methodological
advances to study the coordinated activity of hundreds of neurons
in head-fixed \celegans. Across multiple organisms, they found that
neural activity reliably traced out smooth trajectories in the subspace
of the first three principal components.
% TODO: Figure showing these trajectories
Upon closer evaluation, they found that different components of these
trajectories corresponded to different elements of behavior, like
forward and reverse crawling, dorsal and ventral turns, etc. These
findings raise a number of interesting questions: can we learn a
(potentially nonlinear) dynamical system that approximates these
low-dimensional dynamics; is there a more appropriate subspace, or
collection of subspaces, for capturing these dynamics; can we gain
statistical power by partial, noisy recordings across separate worms
and trials; and, can we segment these trajectories in an unsupervised
manner to glean further insight into the relationship between
low-dimensional state and behavior?

% Then one day:
We develop a probabilistic framework for investigating these
questions.  We model these multi-neuronal recordings as a
high-dimensional projection of a low-dimensional, dynamical latent
state.  We allow the dynamics of this latent state change over time,
intuitively capturing different patterns of brain activity for
different types of behavior.  The \emph{switching linear dynamical
  system} (SLDS) \citep{chang1978state, ackerson1970state,
  hamilton1990analysis, ghahramani1996switching, murphy1998switching}
captures both of these properties: low-dimensional continuous latent
states with different dynamical regimes.  The SLDS is a probabilistic
time series model characterized by co-evolving discrete and continuous
latent states.  We extend this class of models with a hierarchical
Bayesian framework that captures many of the nuances of these
whole-brain recordings. In fitting these models to recordings from
multiple \celegans, we characterize the globally nonlinear dynamics
that govern neural activity, and we parse these recordings into an
interpretable sequence of behavioral segments.

% Raising the stakes:
% Moment of truth:
% World as it is now:
% The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
% we introduce the class of switching linear dynamical systems and provide
% some insight into the latent states and model parameters. Section~\ref{sec:hslds}
% develops a hierarchical extension of the SLDS to share statistical
% strength by combining recordings from multiple organisms with partially
% overlapping sets of observed neurons.  Then, Section~\ref{sec:results}
% presents results of applying these models to a set of recordings from
% five separate \celegans. We conclude with directions for further research.

\section{Data}
\label{sec:data}

The data comes in the form of a collection of matrices~$\{\widetilde{Y}^{(w)}\}_{w=1}^W$
for each of~$W$ worms.  The matrix~${\widetilde{Y}^{(w)} \in \reals^{T_w \times N_w}}$
representes the calcium activity of the~$w$-th worm, which consists of~$T_w$
time frames and~$N_w$ neurons.\footnote{Technically, this calcium activity is
a matrix of first-order temporal differences of a
bleaching-corrected~$\Delta F /F$ signal. Unlike~\citet{kato2015global}, we
do not smooth these derivates. We work with the raw first-order
temporal differences.}  \celegans~is special in that each neuron has a unique
name.  Assigning names to neurons in calcium imaging data is a challenging
task, but typically, out of the~${\sim 100}$ neurons observed in any worm,
we can label about~$20-30$ with high certainty.  As such, we choose to
represent each matrix~$\widetilde{Y}^{(w)}$ in \emph{canonical form} as a
tuple~$(Y^{(w)}, M^{(w)})$, where~${Y^{(w)} \in \reals^{T_w \times N}}$ is
a matrix where each column corresponds to one of the~$N$ neurons that is
identified in at least one of the~$W$ worms, and~${M^{(w)} \in \{0,1\}^{T_w \times N}}$
is a corresponding \emph{mask} matrix that indicates which neurons were
observed in worm~$w$.  If~$M_{:,n}^{(w)} = 1$, neuron~$n$ was labeled in
worm~$w$ and~$Y_{:,n}^{(w)}$ was its observed activity.  If~$M_{:,n}^{(w)}=0$,
this neuron was not observed in worm~$w$, and the corresponding activity
is missing.\footnote{The mask is defined in such a way that neurons can
  be missing for only subsets of time frames, but in our data neurons are
  either labeled or missing.}
% todo: note that we could also (and have also) model the activity of
% unlabeled neurons.  For the final analyses, I'm not doing this though.

\section{Hierarchical Generative Model}

We model the neural activity with a switching linear
dynamical system (SLDS) with three new extensions: (i)~a \emph{hierarchical} model
to share parameters across worms while also allowing for worm-to-worm
variability; (ii)~a \emph{robust} model for dynamics noise to help address
model misspecification; and (iii)~a \emph{recurrent} model to capture
how continuous latent states influence discrete state transition probabilities.
We will introduce the SLDS first and then present each of these extensions
in turn.

\subsection{Switching linear dynamical systems}
\label{sec:slds}
Switching linear dynamical system models (SLDS) break down complex, nonlinear
time series data into sequences of simpler, reused dynamical modes.
By fitting an SLDS to data, we not only learn a flexible nonlinear generative
model, but also learn to parse data sequences into coherent discrete units.

The generative model is as follows. At each time ${t=1,2,\ldots,T}$,
for each worm~$w$,
there is a discrete latent state ${z_t^{(w)} \in \{1,2, \ldots,K\}}$ that
follows Markovian dynamics,
\begin{equation}
  z_{t+1}^{(w)} \given z_t^{(w)}, \{\pi_k\}_{k=1}^K  \sim \pi_{z_t^{(w)}}
  \label{eq:markov_z}
\end{equation}
where ${\{\pi_k\}_{k=1}^K}$ is the Markov transition matrix and
${\pi_k \in [0,1]^K}$ is its $k$th row. 
In addition, a continuous latent state ${x_t^{(w)} \in \reals^D}$ follows
conditionally linear (or affine) dynamics, where the discrete state $z_t^{(w)}$
determines the linear dynamical system used at time $t$:
\begin{align}
  x_{t+1}^{(w)} &= A_{z_{t+1}^{(w)}} x_{t}^{(w)} + b_{z_{t+1}^{(w)}} +  u_t^{(w)},
  &
  u_t^{(w)} &\iid\sim \cN(0,Q_{z_{t+1}^{(w)}}),
  \label{eq:slds_start}
\end{align}
for matrices ${A_k, Q_k \in \reals^{D \times D}}$ and vectors~${b_k\in
\reals^D}$ for~${k=1,2,\ldots,K}$.
Finally, at each time $t$ a linear Gaussian observation $y_t^{(w)} \in \reals^N$
(some entries of~$y_t^{(w)}$ are masked off; recall Section~\ref{sec:data}) is
generated from the corresponding latent continuous state,
\begin{align}
  y_t^{(w)} &= C x_t^{(w)} + d + v_t^{(w)}, & v_t^{(w)} &\iid\sim \cN(0,S),
    \label{eq:slds_end}
\end{align}
for $C \in \reals^{N \times D}$, $S \in \reals^{N \times N}$,
and~$d \in \reals^N$. We denote the rows of~$C$ by vectors~$c_n$.
For simplicity, we assume~$C$,~$d$, and~$S$ to
be shared among all discrete states in our model.  Moreover, we assume
the observation noise is diagonal,
\begin{align*}
  S &= \diag\left(\left[s_{1}, \ldots, s_{N} \right] \right). 
\end{align*}
The system parameters comprise the discrete Markov transition matrix, the
library of linear dynamical system matrices, and the neuron-specific
emission parameters, which we write as
\begin{equation*}
  \theta = \{(\pi_k, A_k, Q_k, b_k)\}_{k=1}^K \cup \{c_n, d_n, s_n\}_{n=1}^N.
\end{equation*}

To learn an SLDS using Bayesian inference, we place conjugate Dirichlet priors
on each row of the transition matrix and conjugate matrix normal
inverse Wishart (MNIW) priors on the linear dynamical system parameters,
Gaussian priors on the rows of the emission matrix,
and inverse gamma priors on the emission noise.
We write this as,
\begin{align*}
  \pi_k &\given \alpha \iid\sim \distDirichlet(\alpha),
  &
  [A_k, b_k], Q_k &\given \lambda \iid\sim \distMNIW(\lambda),
  \\
  [c_n, d_n] &\given \eta \iid\sim \distNormal(\eta),
  &
  s_{n} &\given \eta \iid\sim \mathrm{IG}(\eta).
\end{align*}
where~$[\cdot, \cdot]$ denotes column concatenation, and $\alpha$, $\lambda$,
and~$\eta$ denote appropriate hyperparameters of the transitions, dynamics, and
emissions, respectively.

\begin{figure}[t]
\centering%
\includegraphics[width=\textwidth]{figures/hslds.png} 
\caption{A schematic of the hierarchical SLDS for combining information
  across multiple worms in order to learn a shared dynamical system
  for \celegans.  We have a set of shared set of model parameters,~$\theta$,
  represented by the large node at the top of the graphical model. Each
  worm,~$w$ has its own set of discrete latent states (colored nodes)
  and continuous latent states (nodes with colored outlines) and observed neurons.
  For example, neuron~1 may appear in all neurons whereas neuron~2
  may only be seen in worm~2. Thus, each worm provides information
  about the mapping from latent states to observations for only a
  subset of neurons. In terms of the model, this corresponds to a
  subset of rows in the matrix~$C$ and the vector~$d$. Finally,
  we allow each worm to have local parameters that capture, for example,
  the amount of fluorescent protein expressed in a particular neuron
  for a given worm. 
}
\label{fig:hslds}
\end{figure}

\subsection{Extensions of the standard SLDS}
\label{sec:extensions}

To better model the \celegans~data, we introduce three extensions
to the standard SLDS.

\paragraph{Hierarchical model to allow limited variability across worms.}
The first extension captures shared patterns across worms while
still allowing for worm-to-worm variability.  We do this in
two ways. First, we allow the dynamics parameters to vary
slightly from worm to worm with the following model,
\begin{align*}
  \mathrm{vec}([A_k, b_k]) &\sim \distNormal(\lambda), \\
  \mathrm{vec}([A_k^{(w)}, b_k^{(w)}]) &\sim \distNormal(\mathrm{vec}([A_k, b_k]), \sigma^2 I), \\
  Q_k^{(w)} &\sim \mathrm{IW}(\lambda).
\end{align*}
This allows worm-specific dynamics parameters~$(A_k^{(w)}, b_k^{(w)})$ but
requires that they not deviate too far from the global mean~$(A_k, b_k)$.
It also allows for worm-specific dynamics noise.

Second, we allow each worm to have neuron-specific emission variances,
\begin{align*}
  s_{n}^{(w)} & \iid\sim \mathrm{IG}(\eta).
\end{align*}
We may expect, for example, that calcium indicator expression
varies from worm to worm, which in turn could give rise to
different levels of observed activity.  While this could be
modeled as a scaling of the emission matrix, a simpler approach
is to let the noise term capture these different fluctuations.
We could share the global mean of the worm-specific variances,
but we have little reason to expect the~$s_{n}^{(w)}$ and~$s_{n}^{(w')}$
to be correlated under the prior.

Figure~\ref{fig:hslds} illustrates the hierarchical model.  Each
worm has a set of discrete states~$z_{1:T}$, continuous states~$x_{1:T}$,
and observations for a subset of neurons.  The worms share a set of
global dynamics parameters and emission matrices, but they also
have worm-specific perturbations of the global dynamics, as well as
worm-specific observation variances.

\paragraph{Robust dynamics model.}  While the SLDS can capture nonlinear
dynamics by composing simple linear pieces, it is still an approximation
to the true data-generating process.  To account for some of this
model misspecification, we introduce a \emph{robust} extension of the
SLDS by allowing for heavy-tailed noise in the dynamics.  Specifically,
we replace~\eqref{eq:slds_start} with,
\begin{align}
  x_{t+1}^{(w)} &= A_{z_{t+1}^{(w)}} x_{t}^{(w)} + b_{z_{t+1}^{(w)}} +  u_t^{(w)},
  &
  u_t^{(w)} &\iid\sim \mathrm{t}(0, Q_{z_{t+1}^{(w)}}, \nu_{z_{t+1}^{(w)}}),
\end{align}
where~$\mathrm{t}(\cdot, \cdot, \cdot)$ denotes the multivariate-t
distribution.  This distribution has heavier tails than its Gaussian
counterpart; indeed, it can be derived by a mixture of multivariate
Gaussians with a~$\chi^2$-distributed scaling of the covariance matrix.

\paragraph{Recurrent model of discrete state transition probabilities.}
Finally, we consider a third extension that allows for more complex
transition probabilities.  We replace~\eqref{eq:markov_z} with,
\begin{align}
  z_{t+1}^{(w)} \given z_t^{(w)}, x_t^{(w)}
  &\sim \mathrm{softmax}(r_{z_t^{(w)}} + R x_t^{(w)}),
    \label{eq:recurrent}
\end{align}
where~${r_k \in \reals^K}$ and~${R \in \reals^{K \times D}}$
parameterize a map from previous discrete and continuous states to a
distribution over next discrete states.  The link function is the
softmax function, which exponentiates and noramlizes
the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.

\section{Model Fitting}

We have proposed a variety of Bayesian algorithms for learning the
model parameters~$\theta$ and performing joint inference of the
latent states~$z$ and~$x$~\citep{linderman2017recurrent, linderman2017structure},
but for simplicity, here we treat inference of~$x$ and~$z$ separately.
That is, first we find low-dimensional continuous states to model the
observations~$y$, then we find a set of discrete states~$z$ that best
explain the dynamics in~$x$. We learn the corresponding parameters
at both stages.

\todo[inline, size=\normalsize]{We will present full details of these inference
procedures in the appendix, but for now, here are a few highlights.}

To be more specific, first we fit linear dynamical systems to infer
the continuous latent states~$x$ given the observed values of~$y$.
Since many entries of~$y$ are masked off, this requires inference
with missing data.  Under the assumption of diagonal observation
covariance, this inference is straightforward.  We will compare the
standard LDS with the hierarchical LDS, which has worm- and neuron-specific
variances.

Once we have inferred the continuous latent states, we will fit
an autoregressive hidden Markov model (AR-HMM) to~$x$ in order
to infer~$z$ and learn the dynamics parameters~$(A_k, b_k, Q_k)$.
Here, again, we will consider all possible variants of hierarchical,
robust, and recurrent models.

Both fitting procedures use Markov chain Monte Carlo, specifically
block Gibbs sampling, to sample from the conditional distribution
of latent states given parameters and of parameters given latent
states.  We run these algorithms for 1000 iterations, at which
point we find them to have converged on the basis of held-out
log likelihood.

\section{Results}
\label{sec:results}

\todo[inline, size=\normalsize]{For this draft, I will present results one at a time,
  each on a separate page.  Once we have settled on the story, we
  can add more exposition.}
\clearpage

\subsection{Neural activity is low-dimensional and has worm-specific variance}

\begin{figure}[h]
  \centering
  \includegraphics[width=6in]{figures/lds/dimensionality} \\
  \includegraphics[width=5.5in]{figures/lds/observation_variance} 
  \caption{\textit{(top)} Training and test likelihoods for the linear
    dynamical system and its hierarchical extension as a function of
    the dimensionality of the latent state,~$D$ (units:
    nats/observation over Gaussian baseline). On the basis of held-out
    test likelihood, we choose the hierarchical LDS with
    10-dimensional latent states.  \textit{(bottom)} Inferred standard
    deviaion~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
    indicate neurons that were not observed in that worm.}
  \label{fig:lds}
\end{figure}

We fit linear dynamical systems (LDS) and their hierarchical
extensions with worm-specific variance (see Section~\ref{sec:extensions})
for a range of latent dimensions,~$D$.  We found that the
10-dimensional hierarchical LDS yielded the highest test
likelihood (holding out the final 20\% of time frames for
each worm for testing, assuming stationarity). This is
shown in Figure~\ref{fig:lds}(top).

The hierarchical model has worm-specific observation variances.
These are shown in Figure~\ref{fig:lds}(bottom).  For neurons that
were not observed in a given worm, the corresponding cells are
colored gray.

\clearpage

\subsection{The hierarchical LDS smooths the data and predicts unobserved activity}

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_0}
  \caption{The observed activity~$y$ (i.e. the first-order differences
    of the calcium signal) are shown in black; the smoothed activity
    with the hierarchical LDS is shown in blue.  The hierarchical LDS
    makes predictions about unobserved neurons---those with dotted
    black lines---based on the correlations found in other worms. We
    show only one minute of data and the first 20 neurons
    (alphabetically) here.}
  \label{fig:activity_1}
\end{figure}

Figures~\ref{fig:activity_1}-\ref{fig:activity_5} show the observed
activity~$y$ along with the smoothed activity under the hierarchical
LDS.  Formally, the blue lines are given by~$C \E[x \given y] +
d$. Unobserved neurons are shown as dotted black lines.  Though their
activity is not observed, the hierarchical model can predict it based
on correlations observed in other neurons. We note that some of the
highly active neurons (e.g. \textsf{AVFL} and \textsf{AVFR}) are
largely smoothed out, as the model attributes much of their activity
to noise.

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_1}
  \caption{See caption of Figure~\ref{fig:activity_1}}
  \label{fig:activity_2}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_2}
  \caption{See caption of Figure~\ref{fig:activity_1}}
  \label{fig:activity_3}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_3}
  \caption{See caption of Figure~\ref{fig:activity_1}}
  \label{fig:activity_4}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_4}
  \caption{See caption of Figure~\ref{fig:activity_1}}
  \label{fig:activity_5}
\end{figure}

\clearpage

\subsection{The emission parameters aid in neural identification}

One of the challenges in modeling this data is that most neurons
cannot be labeled with certainty.  In other words, we can detect a
neuron in the calcium imaging data, but we cannot determine its
identity (\textsf{AVAL}, \textsf{AVAR}, etc.).  The emission
parameters of the SLDS are a useful aid for determining these
identities, as we show here.

Across the dataset of five worms, 61 unique neurons were identified in
at least one worm.  In any given worm, we also observe many other
neurons that we cannot accurately label. Our goal here is to assign
labels to the unlabeled neurons.  Of course, if a label, say \textsf{AVAL}, has
already been identified in worm 1, then we cannot assign an unlabeled
neuron to \textsf{AVAL}.  So the problem is to find the most likely matching
of unlabeled neurons to remaining, available labels.  Since there are
more unlabeled neurons than possible labels, our result will only be
a partial matching.

The rows of the emission matrix~$c_n$ serve as ``embeddings'' of the
neurons. These embeddings can aid in this matching.  Specifically,
given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabled
neuron~$n^*$ in worm~$w$, along with the set of continuous latent
states~$x^{(w)} \in \reals^{T \times D}$ inferred from the labeled
neurons, we perform a linear regression to
estimate~$c_{n^*} \in \reals^D$, the ``embedding'' of this unlabeled
neuron.  We then compare this vector~$c_{n^*}$ to the embeddings of
the labeled neurons,~$\{c_n\}_{n=1}^N$, which were learned with the
SLDS.  We use the cosine similarity between two vectors as a measure
of likeness.

\begin{figure}[h]
\centering%
\includegraphics[width=3in]{figures/lds/similarity_comparison} 
\caption{The emission parameters aid in neural identification.  In
  this task, we view unlabeled neurons as \emph{candidates} for a
  particular label. For example, suppose we are trying to determine
  the candidate neuron that is most likely to be~\textsf{AVAL}. For
  this task, we have artifically held-out this label so that we know
  which candidates truly is~\textsf{AVAL}. We use the neural
  embeddings~$c_{\mathsf{AVAL}}$, which was inferred from other worms,
  to aid in this matching.  We measure the cosine similarity
  between~$c_{\mathsf{AVAL}}$ and the coefficients of linear
  regressions between~$x$ and the activity of the unlabeled neurons'
  activity. This histogram shows the results of performing this
  experiment with 10 held-out identities for each of the five worms.
  We see that the correct candidates frequently have much higher
  cosine similarity than incorrect candidates.  This means that we can
  use this similarity measure to infer neuron identities.  }
\label{fig:neuron_id}
\end{figure}

We demonstrate this identification in a synthetic task where we
artificially withheld the labels of ten neurons per worm.  Suppose one
of the held-out neurons in worm 1 was known to be \textsf{AVAL}.  We
iterate through all the unlabeled neurons in worm
1---\emph{candidates} for the label~\textsf{AVAL}---and measure the
cosine similarity between their embeddings~$c_{n^*}$ and the
embedding~$c_{\mathsf{AVAL}}$, which was inferred from worms 2-5.
Figure~\ref{fig:neuron_id} shows the results of this task.  The
embeddings of the held-out neurons were much more similar to the
embedding of their true identity than were the other unlabeled
neurons.

To further quantify these results, we sort these similarities
and find the index of the true held-out neuron in the sorted list.
If the held-out AVAL neuron was the most similar to~$c_{\mathsf{AVAL}}$,
its rank would be 1; if it was the least similar, its rank would
be~$N_{\mathsf{unlabeled}}$, the number of unlabeled neurons for
this worm.  Table~\ref{tab:neuron_id} shows the ranks of the
similarities of held-out neurons to their true identities, compared
to all other unlabeled neurons.  If the held-out
neurons were always the most similar to their true embeddings, the
matching columns would all be 1's, and we would easily be able to
identify neurons with this information.  We see that in most cases
the held-out neurons are within the top three most similar candidates,
demonstrating that embeddings are useful for identification.

Finally, we perform an even harder task. Using the Hungarian algorithm,
we find the partial matching that maximizes the sum of similarities.
We compute the accuracy of this matching by counting the fraction of held-out
neurons that were assigned to their true label. (Since the Hungarian
algorithm is assigning all the neurons at once, in general the number
of correctly assigned neurons is not equal to the sum of 1's in the
rank columns.) Still, we see that in this challenging task we assign
many neurons to their true identities.

\begin{table}[h]
  \caption{\textit{Artificial neuron identification task results.}  In
    each worm we have~$N_{\mathsf{labeled}}$ labeled neurons
    and~$N_{\mathsf{unlabeled}}$ unlabeled neurons. Of the unlabeled
    neurons, 10 had known labels that were held out for this task.  We
    compare the cosine similarity of the unlabeled neuron embeddings
    to the embeddings of the labeled neurons, sort the similarities,
    and compute the rank of each held-out neuron to its true label
    (most similar~=~1; least similar~=~$N_{\mathsf{unlabeled}}$).  We
    use the Hungarian algorithm to find the most likely matching of
    unlabeled neurons to possible labels, and we report the fraction
    of the 10 held-out neurons that were correctly matched~(``Acc.'').
  }
  \vspace{1em}
  \label{tab:neuron_id}
  \centering
  \begin{tabular}{cccccccccccccc}
    \toprule
    & $N_{\mathsf{labeled}}$ & $N_{\mathsf{unlabeled}}$ & \multicolumn{10}{c}{Matching Rank (10 Held-out Neurons)} & Acc. \\
    \cmidrule(r){2-2}
    \cmidrule(rl){3-3}
    \cmidrule(rl){4-13}
    \cmidrule(l){14-14}
    % \midrule
    Worm 1 & 14 & 95 & 10 & 25 & 1 & 1 & 66 & 2 & 1 & 1 & 2 & 1 & 0.30 \\
    Worm 2 & 31 & 76 & 3 & 1 & 1 & 1 & 1 & 39 & 1 & 1 & 1 & 4 & 0.50 \\
    Worm 3 & 20 & 111 & 5 & 53 & 3 & 22 & 1 & 2 & 1 & 1 & 1 & 1 & 0.50 \\
    Worm 4 &33 & 93 & 2 & 1 & 26 & 2 & 1 & 56 & 1 & 1 & 1 & 1 & 0.30 \\
    Worm 5 & 38 & 91 & 1 & 1 & 1 & 1 & 1 & 2 & 1 & 26 & 1 & 67 & 0.70 \\
    \bottomrule
  \end{tabular}
\end{table}

\clearpage

\subsection{Neural embeddings reveal clusters of similarly tuned neurons}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/lds/permuted_similarity}
\vspace{-3em}
\caption{Cosine similarity between embedding vectors~$c_n$
  suggest clusters of similarly tuned neurons. }
\label{fig:clusters}
\end{figure}

The rows of the emission matrix~$c_n$ (i.e. the embedding vectors)
also suggest that neurons cluster into groups of similarly tuned
cells.  Figure~\ref{fig:clusters} shows the~${N \times N}$ matrix of
cosine similarity for each pair of embedding vectors.  We used k-means
to find clusters of embedding vectors and have permuted the
similarity matrix to visualize these groups.  We see that many
clusters contain lateral pairs (e.g. {\sf AVAL, AVAR} in
the first cluster). 
We will use these clusters to better understand the dynamics of
the SLDS.

Note that this clustering is not so simple without a hierarchical model.
Since we never see all 61 neurons in a single worm, we need some means
of combining information across worms in order to cluster their activity.
The hierarchical LDS does exactly this.  We could consider simpler
hierarchical models, e.g. a hierarchical factor analysis model that does
not include assumptions about temporal dynamics; such a model would
probably yield similar clusters.  The key point is that we must combine
information across worms.

\clearpage

\subsection{Latent dynamics are hierarchical, robust, and recurrent}

\begin{figure}[h]
\centering%
\includegraphics[width=\textwidth]{figures/arhmm/dimensionality.pdf} 
\caption{Training and test likelihood of the hierarchical
  autoregressive hidden Markov models (a special case of the SLDS) for
  all combinations of hierarchical, robust, and recurrent extensions.
  We evaluate test likelihoods for a range of numbers of latent
  states, and we find that the hierarchical, robust, recurrent model
  with~$K=8$ latent states performs best.}
\label{fig:arhmm_lls}
\end{figure}

As with the LDS, we fit autoregressive hidden Markov models (AR-HMMs)
with all combinations of hierarchical, robust, and recurrent extensions
described in Section~\ref{sec:extensions}.  For each possible model, we
swept from~$K=2$ to~$K=20$ discrete latent states in increments of two.

Overall, the hierarchical models (red, blue, yellow, green) performed
best across all~$K$ in training data.  This is not surprising as the
hierarchical models effectively have~$W \cdot K$ dynamics
parameters---one version of each state for each worm---whereas the
non-hierarchical models have only~$K$.  Of these hierarchical models,
only those with the robust extension showed good generalization.  The
non-robust models (with standard Gaussian noise; green, yellow)
severely overfit the training data.  Of the robust hierarchical models,
the recurrent model showed substantial improvement over the non-recurrent
model, and obtained peak performance with~$K=8$ states.  Likelihoods
are measured in units of nats/time step.

\clearpage

\subsection{The AR-HMM automatically segments low-dimensional states}

\begin{figure}[h]
\centering%
\includegraphics[width=2.7in]{figures/arhmm/x_3d_1.pdf}
\includegraphics[width=2.7in]{figures/arhmm/x_3d_2.pdf}
\\
\includegraphics[width=2.7in]{figures/arhmm/x_3d_3.pdf}
\includegraphics[width=2.7in]{figures/arhmm/x_3d_4.pdf}
\\
\includegraphics[width=2.7in]{figures/arhmm/x_3d_5.pdf} 
\caption{Three dimensional projections of the 10 dimensional
  latent states~$x$. The trajectories are color coded
  according to the inferred discrete states~$z$.  We find
  that discrete states correspond to different ``loops''
  in latent space, and that they are, to large extent,
  used in separate regions of~$x$-space.  This is consistent
  with the recurrent model.}
\end{figure}

Since the continuous latent states are 10 dimensional, we use only the
three dimensions of activity for visualization.  We choose the three
dimensions that are most correlated with the observed neural activity.
We color code these three-dimensional trajectories according to the
discrete states~$z$, which correspond to different linear dynamical
regimes of the continuous states~$x$. Compare these plots to the state
space plots of~\citet{kato2015global}. These trajectories are not as
smooth (which is expected given that we are modeling the first-order
differences directly), but they capture the same qualitative patterns.
We will quantify this correspondence shortly.

\clearpage

\subsection{The AR-HMM learns dynamical systems for each discrete state}

\begin{figure}[h]
\centering%
\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_0.pdf}
\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_1.pdf}
\\
\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_2.pdf}
\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_3.pdf}
\caption{Each discrete state has a corresponding set of linear dynamics
  parameters~$(A_k, b_k)$, which we visualize here as a vector field.
  Each arrow shows where a point~$x_t$ would be mapped to by the 
  linear dynamics of a given state. We visualize points~$x_t$ where
  the states were actually used. We see a number of states related
  to ``returning to the origin'' (e.g. states 1, 2, 3, 5), as well
  as states corresponding to loops (e.g. states 4, 6, 7).}
\label{fig:dynamics}
\end{figure}

\begin{figure}[h]
  \centering%
  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_4.pdf}
  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_5.pdf}
  \\
  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_6.pdf}
  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_7.pdf}
  \caption{See caption for Figure~\ref{fig:dynamics}}
  \label{fig:dynamics2}
\end{figure}
Figures~\ref{fig:dynamics} and~\ref{fig:dynamics2} illustrate the
global dynamics parameters for each of the~$K$ discrete states.
Specifically, we show where a point~$x_t$ would be mapped to by the
operation~$A_k x_t + b_k$.  We choose a subset of starting
points~$x_t$ where the discrete state was used. Many of the states
correspond to returning to the origin, which in turn corresponds to
constant neural activity (since we are modeling the first-order
differences in the calcium signal).  Other states correspond to loops,
which give rise to increases in the activity of some clusters of
neurons and decreases in others.

\clearpage

\subsection{Inferred states correspond to activation of clusters of neurons}

\begin{figure}[h]
\centering%
\includegraphics[width=5in]{figures/arhmm/cluster_activity_0.pdf}
\includegraphics[width=5in]{figures/arhmm/cluster_activity_1.pdf}
\caption{(top) Colored traces show a number of real-data trajectories
  that were inferred to be in states 1 (blue) and 2 (red).  Then we
  simulate from our model's inferred linear dynamical system for these
  states, and show simulated trajectories with four different starting
  points (circle, triangle, square, and pentagon) in black lines.
  The top left panel shows the top 3 dimensions of these trajectories
  in latent state space. The
  top right panels show 2D views of the same trajectories.  (bottom)
  These four simulated trajectories give rise to different patterns of
  neural activity in the five inferred clusters.  For example, the
  first trajectory of the second state (red circle) corresponds to
  activity that decays toward the baseline for each of the clusters,
  with a starting value of below-baseline activity in the first
  cluster.}
\label{fig:cluster_activity_12}
\end{figure}

Figures~\ref{fig:cluster_activity_12}, \ref{fig:cluster_activity_34},
\ref{fig:cluster_activity_56}, and~\ref{fig:cluster_activity_78} show
how trajectories in continuous latent state space for each of the~$K$
states manifest in the activity of the neural clusters identified in
Figure~\ref{fig:clusters}.  The black lines with circle, triangle,
square, and pentagon starting points are forward simulated
trajectories using the inferred dynamics of state~$k$ with four
different initial conditions. The initial conditions are values
of~$x_t$ where we had inferred an entry into state~$k$.
The bottom rows show how the activity the forward
simulated trajectories activate each of the five clusters of
neurons.

The colored lines in the top panel are segments of~$x$ that were
inferred to be in state~$k$.  Recall that these continuous latent
states were found by first fitting a hierarchical LDS. As such, they
correspond to the conditional mean of the posterior
distribution under the hierarchical LDS,~$\E[x \given y]$.

\begin{figure}[h]
\centering%
\includegraphics[width=5in]{figures/arhmm/cluster_activity_2.pdf}
\includegraphics[width=5in]{figures/arhmm/cluster_activity_3.pdf}
\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
\label{fig:cluster_activity_34}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=5in]{figures/arhmm/cluster_activity_4.pdf}
\includegraphics[width=5in]{figures/arhmm/cluster_activity_5.pdf}
\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
\label{fig:cluster_activity_56}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=5in]{figures/arhmm/cluster_activity_6.pdf}
\includegraphics[width=5in]{figures/arhmm/cluster_activity_7.pdf}
\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
\label{fig:cluster_activity_78}
\end{figure}

\clearpage

\subsection{States are employed roughly equally by these worms}

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figures/arhmm/state_usage_per_worm2.pdf}
\caption{Proportion of time frames spent in each state, broken down per-worm.}
\label{fig:usage}
\end{figure}

The worms appear to use these states with roughly equal proportion,
as shown in Figure~\ref{fig:usage}.  The states have been ordered to
overall population usage, but within worms we see that some states
are used slightly more often than others.  For example, worm 3 uses
state 1 less often than the others, and instead seems to have a higher
usage of state 3.  This particular difference is probably not important, but it
points to how these types of methods could be useful for distinguishing
between, for example, different genetic variants of worms.

\clearpage

\subsection{Transition structure is largely stereotyped across worms as well}

\begin{figure}[h]
\centering
\includegraphics[width=5.5in]{figures/arhmm/trans_matrices.pdf}
\caption{Empirical transition probabilities for each of the worms.  By far
  the most common transition is into the same state.  For visualization, we
  have removed this diagonal component and renormalized so that the~$k$-th row of
  each transition matrix shows the conditional distribution over~$z_{t+1}$ given
  that~$z_{t}=k$ and~$z_{t+1} \neq z_t$. }
\label{fig:trans}
\end{figure}

Figure~\ref{fig:trans} shows the empirical transition matrices for each of the
worms.  We see a few transitions appear across all worms:
\begin{itemize}
\item Pink transitions to red with high probability in all worms. This is evident
  from the state trajectories above.
\item Brown transitions to pink or, with lower probability, to purple.
\item Orange transitions to purple, pink, or blue.
\item Purple and orange transition to blue for most worms, except in worm three,
  which uses yellow in place of blue (see above).
\item Green transitions to brown with high probability across all worms. 
\item For worms 2, 4, and 5, red often transitions into yellow and yellow to green or orange
  for most worms.
\item Blue (the ``origin'' state), tends to transition to purple or orange in
  worms 2, 4, and 5.
\end{itemize}

These aren't meaningful yet, but as we show next, these discrete states closely
match the states identified in~\citep{kato2015global}.

\clearpage

\subsection{Recurrent weights influence transition probabilities}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/recurrent_weights.pdf}
\caption{Weights of the recurrent transition model.  The largest
  weights come from the Markov terms for~$z_{t}$, but the weights
  for the continuous location~$x_t$ are non-trivial.  When we
  compute the effective weights that come from~$r_{z_t}$ and
  the ``recurrent'' terms~$R^\trans x_t$, we see that the recurrent terms
  are a significant fraction of the Markovian terms. }
\label{fig:rec_weights}
\end{figure}

Figure~\ref{fig:rec_weights} shows the two terms that go into the
distribution over next states: the ``Markov'' term~$\{r_{k}\}_{k=1}^K$
and the ``recurrent'' weights~$R$, as defined in
Eq.~\eqref{eq:recurrent}. The largest weights come from the
self-transitions, i.e. the diagonal of the Markov terms. Still, the
recurrent weights are non-negligible. The histogram of Markov
weights~($r_{z_t}$) and the recurrent weights~($R^\trans x_t$) at the
times of transitions (right panel) shows that the recurrent weights
are comparable to the Markov weights, and therefore must play a role
in determining transition probabilities.  This must be the case as the
recurrent models yield significant improvement in held-out log
likelihoods, as shown in Figure~\ref{fig:arhmm_lls}.

\clearpage

\subsection{Some state durations are non-Markovian}

\begin{figure}[h]
\centering%
\includegraphics[width=1.8in]{figures/arhmm/durations_0.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_1.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_2.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_3.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_4.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_5.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_6.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_7.pdf}
\caption{Empirical distribution of state durations (histogram)
  compared to the geometric distribution implied by the Markovian
  weights of the transition model.  Some state durations are
  well-approximated by the geometric distribution, but others
  are peaked away from zero (e.g. states 2, 4, 6, and 7).  The
  recurrent transition model allows the AR-HMM to capture this
  non-Markovianity by modeling transitions as a function of location
  in latent state space~$x_t$.}
\label{fig:durations}
\end{figure}

Figure~\ref{fig:durations} shows histograms of the distribution of inferred
state distributions for each state.  For comparison, we show
the geometric distribution of state durations implied by the
Markov weights as a black line.  Specifically, the distribution
over durations~$d_k$ of state~$k$ under the Markov model (ignoring the
recurrent weights) is given by,
\begin{align*}
  p_{k \to k} &= [\mathrm{softmax}(r_{k})]_k, \\
  d_k &\sim \mathrm{Geom}(1-p_{k \to k}).
\end{align*}
We see that some states (e.g. states 1, 5, and 8) seem to follow the
Markov assumption quite well, whereas others (states 2, 4, and especially 7),
are far from Markovian.  The recurrent model incorporates the continuous
latent state~$x_t$ into the transition probabilities via the additive term~$R x_t$
in the softmax.  It is likely that this term captures the non-Markovianity.

\clearpage

\subsection{Inferred states overlap with those of~\citet{kato2015global}}

\begin{figure}[h]
\centering%
\includegraphics[width=1.9in]{figures/arhmm/overlap_0.pdf}
\includegraphics[width=1.9in]{figures/arhmm/overlap_1.pdf}
\includegraphics[width=1.9in]{figures/arhmm/overlap_2.pdf}
\\
\includegraphics[width=1.9in]{figures/arhmm/overlap_3.pdf}
\includegraphics[width=1.9in]{figures/arhmm/overlap_4.pdf} 
\caption{Overlap between the manually labeled states of~\citet{kato2015global}
  and the inferred states of the AR-HMM. Each row is a normalized probability
  distribution over which of the~$K$ discrete states each worm was inferred to
  be in for the corresponding manual label.}
\label{fig:overlap}
\end{figure}

Figure~\ref{fig:overlap} shows a close correspondence between manually
labeled states and inferred discrete states, indicating that the
switching linear dynamics assumption is a formal specification of the
different dynamical modes reported by~\citet{kato2015global}.
Specifically, different ``behaviors'' correspond to different linear
dynamical regimes in continuous latent space.

A few patterns emerge here:
\begin{itemize}
\item Worm 3 differs from the rest, and seems to have a separate
  matching of inferred states and true states.  This list focuses
  on patterns in Worms 1, 2, 4, and 5. 
\item \textsf{REVSUS} --- the most common manually labeled state ---
  tends to overlap with the red and yellow inferred states.
\item \textsf{SLOW} --- the second most common --- and \textsf{FWD}
  often map onto the blue state.
\item \textsf{VT} (ventral turns) map to the green state, which
  has a rise and fall of activity in cluster 3 (neurons
  {\sf ALA, ASKL, ASKR, RIVL, RIVR, SMDVL} and \textsf{SMDVR}).
  This seems consistent with the roles of sensory neurons~\textsf{ASK*},
  interneurons \textsf{RIV*}, and motor neurons \textsf{SMDV*} in
  local search behavior.
\item Skipping ahead, \textsf{DT} (dorsal turns) often map onto
  the orange and, to a much lesser extent, the purple state. Both
  states encourage upswings in cluster 4 ({\sf SMBDL, SMBDR, SMBVL, SMBVR}, and \textsf{URYVR}).
  The first four of these neurons are motor neurons involved in
  local search behavior. 
\item In addition to the blue state, \textsf{FWD} maps onto the
  green, purple, and orange states as well, though much less significantly.
\item \textsf{REV2} and \textsf{REV1} map primarily onto the pink state.
  This state is characterized by an excursion of activity along cluster
  1 ({\sf AVAL, AVAR, BAGL, RIML, RIMR, SAAVR, SABVL, SABVR, VA01}.
  The \textsf{AVA*} interneurons innvervate motor neurons.
  \textsf{RIM*} are motor neurons that modulate reversals.
  \textsf{SABV*} are interneurons and sublateral motor neurons.
  \textsf{VA01} is a motor neuron.

  We also note that \textsf{REV1} maps onto the brown state, which
  is also associated with ventral turns. Judging by the dynamics plots above,
  it seems ventral turns are often followed by type-1 reversals.
\item \textsf{NOSTATE} seems to often map in different ways across worms.
  This is by far the least common of the manually-labeled states.
\end{itemize}

\todo[inline, size=\normalsize]{Combining these overlap figures
  with the transition matrices above, I find a slightly different
  transition structure from what was reported in Kato et al Figure 4E.
  I find that the transitions between \textsf{REVSUS} and \textsf{FWD}
  are largely mediated by the \textsf{SLOW} state without an intervening
  turn.  I find that ventral turns are often exercised in between
  \textsf{REVSUS} and \textsf{REV1}, and that dorsal turns are often
  used between episodes of forward crawling.  Does this make any sense?
  }

\clearpage

\subsection{Inferred changepoints also match manual labels}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
\caption{(top) Posterior distribution over latent states~$z_t$.
  Each column corresponds to a time bin and sums to one.
  (middle) The posterior probability of changepoints.
  (bottom) For reference,
  the manually labeled states of~\citet{kato2015global}. Colors match those
  of Figure~\ref{fig:overlap}.}
\label{fig:z_cps_12}
\end{figure}

Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
show the posterior distribution of latent states~$z$ inferred from
neural activity alone. From this posterior, we also compute the probability
of a changepoint, where~$z_{t+1} \neq z_t$.  For reference, we show the
manually labeled states of~\citet{kato2015global} below. Here, patterns
like the cascade from states 2 to 3 to 4 (red to yellow to green) are
apparent.

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
\caption{See Figure~\ref{fig:z_cps_12} caption.}
\label{fig:z_cps_34}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
\caption{See Figure~\ref{fig:z_cps_12} caption.}
\label{fig:z_cps_5}
\end{figure}

\clearpage

\subsection{Inferred states changepoints correspond with changes in dynamics}

Figures~\ref{fig:segmentation_1} through~\ref{fig:segmentation_5} show an
interval of latent state trajectories~$x$ along with the segmentation from
the inferred discrete states~$z$ (top).  For comparison, we also show the
manual segmentation of~\citep{kato2015global} (bottom).  The inferred
segmentation seems to identify changepoints when one or more of the latent
continuous states changes dynamics.  Indeed, this is the assumption of the
model.  By contrast, the segmentation of~\citep{kato2015global} is based on
a read-out of a single neuron, and while it is largely consistent with the
inferred segmentation at the population level, it misses some salient
changes in dynamics.  For example, see the inferred changepoint around
time 700 in worm 1.  This changepoint was not identified in the manual
data.

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_0.pdf}
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_0.pdf}
\caption{Latent state trajectories superimposed on the inferred segmentation (top)
  and the manual segmentation (bottom).}
\label{fig:segmentation_1}
\end{figure}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_1.pdf}
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_1.pdf}
\caption{See caption of Figure~\ref{fig:segmentation_1}}
\label{fig:segmentation_2}
\end{figure}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_2.pdf}
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_2.pdf}
\caption{See caption of Figure~\ref{fig:segmentation_1}}
\label{fig:segmentation_3}
\end{figure}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_3.pdf}
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_3.pdf}
\caption{See caption of Figure~\ref{fig:segmentation_1}}
\label{fig:segmentation_4}
\end{figure}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_4.pdf}
\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_4.pdf}
\caption{See caption of Figure~\ref{fig:segmentation_1}}
\label{fig:segmentation_5}
\end{figure}
\clearpage

\clearpage
\bibliography{refs}
\bibliographystyle{abbrvnat}

\end{document}

### 2017-11-09 examples/hierarchical_factor_analysis.py

@@ -1,0 +1,131 @@
+import numpy as np
+np.random.seed(1)
+
+from pybasicbayes.util.text import progprint_xrange
+import matplotlib.pyplot as plt
+from matplotlib.cm import get_cmap
+from pybasicbayes.util.stats import sample_mniw
+
+from zimmer.models import HierarchicalFactorAnalysis
+
+N = 2000
+D_obs = 20
+D_latent = 2
+
+def principal_angle(A,B):
+    """
+    Find the principal angle between two subspaces
+    spanned by columns of A and B
+    """
+    from numpy.linalg import qr, svd
+    qA, _ = qr(A)
+    qB, _ = qr(B)
+    U,S,V = svd(qA.T.dot(qB))
+    return np.arccos(min(S.min(), 1.0))
+
+
+def generate_synth_data(N_groups=5):
+
+    # Create a true model and sample from it
+    mask = np.random.rand(N,D_obs) < 0.9
+    true_model = HierarchicalFactorAnalysis(D_obs, D_latent, N_groups)
+    true_datas = []
+    for group in range(N_groups):
+        true_datas.append(true_model.generate(N=N, mask=mask, group=group, keep=True))
+    return true_model, true_datas
+
+
+def plot_results(lls, angles, Ztrue, Zinf, sigmasq_true, sigmasq_inf):
+    # Plot log probabilities
+    plt.figure()
+    plt.plot(lls)
+    plt.ylabel("Log Likelihood")
+    plt.xlabel("Iteration")
+
+    plt.figure()
+    plt.plot(np.array(angles) / np.pi * 180.)
+    plt.ylabel("Principal Angle")
+    plt.xlabel("Iteration")
+
+    # Plot locations, color by angle
+    N = Ztrue.shape[0]
+    inds_to_plot = np.random.randint(0, N, min(N, 500))
+    th = np.arctan2(Ztrue[:,1], Ztrue[:,0])
+    nperm = np.argsort(np.argsort(th))
+    cmap = get_cmap("jet")
+
+    plt.figure()
+    plt.subplot(121)
+    for n in inds_to_plot:
+        plt.plot(Ztrue[n,0], Ztrue[n,1], 'o', markerfacecolor=cmap(nperm[n] / float(N)), markeredgecolor="none")
+    plt.title("True Embedding")
+    plt.xlim(-4,4)
+    plt.ylim(-4,4)
+
+    plt.subplot(122)
+    for n in inds_to_plot:
[…]

### 2017-11-09 experiments/fit_arhmm.py

@@ -1,0 +1,651 @@
+import os
+import pickle
+
+from functools import partial
+import itertools as it
+
+import numpy as np
+np.random.seed(0)
+
+from tqdm import tqdm
+
+# Plotting stuff
+import matplotlib.pyplot as plt
+from hips.plotting.colormaps import gradient_cmap
+import seaborn as sns
+sns.set_style("white")
+sns.set_context("paper")
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+colors = sns.xkcd_palette(color_names)
+cmap = gradient_cmap(colors)
+
+from matplotlib.cm import get_cmap
+cm = get_cmap("cubehelix")
+zimmer_colors = [cm(i) for i in np.linspace(0.05, 0.95, 8)]
+
+# Modeling stuff
+from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
+
+from autoregressive.models import ARWeakLimitStickyHDPHMM
+from pyslds.util import get_empirical_ar_params
+from pybasicbayes.distributions import AutoRegression, RobustAutoRegression
+from rslds.models import SoftmaxRecurrentARHMM
+from zimmer.models import HierarchicalARWeakLimitStickyHDPHMM, HierarchicalRecurrentARHMM
+from zimmer.dynamics import HierarchicalAutoRegression, HierarchicalRobustAutoRegression
+
+import importlib
+import zimmer.plotting
+importlib.reload(zimmer.plotting)
+from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states, plot_expected_states, \
+    plot_3d_dynamics, plot_2d_dynamics, plot_state_overlap, plot_state_usage_by_worm, plot_all_transition_matrices, \
+    plot_simulated_trajectories, make_state_predictions_3d_movie, plot_simulated_trajectories2, \
+    plot_recurrent_transitions, plot_x_at_changepoints, plot_latent_trajectories_vs_time, \
+    plot_state_usage_by_worm_matrix, plot_duration_histogram
+
+# LDS Results
+lds_dir = os.path.join("results", "2017-11-03-hlds", "run002")
+assert os.path.exists(lds_dir)
+
+# AR-HMM RESULTS
[…]

### 2017-11-09 experiments/fit_hlds.py

@@ -1,0 +1,727 @@
+import os
+import pickle
+
+import numpy as np
+np.random.seed(1234)
+
+from tqdm import tqdm
+from functools import partial
+
+# Plotting stuff
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+from hips.plotting.colormaps import gradient_cmap
+import seaborn as sns
+sns.set_style("white")
+sns.set_context("paper")
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+colors = sns.xkcd_palette(color_names)
+cmap = gradient_cmap(colors)
+
+
+# Modeling stuff
+from pyhsmm.util.general import relabel_by_usage
+
+# Load worm modeling specific stuff
+from zimmer.io import WormData, load_key
+
+from zimmer.models import HierarchicalLDS
+from zimmer.emissions import HierarchicalDiagonalRegression
+from zimmer.plotting import plot_3d_continuous_states
+from pybasicbayes.distributions import Regression, DiagonalRegression
+from pylds.models import MissingDataLDS
+
+# IO
+run_num = 2
+results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
+assert os.path.exists(results_dir)
+fig_dir = os.path.join(results_dir, "figures")
+
+N_worms = 5
+N_clusters = 5
+
+
+def cached(results_name):
+    def _cache(func):
+        def func_wrapper(*args, **kwargs):
+            results_file = os.path.join(results_dir, results_name)
+            if not results_file.endswith(".pkl"):
+                results_file += ".pkl"
+
+            if os.path.exists(results_file):
+                with open(results_file, "rb") as f:
+                    results = pickle.load(f)
+            else:
+                assert os.path.exists(results_dir)
+                results = func(*args, **kwargs)
+                with open(results_file, "wb") as f:
+                    pickle.dump(results, f)
+
+            return results
+        return func_wrapper
+    return _cache
+
+
[…]

### 2017-11-09 experiments/fit_rhslds_final.py

@@ -1,0 +1,1065 @@
+import os, pickle, copy
+
+import numpy as np
+np.random.seed(0)
+
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
+import matplotlib.gridspec as gridspec
+import matplotlib.animation as manimation
+
+from hips.plotting.colormaps import gradient_cmap
+from hips.plotting.sausage import sausage_plot
+from hips.plotting.layout import create_axis_at_location
+
+# Come up with a set of colors
+import seaborn as sns
+sns.set_style("white")
+# sns.set_context("paper")
+
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+
+colors = sns.xkcd_palette(color_names)
+cmap = gradient_cmap(colors)
+
+import importlib
+from sklearn.decomposition import PCA
+
+from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
+from pybasicbayes.util.text import progprint_xrange
+from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
+from pyslds.util import get_empirical_ar_params
+
+import zimmer.io
+importlib.reload(zimmer.io)
+from zimmer.io import WormData, N_worms, find_shared_neurons, load_key
+
+import zimmer.states
+importlib.reload(zimmer.states)
+
+import zimmer.models
+importlib.reload(zimmer.models)
+from zimmer.models import HierarchicalWeakLimitStickyHDPHMMSLDS, HierarchicalRecurrentSLDS
+from zimmer.emissions import HierarchicalDiagonalRegression
+
+import zimmer.plotting
+importlib.reload(zimmer.plotting)
+from zimmer.plotting import plot_1d_continuous_states, plot_3d_continuous_states, plot_vector_field_3d
+
+
+from zimmer.util import states_to_changepoints
+
+# IO
+run_num = 1
+results_dir = os.path.join("results", "08_16_17", "run{:03d}".format(run_num))
+assert os.path.exists(results_dir)
+
+# Hyperparameters
[…]

### 2017-11-09 experiments/nonconj_nascar.py

@@ -341,7 +341,7 @@ def make_rslds_parameters(C_init):
 
 
 # @cached(RESULTS_DIR, "slds")
-def fit_slds(inputs, z_inits, x_inits, ys, masks, groups, C_init,
+def fit_hslds(inputs, z_inits, x_inits, ys, masks, groups, C_init,
               N_iters=1000):
     print("Fitting standard SLDS")
     init_dynamics_distns, dynamics_distns, emission_distns = \
@@ -382,7 +382,7 @@ def fit_slds(inputs, z_inits, x_inits, ys, masks, groups, C_init,
 
 
 # @cached(RESULTS_DIR, "rslds_vbem")
-def fit_rslds_vbem(
+def fit_hrslds_vbem(
         inputs, ys, masks, groups,
         x_inits=None, z_inits=None, C_init=None, initialization="none",
         true_model=None,
@@ -477,7 +477,7 @@ if __name__ == "__main__":
 
     ## Fit a standard SLDS
     slds, slds_lps, slds_z_smpls, slds_x = \
-        fit_slds(inputs, z_inits, x_inits, ys, masks, groups, C_init, N_iters=1000)
+        fit_hslds(inputs, z_inits, x_inits, ys, masks, groups, C_init, N_iters=1000)
 
     ## Fit a recurrent SLDS
     # rslds, rslds_lps, rslds_z_smpls, rslds_x = \
@@ -492,10 +492,10 @@ if __name__ == "__main__":
     #                    initialization="true")
 
     rslds, rslds_lps, rslds_z_smpls, rslds_x = \
-        fit_rslds_vbem(inputs, ys, masks, groups,
-                       z_inits=slds_z_smpls[-1], x_inits=slds_x, C_init=C_init,
-                       initialization="given",
-                       true_model=true_model, N_iters=500)
+        fit_hrslds_vbem(inputs, ys, masks, groups,
+                        z_inits=slds_z_smpls[-1], x_inits=slds_x, C_init=C_init,
+                        initialization="given",
+                        true_model=true_model, N_iters=500)
 
     # rslds, rslds_lps, rslds_z_smpls, rslds_x = \
     #     fit_rslds_vbem(inputs, y, mask,

### 2017-11-09 zimmer/dynamics.py

@@ -1,0 +1,231 @@
+"""
+Hierarchical model for dynamics distributions
+"""
+import numpy as np
+
+from pybasicbayes.distributions import RegressionNonconj, AutoRegression, RobustAutoRegression, GaussianFixedCov
+from pybasicbayes.distributions.regression import _ARMixin
+from pybasicbayes.util.stats import sample_gaussian
+
+
+class NonConjAutoRegression(_ARMixin, RegressionNonconj):
+    pass
+
+
+class _HierarchicalAutoRegressionMixin(object):
+
+    def __init__(self, N_groups, M_0, Sigma_0, nu_0, Q_0, etasq, affine=True, **kwargs):
+        self.N_groups, self.M_0, self.Sigma_0, self.nu_0, self.Q_0, self.etasq, self.affine = \
+            N_groups, M_0, Sigma_0, nu_0, Q_0, etasq, affine
+
+        assert isinstance(N_groups, int) and N_groups >= 1
+
+        assert M_0.ndim == 2
+        self.D_out, self.D_in = M_0.shape
+
+        if np.isscalar(Sigma_0):
+            self.Sigma_0 = Sigma_0 * np.eye(self.D_out * self.D_in)
+        assert Q_0.shape == (self.D_out, self.D_out)
+        assert np.isscalar(etasq) and etasq > 0
+
+        self.A_0 = M_0.copy()
+        self.regressions = []
+        self._initialize_regressions(**kwargs)
+
+    def _initialize_regressions(self, **kwargs):
+        raise NotImplementedError
+
+    def _resample_A(self, groups):
+        raise NotImplementedError
+
+    @property
+    def D(self):
+        return self.regressions[0].D
+
+    @property
+    def nlags(self):
+        return self.regressions[0].nlags
+
+    @property
+    def As(self):
+        return np.array([r.A for r in self.regressions])
+
+    @property
+    def sigmas(self):
+        return np.array([r.sigma for r in self.regressions])
+
+
+    def predict(self, x, group=None):
+        A = self.A_0 if group is None else self.As[group]
+
+        if self.affine:
+            A, b = A[:, :-1], A[:, -1]
+            y = x.dot(A.T) + b.T
+        else:
+            y = x.dot(A.T)
+
+        return y
+
+
+    def rvs(self, x=None, size=1, return_xy=True, group=None):
+        if group is None:
+            A, sigma = self.A_0, self.Q_0
+        else:
[…]

### 2017-11-09 zimmer/emissions.py

@@ -3,7 +3,7 @@ More sophisticated emission models than just simple regressions.
 """
 import numpy as np
 
-from pybasicbayes.distributions import DiagonalRegression
+from pybasicbayes.distributions import DiagonalRegression, Regression
 from pybasicbayes.util.stats import sample_gaussian, sample_invgamma
 
 class HierarchicalDiagonalRegression(DiagonalRegression):
@@ -192,3 +192,4 @@ class HierarchicalDiagonalRegression(DiagonalRegression):
             .dot(np.linalg.cholesky(self.sigma[group]).T)
 
         return np.hstack((x,y)) if return_xy else y
+

### 2017-11-09 zimmer/io.py

@@ -71,6 +71,23 @@ class WormData(object):
         self.dff_bc = interp_data(zimmer_data["wbData"]['deltaFOverF_bc'][0, 0])
         self.dff_deriv = interp_data(zimmer_data["wbData"]['deltaFOverF_deriv'][0, 0])
 
+        # Kato et al smoothed the derivative.  Let's just work with the first differences
+        # of the bleaching corrected and normalized dF/F
+        self.dff_bc_zscored = (self.dff_bc - self.dff_bc.mean(0)) / self.dff_bc.std(0)
+        self.dff_diff = np.vstack((np.zeros((1, self.dff_bc_zscored.shape[1])),
+                                   np.diff(self.dff_bc_zscored, axis=0)))
+
+        # # Let's try our hand at our own smoothing of the derivative
+        # from pylds.models import DefaultLDS
+        # lds = DefaultLDS(D_obs=1, D_latent=1,
+        #                  A=np.eye(1), sigma_states=0.01 * np.eye(1),
+        #                  C=np.eye(1), sigma_obs=np.eye(1))
+        #
+        # self.smoothed_diff = np.zeros_like(self.dff_diff)
+        # self.smoothed_y = np.zeros_like(self.dff_bc_zscored)
+        # for n in range(self.dff.shape[1]):
+        #     self.smoothed_diff[:,n] = lds.smooth(self.dff_diff[:, n:n + 1]).ravel()
+        #     self.smoothed_y[:,n] = self.dff_bc_zscored[0,n] + np.cumsum(self.smoothed_diff[:,n])
 
         # Get the state sequence as labeled in Kato et al
         # Interpolate to get at new time points

### 2017-11-09 zimmer/models.py

@@ -1,8 +1,145 @@
 import numpy as np
 
-from zimmer.states import HierarchicalSLDSStates, HierarchicalRecurrentSLDSStates
+from zimmer.emissions import HierarchicalDiagonalRegression
+from zimmer.dynamics import _HierarchicalAutoRegressionMixin
+from zimmer.states import HierarchicalFactorAnalysisStates, HierarchicalSLDSStates, \
+    HierarchicalRecurrentSLDSStates, HierarchicalLDSStates, HierarchicalARHMMStates, \
+    HierarchicalRecurrentARHMMStates
+from pybasicbayes.models.factor_analysis import FactorAnalysis
+from pyhsmm.models import _HMMGibbsSampling
+from pylds.models import MissingDataLDS
+from autoregressive.models import ARWeakLimitStickyHDPHMM, ARHMM
 from pyslds.models import HMMSLDS, WeakLimitStickyHDPHMMSLDS
-from rslds.models import SoftmaxRecurrentSLDS, SoftmaxRecurrentOnlySLDS
+from rslds.models import SoftmaxRecurrentSLDS, SoftmaxRecurrentOnlySLDS, _RecurrentARHMMMixin
+
+
+class HierarchicalFactorAnalysis(FactorAnalysis):
+    _states_class = HierarchicalFactorAnalysisStates
+
+    def __init__(self, D_obs, D_latent, N_groups,
+                 W=None, sigmasq=None,
+                 sigmasq_W_0=1.0, mu_W_0=0.0,
+                 alpha_0=3.0, beta_0=2.0):
+
+        self.D_obs, self.D_latent, self.N_groups = D_obs, D_latent, N_groups
+
+        self.regression = \
+            HierarchicalDiagonalRegression(
+                self.D_obs, self.D_latent, self.N_groups,
+                mu_0=mu_W_0 * np.ones(self.D_latent),
+                Sigma_0=sigmasq_W_0 * np.eye(self.D_latent),
+                alpha_0=alpha_0, beta_0=beta_0,
+                A=W, sigmasq=sigmasq)
+
+        self.data_list = []
+
+    def generate(self, keep=True, N=1, mask=None, group=0, **kwargs):
+        # Sample from the factor analysis model
+        W, sigmasq = self.W, self.sigmasq[group]
+        Z = np.random.randn(N, self.D_latent)
+        X = np.dot(Z, W.T) + np.sqrt(sigmasq) * np.random.randn(N, self.D_obs)
+
[…]

### 2017-11-09 zimmer/plotting.py

@@ -11,9 +11,43 @@ import numpy as np
 from scipy.stats import multivariate_normal
 
 import matplotlib.pyplot as plt
+import matplotlib.gridspec as gridspec
 from mpl_toolkits.axes_grid1 import make_axes_locatable
+from mpl_toolkits.mplot3d import Axes3D
+import matplotlib.animation as manimation
+
+from hips.plotting.colormaps import gradient_cmap
+from hips.plotting.sausage import sausage_plot
+from hips.plotting.layout import create_axis_at_location
+
+# Come up with a set of colors
+import seaborn as sns
+sns.set_style("white")
+# sns.set_context("paper")
+
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+
+default_colors = sns.xkcd_palette(color_names)
+default_cmap = gradient_cmap(default_colors)
 
 from zimmer.util import states_to_changepoints
+from tqdm import tqdm
+
 
 def plot_1d_continuous_states(x_inf, z_inf, z_zimmer, colors,
                               x_index=0, plt_slice=(0,1000),
@@ -56,8 +90,15 @@ def plot_1d_continuous_states(x_inf, z_inf, z_zimmer, colors,
 
     plt.savefig(os.path.join(results_dir, filename))
 
-def plot_2d_continuous_states(ax, x, z, colors,
-                              inds=(0,1)):
+def plot_2d_continuous_states(x, z, colors,
+                              ax=None,
+                              inds=(0,1),
+                              figsize=(2.5, 2.5),
+                              results_dir=".", filename=None,):
+
+    if ax is None:
+        fig = plt.figure(figsize=figsize)
+        ax = fig.add_subplot(111)
 
     cps = states_to_changepoints(z)
 
@@ -67,11 +108,16 @@ def plot_2d_continuous_states(ax, x, z, colors,
[…]

### 2017-11-09 zimmer/states.py

@@ -1,8 +1,123 @@
 # Build specific models for joint observations
 import numpy as np
 
-from pyslds.states import _SLDSStatesMaskedData, _SLDSStatesGibbs,\
-    _SLDSStatesVBEM, HMMStatesEigen
+from pybasicbayes.models.factor_analysis import FactorAnalysisStates
+from pylds.states import LDSStatesMissingData
+from pyslds.states import _SLDSStatesMaskedData, _SLDSStatesGibbs, _SLDSStatesVBEM
+from pyhsmm.internals.hmm_states import HMMStatesEigen
+from rslds.states import InputHMMStates
+
+
+class HierarchicalFactorAnalysisStates(FactorAnalysisStates):
+    def __init__(self, model, data, group=0, mask=None):
+        self.group = group
+        super(HierarchicalFactorAnalysisStates, self).__init__(model=model, data=data, mask=mask)
+
+    @property
+    def sigmasq(self):
+        return self.model.regression.sigmasq_flat[self.group]
+
+
+class HierarchicalLDSStates(LDSStatesMissingData):
+    def __init__(self, model, data=None, mask=None, group=None, **kwargs):
+        self.group = group
+        super(HierarchicalLDSStates, self).__init__(model,data=data, mask=mask, **kwargs)
+
+    @property
+    def info_emission_params(self):
+        if self.mask is None:
+            return super(LDSStatesMissingData, self).info_emission_params
+
+        assert self.diagonal_noise
+        return self._info_emission_params_diag
+
+    @property
+    def _info_emission_params_diag(self):
+        C, D = self.C, self.D
+        sigmasq = self.emission_distn.sigmasq_flat[self.group]
+        J_obs = self.mask / sigmasq
+        centered_data = self.data - self.inputs.dot(D.T)
+
+        CCT = np.array([np.outer(cp, cp) for cp in C]). \
+            reshape((self.D_emission, self.D_latent ** 2))
+
+        J_node = np.dot(J_obs, CCT)
+        J_node = J_node.reshape((self.T, self.D_latent, self.D_latent))
+
+        # h_node = y^T R^{-1} C - u^T D^T R^{-1} C
+        h_node = (centered_data * J_obs).dot(C)
+
+        log_Z_node = -self.mask.sum(1) / 2. * np.log(2 * np.pi)
+        log_Z_node -= 1. / 2 * np.sum(self.mask * np.log(sigmasq), axis=1)
[…]

### 2017-11-09 drafts/2017_11_06/zimmer.tex
final draft of what was sent to zimmer

@@
   behavior that drive dynamics in this latent state space.  We find
-  characeristic transition patterns between these discrete modes, and
+  characteristic transition patterns between these discrete modes, and
   we see that transition probabilities are determined, in part, by the
@@
 for each of~$W$ worms.  The matrix~${\widetilde{Y}^{(w)} \in \reals^{T_w \times N_w}}$
-representes the calcium activity of the~$w$-th worm, which consists of~$T_w$
+represents the calcium activity of the~$w$-th worm, which consists of~$T_w$
 time frames and~$N_w$ neurons.\footnote{Technically, this calcium activity is
@@
 bleaching-corrected~$\Delta F /F$ signal. Unlike~\citet{kato2015global}, we
-do not smooth these derivates. We work with the raw first-order
+do not smooth these derivatives. We work with the raw first-order
 temporal differences.}  \celegans~is special in that each neuron has a unique
@@
 distribution over next discrete states.  The link function is the
-softmax function, which exponentiates and noramlizes
+softmax function, which exponentiates and normalizes
 the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.
@@
     10-dimensional latent states.  \textit{(bottom)} Inferred standard
-    deviaion~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
+    deviation~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
     indicate neurons that were not observed in that worm.}
@@
     black lines---based on the correlations found in other worms. We
-    show only one minute of data and the first 20 neurons
-    (alphabetically) here.}
+    show two minutes of data here.  The neurons have been grouped into
+    five clusters, as discussed below.  Within each cluster, the
+    neurons are sorted alphabetically. Each neuron is individually
+    scaled by the standard deviation of its observed activity to
+    illustrate dynamics of low-amplitude neurons. }
   \label{fig:activity_1}
@@
 activity is not observed, the hierarchical model can predict it based
-on correlations observed in other neurons. We note that some of the
-highly active neurons (e.g. \textsf{AVFL} and \textsf{AVFR}) are
-largely smoothed out, as the model attributes much of their activity
-to noise.
+on correlations observed in other worms.
 
@@
 neurons. These embeddings can aid in this matching.  Specifically,
-given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabled
+given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabeled
 neuron~$n^*$ in worm~$w$, along with the set of continuous latent
@@
   the candidate neuron that is most likely to be~\textsf{AVAL}. For
-  this task, we have artifically held-out this label so that we know
+  this task, we have artificially held-out this label so that we know
   which candidates truly is~\textsf{AVAL}. We use the neural
@@
 and the ``recurrent'' weights~$R$, as defined in
-Eq.~\eqref{eq:recurrent}. The largest weights come from the
+Eq.~\eqref{eq:recurrent}.  Since the impact of the recurrent weights
+depends on the scale of the continuous latent states, we multiply the
+weight matrix~$R$ by the standard deviation of~$x$ (since~$x$ is
+roughly zero mean, this is like z-scoring~$x$ so that it is comparable
+with the Markov term).  The largest weights come from the
 self-transitions, i.e. the diagonal of the Markov terms. Still, the
@@
 
-\subsection{Some state durations are non-Markovian}
+\subsection{The recurrent model captures non-Markovian state durations}
 
@@
   non-Markovianity by modeling transitions as a function of location
-  in latent state space~$x_t$.}
+  in latent state space~$x_t$. Solid lines show the distribution
+  of simulated durations under the recurrent AR-HMM; dashed lines
+  show the geometric duration distribution under the Markovian model.}
 \label{fig:durations}
@@
 
-Figure~\ref{fig:durations} shows histograms of the distribution of inferred
-state distributions for each state.  For comparison, we show
-the geometric distribution of state durations implied by the
-Markov weights as a black line.  Specifically, the distribution
-over durations~$d_k$ of state~$k$ under the Markov model (ignoring the
-recurrent weights) is given by,
+Figure~\ref{fig:durations} shows histograms of the
+empirical distribution of inferred state durations for each state. For
+comparison, the solid lines show the distribution of
+\emph{simulated} state distributions from our recurrent AR-HMM.  We
+see a close correspondence.
+
+This correspondence is a direct result of the recurrent nature of the
+model.  Without the recurrent terms, the state duration distributions
+would necessarily follow the geometric distribution shown in dotted
+black lines.  These distributions decay monotonically and cannot
+capture the peaks away from zero that we find in the data.  This is
+due to the fact that the distribution over durations~$d_k$ of
+state~$k$ under a Markov model (ignoring the recurrent weights) is
+given by,
 \begin{align*}
@@
 \end{align*}
-We see that some states (e.g. states 1, 5, and 8) seem to follow the
-Markov assumption quite well, whereas others (states 2, 4, and especially 7),
-are far from Markovian.  The recurrent model incorporates the continuous
-latent state~$x_t$ into the transition probabilities via the additive term~$R x_t$
-in the softmax.  It is likely that this term captures the non-Markovianity.
+While some states (e.g. states 1, 5, and 8) seem to follow the Markov
+assumption quite well, others (states 2, 4, and especially 7), are far
+from Markovian.  The recurrent model incorporates the continuous
+latent state~$x_t$ into the transition probabilities via the additive
+term~$R x_t$ in the softmax and thereby captures the non-Markovianity.
 
@@
   1 ({\sf AVAL, AVAR, BAGL, RIML, RIMR, SAAVR, SABVL, SABVR, VA01}.
-  The \textsf{AVA*} interneurons innvervate motor neurons.
+  The \textsf{AVA*} interneurons innervate motor neurons.
   \textsf{RIM*} are motor neurons that modulate reversals.
@@
 
-\subsection{Inferred changepoints also match manual labels}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
-\caption{(top) Posterior distribution over latent states~$z_t$.
-  Each column corresponds to a time bin and sums to one.
-  (middle) The posterior probability of changepoints.
-  (bottom) For reference,
-  the manually labeled states of~\citet{kato2015global}. Colors match those
-  of Figure~\ref{fig:overlap}.}
-\label{fig:z_cps_12}
-\end{figure}
-
-Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
-show the posterior distribution of latent states~$z$ inferred from
-neural activity alone. From this posterior, we also compute the probability
-of a changepoint, where~$z_{t+1} \neq z_t$.  For reference, we show the
-manually labeled states of~\citet{kato2015global} below. Here, patterns
-like the cascade from states 2 to 3 to 4 (red to yellow to green) are
-apparent.
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
-\caption{See Figure~\ref{fig:z_cps_12} caption.}
-\label{fig:z_cps_34}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
-\caption{See Figure~\ref{fig:z_cps_12} caption.}
-\label{fig:z_cps_5}
-\end{figure}
-
-\clearpage
-
-\subsection{Inferred states changepoints correspond with changes in dynamics}
+% \subsection{Inferred change-points also match manual labels}
+
+% \begin{figure}[h]
+% \centering%
+% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
+% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
+% \caption{(top) Posterior distribution over latent states~$z_t$.
+%   Each column corresponds to a time bin and sums to one.
+%   (middle) The posterior probability of change-points.
+%   (bottom) For reference,
+%   the manually labeled states of~\citet{kato2015global}. Colors match those
+%   of Figure~\ref{fig:overlap}.}
+% \label{fig:z_cps_12}
+% \end{figure}
+
+% Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
+% show the posterior distribution of latent states~$z$ inferred from
+% neural activity alone. From this posterior, we also compute the probability
+% of a change-point, where~$z_{t+1} \neq z_t$.  For reference, we show the
+% manually labeled states of~\citet{kato2015global} below. Here, patterns
+% like the cascade from states 2 to 3 to 4 (red to yellow to green) are
+% apparent.
+
+% \begin{figure}[h]
+% \centering%
+% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
+% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
+% \caption{See Figure~\ref{fig:z_cps_12} caption.}
+% \label{fig:z_cps_34}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
+% \caption{See Figure~\ref{fig:z_cps_12} caption.}
+% \label{fig:z_cps_5}
+% \end{figure}
+
+% \clearpage
+
+\subsection{Inferred states change-points correspond with changes in dynamics}
 
@@
 manual segmentation of~\citep{kato2015global} (bottom).  The inferred
-segmentation seems to identify changepoints when one or more of the latent
+segmentation seems to identify change-points when one or more of the latent
 continuous states changes dynamics.  Indeed, this is the assumption of the
@@
 inferred segmentation at the population level, it misses some salient
-changes in dynamics.  For example, see the inferred changepoint around
-time 700 in worm 1.  This changepoint was not identified in the manual
+changes in dynamics.  For example, see the inferred change-point around
+time 700 in worm 1.  This change-point was not identified in the manual
 data.

### 2017-11-09 experiments/fit_arhmm.py

@@ -59,10 +59,12 @@ from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states
 
 # LDS Results
 lds_dir = os.path.join("results", "2017-11-03-hlds", "run002")
+# lds_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
 assert os.path.exists(lds_dir)
 
 # AR-HMM RESULTS
 results_dir = os.path.join("results", "2017-11-04-arhmm", "run003")
+# results_dir = os.path.join("results", "2017-11-04-arhmm", "run004_dff_bc")
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
@@ -71,7 +73,7 @@ fig_dir = os.path.join(results_dir, "figures")
 N_lags = 1
 N_samples = 1000
 N_worms = 5
-T_sim = 10 * 3
+
 
 def cached(results_name):
     def _cache(func):
@@ -285,24 +287,14 @@ def fit_best_model(K=8,
     return best_model, lls, hll
 
 
-def simulate_trajectories(N_trajs=100, N_sims=4, worm=1):
-    # Simulate a canonical trajectory
-    def _simulate_dynamics(dd, x0, T):
-        X = np.zeros((T, dd.D_out))
-        X[0] = x0
-        for t in range(1, T):
-            # xp = np.concatenate((X[t - 1], [1]))
-            xp = X[t-1]
-            X[t] = dd.predict(xp)
-        return X
-
+def simulate_trajectories(N_trajs=100, T_sim=30, N_sims=4, worm=1, min_sim_dur=6):
     from pyhsmm.util.general import rle
     z_finals_rles = [rle(z) for z in z_finals]
 
     x_trajss = []
     x_simss = []
     for k in range(best_model.num_states):
-
+        print("Simulating state ", k)
         x_trajs = []
         for worm in range(N_worms):
             x = xs[worm][N_lags:]
@@ -339,12 +331,15 @@ def simulate_trajectories(N_trajs=100, N_sims=4, worm=1):
                 dur_k = T_sim
 
             # Make sure the simulation is long enough
-            if dur_k < 6:
+            if dur_k < min_sim_dur:
                 continue
 
             # If the simulation passes, keep it
             x_sims.append(x_sim[:dur_k])
             j += 1
+
+            if j % 50 == 0:
+                print("{} / {}".format(j, N_sims))
 
         x_simss.append(x_sims)
 
[…]

### 2017-11-09 experiments/fit_hlds.py

@@ -47,8 +47,12 @@ from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
 
 # IO
-run_num = 2
-results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
+# run_num = 2
+# results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
+# signal = "dff_diff"
+results_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
+signal = "dff_bc"
+
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
@@ -77,7 +81,7 @@ def cached(results_name):
     return _cache
 
 
-def load_data(signal="dff_diff", include_unnamed=True):
+def load_data(include_unnamed=True):
     # Load the data
     worm_datas = [WormData(i, name="worm{}".format(i)) for i in range(N_worms)]
 
@@ -306,22 +310,24 @@ def plot_best_model_results(best_model,
             plot_3d_continuous_states(xtrains[i], z_true_trains[i], colors,
                                       figsize=(4, 4),
                                       title="LDS Worm {} States (Zimmer Lables)".format(i + 1),
-                                      results_dir=results_dir,
+                                      results_dir=fig_dir,
                                       filename="xtr_zimmer_{}.pdf".format(i + 1),
-                                      lim=1.5,
+                                      # lim=1.5,
                                       lw=1)
         plt.close("all")
 
+        # for i in range(N_worms):
+        #     plot_3d_continuous_states(xtests[i], z_true_tests[i], colors,
+        #                               figsize=(4, 4),
+        #                               title="LDS Worm {} States (Zimmer Lables)".format(i + 1),
+        #                               results_dir=fig_dir,
+        #                               filename="xte_zimmer_{}.pdf".format(i + 1),
+        #                               lim=1.5,
+        #                               lw=1)
+        # plt.close("all")
+
     if do_plot_x_2d:
-        for i in range(N_worms):
-            plot_3d_continuous_states(xtests[i], z_true_tests[i], colors,
-                                      figsize=(4, 4),
[…]

### 2017-11-09 zimmer/plotting.py

@@ -1259,7 +1259,7 @@ def plot_recurrent_transitions(trans_distn, xs, zs,
 
 
 
-def plot_duration_histogram(trans_distn, zs,
+def plot_duration_histogram(trans_distn, zs, sim_durs,
                             colors=None,
                             results_dir=None):
 
@@ -1283,12 +1283,27 @@ def plot_duration_histogram(trans_distn, zs,
         dk = durs[states == k]
         dmax = dk.max()
         bins = np.linspace(0, dmax+1, 15)
+        width = (bins[1] - bins[0]) / 2.0
 
         print(len(dk))
 
         plt.figure(figsize=(1.8, 1.8))
-        plt.hist(dk / 3.0, bins/ 3.0, color=colors[k], edgecolor='k', normed=True, label="Empirical")
-        plt.plot(np.arange(1, dmax) / 3.0, g.pmf(np.arange(1, dmax)) * 3.0, '-k', label="Markov")
+
+        # Plot the histogram of inferred transitions
+        tmp, _ = np.histogram(dk, bins, density=True)
+        # plt.bar(bins[:-1] / 3.0, tmp * 3.0, width=width / 3.0,
+        #         color=colors[k], edgecolor='k', label="Empirical")
+        plt.bar(bins[:-1] / 3.0, tmp * 3.0, width=width * 2.0 / 3.0,
+                color=colors[k], alpha=0.75, edgecolor='k', label="Emp")
+
+
+        # Plot the histogram of simulatedtransitions
+        tmp, _ = np.histogram(sim_durs[k], bins, density=True)
+        # plt.bar((bins[:-1] + width) / 3.0, tmp * 3.0, width=width / 3.0,
+        #          color=colors[k], alpha=0.5, edgecolor='k', label="Simulated")
+        plt.plot(bins[:-1] / 3.0, tmp * 3.0, '-k', label="Rec")
+
+        plt.plot(np.arange(1, dmax) / 3.0, g.pmf(np.arange(1, dmax)) * 3.0, ':k', label="Mkv")
         plt.xlabel("duration (s)")
         plt.ylabel("probability")
         plt.legend(loc="upper right", fontsize=6)

### 2017-12-19 experiments/fit_arhmm_nichols.py
updating fitting script

@@ -10,6 +10,8 @@ np.random.seed(0)
 from tqdm import tqdm
 
 # Plotting stuff
+import matplotlib
+matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from hips.plotting.colormaps import gradient_cmap
 import seaborn as sns
@@ -56,7 +58,8 @@ from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states
     plot_3d_dynamics, plot_2d_dynamics, plot_state_overlap, plot_state_usage_by_worm, plot_all_transition_matrices, \
     plot_simulated_trajectories, make_state_predictions_3d_movie, plot_simulated_trajectories2, \
     plot_recurrent_transitions, plot_x_at_changepoints, plot_latent_trajectories_vs_time, \
-    plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_driven_transition_matrices
+    plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_driven_transition_matrices, \
+    plot_driven_transition_mod
 
 # LDS Results
 lds_dir = os.path.join("results", "nichols", "2017-11-13-hlds", "run001")
@@ -178,10 +181,13 @@ def _fit_model_wrapper(K, alpha=3, gamma=100., kappa=100.,
         data_kwargs = {}
         if is_hierarchical:
             data_kwargs["group"] = i
-        if is_separate_trans:
+        if is_separate_trans and is_recurrent and is_hierarchical:
             data_kwargs["trans_group"] = worms_groups_conditions[i][1]
 
-        model.add_data(x, covariates=u, stateseq=z.astype(np.int32), **data_kwargs)
+        if is_driven and is_recurrent:
+            data_kwargs["covariates"] = u
+            
+        model.add_data(x, stateseq=z.astype(np.int32), **data_kwargs)
 
     # Fit the model with Gibbs
     lls = []
@@ -195,7 +201,7 @@ def _fit_model_wrapper(K, alpha=3, gamma=100., kappa=100.,
         data_kwargs = {}
         if is_hierarchical:
             data_kwargs["group"] = i
-        if is_separate_trans:
+        if is_separate_trans and is_recurrent and is_hierarchical:
             data_kwargs["trans_group"] = worms_groups_conditions[i][1]
 
         if is_driven:
@@ -203,7 +209,10 @@ def _fit_model_wrapper(K, alpha=3, gamma=100., kappa=100.,
         else:
[…]

### 2017-12-19 zimmer/plotting.py

@@ -695,11 +695,15 @@ def plot_state_overlap(z_finals, z_trues,
                        z_key=None,
                        colors=None,
                        z_colors=None,
+                       filename=None,
+                       titles=None,
                        results_dir=None):
     colors = default_colors if colors is None else colors
     z_colors = default_colors if z_colors is None else z_colors
     Kmax = np.max(np.concatenate(z_finals)) + 1
     K_zimmer = np.concatenate(z_trues).max() + 1
+    titles = ["State Overlap (Worm {})".format(i+1) for i in range(len(z_trues))] \
+      if titles is None else titles
 
     # Use the Hungarian algorithm to find a permutation of states that
     # yields the highest overlap
@@ -728,8 +732,8 @@ def plot_state_overlap(z_finals, z_trues,
         overlap /= overlap.sum(1)[:, None]
 
         fig = plt.figure(figsize=(2., 2))
-        ax1 = create_axis_at_location(fig, .6, .5, 1., 1)
-        im = ax1.imshow(overlap[:, perm], vmin=0, vmax=.75, interpolation="nearest", aspect="auto")
+        ax1 = create_axis_at_location(fig, .5, .5, 1., 1)
+        im = ax1.imshow(overlap[:, perm], vmin=0, vmax=1.0, cmap="Greys", interpolation="nearest", aspect="auto")
         ax1.set_xticks([])
         if z_key is None:
             ax1.set_yticks([])
@@ -737,9 +741,9 @@ def plot_state_overlap(z_finals, z_trues,
             ax1.set_yticks(np.arange(K_zimmer))
             ax1.set_yticklabels(z_key, fontdict=dict(size=6))
             ax1.tick_params(axis='y', which='major', pad=11)
-        ax1.set_title("State Overlap (Worm {})".format(worm + 1))
+        ax1.set_title(titles[worm])
 
-        lax = create_axis_at_location(fig, .5, .5, .06, 1)
+        lax = create_axis_at_location(fig, .4, .5, .06, 1)
         lax.imshow(np.arange(K_zimmer)[:, None], cmap=gradient_cmap(z_colors[:K_zimmer]), interpolation="nearest",
                    aspect="auto")
         lax.set_xticks([])
@@ -748,17 +752,18 @@ def plot_state_overlap(z_finals, z_trues,
         if z_key is None:
             lax.set_ylabel("Zimmer State", fontsize=8)
 
[…]

### 2017-12-20 drafts/elife/draft.tex
adding some plotting scripts, starting journal manuscript

--- 2017-11-09 drafts/2017_11_06/zimmer.tex
@@
-% \documentclass{article}
-
-% \usepackage{natbib}
-
-% \usepackage[utf8]{inputenc} % allow utf-8 input
-% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-% \usepackage{hyperref}       % hyperlinks
-% \usepackage{url}            % simple URL typesetting
-% \usepackage{booktabs}       % professional-quality tables
-% \usepackage{amsfonts}       % blackboard math symbols
-% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
-% \usepackage{microtype}      % microtypography
-
-% \usepackage{amsthm,amsmath,amssymb}
-% \usepackage{macros}
-% \usepackage{subcaption}
-% \usepackage[textfont=small, labelfont=small]{caption}
-% \usepackage{graphicx}
-% \DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
-
-% \usepackage{algorithm}
-% \usepackage{algorithmic}
-
-% \usepackage[color=yellow]{todonotes}
-% \usepackage{booktabs}
-% \usepackage[inline]{enumitem}
-% \usepackage{verbatim}
-
-% \usepackage[left=1.5in, right=1.5in, top=1.25in, bottom=1.25in]{geometry}
-
-% \usepackage{setspace}
-
 \documentclass{article}
@@
 
-\title{Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
-
-\author{Scott W. Linderman,
-  David M. Blei,
+% \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
+\title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
+\author{Scott Linderman$^{\text{1}}$,
+  Annika Nichols$^{\text{2}}$,
+  David Blei$^{\text{1}}$,
+  Manuel Zimmer$^{\text{2}}$,
   and
-  Liam Paninski
+  Liam Paninski$^{\text{1}}$
   \\
-  Columbia University
+  $^{\text{1}}$Columbia University,
+  $^{\text{2}}$Research Institute of Molecular Pathology (IMP), Vienna Biocenter
 }
@@
 
-%\singlespacing
+\doublespacing
+
 \maketitle
@@
   simultaneous measurements of the majority of head ganglia neurons in
-  both immobilized and freely-behaving \celegans.  The dynamics of
-  neural activity shed light on how \celegans~processes sensory
-  information and generates motor activity.  To understand these
-  dynamics, we leverage \emph{recurrent switching linear dynamical
-    systems}~\citep{linderman2017recurrent}---probabilistic models
-  that decompose complex time-series into segments with simple, linear
-  dynamics. We incorporate these models into a robust, hierarchical
-  framework for combining information across whole-brain recordings of
-  multiple worms.  We use this framework to reveal latent states of
-  population neural activity, along with the discrete modes of
-  behavior that drive dynamics in this latent state space.  We find
-  characeristic transition patterns between these discrete modes, and
-  we see that transition probabilities are determined, in part, by the
-  current brain state.  This probabilistic framework is a useful aid
-  for neural identification, currently a laborious, manual task.
-  Moreover, we find clusters of neurons that are similarly tuned in
-  latent state space, and we show how the different dynamical modes
-  activate these clusters.  Finally, we find a significant overlap
-  between our inferred modes and the manually-labeled states
-  of~\citet{kato2015global}, which were shown to correspond to
-  different worm behaviors. Our framework automatically discovers
-  these behaviorally-meaningful regimes directly from neural activity.
+  both immobilized and freely-behaving
+  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
+    nguyen2016whole}.  The dynamics of neural activity shed light on
+  how \celegans~processes sensory information and generates motor
+  activity.  To understand these dynamics, we develop recurrent state
+  space models that decompose complex time-series into segments with
+  simple, linear dynamics. We incorporate these models into a robust,
+  hierarchical framework for combining information across whole-brain
+  recordings of many worms.  Using this framework, we reveal latent
+  states of population neural activity, along with the discrete
+  behavioral modes that drive dynamics in this latent state space.  We
+  find stochastic transition patterns between these discrete
+  behavioral modes, and we see that transition probabilities are
+  determined by a combination of current brain state and environmental
+  cues.  In addition to quantifying neural dynamics, this
+  probabilistic framework aids in neural identification---currently a
+  laborious, manual task---and reveals clusters of neurons that are
+  similarly tuned in latent state space.  Finally, we find a
+  significant overlap between our inferred modes and the
+  manually-labeled modes of~\citet{kato2015global}
+  and~\citet{nichols2017global}, which were shown to correspond to
+  different behaviors, like forward crawling, reversals, and
+  turns. Our methods automatically discover and quantify these
+  behaviorally-meaningful states directly from neural activity,
+  yielding powerful new tools for neuro-behavioral analysis.
 \end{abstract}
 
-\section{Introduction}
-% The world as it was:
-The nematode \celegans~is a uniquely accessible model organism of
-neuroscience. With recent advances in optical recording
-technologies~\citep{schrodel2013brain, prevedel2014simultaneous,
-  nguyen2016whole}, we now possess a mounting set of tools for
-measuring the activity of large fractions of these neurons
-simultaneously. These technologies provide exciting opportunities to
-probe neural dynamics and their link to sensory
-processing and behavior.
+\clearpage
 
-Recently, \citet{kato2015global} harnessed these methodological
-advances to study the coordinated activity of hundreds of neurons
-in head-fixed \celegans. Across multiple organisms, they found that
-neural activity reliably traced out smooth trajectories in the subspace
-of the first three principal components.
-% TODO: Figure showing these trajectories
-Upon closer evaluation, they found that different components of these
-trajectories corresponded to different elements of behavior, like
-forward and reverse crawling, dorsal and ventral turns, etc. These
-findings raise a number of interesting questions: can we learn a
-(potentially nonlinear) dynamical system that approximates these
-low-dimensional dynamics; is there a more appropriate subspace, or
-collection of subspaces, for capturing these dynamics; can we gain
-statistical power by partial, noisy recordings across separate worms
-and trials; and, can we segment these trajectories in an unsupervised
-manner to glean further insight into the relationship between
-low-dimensional state and behavior?
+\section*{Introduction}
 
-% Then one day:
-We develop a probabilistic framework for investigating these
-questions.  We model these multi-neuronal recordings as a
-high-dimensional projection of a low-dimensional, dynamical latent
-state.  We allow the dynamics of this latent state change over time,
-intuitively capturing different patterns of brain activity for
-different types of behavior.  The \emph{switching linear dynamical
-  system} (SLDS) \citep{chang1978state, ackerson1970state,
-  hamilton1990analysis, ghahramani1996switching, murphy1998switching}
-captures both of these properties: low-dimensional continuous latent
-states with different dynamical regimes.  The SLDS is a probabilistic
-time series model characterized by co-evolving discrete and continuous
-latent states.  We extend this class of models with a hierarchical
-Bayesian framework that captures many of the nuances of these
-whole-brain recordings. In fitting these models to recordings from
-multiple \celegans, we characterize the globally nonlinear dynamics
-that govern neural activity, and we parse these recordings into an
-interpretable sequence of behavioral segments.
+\clearpage
 
-% Raising the stakes:
-% Moment of truth:
-% World as it is now:
-% The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
-% we introduce the class of switching linear dynamical systems and provide
-% some insight into the latent states and model parameters. Section~\ref{sec:hslds}
-% develops a hierarchical extension of the SLDS to share statistical
-% strength by combining recordings from multiple organisms with partially
-% overlapping sets of observed neurons.  Then, Section~\ref{sec:results}
-% presents results of applying these models to a set of recordings from
-% five separate \celegans. We conclude with directions for further research.
+\section*{Results}
+
+\subsection*{Hierarchical models pool information while allowing individual variability}
+% Introduce the probabilistic model
+
+\begin{figure}[t]
+\centering%
+\includegraphics[width=6in]{figures/figure1} 
+\caption{A schematic of the hierarchical model for combining
+  information across multiple worms in order to learn a shared
+  dynamical system for \celegans.  \textbf{a.}~All worms share a
+  canonical dynamics library. Here, we have three dynamics functions
+  (blue, red, gold), each of which is a simple linear mapping from
+  $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
+  assigns probabilities to the different dynamics functions based on
+  the current location in $x$-space.  \textbf{b.}~The worms also share
+  an emission matrix, $C$, which can be viewed as an embedding of
+  neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
+  library, which allows for slight deviations from the canonical
+  dynamics.  \textbf{d.}~The model segments neural activity into a
+  sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
+  into the dynamics library.  \textbf{e.}~The discrete states specify
+  the dynamics that are used to propagate the continuous latent
+  states~$x_t$ forward in time. Here, the two dimensional latent
+  states are colored according to the instantaneous discrete
+  state. \textbf{f.}~The observed neural activity is a linear
+  projection of the underlying continuous latent states.  On any given
+  trial, we observe only a fraction of all the neurons.  The
+  hierarchical model combines these partial recordings to infer the
+  underlying states, the dynamics libraries, and the shared neural
+  embedding. }
+\label{fig:hslds}
+\end{figure}
+
+\subsection*{Emission models reveal clusters of neurons and aid in identification}
+
+\subsection*{AR-HMM's automatically parse data into discrete states with simple dynamics}
+
+\subsection*{Recurrent models learn spatially localized states and transition boundaries}
+
+\subsection*{Oxygen level modulates transition probabilities but not dynamics or emissions}
+
+\clearpage
+
+\section*{Discussion}
+
+\clearpage
+
+\bibliography{refs}
+\bibliographystyle{abbrvnat}
+
+\clearpage
+
+\appendix
 
@@
 
-\begin{figure}[t]
-\centering%
-\includegraphics[width=\textwidth]{figures/hslds.png} 
-\caption{A schematic of the hierarchical SLDS for combining information
-  across multiple worms in order to learn a shared dynamical system
-  for \celegans.  We have a set of shared set of model parameters,~$\theta$,
-  represented by the large node at the top of the graphical model. Each
-  worm,~$w$ has its own set of discrete latent states (colored nodes)
-  and continuous latent states (nodes with colored outlines) and observed neurons.
-  For example, neuron~1 may appear in all neurons whereas neuron~2
-  may only be seen in worm~2. Thus, each worm provides information
-  about the mapping from latent states to observations for only a
-  subset of neurons. In terms of the model, this corresponds to a
-  subset of rows in the matrix~$C$ and the vector~$d$. Finally,
-  we allow each worm to have local parameters that capture, for example,
-  the amount of fluorescent protein expressed in a particular neuron
-  for a given worm. 
-}
-\label{fig:hslds}
-\end{figure}
-
-\subsection{Extensions of the standard SLDS}
-\label{sec:extensions}
-
-To better model the \celegans~data, we introduce three extensions
-to the standard SLDS.
-
-\paragraph{Hierarchical model to allow limited variability across worms.}
-The first extension captures shared patterns across worms while
-still allowing for worm-to-worm variability.  We do this in
-two ways. First, we allow the dynamics parameters to vary
-slightly from worm to worm with the following model,
-\begin{align*}
-  \mathrm{vec}([A_k, b_k]) &\sim \distNormal(\lambda), \\
-  \mathrm{vec}([A_k^{(w)}, b_k^{(w)}]) &\sim \distNormal(\mathrm{vec}([A_k, b_k]), \sigma^2 I), \\
-  Q_k^{(w)} &\sim \mathrm{IW}(\lambda).
-\end{align*}
-This allows worm-specific dynamics parameters~$(A_k^{(w)}, b_k^{(w)})$ but
-requires that they not deviate too far from the global mean~$(A_k, b_k)$.
-It also allows for worm-specific dynamics noise.
-
-Second, we allow each worm to have neuron-specific emission variances,
-\begin{align*}
-  s_{n}^{(w)} & \iid\sim \mathrm{IG}(\eta).
-\end{align*}
-We may expect, for example, that calcium indicator expression
-varies from worm to worm, which in turn could give rise to
-different levels of observed activity.  While this could be
-modeled as a scaling of the emission matrix, a simpler approach
-is to let the noise term capture these different fluctuations.
-We could share the global mean of the worm-specific variances,
-but we have little reason to expect the~$s_{n}^{(w)}$ and~$s_{n}^{(w')}$
-to be correlated under the prior.
-
-Figure~\ref{fig:hslds} illustrates the hierarchical model.  Each
-worm has a set of discrete states~$z_{1:T}$, continuous states~$x_{1:T}$,
-and observations for a subset of neurons.  The worms share a set of
-global dynamics parameters and emission matrices, but they also
-have worm-specific perturbations of the global dynamics, as well as
-worm-specific observation variances.
-
-\paragraph{Robust dynamics model.}  While the SLDS can capture nonlinear
-dynamics by composing simple linear pieces, it is still an approximation
-to the true data-generating process.  To account for some of this
-model misspecification, we introduce a \emph{robust} extension of the
-SLDS by allowing for heavy-tailed noise in the dynamics.  Specifically,
-we replace~\eqref{eq:slds_start} with,
-\begin{align}
-  x_{t+1}^{(w)} &= A_{z_{t+1}^{(w)}} x_{t}^{(w)} + b_{z_{t+1}^{(w)}} +  u_t^{(w)},
-  &
-  u_t^{(w)} &\iid\sim \mathrm{t}(0, Q_{z_{t+1}^{(w)}}, \nu_{z_{t+1}^{(w)}}),
-\end{align}
-where~$\mathrm{t}(\cdot, \cdot, \cdot)$ denotes the multivariate-t
-distribution.  This distribution has heavier tails than its Gaussian
-counterpart; indeed, it can be derived by a mixture of multivariate
-Gaussians with a~$\chi^2$-distributed scaling of the covariance matrix.
-
-\paragraph{Recurrent model of discrete state transition probabilities.}
-Finally, we consider a third extension that allows for more complex
-transition probabilities.  We replace~\eqref{eq:markov_z} with,
-\begin{align}
-  z_{t+1}^{(w)} \given z_t^{(w)}, x_t^{(w)}
-  &\sim \mathrm{softmax}(r_{z_t^{(w)}} + R x_t^{(w)}),
-    \label{eq:recurrent}
-\end{align}
-where~${r_k \in \reals^K}$ and~${R \in \reals^{K \times D}}$
-parameterize a map from previous discrete and continuous states to a
-distribution over next discrete states.  The link function is the
-softmax function, which exponentiates and noramlizes
-the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.
-
-\section{Model Fitting}
-
-We have proposed a variety of Bayesian algorithms for learning the
-model parameters~$\theta$ and performing joint inference of the
-latent states~$z$ and~$x$~\citep{linderman2017recurrent, linderman2017structure},
-but for simplicity, here we treat inference of~$x$ and~$z$ separately.
-That is, first we find low-dimensional continuous states to model the
-observations~$y$, then we find a set of discrete states~$z$ that best
-explain the dynamics in~$x$. We learn the corresponding parameters
-at both stages.
-
-\todo[inline, size=\normalsize]{We will present full details of these inference
-procedures in the appendix, but for now, here are a few highlights.}
-
-To be more specific, first we fit linear dynamical systems to infer
-the continuous latent states~$x$ given the observed values of~$y$.
-Since many entries of~$y$ are masked off, this requires inference
-with missing data.  Under the assumption of diagonal observation
-covariance, this inference is straightforward.  We will compare the
-standard LDS with the hierarchical LDS, which has worm- and neuron-specific
-variances.
-
-Once we have inferred the continuous latent states, we will fit
-an autoregressive hidden Markov model (AR-HMM) to~$x$ in order
-to infer~$z$ and learn the dynamics parameters~$(A_k, b_k, Q_k)$.
-Here, again, we will consider all possible variants of hierarchical,
-robust, and recurrent models.
-
-Both fitting procedures use Markov chain Monte Carlo, specifically
-block Gibbs sampling, to sample from the conditional distribution
-of latent states given parameters and of parameters given latent
-states.  We run these algorithms for 1000 iterations, at which
-point we find them to have converged on the basis of held-out
-log likelihood.
-
-\section{Results}
-\label{sec:results}
-
-\todo[inline, size=\normalsize]{For this draft, I will present results one at a time,
-  each on a separate page.  Once we have settled on the story, we
-  can add more exposition.}
-\clearpage
-
-\subsection{Neural activity is low-dimensional and has worm-specific variance}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=6in]{figures/lds/dimensionality} \\
-  \includegraphics[width=5.5in]{figures/lds/observation_variance} 
-  \caption{\textit{(top)} Training and test likelihoods for the linear
-    dynamical system and its hierarchical extension as a function of
-    the dimensionality of the latent state,~$D$ (units:
-    nats/observation over Gaussian baseline). On the basis of held-out
-    test likelihood, we choose the hierarchical LDS with
-    10-dimensional latent states.  \textit{(bottom)} Inferred standard
-    deviaion~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
-    indicate neurons that were not observed in that worm.}
-  \label{fig:lds}
-\end{figure}
-
-We fit linear dynamical systems (LDS) and their hierarchical
-extensions with worm-specific variance (see Section~\ref{sec:extensions})
-for a range of latent dimensions,~$D$.  We found that the
-10-dimensional hierarchical LDS yielded the highest test
-likelihood (holding out the final 20\% of time frames for
-each worm for testing, assuming stationarity). This is
-shown in Figure~\ref{fig:lds}(top).
-
-The hierarchical model has worm-specific observation variances.
-These are shown in Figure~\ref{fig:lds}(bottom).  For neurons that
-were not observed in a given worm, the corresponding cells are
-colored gray.
-
-\clearpage
-
-\subsection{The hierarchical LDS smooths the data and predicts unobserved activity}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=5.5in]{figures/lds/y_0}
-  \caption{The observed activity~$y$ (i.e. the first-order differences
-    of the calcium signal) are shown in black; the smoothed activity
-    with the hierarchical LDS is shown in blue.  The hierarchical LDS
-    makes predictions about unobserved neurons---those with dotted
-    black lines---based on the correlations found in other worms. We
-    show only one minute of data and the first 20 neurons
-    (alphabetically) here.}
-  \label{fig:activity_1}
-\end{figure}
-
-Figures~\ref{fig:activity_1}-\ref{fig:activity_5} show the observed
-activity~$y$ along with the smoothed activity under the hierarchical
-LDS.  Formally, the blue lines are given by~$C \E[x \given y] +
-d$. Unobserved neurons are shown as dotted black lines.  Though their
-activity is not observed, the hierarchical model can predict it based
-on correlations observed in other neurons. We note that some of the
-highly active neurons (e.g. \textsf{AVFL} and \textsf{AVFR}) are
-largely smoothed out, as the model attributes much of their activity
-to noise.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=5.5in]{figures/lds/y_1}
-  \caption{See caption of Figure~\ref{fig:activity_1}}
-  \label{fig:activity_2}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=5.5in]{figures/lds/y_2}
-  \caption{See caption of Figure~\ref{fig:activity_1}}
-  \label{fig:activity_3}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=5.5in]{figures/lds/y_3}
-  \caption{See caption of Figure~\ref{fig:activity_1}}
-  \label{fig:activity_4}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=5.5in]{figures/lds/y_4}
-  \caption{See caption of Figure~\ref{fig:activity_1}}
-  \label{fig:activity_5}
-\end{figure}
-
-\clearpage
-
-\subsection{The emission parameters aid in neural identification}
-
-One of the challenges in modeling this data is that most neurons
-cannot be labeled with certainty.  In other words, we can detect a
-neuron in the calcium imaging data, but we cannot determine its
-identity (\textsf{AVAL}, \textsf{AVAR}, etc.).  The emission
-parameters of the SLDS are a useful aid for determining these
-identities, as we show here.
-
-Across the dataset of five worms, 61 unique neurons were identified in
-at least one worm.  In any given worm, we also observe many other
-neurons that we cannot accurately label. Our goal here is to assign
-labels to the unlabeled neurons.  Of course, if a label, say \textsf{AVAL}, has
-already been identified in worm 1, then we cannot assign an unlabeled
-neuron to \textsf{AVAL}.  So the problem is to find the most likely matching
-of unlabeled neurons to remaining, available labels.  Since there are
-more unlabeled neurons than possible labels, our result will only be
-a partial matching.
-
-The rows of the emission matrix~$c_n$ serve as ``embeddings'' of the
-neurons. These embeddings can aid in this matching.  Specifically,
-given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabled
-neuron~$n^*$ in worm~$w$, along with the set of continuous latent
-states~$x^{(w)} \in \reals^{T \times D}$ inferred from the labeled
-neurons, we perform a linear regression to
-estimate~$c_{n^*} \in \reals^D$, the ``embedding'' of this unlabeled
-neuron.  We then compare this vector~$c_{n^*}$ to the embeddings of
-the labeled neurons,~$\{c_n\}_{n=1}^N$, which were learned with the
-SLDS.  We use the cosine similarity between two vectors as a measure
-of likeness.
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=3in]{figures/lds/similarity_comparison} 
-\caption{The emission parameters aid in neural identification.  In
-  this task, we view unlabeled neurons as \emph{candidates} for a
-  particular label. For example, suppose we are trying to determine
-  the candidate neuron that is most likely to be~\textsf{AVAL}. For
-  this task, we have artifically held-out this label so that we know
-  which candidates truly is~\textsf{AVAL}. We use the neural
-  embeddings~$c_{\mathsf{AVAL}}$, which was inferred from other worms,
-  to aid in this matching.  We measure the cosine similarity
-  between~$c_{\mathsf{AVAL}}$ and the coefficients of linear
-  regressions between~$x$ and the activity of the unlabeled neurons'
-  activity. This histogram shows the results of performing this
-  experiment with 10 held-out identities for each of the five worms.
-  We see that the correct candidates frequently have much higher
-  cosine similarity than incorrect candidates.  This means that we can
-  use this similarity measure to infer neuron identities.  }
-\label{fig:neuron_id}
-\end{figure}
-
-We demonstrate this identification in a synthetic task where we
-artificially withheld the labels of ten neurons per worm.  Suppose one
-of the held-out neurons in worm 1 was known to be \textsf{AVAL}.  We
-iterate through all the unlabeled neurons in worm
-1---\emph{candidates} for the label~\textsf{AVAL}---and measure the
-cosine similarity between their embeddings~$c_{n^*}$ and the
-embedding~$c_{\mathsf{AVAL}}$, which was inferred from worms 2-5.
-Figure~\ref{fig:neuron_id} shows the results of this task.  The
-embeddings of the held-out neurons were much more similar to the
-embedding of their true identity than were the other unlabeled
-neurons.
-
-To further quantify these results, we sort these similarities
-and find the index of the true held-out neuron in the sorted list.
-If the held-out AVAL neuron was the most similar to~$c_{\mathsf{AVAL}}$,
-its rank would be 1; if it was the least similar, its rank would
-be~$N_{\mathsf{unlabeled}}$, the number of unlabeled neurons for
-this worm.  Table~\ref{tab:neuron_id} shows the ranks of the
-similarities of held-out neurons to their true identities, compared
-to all other unlabeled neurons.  If the held-out
-neurons were always the most similar to their true embeddings, the
-matching columns would all be 1's, and we would easily be able to
-identify neurons with this information.  We see that in most cases
-the held-out neurons are within the top three most similar candidates,
-demonstrating that embeddings are useful for identification.
-
-Finally, we perform an even harder task. Using the Hungarian algorithm,
-we find the partial matching that maximizes the sum of similarities.
-We compute the accuracy of this matching by counting the fraction of held-out
-neurons that were assigned to their true label. (Since the Hungarian
-algorithm is assigning all the neurons at once, in general the number
-of correctly assigned neurons is not equal to the sum of 1's in the
-rank columns.) Still, we see that in this challenging task we assign
-many neurons to their true identities.
-
-\begin{table}[h]
-  \caption{\textit{Artificial neuron identification task results.}  In
-    each worm we have~$N_{\mathsf{labeled}}$ labeled neurons
-    and~$N_{\mathsf{unlabeled}}$ unlabeled neurons. Of the unlabeled
-    neurons, 10 had known labels that were held out for this task.  We
-    compare the cosine similarity of the unlabeled neuron embeddings
-    to the embeddings of the labeled neurons, sort the similarities,
-    and compute the rank of each held-out neuron to its true label
-    (most similar~=~1; least similar~=~$N_{\mathsf{unlabeled}}$).  We
-    use the Hungarian algorithm to find the most likely matching of
-    unlabeled neurons to possible labels, and we report the fraction
-    of the 10 held-out neurons that were correctly matched~(``Acc.'').
-  }
-  \vspace{1em}
-  \label{tab:neuron_id}
-  \centering
-  \begin{tabular}{cccccccccccccc}
-    \toprule
-    & $N_{\mathsf{labeled}}$ & $N_{\mathsf{unlabeled}}$ & \multicolumn{10}{c}{Matching Rank (10 Held-out Neurons)} & Acc. \\
-    \cmidrule(r){2-2}
-    \cmidrule(rl){3-3}
-    \cmidrule(rl){4-13}
-    \cmidrule(l){14-14}
-    % \midrule
-    Worm 1 & 14 & 95 & 10 & 25 & 1 & 1 & 66 & 2 & 1 & 1 & 2 & 1 & 0.30 \\
-    Worm 2 & 31 & 76 & 3 & 1 & 1 & 1 & 1 & 39 & 1 & 1 & 1 & 4 & 0.50 \\
-    Worm 3 & 20 & 111 & 5 & 53 & 3 & 22 & 1 & 2 & 1 & 1 & 1 & 1 & 0.50 \\
-    Worm 4 &33 & 93 & 2 & 1 & 26 & 2 & 1 & 56 & 1 & 1 & 1 & 1 & 0.30 \\
-    Worm 5 & 38 & 91 & 1 & 1 & 1 & 1 & 1 & 2 & 1 & 26 & 1 & 67 & 0.70 \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\clearpage
-
-\subsection{Neural embeddings reveal clusters of similarly tuned neurons}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/lds/permuted_similarity}
-\vspace{-3em}
-\caption{Cosine similarity between embedding vectors~$c_n$
-  suggest clusters of similarly tuned neurons. }
-\label{fig:clusters}
-\end{figure}
-
-The rows of the emission matrix~$c_n$ (i.e. the embedding vectors)
-also suggest that neurons cluster into groups of similarly tuned
-cells.  Figure~\ref{fig:clusters} shows the~${N \times N}$ matrix of
-cosine similarity for each pair of embedding vectors.  We used k-means
-to find clusters of embedding vectors and have permuted the
-similarity matrix to visualize these groups.  We see that many
-clusters contain lateral pairs (e.g. {\sf AVAL, AVAR} in
-the first cluster). 
-We will use these clusters to better understand the dynamics of
-the SLDS.
-
-Note that this clustering is not so simple without a hierarchical model.
-Since we never see all 61 neurons in a single worm, we need some means
-of combining information across worms in order to cluster their activity.
-The hierarchical LDS does exactly this.  We could consider simpler
-hierarchical models, e.g. a hierarchical factor analysis model that does
-not include assumptions about temporal dynamics; such a model would
-probably yield similar clusters.  The key point is that we must combine
-information across worms.
-
-\clearpage
-
-\subsection{Latent dynamics are hierarchical, robust, and recurrent}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=\textwidth]{figures/arhmm/dimensionality.pdf} 
-\caption{Training and test likelihood of the hierarchical
-  autoregressive hidden Markov models (a special case of the SLDS) for
-  all combinations of hierarchical, robust, and recurrent extensions.
-  We evaluate test likelihoods for a range of numbers of latent
-  states, and we find that the hierarchical, robust, recurrent model
-  with~$K=8$ latent states performs best.}
-\label{fig:arhmm_lls}
-\end{figure}
-
-As with the LDS, we fit autoregressive hidden Markov models (AR-HMMs)
-with all combinations of hierarchical, robust, and recurrent extensions
-described in Section~\ref{sec:extensions}.  For each possible model, we
-swept from~$K=2$ to~$K=20$ discrete latent states in increments of two.
-
-Overall, the hierarchical models (red, blue, yellow, green) performed
-best across all~$K$ in training data.  This is not surprising as the
-hierarchical models effectively have~$W \cdot K$ dynamics
-parameters---one version of each state for each worm---whereas the
-non-hierarchical models have only~$K$.  Of these hierarchical models,
-only those with the robust extension showed good generalization.  The
-non-robust models (with standard Gaussian noise; green, yellow)
-severely overfit the training data.  Of the robust hierarchical models,
-the recurrent model showed substantial improvement over the non-recurrent
-model, and obtained peak performance with~$K=8$ states.  Likelihoods
-are measured in units of nats/time step.
-
-\clearpage
-
-\subsection{The AR-HMM automatically segments low-dimensional states}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=2.7in]{figures/arhmm/x_3d_1.pdf}
-\includegraphics[width=2.7in]{figures/arhmm/x_3d_2.pdf}
-\\
-\includegraphics[width=2.7in]{figures/arhmm/x_3d_3.pdf}
-\includegraphics[width=2.7in]{figures/arhmm/x_3d_4.pdf}
-\\
-\includegraphics[width=2.7in]{figures/arhmm/x_3d_5.pdf} 
-\caption{Three dimensional projections of the 10 dimensional
-  latent states~$x$. The trajectories are color coded
-  according to the inferred discrete states~$z$.  We find
-  that discrete states correspond to different ``loops''
-  in latent space, and that they are, to large extent,
-  used in separate regions of~$x$-space.  This is consistent
-  with the recurrent model.}
-\end{figure}
-
-Since the continuous latent states are 10 dimensional, we use only the
-three dimensions of activity for visualization.  We choose the three
-dimensions that are most correlated with the observed neural activity.
-We color code these three-dimensional trajectories according to the
-discrete states~$z$, which correspond to different linear dynamical
-regimes of the continuous states~$x$. Compare these plots to the state
-space plots of~\citet{kato2015global}. These trajectories are not as
-smooth (which is expected given that we are modeling the first-order
-differences directly), but they capture the same qualitative patterns.
-We will quantify this correspondence shortly.
-
-\clearpage
-
-\subsection{The AR-HMM learns dynamical systems for each discrete state}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_0.pdf}
-\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_1.pdf}
-\\
-\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_2.pdf}
-\includegraphics[width=2.7in]{figures/arhmm/dynamics_123_3.pdf}
-\caption{Each discrete state has a corresponding set of linear dynamics
-  parameters~$(A_k, b_k)$, which we visualize here as a vector field.
-  Each arrow shows where a point~$x_t$ would be mapped to by the 
-  linear dynamics of a given state. We visualize points~$x_t$ where
-  the states were actually used. We see a number of states related
-  to ``returning to the origin'' (e.g. states 1, 2, 3, 5), as well
-  as states corresponding to loops (e.g. states 4, 6, 7).}
-\label{fig:dynamics}
-\end{figure}
-
-\begin{figure}[h]
-  \centering%
-  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_4.pdf}
-  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_5.pdf}
-  \\
-  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_6.pdf}
-  \includegraphics[width=2.7in]{figures/arhmm/dynamics_123_7.pdf}
-  \caption{See caption for Figure~\ref{fig:dynamics}}
-  \label{fig:dynamics2}
-\end{figure}
-Figures~\ref{fig:dynamics} and~\ref{fig:dynamics2} illustrate the
-global dynamics parameters for each of the~$K$ discrete states.
-Specifically, we show where a point~$x_t$ would be mapped to by the
-operation~$A_k x_t + b_k$.  We choose a subset of starting
-points~$x_t$ where the discrete state was used. Many of the states
-correspond to returning to the origin, which in turn corresponds to
-constant neural activity (since we are modeling the first-order
-differences in the calcium signal).  Other states correspond to loops,
-which give rise to increases in the activity of some clusters of
-neurons and decreases in others.
-
-\clearpage
-
-\subsection{Inferred states correspond to activation of clusters of neurons}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_0.pdf}
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_1.pdf}
-\caption{(top) Colored traces show a number of real-data trajectories
-  that were inferred to be in states 1 (blue) and 2 (red).  Then we
-  simulate from our model's inferred linear dynamical system for these
-  states, and show simulated trajectories with four different starting
-  points (circle, triangle, square, and pentagon) in black lines.
-  The top left panel shows the top 3 dimensions of these trajectories
-  in latent state space. The
-  top right panels show 2D views of the same trajectories.  (bottom)
-  These four simulated trajectories give rise to different patterns of
-  neural activity in the five inferred clusters.  For example, the
-  first trajectory of the second state (red circle) corresponds to
-  activity that decays toward the baseline for each of the clusters,
-  with a starting value of below-baseline activity in the first
-  cluster.}
-\label{fig:cluster_activity_12}
-\end{figure}
-
-Figures~\ref{fig:cluster_activity_12}, \ref{fig:cluster_activity_34},
-\ref{fig:cluster_activity_56}, and~\ref{fig:cluster_activity_78} show
-how trajectories in continuous latent state space for each of the~$K$
-states manifest in the activity of the neural clusters identified in
-Figure~\ref{fig:clusters}.  The black lines with circle, triangle,
-square, and pentagon starting points are forward simulated
-trajectories using the inferred dynamics of state~$k$ with four
-different initial conditions. The initial conditions are values
-of~$x_t$ where we had inferred an entry into state~$k$.
-The bottom rows show how the activity the forward
-simulated trajectories activate each of the five clusters of
-neurons.
-
-The colored lines in the top panel are segments of~$x$ that were
-inferred to be in state~$k$.  Recall that these continuous latent
-states were found by first fitting a hierarchical LDS. As such, they
-correspond to the conditional mean of the posterior
-distribution under the hierarchical LDS,~$\E[x \given y]$.
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_2.pdf}
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_3.pdf}
-\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
-\label{fig:cluster_activity_34}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_4.pdf}
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_5.pdf}
-\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
-\label{fig:cluster_activity_56}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_6.pdf}
-\includegraphics[width=5in]{figures/arhmm/cluster_activity_7.pdf}
-\caption{See Figure~\ref{fig:cluster_activity_12} caption.}
-\label{fig:cluster_activity_78}
-\end{figure}
-
-\clearpage
-
-\subsection{States are employed roughly equally by these worms}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=3in]{figures/arhmm/state_usage_per_worm2.pdf}
-\caption{Proportion of time frames spent in each state, broken down per-worm.}
-\label{fig:usage}
-\end{figure}
-
-The worms appear to use these states with roughly equal proportion,
-as shown in Figure~\ref{fig:usage}.  The states have been ordered to
-overall population usage, but within worms we see that some states
-are used slightly more often than others.  For example, worm 3 uses
-state 1 less often than the others, and instead seems to have a higher
-usage of state 3.  This particular difference is probably not important, but it
-points to how these types of methods could be useful for distinguishing
-between, for example, different genetic variants of worms.
-
-\clearpage
-
-\subsection{Transition structure is largely stereotyped across worms as well}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=5.5in]{figures/arhmm/trans_matrices.pdf}
-\caption{Empirical transition probabilities for each of the worms.  By far
-  the most common transition is into the same state.  For visualization, we
-  have removed this diagonal component and renormalized so that the~$k$-th row of
-  each transition matrix shows the conditional distribution over~$z_{t+1}$ given
-  that~$z_{t}=k$ and~$z_{t+1} \neq z_t$. }
-\label{fig:trans}
-\end{figure}
-
-Figure~\ref{fig:trans} shows the empirical transition matrices for each of the
-worms.  We see a few transitions appear across all worms:
-\begin{itemize}
-\item Pink transitions to red with high probability in all worms. This is evident
-  from the state trajectories above.
-\item Brown transitions to pink or, with lower probability, to purple.
-\item Orange transitions to purple, pink, or blue.
-\item Purple and orange transition to blue for most worms, except in worm three,
-  which uses yellow in place of blue (see above).
-\item Green transitions to brown with high probability across all worms. 
-\item For worms 2, 4, and 5, red often transitions into yellow and yellow to green or orange
-  for most worms.
-\item Blue (the ``origin'' state), tends to transition to purple or orange in
-  worms 2, 4, and 5.
-\end{itemize}
-
-These aren't meaningful yet, but as we show next, these discrete states closely
-match the states identified in~\citep{kato2015global}.
-
-\clearpage
-
-\subsection{Recurrent weights influence transition probabilities}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/recurrent_weights.pdf}
-\caption{Weights of the recurrent transition model.  The largest
-  weights come from the Markov terms for~$z_{t}$, but the weights
-  for the continuous location~$x_t$ are non-trivial.  When we
-  compute the effective weights that come from~$r_{z_t}$ and
-  the ``recurrent'' terms~$R^\trans x_t$, we see that the recurrent terms
-  are a significant fraction of the Markovian terms. }
-\label{fig:rec_weights}
-\end{figure}
-
-Figure~\ref{fig:rec_weights} shows the two terms that go into the
-distribution over next states: the ``Markov'' term~$\{r_{k}\}_{k=1}^K$
-and the ``recurrent'' weights~$R$, as defined in
-Eq.~\eqref{eq:recurrent}. The largest weights come from the
-self-transitions, i.e. the diagonal of the Markov terms. Still, the
-recurrent weights are non-negligible. The histogram of Markov
-weights~($r_{z_t}$) and the recurrent weights~($R^\trans x_t$) at the
-times of transitions (right panel) shows that the recurrent weights
-are comparable to the Markov weights, and therefore must play a role
-in determining transition probabilities.  This must be the case as the
-recurrent models yield significant improvement in held-out log
-likelihoods, as shown in Figure~\ref{fig:arhmm_lls}.
-
-\clearpage
-
-\subsection{Some state durations are non-Markovian}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=1.8in]{figures/arhmm/durations_0.pdf}
-\includegraphics[width=1.8in]{figures/arhmm/durations_1.pdf}
-\includegraphics[width=1.8in]{figures/arhmm/durations_2.pdf}
-\\
-\includegraphics[width=1.8in]{figures/arhmm/durations_3.pdf}
-\includegraphics[width=1.8in]{figures/arhmm/durations_4.pdf}
-\includegraphics[width=1.8in]{figures/arhmm/durations_5.pdf}
-\\
-\includegraphics[width=1.8in]{figures/arhmm/durations_6.pdf}
-\includegraphics[width=1.8in]{figures/arhmm/durations_7.pdf}
-\caption{Empirical distribution of state durations (histogram)
-  compared to the geometric distribution implied by the Markovian
-  weights of the transition model.  Some state durations are
-  well-approximated by the geometric distribution, but others
-  are peaked away from zero (e.g. states 2, 4, 6, and 7).  The
-  recurrent transition model allows the AR-HMM to capture this
-  non-Markovianity by modeling transitions as a function of location
-  in latent state space~$x_t$.}
-\label{fig:durations}
-\end{figure}
-
-Figure~\ref{fig:durations} shows histograms of the distribution of inferred
-state distributions for each state.  For comparison, we show
-the geometric distribution of state durations implied by the
-Markov weights as a black line.  Specifically, the distribution
-over durations~$d_k$ of state~$k$ under the Markov model (ignoring the
-recurrent weights) is given by,
-\begin{align*}
-  p_{k \to k} &= [\mathrm{softmax}(r_{k})]_k, \\
-  d_k &\sim \mathrm{Geom}(1-p_{k \to k}).
-\end{align*}
-We see that some states (e.g. states 1, 5, and 8) seem to follow the
-Markov assumption quite well, whereas others (states 2, 4, and especially 7),
-are far from Markovian.  The recurrent model incorporates the continuous
-latent state~$x_t$ into the transition probabilities via the additive term~$R x_t$
-in the softmax.  It is likely that this term captures the non-Markovianity.
-
-\clearpage
-
-\subsection{Inferred states overlap with those of~\citet{kato2015global}}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=1.9in]{figures/arhmm/overlap_0.pdf}
-\includegraphics[width=1.9in]{figures/arhmm/overlap_1.pdf}
-\includegraphics[width=1.9in]{figures/arhmm/overlap_2.pdf}
-\\
-\includegraphics[width=1.9in]{figures/arhmm/overlap_3.pdf}
-\includegraphics[width=1.9in]{figures/arhmm/overlap_4.pdf} 
-\caption{Overlap between the manually labeled states of~\citet{kato2015global}
-  and the inferred states of the AR-HMM. Each row is a normalized probability
-  distribution over which of the~$K$ discrete states each worm was inferred to
-  be in for the corresponding manual label.}
-\label{fig:overlap}
-\end{figure}
-
-Figure~\ref{fig:overlap} shows a close correspondence between manually
-labeled states and inferred discrete states, indicating that the
-switching linear dynamics assumption is a formal specification of the
-different dynamical modes reported by~\citet{kato2015global}.
-Specifically, different ``behaviors'' correspond to different linear
-dynamical regimes in continuous latent space.
-
-A few patterns emerge here:
-\begin{itemize}
-\item Worm 3 differs from the rest, and seems to have a separate
-  matching of inferred states and true states.  This list focuses
-  on patterns in Worms 1, 2, 4, and 5. 
-\item \textsf{REVSUS} --- the most common manually labeled state ---
-  tends to overlap with the red and yellow inferred states.
-\item \textsf{SLOW} --- the second most common --- and \textsf{FWD}
-  often map onto the blue state.
-\item \textsf{VT} (ventral turns) map to the green state, which
-  has a rise and fall of activity in cluster 3 (neurons
-  {\sf ALA, ASKL, ASKR, RIVL, RIVR, SMDVL} and \textsf{SMDVR}).
-  This seems consistent with the roles of sensory neurons~\textsf{ASK*},
-  interneurons \textsf{RIV*}, and motor neurons \textsf{SMDV*} in
-  local search behavior.
-\item Skipping ahead, \textsf{DT} (dorsal turns) often map onto
-  the orange and, to a much lesser extent, the purple state. Both
-  states encourage upswings in cluster 4 ({\sf SMBDL, SMBDR, SMBVL, SMBVR}, and \textsf{URYVR}).
-  The first four of these neurons are motor neurons involved in
-  local search behavior. 
-\item In addition to the blue state, \textsf{FWD} maps onto the
-  green, purple, and orange states as well, though much less significantly.
-\item \textsf{REV2} and \textsf{REV1} map primarily onto the pink state.
-  This state is characterized by an excursion of activity along cluster
-  1 ({\sf AVAL, AVAR, BAGL, RIML, RIMR, SAAVR, SABVL, SABVR, VA01}.
-  The \textsf{AVA*} interneurons innvervate motor neurons.
-  \textsf{RIM*} are motor neurons that modulate reversals.
-  \textsf{SABV*} are interneurons and sublateral motor neurons.
-  \textsf{VA01} is a motor neuron.
-
-  We also note that \textsf{REV1} maps onto the brown state, which
-  is also associated with ventral turns. Judging by the dynamics plots above,
-  it seems ventral turns are often followed by type-1 reversals.
-\item \textsf{NOSTATE} seems to often map in different ways across worms.
-  This is by far the least common of the manually-labeled states.
-\end{itemize}
-
-\todo[inline, size=\normalsize]{Combining these overlap figures
-  with the transition matrices above, I find a slightly different
-  transition structure from what was reported in Kato et al Figure 4E.
-  I find that the transitions between \textsf{REVSUS} and \textsf{FWD}
-  are largely mediated by the \textsf{SLOW} state without an intervening
-  turn.  I find that ventral turns are often exercised in between
-  \textsf{REVSUS} and \textsf{REV1}, and that dorsal turns are often
-  used between episodes of forward crawling.  Does this make any sense?
-  }
-
-\clearpage
-
-\subsection{Inferred changepoints also match manual labels}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
-\caption{(top) Posterior distribution over latent states~$z_t$.
-  Each column corresponds to a time bin and sums to one.
-  (middle) The posterior probability of changepoints.
-  (bottom) For reference,
-  the manually labeled states of~\citet{kato2015global}. Colors match those
-  of Figure~\ref{fig:overlap}.}
-\label{fig:z_cps_12}
-\end{figure}
-
-Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
-show the posterior distribution of latent states~$z$ inferred from
-neural activity alone. From this posterior, we also compute the probability
-of a changepoint, where~$z_{t+1} \neq z_t$.  For reference, we show the
-manually labeled states of~\citet{kato2015global} below. Here, patterns
-like the cascade from states 2 to 3 to 4 (red to yellow to green) are
-apparent.
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
-\caption{See Figure~\ref{fig:z_cps_12} caption.}
-\label{fig:z_cps_34}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
-\caption{See Figure~\ref{fig:z_cps_12} caption.}
-\label{fig:z_cps_5}
-\end{figure}
-
-\clearpage
-
-\subsection{Inferred states changepoints correspond with changes in dynamics}
-
-Figures~\ref{fig:segmentation_1} through~\ref{fig:segmentation_5} show an
-interval of latent state trajectories~$x$ along with the segmentation from
-the inferred discrete states~$z$ (top).  For comparison, we also show the
-manual segmentation of~\citep{kato2015global} (bottom).  The inferred
-segmentation seems to identify changepoints when one or more of the latent
-continuous states changes dynamics.  Indeed, this is the assumption of the
-model.  By contrast, the segmentation of~\citep{kato2015global} is based on
-a read-out of a single neuron, and while it is largely consistent with the
-inferred segmentation at the population level, it misses some salient
-changes in dynamics.  For example, see the inferred changepoint around
-time 700 in worm 1.  This changepoint was not identified in the manual
-data.
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_0.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_0.pdf}
-\caption{Latent state trajectories superimposed on the inferred segmentation (top)
-  and the manual segmentation (bottom).}
-\label{fig:segmentation_1}
-\end{figure}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_1.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_1.pdf}
-\caption{See caption of Figure~\ref{fig:segmentation_1}}
-\label{fig:segmentation_2}
-\end{figure}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_2.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_2.pdf}
-\caption{See caption of Figure~\ref{fig:segmentation_1}}
-\label{fig:segmentation_3}
-\end{figure}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_3.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_3.pdf}
-\caption{See caption of Figure~\ref{fig:segmentation_1}}
-\label{fig:segmentation_4}
-\end{figure}
-
-\begin{figure}[h]
-\centering%
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_4.pdf}
-\includegraphics[width=5.5in]{figures/arhmm/x_segmentation_zimmer_4.pdf}
-\caption{See caption of Figure~\ref{fig:segmentation_1}}
-\label{fig:segmentation_5}
-\end{figure}
-\clearpage
-
-\clearpage
-\bibliography{refs}
-\bibliographystyle{abbrvnat}
-
 \end{document}

### 2017-12-20 drafts/elife/figures/figure1/make_figure_1.py

@@ -1,0 +1,273 @@
+import numpy as np
+np.random.seed(0)
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
+
+import seaborn as sns
+sns.set_style("white")
+sns.set_context("paper")
+
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange"]
+
+colors = sns.xkcd_palette(color_names)
+
+W = 3    # number of worm panels
+N = 10   # number of neurons
+T = 50   # number of time bins
+D = 2    # dimension of latent states
+K = 3    # number of latent states
+
+
+def make_dynamics_library_2D():
+    # First state is return to zero
+    assert D == 2
+    As = np.zeros((K, D, D))
+    bs = np.zeros((K, D))
+
+    # Remaining states are rotations in a plane
+    ths = np.pi / 8 + np.pi / 8 * np.random.rand(K)
+    # ths *= (-1 + 2 * (np.arange(K) % 2 == 0))
+    # ths = np.pi / 8 * np.ones(K)
+
+    for k in range(K):
+        As[k] = np.array([
+            [np.cos(ths[k]), -np.sin(ths[k])],
+            [np.sin(ths[k]),  np.cos(ths[k])]
+        ])
+
+        phase = 2 * np.pi / K * k
+        fp = np.array(
+            [np.cos(phase),
+             np.sin(phase)])
+        bs[k] = (np.eye(D) - As[k]).dot(fp)
+
+    return As, bs
+
+
+def plot_vector_field_2d(k, As, bs,
+                         worm=0,
+                         ax=None,
+                         lims=(-3, 3),
+                         n_pts=7):
+    X, Y = np.meshgrid(np.linspace(lims[0], lims[1], n_pts),
+                     np.linspace(lims[0], lims[1], n_pts))
+    x = np.column_stack((X.ravel(), Y.ravel()))
+
+    Ai, bi = As[k], bs[k]
+    dxdt = x.dot(Ai.T) + bi - x
+
+    # Find the probability at each of the points
+    W = np.array([np.linalg.solve(np.eye(D) - A, b) for A, b in zip(As, bs)])
+    p = np.exp(x.dot(W.T))
[…]

### 2017-12-20 experiments/fit_arhmm.py

@@ -508,6 +508,7 @@ def plot_best_model_results(do_plot_expected_states=True,
         plot_duration_histogram(best_model.trans_distn,
                                 z_finals,
                                 durss,
+                                perm=perm,
                                 results_dir=fig_dir)
         plt.close("all")
 
@@ -559,8 +560,6 @@ if __name__ == "__main__":
     C_norm = C[:, :-1] / np.linalg.norm(C[:, :-1], axis=1)[:, None]
     C_clusters = np.array([C[neuron_clusters == c].mean(0) for c in range(N_clusters)])
     d_clusters = np.array([d[neuron_clusters == c].mean(0) for c in range(N_clusters)])
-
-
 
     # Set the AR-HMM hyperparameters
     ar_params = dict(nu_0=D_latent + 2,
@@ -614,17 +613,17 @@ if __name__ == "__main__":
     plot_best_model_results(
         do_plot_expected_states=False,
         do_plot_x_2d=False,
-        do_plot_x_3d=True,
+        do_plot_x_3d=False,
         do_plot_dynamics_3d=False,
         do_plot_dynamics_2d=False,
         do_plot_state_overlap=False,
         do_plot_state_usage=False,
-        do_plot_transition_matrices=False,
+        do_plot_transition_matrices=True,
         do_plot_simulated_trajs=False,
         do_plot_recurrent_weights=False,
         do_plot_x_at_changepoints=False,
         do_plot_latent_trajectories_vs_time=False,
-        do_plot_duration_histogram=True
+        do_plot_duration_histogram=False
     )
 
     # Rolling predictions

### 2017-12-20 experiments/fit_arhmm_nichols.py

@@ -499,6 +499,16 @@ def plot_best_model_results(do_plot_expected_states=True,
                                titles=[title],
                                filename=filename,
                                results_dir=fig_dir)
+
+        filename = condition + "overlap_total.pdf"
+        plot_state_overlap([np.concatenate([z_finals[i] for i in range(len(z_finals))])],
+                           [np.concatenate([z_trues[i][N_lags:] for i in range(len(z_finals))])],
+                           z_key=z_key,
+                           z_colors=zimmer_colors,
+                           titles=[title],
+                           filename=filename,
+                           results_dir=fig_dir)
+
         plt.close("all")
 
     if do_plot_state_usage:
@@ -564,6 +574,7 @@ def plot_best_model_results(do_plot_expected_states=True,
         plot_duration_histogram(best_model.trans_distn,
                                 z_finals,
                                 durss,
+                                perm=perm,
                                 results_dir=fig_dir)
         plt.close("all")
 

### 2017-12-20 experiments/fit_hlds.py

@@ -47,11 +47,11 @@ from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
 
 # IO
-# run_num = 2
-# results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
-# signal = "dff_diff"
-results_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
-signal = "dff_bc"
+run_num = 2
+results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
+signal = "dff_diff"
+# results_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
+# signal = "dff_bc"
 
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
@@ -83,7 +83,7 @@ def cached(results_name):
 
 def load_data(include_unnamed=True):
     # Load the data
-    worm_datas = [WormData(i, name="worm{}".format(i)) for i in range(N_worms)]
+    worm_datas = [WormData(i, name="worm{}".format(i), version="kato") for i in range(N_worms)]
 
     # Get the "true" discrete states as labeled by Zimmer
     z_trues = [wd.zimmer_states for wd in worm_datas]
@@ -302,6 +302,7 @@ def plot_best_model_results(best_model,
                             do_plot_sigmasq=True,
                             do_plot_similarity=True,
                             do_plot_cluster_embedding=True,
+                            do_plot_cluster_locations=True,
                             do_plot_data=True):
 
     # False color with the "true" label from Zimmer
@@ -488,6 +489,45 @@ def plot_best_model_results(best_model,
 
         plt.close("all")
         # plt.show()
+
+    if do_plot_cluster_locations:
+        import pandas
+        locs = pandas.read_csv("wormatlas_locations.csv").values
+        l_values = np.unique(locs[:,1])
+
+        fig = plt.figure(figsize=(1.5, 2.5))
+        ax = fig.add_subplot(111)
+        spc = 1
+
+        sizes = dict([(lv, 2) for lv in l_values])
+        for name in neuron_names:
+            i2 = np.where(locs[:, 0] == name)[0][0]
+            l = locs[i2, 1]
+            plt.plot(l, 0, 'ko', markersize=sizes[l], alpha=0.5)
+            sizes[l] += 1
+
+        ax.spines["left"].set_visible(False)
[…]

### 2017-12-20 zimmer/plotting.py

@@ -705,31 +705,7 @@ def plot_state_overlap(z_finals, z_trues,
     titles = ["State Overlap (Worm {})".format(i+1) for i in range(len(z_trues))] \
       if titles is None else titles
 
-    # Use the Hungarian algorithm to find a permutation of states that
-    # yields the highest overlap
-    overlap = np.zeros((K_zimmer, Kmax), dtype=float)
-    for z_true, z_inf in zip(z_trues, z_finals):
-        for k1 in range(K_zimmer):
-            for k2 in range(Kmax):
-                overlap[k1, k2] = np.sum((z_true == k1) & (z_inf == k2))
-
-    from scipy.optimize import linear_sum_assignment
-    _, perm = linear_sum_assignment(-overlap)
-
-    # Add any unused inferred states
-    if Kmax > K_zimmer:
-        unused = np.array(list(set(np.arange(Kmax)) - set(perm)))
-        perm = np.concatenate((perm, unused))
-
-    # Compare zimmer labels to inferred labels
-    for worm, (z_true, z_inf) in enumerate(zip(z_trues, z_finals)):
-        overlap = np.ones((K_zimmer, Kmax), dtype=float)
-        for k1 in range(K_zimmer):
-            for k2 in range(Kmax):
-                overlap[k1, k2] += np.sum((z_true == k1) & (z_inf == k2))
-
-        # Normalize the rows
-        overlap /= overlap.sum(1)[:, None]
+    def _plot_overlap(overlap, perm, filename, title):
 
         fig = plt.figure(figsize=(2., 2))
         ax1 = create_axis_at_location(fig, .5, .5, 1., 1)
@@ -741,7 +717,7 @@ def plot_state_overlap(z_finals, z_trues,
             ax1.set_yticks(np.arange(K_zimmer))
             ax1.set_yticklabels(z_key, fontdict=dict(size=6))
             ax1.tick_params(axis='y', which='major', pad=11)
-        ax1.set_title(titles[worm])
+        ax1.set_title(title)
 
         lax = create_axis_at_location(fig, .4, .5, .06, 1)
         lax.imshow(np.arange(K_zimmer)[:, None], cmap=gradient_cmap(z_colors[:K_zimmer]), interpolation="nearest",
@@ -762,10 +738,73 @@ def plot_state_overlap(z_finals, z_trues,
         plt.colorbar(im, cax=axcb)
 
         if results_dir is not None:
[…]

### 2017-12-20 zimmer/plotting.py
add color code to state usage

@@ -770,41 +770,6 @@ def plot_state_overlap(z_finals, z_trues,
         # Normalize the rows
         overlap /= overlap.sum(1)[:, None]
         _plot_overlap(overlap, perm, "overlap_{}.pdf".format(worm), titles[worm])
-        #     fig = plt.figure(figsize=(2., 2))
-    #     ax1 = create_axis_at_location(fig, .5, .5, 1., 1)
-    #     im = ax1.imshow(overlap[:, perm], vmin=0, vmax=1.0, cmap="Greys", interpolation="nearest", aspect="auto")
-    #     ax1.set_xticks([])
-    #     if z_key is None:
-    #         ax1.set_yticks([])
-    #     else:
-    #         ax1.set_yticks(np.arange(K_zimmer))
-    #         ax1.set_yticklabels(z_key, fontdict=dict(size=6))
-    #         ax1.tick_params(axis='y', which='major', pad=11)
-    #     ax1.set_title(titles[worm])
-    #
-    #     lax = create_axis_at_location(fig, .4, .5, .06, 1)
-    #     lax.imshow(np.arange(K_zimmer)[:, None], cmap=gradient_cmap(z_colors[:K_zimmer]), interpolation="nearest",
-    #                aspect="auto")
-    #     lax.set_xticks([])
-    #     lax.set_yticks([])
-    #
-    #     if z_key is None:
-    #         lax.set_ylabel("Zimmer State", fontsize=8)
-    #
-    #     bax = create_axis_at_location(fig, .5, .4, 1., .06)
-    #     bax.imshow(np.arange(Kmax)[perm][None, :], cmap=gradient_cmap(colors[:Kmax]), interpolation="nearest", aspect="auto")
-    #     bax.set_xticks([])
-    #     bax.set_yticks([])
-    #     bax.set_xlabel("Inferred State", fontsize=8)
-    #
-    #     axcb = create_axis_at_location(fig, 1.55, .5, .1, 1)
-    #     plt.colorbar(im, cax=axcb)
-    #
-    #     if results_dir is not None:
-    #         filename = "overlap_{}.pdf".format(worm) if filename is None else filename
-    #         plt.savefig(os.path.join(results_dir, filename))
-    #
-    #     plt.close("all")
 
 
 def plot_state_usage_by_worm(z_finals,
@@ -876,6 +841,13 @@ def plot_state_usage_by_worm_matrix(z_finals,
     # ax.set_xticks(np.arange(N_worms))
[…]

### 2017-12-20 drafts/elife/draft.tex
adding figures for elife version of paper
some text for figure 1

@@
-\documentclass{article}
+\documentclass[11pt]{article}
 
@@
 
-\subsection*{Hierarchical models pool information while allowing individual variability}
-% Introduce the probabilistic model
+\subsection*{Hierarchical state space models combine information across
+  trials to learn canonical dynamics}
+
+% Goal: learn dynamical model of neural activity
+% Challenge 1: how to capture nonlinear dynamics of neural activity with simple model? 
+% Solution 1: recurrent state space models
+% Challenge 2: how to combine information across trials?
+% Solution 2: hierarchical model
+
+We seek a parsimonious characterization of the dynamics of neural
+activity in the head ganglia of \celegans. Recent advances in optical
+imaging offer simultaneous recordings of most of the head ganglia
+neurons, but we must surmount two main challenges in order to extract
+a dynamics model from these data.  We need to design a model that is
+flexible enough to capture nonlinear dynamics yet interpretable enough
+to reveal simplifying structure when it exists. Moreover, our model
+must accept partial recordings from separate organisms to learn shared
+neural dynamics yet still allow for individual variability. We
+construct a hierarchical state space model that satisfies both
+criteria, decomposing complex dynamics into a collection of simpler
+pieces that can be learned from noisy, partial recordings with
+trial-to-trial variability.
+
+We study the data presented in~\citet{kato2015global} and
+\citet{nichols2017global}, which consist of a collection of optical
+recordings of calcium fluorescence in immobilized \celegans. The
+recordings are 18 minutes long with a frame rate of about 3Hz.  On
+average, about 100 single units are extracted from each
+recording. \celegans~is genetically stereotyped and each of its 302
+neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
+units could be unambiguously labeled in each recording.
+
+To model this data, we build on switching linear dynamical systems
+(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
+  ghahramani1996switching, murphy1998switching}, a type of state space
+model that views data as a projection of a low-dimensional latent
+state that switches between linear dynamical regimes.  Formally,
+let~$y_t \in \reals^N$ denote the vector of neural activity observed
+at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
+where~$x_t \in \reals^D$ is a continuous latent state
+and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
+reflecting the low-dimensional nature of neural activity. SLDS
+simplify the dynamics of the continuous latent states by introducing a
+discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
+current dynamical regime.  Each discrete regime~$k$ is associated with
+a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
+specifies which map to use in order to propagate~$x_t$ to the next
+time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
+between different linear dynamical regimes, the SLDS captures highly
+nonlinear dynamics.  We seek to infer the continuous and discrete
+latent states and simultaneously learn the linear maps, the
+distribution of the noise in the observations and the continuous
+states, and the dynamics that govern how the discrete latent states
+evolve.
+
+% Recurrent model
+
+% Robust model
+
+% Hierarchical model
+
+% Describe figure 1
 
@@
   embedding. }
-\label{fig:hslds}
+\label{fig:model}
 \end{figure}
@@
 
-\subsection*{AR-HMM's automatically parse data into discrete states with simple dynamics}
+\begin{figure}[t]
+\centering%
+\includegraphics[width=6in]{figures/figure2} 
+\caption{}
+\label{fig:emissions}
+\end{figure}
+
+\subsection*{SLDS automatically parse data into discrete states with simple dynamics}
+
+\begin{figure}[t]
+\centering%
+\includegraphics[width=6in]{figures/figure3} 
+\caption{}
+\label{fig:syllables}
+\end{figure}
 
@@
 
+\begin{figure}[t]
+\centering%
+\includegraphics[width=6in]{figures/figure4} 
+\caption{}
+\label{fig:recurrent}
+\end{figure}
+
 \subsection*{Oxygen level modulates transition probabilities but not dynamics or emissions}
+
+\begin{figure}[t]
+\centering%
+\includegraphics[width=6in]{figures/figure5} 
+\caption{}
+\label{fig:o2}
+\end{figure}
 

### 2017-12-21 drafts/elife/draft.tex
adding notebooks etc

@@
 
-\subsection*{Hierarchical state space models combine information across
+\subsection*{A probabilistic framework that combines information across
   trials to learn canonical dynamics}
 
-% Goal: learn dynamical model of neural activity
-% Challenge 1: how to capture nonlinear dynamics of neural activity with simple model? 
-% Solution 1: recurrent state space models
-% Challenge 2: how to combine information across trials?
-% Solution 2: hierarchical model
-
-We seek a parsimonious characterization of the dynamics of neural
-activity in the head ganglia of \celegans. Recent advances in optical
-imaging offer simultaneous recordings of most of the head ganglia
-neurons, but we must surmount two main challenges in order to extract
-a dynamics model from these data.  We need to design a model that is
-flexible enough to capture nonlinear dynamics yet interpretable enough
-to reveal simplifying structure when it exists. Moreover, our model
-must accept partial recordings from separate organisms to learn shared
-neural dynamics yet still allow for individual variability. We
-construct a hierarchical state space model that satisfies both
-criteria, decomposing complex dynamics into a collection of simpler
-pieces that can be learned from noisy, partial recordings with
-trial-to-trial variability.
-
-We study the data presented in~\citet{kato2015global} and
-\citet{nichols2017global}, which consist of a collection of optical
-recordings of calcium fluorescence in immobilized \celegans. The
-recordings are 18 minutes long with a frame rate of about 3Hz.  On
-average, about 100 single units are extracted from each
-recording. \celegans~is genetically stereotyped and each of its 302
-neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
-units could be unambiguously labeled in each recording.
-
-To model this data, we build on switching linear dynamical systems
-(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
-  ghahramani1996switching, murphy1998switching}, a type of state space
-model that views data as a projection of a low-dimensional latent
-state that switches between linear dynamical regimes.  Formally,
-let~$y_t \in \reals^N$ denote the vector of neural activity observed
-at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
-where~$x_t \in \reals^D$ is a continuous latent state
-and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
-reflecting the low-dimensional nature of neural activity. SLDS
-simplify the dynamics of the continuous latent states by introducing a
-discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
-current dynamical regime.  Each discrete regime~$k$ is associated with
-a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
-specifies which map to use in order to propagate~$x_t$ to the next
-time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
-between different linear dynamical regimes, the SLDS captures highly
-nonlinear dynamics.  We seek to infer the continuous and discrete
-latent states and simultaneously learn the linear maps, the
-distribution of the noise in the observations and the continuous
-states, and the dynamics that govern how the discrete latent states
-evolve.
-
-% Recurrent model
-
-% Robust model
-
-% Hierarchical model
-
-% Describe figure 1
-
-\begin{figure}[t]
+\begin{figure}[ht]
 \centering%
@@
 \end{figure}
+
+We seek a parsimonious characterization of the dynamics of neural
+activity in the head ganglia of \celegans. Recent advances in optical
+imaging offer simultaneous recordings of most of the head ganglia
+neurons, but we must surmount three main challenges in order to learn
+a dynamics model from these data.  First, we need to design a model
+flexible enough to capture nonlinear dynamics yet interpretable enough
+to reveal simplifying structure when it exists.  Moreover, any model
+will necessarily be an approximation to the true neural dynamics, and
+as such we need to be robust to data that does not quite match model
+predictions.  Finally, our model must combine partial recordings from
+separate organisms to learn shared neural dynamics yet still allow for
+individual variability. We construct a recurrent, robust, and
+hierarchical state space model that addresses these challenges,
+decomposing complex dynamics into a collection of simpler pieces that
+can be learned from noisy, partial recordings with trial-to-trial
+variability.
+
+We study the data presented in~\citet{kato2015global} and
+\citet{nichols2017global}, which consist of a collection of optical
+recordings of calcium fluorescence in immobilized \celegans. The
+recordings are 18 minutes long with a frame rate of about 3Hz.  On
+average, about 100 single units are extracted from each
+recording. \celegans~is genetically stereotyped and each of its 302
+neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
+units could be unambiguously labeled in each
+recording. Fig.~\ref{fig:model}a illustrates this type of data. The
+three panels represent recordings from three separate worms, and the
+rows correspond to different neuron labels.  Neural activity is
+measured for only a subset of neurons in each worm.
+
+% To model this data, we build on switching linear dynamical systems
+% (SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
+%   ghahramani1996switching, murphy1998switching}, a type of state space
+% model that views data as a projection of a low-dimensional latent
+% state that switches between linear dynamical regimes.  Formally,
+% let~$y_t \in \reals^N$ denote the vector of neural activity observed
+% at time~$t$.  The expected activity is given by,~$\E[y_t] = g(x_t)$,
+% where~$x_t \in \reals^D$ is a continuous latent state
+% and~$g: \reals^D \to \reals^N$ is a linear map. We assume~$D \ll N$,
+% reflecting the low-dimensional nature of neural activity. SLDS
+% simplify the dynamics of the continuous latent states by introducing a
+% discrete latent state~$z_t \in \{1, \ldots, K\}$ that indicates the
+% current dynamical regime.  Each discrete regime~$k$ is associated with
+% a linear map~$f_k: \reals^D \to \reals^D$. The discrete state~$z_t$
+% specifies which map to use in order to propagate~$x_t$ to the next
+% time step.  Specifically,~$\E[x_{t+1}] = f_{z_t}(x_t)$. By switching
+% between different linear dynamical regimes, the SLDS captures highly
+% nonlinear dynamics.  We seek to infer the continuous and discrete
+% latent states and simultaneously learn the linear maps, the
+% distribution of the noise in the observations and the continuous
+% states, and the dynamics that govern how the discrete latent states
+% evolve.
+To model this data, we build on switching linear dynamical systems
+(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
+  ghahramani1996switching, murphy1998switching}, a type of state space
+model that views data as a projection of a low-dimensional latent
+state that switches between linear dynamical regimes.  The expected
+activity is a linear function of an underlying, continuous latent
+state, illustrated as two-dimensional time series for each worm in
+Fig.~\ref{fig:model}b.  We typically assume this latent state is lower
+dimensional than the number of observed neurons. SLDS simplify the
+dynamics of the continuous latent states by introducing discrete
+latent states that indicate the current dynamical regime
+(Fig.~\ref{fig:model}c).  Each discrete regime is associated with a
+linear map (Fig.~\ref{fig:model}d), and the discrete state specifies
+which map to use in order to propagate the continuous latent state to
+the next time step.  By switching between different linear dynamical
+regimes, the SLDS captures highly nonlinear dynamics.  We seek to
+infer the continuous and discrete latent states and simultaneously
+learn the linear maps, the distribution of the noise in the
+observations and the continuous states, and the dynamics that govern
+how the discrete latent states evolve.
+
+% Recurrent model
+Standard SLDS treat the discrete latent states as a simple Markov
+process: the next state depends only on the current state. However,
+this neglects the more nuanced ways in which discrete and continuous
+states may co-evolve. For example, the next discrete state may depend
+on both the current discrete state and the current location in
+continuous latent space; intuitively, different dynamical regimes may
+be employed in different regions of continuous latent space. These
+types of dependencies are captured by what are variously called
+hybrid~\citep{paoletti2007identification},
+augmented~\citep{barber2006expectation}, or \emph{recurrent} state
+space models~\citep{linderman2017recurrent}. The maps in
+Fig.~\ref{fig:model}d illustrate how these dependencies may modulate
+the probability of different dynamics depending on the current
+location in continuous latent space. We employ these recurrent models
+to learn dynamical regimes that are selectively employed as a
+function of current continuous brain state.
+
+% Robust model
+While recurrent SLDS can approximate highly nonlinear systems, they
+are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
+ameliorate this concern by lessening the penalty on observations that
+do not quite match the dynamics of the current regime.  We achieve
+this robustness by replacing the standard Gaussian noise model with a
+multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
+of this distribution permit outliers and account for some of the model
+misspecification.
+
+% Hierarchical model
+The third component of our framework is a \emph{hierarchical} model
+for learning canonical dynamics from partial recordings collected from
+many worms.  Since each recording contains a different subset of
+labeled neurons, our first step is to align the data, treating the
+activity of neurons that were not found in a given recording as
+missing data that must be inferred.  We handle this by integrating
+over possible activity of the missing neurons, analytically
+marginalizing out this missing data.  The second challenge is
+combining information from many recordings while also allowing for
+worm-to-worm differences.  For example, the neural activitiy may have
+larger variance in some recordings than in others due to different
+levels of expression of the fluorescent protein.  We allow for this by
+having a separate variance for each worm and neuron, as shown in
+Fig.~\ref{fig:model}f.  More importantly, the dynamics of neural
+activity may be slightly different from worm to worm.  We account for
+this by introducing a canonical library of dynamical regimes
+(Fig.~\ref{fig:model}e) that are shared by all worms as well as a
+unique library for each individual worm (Fig.~\ref{fig:model}d);
+however, we constrain the per-worm dynamics to be close to the
+canonical dynamics.
 

### 2017-12-21 drafts/elife/figures/figure1/make_figure_1.py

@@ -215,12 +215,9 @@ def plot_neural_activity():
         fig = plt.figure(figsize=(1.5, 2))
         ax = fig.add_subplot(111)
 
-        mask = np.zeros(N, dtype=bool)
-        mask[np.random.choice(N, size=6, replace=False)] = True
-
         for n in range(N):
             ax.plot(np.arange(T), n + np.zeros(T), '-k', lw=0.5)
-            if mask[n]:
+            if masks[w, n]:
                 ax.plot(np.arange(T), n + ys[w][:,n], color=colors[3])
 
         ax.set_ylim(N, -1)
@@ -232,6 +229,32 @@ def plot_neural_activity():
         plt.savefig("fig1_y{}.pdf".format(w + 1))
 
     plt.close("all")
+
+
+def plot_observation_variance():
+    sigmasq = np.random.gamma(shape=1, scale=1, size=(W, N))
+    sigmasq[~masks] = np.nan
+
+    from hips.plotting.colormaps import gradient_cmap
+    cmap = gradient_cmap([np.ones(3), colors[0]])
+    cmap.set_bad(0.7 * np.ones(3))
+
+    fig = plt.figure(figsize=(1.5, .5))
+    ax = fig.add_subplot(111)
+
+    im = ax.imshow(sigmasq, vmin=0, aspect="auto", cmap=cmap)
+    ax.set_xticks([])
+    ax.set_xlabel("neuron")
+    ax.set_yticks([])
+    ax.set_ylabel("worm")
+
+    ax.set_title("observation variance")
+
+    plt.colorbar(im, ticks=[0, 2.5])
+
+    plt.tight_layout(pad=0.1)
+    plt.savefig("fig1_variance.pdf")
+    plt.show()
 
 
 if __name__ == "__main__":
@@ -251,23 +274,27 @@ if __name__ == "__main__":
     # simulate observations
     sigma = 0.1
     ys = [x.dot(C.T) + sigma * np.random.randn(T, N) for x in xs]
+    masks = np.zeros((W, N), dtype=bool)
+    for w in range(W):
+        masks[w, np.random.choice(N, size=6, replace=False)] = True
+
     for y in ys:
         y /= (2.1 * np.max(abs(y)))
 
-    # Plot the canonical dynamics
-    for k in range(K):
-        plot_vector_field_2d(k, As, bs, worm=-1)
-    plt.close("all")
-
-    # Plot the individual worm dynamics
-    for k in range(K):
-        for w in range(W):
[…]

### 2017-12-21 drafts/elife/draft.tex
emission model results

@@
 
-\begin{figure}[ht]
-\centering%
-\includegraphics[width=6in]{figures/figure1} 
-\caption{A schematic of the hierarchical model for combining
-  information across multiple worms in order to learn a shared
-  dynamical system for \celegans.  \textbf{a.}~All worms share a
-  canonical dynamics library. Here, we have three dynamics functions
-  (blue, red, gold), each of which is a simple linear mapping from
-  $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
-  assigns probabilities to the different dynamics functions based on
-  the current location in $x$-space.  \textbf{b.}~The worms also share
-  an emission matrix, $C$, which can be viewed as an embedding of
-  neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
-  library, which allows for slight deviations from the canonical
-  dynamics.  \textbf{d.}~The model segments neural activity into a
-  sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
-  into the dynamics library.  \textbf{e.}~The discrete states specify
-  the dynamics that are used to propagate the continuous latent
-  states~$x_t$ forward in time. Here, the two dimensional latent
-  states are colored according to the instantaneous discrete
-  state. \textbf{f.}~The observed neural activity is a linear
-  projection of the underlying continuous latent states.  On any given
-  trial, we observe only a fraction of all the neurons.  The
-  hierarchical model combines these partial recordings to infer the
-  underlying states, the dynamics libraries, and the shared neural
-  embedding. }
-\label{fig:model}
-\end{figure}
-
 We seek a parsimonious characterization of the dynamics of neural
@@
 measured for only a subset of neurons in each worm.
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/figure1} 
+\caption{
+  % A schematic of the hierarchical model for combining
+  % information across multiple worms in order to learn a shared
+  % dynamical system for \celegans.  \textbf{a.}~All worms share a
+  % canonical dynamics library. Here, we have three dynamics functions
+  % (blue, red, gold), each of which is a simple linear mapping from
+  % $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
+  % assigns probabilities to the different dynamics functions based on
+  % the current location in $x$-space.  \textbf{b.}~The worms also share
+  % an emission matrix, $C$, which can be viewed as an embedding of
+  % neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
+  % library, which allows for slight deviations from the canonical
+  % dynamics.  \textbf{d.}~The model segments neural activity into a
+  % sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
+  % into the dynamics library.  \textbf{e.}~The discrete states specify
+  % the dynamics that are used to propagate the continuous latent
+  % states~$x_t$ forward in time. Here, the two dimensional latent
+  % states are colored according to the instantaneous discrete
+  % state. \textbf{f.}~The observed neural activity is a linear
+  % projection of the underlying continuous latent states.  On any given
+  % trial, we observe only a fraction of all the neurons.  The
+  % hierarchical model combines these partial recordings to infer the
+  % underlying states, the dynamics libraries, and the shared neural
+  % embedding.
+}
+\label{fig:model}
+\end{figure}
 
@@
 multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
-of this distribution permit outliers and account for some of the model
-misspecification.
+of this distribution Fig.~\ref{fig:model}e compares a multivariate-t
+distribution to a Gaussian with the same mean; though they have
+the same covariance shape, the tails of the robust model fall off
+more slowly than those of the Gaussian, allowing for outliers and a
+degree of model misspecification.
 
@@
 having a separate variance for each worm and neuron, as shown in
-Fig.~\ref{fig:model}f.  More importantly, the dynamics of neural
+Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
 activity may be slightly different from worm to worm.  We account for
 this by introducing a canonical library of dynamical regimes
-(Fig.~\ref{fig:model}e) that are shared by all worms as well as a
+(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
 unique library for each individual worm (Fig.~\ref{fig:model}d);
@@
 
-\subsection*{Emission models reveal clusters of neurons and aid in identification}
+% Inference
+We fit the the canonical and per-worm dynamics libraries, the
+parameters of the dynamics and observation noise models, and the
+discrete and continuous latent states with approximate Bayesian
+inference as described in the Supplementary Material. Briefly, we
+separate the problem into two steps: learning the continuous latent
+states and the mapping from continuous latent states to observations,
+then learning the dynamics libraries that govern the continuous latent
+states and the correspdonding discrete state sequences.  Both steps are
+performed with a block Gibbs sampler, leveraging the structure of the
+probabilistic model to perform efficient conditional updates of the
+various parameters.
 
-\begin{figure}[t]
+\subsection*{Emission model aids in identification and reveals clusters of neurons}
+
+\begin{figure}[t!]
 \centering%
@@
 \end{figure}
+
+The parameters of the learned model offer novel insights into the
+structure of neural activity in the head ganglia that can only be
+obtained by pooling information across many partial recordings.  With
+this aggregated information, we can smooth noisy observations and
+predict the activity of unobserved neurons. \citet{kato2015global}
+used a smoothing procedure to estimate temporal principal components.
+By contrast, we fit our model directly to the first-order differences
+in calcium fluorescence; our only preprocessing step is to correct for
+photo-bleaching. Fig~\ref{fig:emissions}a show our results for the first
+of the five worms in this dataset. We see that the model smooths out
+noise in the observed activity and, moreover, makes predictions about
+the activity of neurons that could not be identified in this worm.
+
+The predicted activity is based on the correlations between neurons
+learned from many recordings.  These predictions are a useful aid for
+identifying unlabeled neurons.  For example, we see that
+\textsf{SMBDL} was not found in the first worm
+(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
+neurons closely matches the model's predictions for \textsf{SMBDL},
+then this provides some evidence in favor of assigning it this
+label. We tested this idea by witholding the labels of some
+confidently-identified neurons and comparing their activity to the
+model's predicted activity for all unlabeled neurons.  We measured the
+similarity between the actual and predicted activity and found that
+correct label, or ``candidate,'' tends to be much more similar than
+other incorrect candidates, as summarized in
+Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
+the location of the neuron, this provides a powerful aid for identifying
+neurons.  The more neurons that we can identify, the better we will be
+able to learn about the dynamics of neural activity.
+
+Even though each individual recording contains only a subset of neurons,
+the hierarchical model can still estimate the dimensionality of
+neural activity in the entire head ganglia.  We reserve the last 20\% of
+the data (3.6 min) for testing and estimate the dimensionality
+of neural activity based on the marginal log likelihood the model assigns
+to this held-out data. We find that the test likelihood peaks with a
+ten-dimensional latent state (Fig.~\ref{fig:emissions}c).  Moreover, the likelihood is substantially
+higher in the model that includes per-worm observation variances
+compared to the model with shared variances for all worms. Fig.~\ref{fig:emissions}d
+shows the inferred variance for each worm in the \citet{kato2015global}
+dataset.  Neurons that were not identified are marked with a gray box.
+This shows the extent of the missing data and the need for hierarchical
+modeling.
+
+The linear mapping from latent states to observations reveals clusters
+of similarly tuned neurons.  The mapping maybe viewed as either an $N$
+(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
+vectors in $D$-dimensional space, i.e. as an embedding of neurons.
+These two views are shown in Fig.~\ref{fig:emissions}d. We note that
+many vectors point in similar directions, which implies that there are
+groups of correlated neurons.  We clustered the neurons based on the
+cosine similarity between their embedding vecors and found five clusters
+of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
+as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
+looked at their spatial distribution along the head-to-tail axis of the
+worm's body.  We found that clusters 2-4 correspond to activity closer to
+the head and working backward toward the tail.  The last and largest cluster
+contains many neurons in the nerve ring (location 0.15) \todo{check} as well
+as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}
 

### 2017-12-21 drafts/elife/figures/figure1/make_figure_1.py

@@ -257,6 +257,83 @@ def plot_observation_variance():
     plt.show()
 
 
+def plot_noise_density(lims=(-3, 3), n_pts=50):
+
+    X, Y = np.meshgrid(np.linspace(lims[0], lims[1], n_pts),
+                     np.linspace(lims[0], lims[1], n_pts))
+    x = np.column_stack((X.ravel(), Y.ravel()))
+
+    mu = np.array([1, 1])
+    sigma = np.array([[1, 0.5],
+                      [0.5, 1]])
+    sigmainv = np.linalg.inv(sigma)
+
+    from scipy.stats import multivariate_normal
+    p_mvn = multivariate_normal(mu, sigma).pdf(x)
+    lp_mvn = multivariate_normal(mu, sigma).logpdf(x)
+
+    # Compute multivariate t density
+    nu = 4
+    D = 2
+    from scipy.special import gammaln
+    lp_mvt = np.zeros(x.shape[0])
+    lp_mvt += gammaln(0.5 * (nu + D))
+    lp_mvt -= (gammaln(0.5 * nu) + 0.5 * D * np.log(nu) + 0.5 * D * np.log(np.pi))
+    lp_mvt -= 0.5 * np.linalg.slogdet(sigma)[1]
+    lp_mvt += -0.5 * (nu + D) * np.log(1 + 1./nu * ((x-mu).dot(sigmainv) * (x-mu)).sum(1))
+    p_mvt = np.exp(lp_mvt)
+
+    dx = (lims[1] - lims[0]) / (n_pts - 1)
+    print(np.sum(p_mvn * dx**2))
+    print(np.sum(p_mvt * dx**2))
+
+    # Get limits
+    # p_max = max(p_mvn.max(), p_mvt.max())
+    # levels = np.linspace(0, p_max, 8)[1:]
+
+    lp_min = min(lp_mvn.min(), lp_mvt.min())
+    lp_max = max(lp_mvn.max(), lp_mvt.max())
+    levels = np.linspace(lp_min, lp_max, 15)[1:]
+    # import ipdb; ipdb.set_trace()
+
+
+    fig = plt.figure(figsize=(1.25, .5))
+    ax = fig.add_subplot(121, aspect="equal")
+
+    from hips.plotting.colormaps import white_to_color_cmap
+    cmap = white_to_color_cmap(colors[3])
+    ax.contourf(X, Y, lp_mvn.reshape(X.shape), levels, cmap=cmap)
[…]

### 2017-12-21 experiments/fit_hlds.py

@@ -392,34 +392,48 @@ def plot_best_model_results(best_model,
 
         plt.show()
 
-    # if do_plot_cluster_embedding:
-    #     for cluster in range(N_clusters):
-    #         fig = plt.figure(figsize=(1.5, 1.5))
-    #         ax = fig.add_subplot(111, projection="3d")
-    #         for i in np.where(neuron_clusters == cluster)[0]:
-    #             ci = C[i]
-    #             ci /= np.linalg.norm(ci)
-    #             ax.plot([0, ci[0]],
-    #                      [0, ci[1]],
-    #                      [0, ci[2]],
-    #                      '-k', lw=1)
-    #             ax.plot([ci[0]],
-    #                     [ci[1]],
-    #                     [ci[2]],
-    #                     'ok', markersize=4)
-    #             # ax.text(1.1 * ci[0], 1.1 * ci[1], 1.1 * ci[2], neuron_names[i], fontsize=6)
-    #
-    #         ax.set_xlabel("$x_1$", labelpad=-10)
-    #         ax.set_ylabel("$x_2$", labelpad=-10)
-    #         ax.set_zlabel("$x_3$", labelpad=-10)
-    #         ax.set_xticklabels([])
-    #         ax.set_yticklabels([])
-    #         ax.set_zticklabels([])
-    #         ax.set_xlim(-.75, .75)
-    #         ax.set_ylim(-.75, .75)
-    #         ax.set_zlim(-.75, .75)
-    #
-    #     plt.show()
+    if do_plot_cluster_embedding:
+
+        fig = plt.figure(figsize=(.5, .5))
+        ax = fig.add_subplot(111, projection="3d")
+        for cluster in range(N_clusters):
+            for i in np.where(neuron_clusters == cluster)[0]:
+                ci = C[i]
+                ci /= np.linalg.norm(ci)
+                ax.plot([0, ci[0]], [0, ci[1]], [0, ci[2]],
+                         '-k', lw=.5)
+                ax.plot([ci[0]], [ci[1]], [ci[2]],
+                        'ok', markersize=1)
+                # ax.text(1.1 * ci[0], 1.1 * ci[1], 1.1 * ci[2], neuron_names[i], fontsize=6)
+
+            # ax.set_xlabel("$x_1$", labelpad=-10)
+            # ax.set_ylabel("$x_2$", labelpad=-10)
+            # ax.set_zlabel("$x_3$", labelpad=-10)
+            ax.set_xticklabels([])
+            ax.set_yticklabels([])
[…]

### 2017-12-21 drafts/elife/draft.tex
adding results on discrete states

@@
 
+Having learned a mapping from latent state space to observed activity,
+we can now reason about the dynamics of these continuous latent
+states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
+worm's neural activity in the first three dimensions of the continuous
+state space, color coded by the discrete regime inferred under the
+SLDS.  We see that different regimes correspond to distinct loops
+through latent space.  Compare this with Fig. 4b
+of~\citet{kato2015global}---they achieved a similar result by manually
+segmenting smoothed trajectories in the space spanned by the first
+three principal components.  In contrast, our probabilistic framework
+is automatic, operates on ten-dimensional latent states, and requires
+no additional smoothing.
+
+Moreover, we can use the likelihood of held-out test data to determine
+the number of discrete regimes and assess the value of the robust and
+recurrent model extensions, as shown in Fig.~\ref{fig:syllables}b.
+While the non-robust models with Gaussian dynamics noise achieve
+higher training likelihood, the robust models generalize better to
+test data.  Moreover, the recurrent models provide further improvement
+over the standard Markovian models.  We find that test likelihood is
+maximized with eight latent states, which matches the number manually
+chosen by~\citet{kato2015global}.
+
+Not only does the SLDS find the same number of states, it also finds a
+close correspondence between the manually labeled and inferred states.
+Fig.~\ref{fig:syllables}c shows ten-dimensional latent states and the
+manual and inferred segmentations for a 5.5 minute section of data.
+The manual and inferred states often change at similar times. We also
+see instances where there appears to be a change in continuous state
+dynamics that are identified by the SLDS but missed in the manual
+segmentation (e.g. around $t=700$ seconds).  Fig.~\ref{fig:syllables}d
+quantifies the overlap in states for the first worm and for all worms.
+Many inferred states are in one-to-one correspondence with manually
+labeled states, and the remainder tend to be split between two manual
+states.  This confirms that the SLDS is a useful tool for automatically
+segmenting brain data.
+
+Beyond simply automating a manual process, the SLDS also provides a
+quantitative description of the different states.  In particular, each
+state is associated with a set of linear dynamics.
+Fig.~\ref{fig:syllables}e shows the canonical dynamics associated with
+each of the eight discrete states. The arrows in these vector fields
+show where the continuous state is mapped to under the corresponding
+discrete state (for the first three latent dimensions).  Arrows are
+only drawn at those locations where the discrete state was actually
+employed.  We see that these dynamics capture the rotational patterns
+of the various loops.
+
+We use these dynamics to simulate how brain activity would evolve
+under the different states. Fig.~\ref{fig:syllables}f shows 200
+simulations of state 4 from various starting locations.  Four
+simulations are highlighted in particular, as denoted by the outlined
+circle, triangle, square, and pentagon.  We then map these
+trajectories into the space of neural activity and show how they would
+activate the five clusters shown in Fig.~\ref{fig:emissions}.  We see
+that state 4 corresponds to a pronounced dip followed by a rise in the
+activity of cluster 3, alongside a slow ramp in the activity of
+clusters 1 and 2. Referring back to Fig~\ref{fig:emissions}a, we see
+that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
+neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
+pair are sensory neurons responsible for initiating local search behavior,
+the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
+the ventral bias of turns, and the \textsf{SMDV} are motor neurons
+that initiate ventral turns. As state 4 is in close correspondence with
+the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
+sense that these neurons would be activated.
+
 \subsection*{Recurrent models learn spatially localized states and transition boundaries}

### 2017-12-21 drafts/elife/draft.tex
recurrent results section

@@
 
+The final and most critical component of the dynamical model for
+neural activity in the head ganglia is the discrete transition
+distribution.  In contrast to standard SLDS, we allow these transition
+probabilities to depend on the location in continuous state space (the
+\emph{recurrent} dependency) as well as the preceding discrete state
+(the \emph{Markovian} dependency).  Fig.~\ref{fig:recurrent}a shows
+the contribution of the Markovian term.  Each cell in this matrix
+denotes the number of transitions from one state to another. We are
+only counting transitions in which the state changes, hence the
+diagonal is zero.  We have permuted the states to emphasize the
+largely feed-forward structure.
+
+Fig.~\ref{fig:recurrent}b shows the strongest entries in this matrix
+in an alternative form.  We have illustrated each state as a node in a
+graph, and labeled the states based on their correspondence with the
+manually labeled states of~\citet{kato2015global} (see
+Fig.~\ref{fig:syllables}d).  This reveals a chain-like transition
+structure in which slow and forward (\textsf{FWD}) crawling are linked
+to reverse (\textsf{REV1}, \textsf{REV2}) and sustained reverse
+(\textsf{REVSUS}) via dorsal turns (\textsf{DT}). From sustained
+reverse crawling, the worm may execute either dorsal or ventral turns
+(\textsf{VT}).
+
+To accurately model these dynamics, the changes must occur at precise
+locations in continuous latent space.  Namely,
+Fig.~\ref{fig:syllables}a shows that the loops tend to start and end
+at the origin, and we rarely see transitions when the continuous
+latent state is far from zero. This manifests in signature duration
+distributions for the various states.  Fig.~\ref{fig:recurrent}c plots
+these distributions for each of the eight states.  Superimposed on
+top, we show the predicted distributions with the recurrent SLDS
+(solid black lines) compared with the predictions of the standard,
+Markovian SLDS (dotted black lines). The Markovian model knows nothing
+of the continuous latent state and implies a geometric duration
+distribution with monotonically decreasing probability. By contrast,
+the recurrent model learns to transition near the origin, which
+gives rise to characteristic duration distributions that closely
+match the empirical distributions.
+
 \subsection*{Oxygen level modulates transition probabilities but not dynamics or emissions}
@@
 \end{figure}
+
+Having demonstrated the utility of
 

### 2017-12-22 drafts/elife/draft.tex
figure captions

@@
   similarly tuned in latent state space.  Finally, we find a
-  significant overlap between our inferred modes and the
-  manually-labeled modes of~\citet{kato2015global}
+  significant overlap between our inferred states and the
+  manually-labeled states of~\citet{kato2015global}
   and~\citet{nichols2017global}, which were shown to correspond to
@@
 \caption{
-  % A schematic of the hierarchical model for combining
-  % information across multiple worms in order to learn a shared
-  % dynamical system for \celegans.  \textbf{a.}~All worms share a
-  % canonical dynamics library. Here, we have three dynamics functions
-  % (blue, red, gold), each of which is a simple linear mapping from
-  % $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
-  % assigns probabilities to the different dynamics functions based on
-  % the current location in $x$-space.  \textbf{b.}~The worms also share
-  % an emission matrix, $C$, which can be viewed as an embedding of
-  % neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
-  % library, which allows for slight deviations from the canonical
-  % dynamics.  \textbf{d.}~The model segments neural activity into a
-  % sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
-  % into the dynamics library.  \textbf{e.}~The discrete states specify
-  % the dynamics that are used to propagate the continuous latent
-  % states~$x_t$ forward in time. Here, the two dimensional latent
-  % states are colored according to the instantaneous discrete
-  % state. \textbf{f.}~The observed neural activity is a linear
-  % projection of the underlying continuous latent states.  On any given
-  % trial, we observe only a fraction of all the neurons.  The
-  % hierarchical model combines these partial recordings to infer the
-  % underlying states, the dynamics libraries, and the shared neural
-  % embedding.
+  \textit{A probabilistic framework for combining
+  information across multiple worms in order to learn a shared
+  dynamical system for \celegans.}
+  \textbf{a.}~The data consists of observed neural activity collected
+  from many different worms.  In each recording we observe a different
+  subset of labeled neurons in the head ganglia.
+  \textbf{b.}~We model the observed activity as a linear function of
+  an underlying, continuous, and typically low-dimensional latent state,
+  here shown as a two-dimensional signal for each recording.
+  \textbf{c.}~The dynamics of these continuous states are governed by
+  a co-evolving discrete state.
+  \textbf{d.}~These discrete states index into a library of linear
+  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
+  Each dynamics function is selectively used in different regions of
+  continuous latent space, as shown by the different intensities of
+  the vector field.
+  \textbf{e.}~To increase robustness to model misspecification, we
+  use a multivariate-t distribution for the noise in the continuous
+  state updates.
+  \textbf{f.}~Each worm has its own dynamics library, but they are
+  constrained to be close to the shared, canonical library.  This
+  combines information to learn shared structure while also allowing
+  for worm-to-worm variability.
+  \textbf{g.} Likewise, we allow each worm to have its own observation
+  variances to account for individual differences that may arise from,
+  for example, variable levels of fluorescent protein expression.
+  Missing neurons are denoted by gray boxes. 
 }
@@
 \begin{figure}[t!]
-\centering%
+\centering
 \includegraphics[width=6in]{figures/figure2} 
-\caption{}
+\caption{ \textit{The mapping from continuous latent states to
+    observations aids in neural identification and reveals clusters of
+    neurons.}
+  \textbf{a.}~The observed neural activity (solid black lines) in the
+  first worm of the \citet{kato2015global} dataset along with the
+  smoothed activity given by the model (blue lines). We use
+  correlations learned across the population to predict the activity
+  of unobserved neurons (those with dotted black lines).
+  \textbf{b.}~These predictions are a useful aid for neural identification.
+  We withheld the labels of a fraction of neurons and measured the similarity
+  between their activity and the models predictions.  The similarity was
+  significantly higher for the correct assignments as compared to incorrect
+  assignments.
+  \textbf{c.}~The probabilistic model offers a principled means of assessing
+  the dimensionality of neural activity, even though only a subset of neurons
+  are labeled on any given trial.  We find that ten latent dimensions yields
+  the highest test likelihood, and that allowing per-worm observation variances
+  yields substantial improvements in test performance.
+  \textbf{d.}~The per-worm variances show that observation noise can
+  vary substantially from one worm to another. Gray squares denote missing
+  neurons.
+  \textbf{e.}~We can view the linear mapping from latent states to observations
+  as either a matrix or as an embedding of neurons.
+  \textbf{f.}~Viewed as an embedding, we see that neurons cluster into
+  correlated groups with similar tuning to the continuous latent state.
+  \textbf{g.}~These groups roughly correspond to neurons at different
+  locations along the worm's body.  }
 \label{fig:emissions}
@@
 
-\begin{figure}[t]
-\centering%
+\begin{figure}[t!]
+\centering
 \includegraphics[width=6in]{figures/figure3} 
-\caption{}
+\caption{ \textit{The hierarchical SLDS decomposes continuous states
+    into segments with simple linear dynamics.}
+  \textbf{a.}~Three dimensions of the inferred continuous latent states
+  for the first worm, color coded by the corresponding discrete latent
+  state.
+  \textbf{b.}~We select the number of discrete states and justify the
+  recurrent and robust extensions by measuring the marginal likelihood
+  of held-out test data. We find that the robust, recurrent model
+  with eight states performs best.
+  \textbf{c.}~The inferred segmentations align with the manual segmentations
+  of~\citet{kato2015global}.
+  \textbf{d.}~Moreover, the inferred states are in close correspodence with
+  the manually labeled states, which were shown to indicate different behaviors.
+  \textbf{e.}~Each discrete state is associated with a canonical set of linear
+  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
+  We show the mappings for the first three dimensions here, with vectors placed
+  at locations where each state was used.  We see a variety of rotational
+  states that can be combined to yield dynamics like those in~\textbf{a.}.
+  \textbf{f.}~We simulate these linear dynamics from a variety of starting
+  locations for the fourth (green) and highlight four for closer inspection.
+  \textbf{g.}~These four trajectories give rise to different activations of
+  the neural clusters. We see see a dip then rise in cluster 3, which consists
+  of neurons related to ventral turns, consistent with the manual labeling
+  of the green state in~\textbf{d}.
+}
 \label{fig:syllables}
@@
 
-\begin{figure}[t]
-\centering%
+\begin{figure}[t!]
+\centering
 \includegraphics[width=6in]{figures/figure4} 
-\caption{}
+\caption{ \textit{Discrete state transitions show a non-Markovian,
+    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
+  from one state to another. The states are permuted to emphasize the
+  feed-forward structure.  \textbf{b.}~An alternative view of the
+  transitions; each state is a node in a graph and edges denote
+  high-probability transitions.  This shows transitions between
+  forward and reverse crawling linked by dorsal and ventral turns.
+  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
+  imply a geometric distribution of state durations (dotted black
+  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
+  transitions occur near the origin.  The recurrent SLDS incorporates
+  this dependency and generates state durations (solid black lines)
+  that are a much better fit to those inferred from the data (colored
+  bars).  }
 \label{fig:recurrent}
@@
 
-\begin{figure}[t]
-\centering%
+\begin{figure}[t!]
+\centering
 \includegraphics[width=6in]{figures/figure5} 
-\caption{}
+\caption{\textit{Incorporating exogenous factors like oxygen level
+    into the dynamics model yield differential responses in worms from
+    different genetic strains and developmental stages.}
+  \textbf{a.}~The observed neurons and their inferred observation variances
+  for each of the 41 worms in the~\citet{nichols2017global} dataset.
+  \textbf{b.}~Worms of different genetic strains (N2, npr1) and developmental
+  stages (pre-lethargus, lethargus) exhibit different levels of overall
+  state usage, with blue and yellow much more likely in the lethargus stage.
+  \textbf{c.}~As above, the inferred states closely match the manually
+  labeled states of~\citet{nichols2017global}, with blue and yellow mapping
+  onto the forward (\textsf{Fwd}) and quiescent (\textsf{Q}) states.
+  \textbf{d.}~We extended the SLDS to allow exogenous factors like oxygen
+  level influence discrete transition probabilities.
+  \textbf{e., f.}~Upon fitting the model, we find that oxygen has a limited
+  effect on N2 worms and a much stronger effect on npr1 worms. In npr1 worms,
+  increasing the oxygen level decreases the probability of quiescent and
+  forward states and elicits reversals and turns.
+}
 \label{fig:o2}
@@
 
-Having demonstrated the utility of
+The preceding results demonstrated the efficacy of hierarchical,
+recurrent, and robust SLDS for learning interpretable dynamics of
+neural activity from the noisy, partial recordings
+of~\citet{kato2015global}.  We now turn to the dataset studied
+in~\citet{nichols2017global}, which contains many more worms with
+roughly the same number of neurons per recording but is complicated by
+additional factors.  First, there are two strains of worms (N2 and
+npr-1) at two different stages of development (pre-lethargus and
+lethargus); second, the oxygen level is manipulated over the course of
+the recordings to elicit different responses.  Here we show how our
+framework can be extended to study the effects of these manipulations.
+
+Fig.~\ref{fig:o2}a shows which neurons were labeled in each of the
+recordings along with the inferred observation variance. There are 41
+worms, roughly ten from each strain and developmental stage
+combination (N2 + pre-lethargus, N2 + lethargus, npr1 + pre-lethargus,
+and npr1 + lethargus). Some neurons are consistently identified in
+almost all worms; others are identified in only a handful.  As before,
+our hierarchical framework combines all this information.
+
+Upon fitting our model, we immediately see a difference in the overall
+usage of the eight states (Fig.~\ref{fig:o2}b). The worms in the
+lethargus stage predominantly use the blue and yellow states.  Again,
+we found a close correspondence between our inferred states and the
+manually labeled states of~\citet{nichols2017global}
+(Fig.~\ref{fig:o2}c). Here, blue and yellow both map strongly onto the
+quiescent (\textsf{Q}) state. This is consistent with the
+characterization of the lethargus stage---at this stage, worms exhibit
+much less crawling behavior. The pre-lethargus worms execute a broader
+variety of behaviors, including reversals (\textsf{Rev.}) and dorsal
+and ventral turns  (\textsf{DT} and \textsf{VT}, respectively).
+
+A key finding of~\citet{nichols2017global} was that increasing the
+oxygen level can elicit activity in lethargus npr1 worms but not in
+lethargus N2 worms. We can quantify this effect with our models by
+including oxygen level as an exogenous input.  Fig.~\ref{fig:o2}d
+shows how the graphical model of the recurrent SLDS changes: now the
+discrete state at time~$t$ (red circle) depends on the preceding discrete
+state, the preceding continuous state, and the current oxygen level.
+\todo[inline]{consider alternative models in which oxygen level
+  influences continuous states and observations?}
+
+Fig.~\ref{fig:o2}e and~\ref{fig:o2}f show the inferred effect of
+oxygen level for each of the four strain/developmental stage pairs.
+As predicted, only the npr1 worms show a much larger response to
+oxygen. Increasing the oxygen level increases the probability
+of reversals and turning related states (red, orange, purple) and
+decreases the probability of forward and quiescent (blue, yellow)
+states. Somewhat surprisingly, the magnitude of the effect is
+comparable for both pre-lethargus and lethargus npr1 worms, even
+though the baseline level of reversals and turns is much higher in
+the pre-lethargus worms.
 

### 2018-01-02 drafts/elife/figures/figure1/make_figure_1.py
many changes to figure plotting code

@@ -294,8 +294,6 @@ def plot_noise_density(lims=(-3, 3), n_pts=50):
     lp_min = min(lp_mvn.min(), lp_mvt.min())
     lp_max = max(lp_mvn.max(), lp_mvt.max())
     levels = np.linspace(lp_min, lp_max, 15)[1:]
-    # import ipdb; ipdb.set_trace()
-
 
     fig = plt.figure(figsize=(1.25, .5))
     ax = fig.add_subplot(121, aspect="equal")

### 2018-01-02 experiments/fit_hlds.py

@@ -11,6 +11,7 @@ from functools import partial
 import matplotlib.pyplot as plt
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 from hips.plotting.colormaps import gradient_cmap
+from hips.plotting.layout import create_axis_at_location
 import seaborn as sns
 sns.set_style("white")
 sns.set_context("paper")
@@ -57,7 +58,7 @@ assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
 N_worms = 5
-N_clusters = 5
+N_clusters = 6
 
 
 def cached(results_name):
@@ -122,6 +123,19 @@ def load_data(include_unnamed=True):
 
     return ys, masks, z_trues, z_key, neuron_names
 
+
+def comput_xcorr(ys, masks):
+    N = D_obs
+    xcorr = np.nan * np.ones((N, N))
+    for n1 in range(N):
+        for n2 in range(n1+1, N):
+            valid = [(np.all(ms[i][:,n1]) and np.all(ms[i][:,n2])) for i in range(N_worms)]
+            if np.any(valid):
+                y1s = np.concatenate([ys[i][:, n1] for i in range(N_worms) if valid[i]])
+                y2s = np.concatenate([ys[i][:, n2] for i in range(N_worms) if valid[i]])
+                xcorr[n1, n2] = np.corrcoef(y1s, y2s)[0, 1]
+                xcorr[n2, n1] = xcorr[n1, n2]
+    return xcorr
 
 def _split_test_train(y, train=None, train_frac=0.8):
     T = y.shape[0]
@@ -300,10 +314,14 @@ def plot_best_model_results(best_model,
                             do_plot_x_3d=True,
                             do_plot_x_2d=True,
                             do_plot_sigmasq=True,
+                            do_plot_xcorr=True,
                             do_plot_similarity=True,
                             do_plot_cluster_embedding=True,
                             do_plot_cluster_locations=True,
-                            do_plot_data=True):
+                            do_plot_data=True,
+                            do_plot_data_zoom=True,
+                            do_plot_data_as_matrix=True,
[…]

### 2018-01-02 zimmer/plotting.py

@@ -136,9 +136,9 @@ def plot_3d_continuous_states(x, z, colors,
                 color=colors[z[cp_start]],
                 **kwargs)
 
-    # ax.set_xlabel("$x_1$", labelpad=-10)
-    # ax.set_ylabel("$x_2$", labelpad=-10)
-    # ax.set_zlabel("$x_3$", labelpad=-10)
+    ax.set_xlabel("dim {}".format(inds[0]+1), labelpad=-18, fontsize=6)
+    ax.set_ylabel("dim {}".format(inds[1]+1), labelpad=-18, fontsize=6)
+    ax.set_zlabel("dim {}".format(inds[2]+1), labelpad=-18, fontsize=6)
     ax.set_xticklabels([])
     ax.set_yticklabels([])
     ax.set_zticklabels([])
@@ -148,7 +148,7 @@ def plot_3d_continuous_states(x, z, colors,
         ax.set_zlim(-lim, lim)
 
     if title is not None:
-        ax.set_title(title)
+        ax.set_title(title, fontsize=8)
 
     plt.tight_layout(pad=0.1)
 
@@ -298,10 +298,10 @@ def plot_vector_field_3d(ii, z, x, perm_dynamics_distns, colors,
                          ax=None, lims=(-3,3), N_plot=500,
                          **kwargs):
 
-    qargs = dict(arrow_length_ratio=0.5,
+    qargs = dict(arrow_length_ratio=0.66,
                  alpha=1.0,
-                 length=1.5,
-                 lw=1,
+                 length=2,
+                 lw=.5,
                  pivot="middle",)
     qargs.update(kwargs)
 
@@ -332,9 +332,9 @@ def plot_vector_field_3d(ii, z, x, perm_dynamics_distns, colors,
               color=colors[ii],
               **qargs)
 
-    ax.set_xlabel('$x_1$', labelpad=-10)
-    ax.set_ylabel('$x_2$', labelpad=-10)
-    ax.set_zlabel('$x_3$', labelpad=-10)
+    ax.set_xlabel('dim 1', labelpad=-18, fontsize=6)
+    ax.set_ylabel('dim 2', labelpad=-18, fontsize=6)
+    ax.set_zlabel('dim 3', labelpad=-18, fontsize=6)
     ax.set_xticklabels([])
     ax.set_yticklabels([])
     ax.set_zticklabels([])
@@ -349,19 +349,18 @@ def plot_vector_field_3d(ii, z, x, perm_dynamics_distns, colors,
[…]

### 2018-01-02 experiments/fit_arhmm_nichols.py
minor changes

@@ -708,14 +708,14 @@ if __name__ == "__main__":
         do_plot_dynamics_3d=False,
         do_plot_dynamics_2d=False,
         do_plot_state_overlap=False,
-        do_plot_state_usage=False,
+        do_plot_state_usage=True,
         do_plot_transition_matrices=False,
         do_plot_simulated_trajs=False,
         do_plot_recurrent_weights=False,
         do_plot_x_at_changepoints=False,
         do_plot_latent_trajectories_vs_time=False,
         do_plot_duration_histogram=False,
-        do_plot_driven_trans_matrices=True
+        do_plot_driven_trans_matrices=False
     )
 
     # Rolling predictions

### 2018-01-02 zimmer/plotting.py

@@ -834,9 +834,7 @@ def plot_state_usage_by_worm_matrix(z_finals,
     fig = plt.figure(figsize=(3, 2))
     ax = fig.add_subplot(111, aspect="auto")
     im = ax.imshow(usage.T, cmap="Greys")
-    ax.set_ylabel("State")
-    ax.set_yticks([0, Kmax-1])
-    ax.set_yticklabels((ax.get_yticks() + 1).astype(int))
+    ax.set_yticks([])
     ax.set_xlabel("Worm")
     # ax.set_xticks(np.arange(N_worms))
     ax.set_xticklabels((ax.get_xticks() + 1).astype(int))
@@ -848,8 +846,8 @@ def plot_state_usage_by_worm_matrix(z_finals,
                aspect="auto")
     lax.set_xticks([])
     lax.set_yticks([])
-
-    divider = make_axes_locatable(ax)
+    lax.set_ylabel("State")
+    
     cax = divider.append_axes("right", size="5%", pad=0.05)
     plt.colorbar(im, cax=cax)
 

### 2018-01-02 drafts/elife/draft.tex

@@
 combining information from many recordings while also allowing for
-worm-to-worm differences.  For example, the neural activitiy may have
+worm-to-worm differences.  For example, the neural activity may have
 larger variance in some recordings than in others due to different
@@
 then learning the dynamics libraries that govern the continuous latent
-states and the correspdonding discrete state sequences.  Both steps are
+states and the corresponding discrete state sequences.  Both steps are
 performed with a block Gibbs sampler, leveraging the structure of the
@@
 then this provides some evidence in favor of assigning it this
-label. We tested this idea by witholding the labels of some
+label. We tested this idea by withholding the labels of some
 confidently-identified neurons and comparing their activity to the
@@
 groups of correlated neurons.  We clustered the neurons based on the
-cosine similarity between their embedding vecors and found five clusters
+cosine similarity between their embedding vectors and found five clusters
 of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
@@
   of~\citet{kato2015global}.
-  \textbf{d.}~Moreover, the inferred states are in close correspodence with
+  \textbf{d.}~Moreover, the inferred states are in close correspondence with
   the manually labeled states, which were shown to indicate different behaviors.

### 2018-01-03 drafts/elife/draft.tex
plotting for o2 figures. adding feature for change in o2

@@
 \centering%
-\includegraphics[width=6in]{figures/figure1} 
+\includegraphics[width=6in]{figures/v2/figure1} 
 \caption{
@@
 
-\subsection*{Emission model aids in identification and reveals clusters of neurons}
+\subsection*{Hierarchical state space models smooth noisy recordings and predict missing data}
 
@@
 \centering
-\includegraphics[width=6in]{figures/figure2} 
-\caption{ \textit{The mapping from continuous latent states to
-    observations aids in neural identification and reveals clusters of
-    neurons.}
-  \textbf{a.}~The observed neural activity (solid black lines) in the
-  first worm of the \citet{kato2015global} dataset along with the
-  smoothed activity given by the model (blue lines). We use
-  correlations learned across the population to predict the activity
-  of unobserved neurons (those with dotted black lines).
-  \textbf{b.}~These predictions are a useful aid for neural identification.
-  We withheld the labels of a fraction of neurons and measured the similarity
-  between their activity and the models predictions.  The similarity was
-  significantly higher for the correct assignments as compared to incorrect
-  assignments.
-  \textbf{c.}~The probabilistic model offers a principled means of assessing
-  the dimensionality of neural activity, even though only a subset of neurons
-  are labeled on any given trial.  We find that ten latent dimensions yields
-  the highest test likelihood, and that allowing per-worm observation variances
-  yields substantial improvements in test performance.
-  \textbf{d.}~The per-worm variances show that observation noise can
-  vary substantially from one worm to another. Gray squares denote missing
-  neurons.
-  \textbf{e.}~We can view the linear mapping from latent states to observations
-  as either a matrix or as an embedding of neurons.
-  \textbf{f.}~Viewed as an embedding, we see that neurons cluster into
-  correlated groups with similar tuning to the continuous latent state.
-  \textbf{g.}~These groups roughly correspond to neurons at different
-  locations along the worm's body.  }
-\label{fig:emissions}
+\includegraphics[width=6in]{figures/v2/figure2part1} 
+\caption{ \textit{Hierarchical state space models smooth noisy recordings and
+  make predictions about activity of unseen neurons.}
+  \textbf{a.}~The observed neural activity for the
+  five worms of the \citet{kato2015global} dataset along with the
+  smoothed activity under the hierarchical model.  Only a subset of neurons
+  are labeled in each worm; the rest are treated as missing data.
+  The model uses correlations learned across the population to predict the activity
+  of unobserved neurons.
+  The neurons have been clustered based on similarity, as shown in Fig.~\ref{fig:clustering}.
+  \textbf{b.} The observed activity (green lines) and smoothed activity (black lines)
+  for 3 minutes of activity from worm 5, as denoted by the dotted black box in panel a.
+  Neurons are in the same order as panels a and c.
+  \textbf{c.} Zooming in on 30 seconds of activity denoted by the box in panel b.
+}
+\label{fig:smoothing}
 \end{figure}
@@
 in calcium fluorescence; our only preprocessing step is to correct for
-photo-bleaching. Fig~\ref{fig:emissions}a show our results for the first
-of the five worms in this dataset. We see that the model smooths out
+photo-bleaching. Fig~\ref{fig:smoothing}a show our results for the 
+five worms in this dataset. We see that the model smooths out
 noise in the observed activity and, moreover, makes predictions about
@@
 
-The predicted activity is based on the correlations between neurons
-learned from many recordings.  These predictions are a useful aid for
-identifying unlabeled neurons.  For example, we see that
-\textsf{SMBDL} was not found in the first worm
-(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
-neurons closely matches the model's predictions for \textsf{SMBDL},
-then this provides some evidence in favor of assigning it this
-label. We tested this idea by withholding the labels of some
-confidently-identified neurons and comparing their activity to the
-model's predicted activity for all unlabeled neurons.  We measured the
-similarity between the actual and predicted activity and found that
-correct label, or ``candidate,'' tends to be much more similar than
-other incorrect candidates, as summarized in
-Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
-the location of the neuron, this provides a powerful aid for identifying
-neurons.  The more neurons that we can identify, the better we will be
-able to learn about the dynamics of neural activity.
+Fig.~\ref{fig:smoothing}b and shows the segment of neural activity
+denoted by the dotted black box in a. The neurons are clustered based
+on similarity, as discussed in the next section.  Here, the green lines
+show the observed activity and the black lines show the smoothed activity.
+Neurons that were not identified in this worm do not have observed activity,
+but the model still makes predictions about their activity. Fig.~\ref{fig:smoothing}c
+zooms in further to show a 30 sec window of activity. This shows the
+degree of noise in the raw data.
 
@@
 to this held-out data. We find that the test likelihood peaks with a
-ten-dimensional latent state (Fig.~\ref{fig:emissions}c).  Moreover, the likelihood is substantially
+ten-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).  Moreover, the likelihood is substantially
 higher in the model that includes per-worm observation variances
-compared to the model with shared variances for all worms. Fig.~\ref{fig:emissions}d
+compared to the model with shared variances for all worms. Fig.~\ref{fig:supp_emissions}d
 shows the inferred variance for each worm in the \citet{kato2015global}
@@
 This shows the extent of the missing data and the need for hierarchical
-modeling.
+modeling.\todo{make supplement figure}
+
+\subsection*{Predictions aid in neural identification}
+
+\todo[inline]{Make a figure for this result}
+
+\subsection*{Emission model reveals clusters of functionally similar neurons}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=6in]{figures/v2/figure2part2} 
+\caption{ \textit{The mapping from continuous latent states reveals clusters of
+    functionally similar neurons.}
+  We can view the linear mapping from latent states to observations
+  as either a matrix (\textbf{a.}) or as an embedding of neurons (\textbf{b.}).
+  \textbf{c.}~Viewed as an embedding, we see that neurons cluster into
+  correlated groups with similar tuning to the continuous latent state.
+  \textbf{d.}~The embedding similarity reflects correlations between
+  neurons, but since it is learned as part of a hierarchical model, it
+  predicts similarities for all pairs of neurons, even those that are
+  never simultaneously recorded.  These pairs are denoted by gray boxes
+  in the empirical correlation matrix.
+  \textbf{e.}~Neurons in the same cluster are not necessarily close
+  to one another in space, though we do see that, on average, neurons in cluster 2
+  are more anterior whereas neurons in clusters 5 and 6 are more posterior.}
+\label{fig:clustering}
+\end{figure}
+
+% --------- neural identification ----------------
+% The predicted activity is based on the correlations between neurons
+% learned from many recordings.  These predictions are a useful aid for
+% identifying unlabeled neurons.  For example, we see that
+% \textsf{SMBDL} was not found in the first worm
+% (Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
+% neurons closely matches the model's predictions for \textsf{SMBDL},
+% then this provides some evidence in favor of assigning it this
+% label. We tested this idea by withholding the labels of some
+% confidently-identified neurons and comparing their activity to the
+% model's predicted activity for all unlabeled neurons.  We measured the
+% similarity between the actual and predicted activity and found that
+% correct label, or ``candidate,'' tends to be much more similar than
+% other incorrect candidates, as summarized in
+% Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
+% the location of the neuron, this provides a powerful aid for identifying
+% neurons.  The more neurons that we can identify, the better we will be
+% able to learn about the dynamics of neural activity.
 
@@
 
-\subsection*{SLDS automatically parse data into discrete states with simple dynamics}
+\subsection*{SLDS automatically parse data into discrete, behaviorally meaningful states}
 
@@
 \centering
-\includegraphics[width=6in]{figures/figure3} 
-\caption{ \textit{The hierarchical SLDS decomposes continuous states
-    into segments with simple linear dynamics.}
+\includegraphics[width=6in]{figures/v2/figure3part1} 
+\caption{ \textit{The hierarchical SLDS automatically segments neural
+  activity based on its dynamics.}
   \textbf{a.}~Three dimensions of the inferred continuous latent states
-  for the first worm, color coded by the corresponding discrete latent
+  for each of the five worms, color coded by the corresponding discrete latent
   state.
   \textbf{b.}~We select the number of discrete states and justify the
-  recurrent and robust extensions by measuring the marginal likelihood
-  of held-out test data. We find that the robust, recurrent model
-  with eight states performs best.
+  hierarchcial, robust, and recurrent extensions by measuring the marginal likelihood
+  of held-out test data. We find that the hierarchical, robust, recurrent model
+  with~$K=8$ states performs best.
   \textbf{c.}~The inferred segmentations align with the manual segmentations
@@
   the manually labeled states, which were shown to indicate different behaviors.
-  \textbf{e.}~Each discrete state is associated with a canonical set of linear
-  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
-  We show the mappings for the first three dimensions here, with vectors placed
-  at locations where each state was used.  We see a variety of rotational
-  states that can be combined to yield dynamics like those in~\textbf{a.}.
-  \textbf{f.}~We simulate these linear dynamics from a variety of starting
-  locations for the fourth (green) and highlight four for closer inspection.
-  \textbf{g.}~These four trajectories give rise to different activations of
-  the neural clusters. We see see a dip then rise in cluster 3, which consists
-  of neurons related to ventral turns, consistent with the manual labeling
-  of the green state in~\textbf{d}.
 }
@@
 
+\subsection*{Quantifying the dynamics of neural activity in each discrete state}
+\begin{figure}[h!]
+\centering
+\includegraphics[width=6in]{figures/v2/figure3part2} 
+\caption{ \textit{Each state is characterized by simple linear dynamics.}
+    \textbf{a.}~Each discrete state is associated with a canonical set of linear
+  dynamics, which can be viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.
+  We show the mappings for the first three dimensions here, with vectors placed
+  at locations where each state was used.
+  \textbf{b.}~We simulate these linear dynamics from a variety of starting
+  locations and compute the time course of resulting activations for each
+  cluster of neurons.
+  \textbf{c.}~The eigenspectra of the dynamics matrices highlight the rotational
+  vs. contractive nature of the states and illustrate worm-to-worm variability. 
+}
+\label{fig:dynamics}
+\end{figure}
+
 Beyond simply automating a manual process, the SLDS also provides a
@@
 state is associated with a set of linear dynamics.
-Fig.~\ref{fig:syllables}e shows the canonical dynamics associated with
+Fig.~\ref{fig:dynamics}a shows the canonical dynamics associated with
 each of the eight discrete states. The arrows in these vector fields
@@
 We use these dynamics to simulate how brain activity would evolve
-under the different states. Fig.~\ref{fig:syllables}f shows 200
-simulations of state 4 from various starting locations.  Four
-simulations are highlighted in particular, as denoted by the outlined
-circle, triangle, square, and pentagon.  We then map these
-trajectories into the space of neural activity and show how they would
-activate the five clusters shown in Fig.~\ref{fig:emissions}.  We see
+under the different states. Fig.~\ref{fig:dynamics}b shows how four
+simulations of each state from various starting locations  map 
+into activations in the space of neural activity. In particular, we show how they would
+activate the six clusters shown in Fig.~\ref{fig:clustering}.  We see
 that state 4 corresponds to a pronounced dip followed by a rise in the
 activity of cluster 3, alongside a slow ramp in the activity of
-clusters 1 and 2. Referring back to Fig~\ref{fig:emissions}a, we see
-that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
-neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
-pair are sensory neurons responsible for initiating local search behavior,
-the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
-the ventral bias of turns, and the \textsf{SMDV} are motor neurons
-that initiate ventral turns. As state 4 is in close correspondence with
-the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
-sense that these neurons would be activated.
+clusters 1 and 2.
+% Referring back to Fig~\ref{fig:emissions}a, we see
+% that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
+% neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
+% pair are sensory neurons responsible for initiating local search behavior,
+% the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
+% the ventral bias of turns, and the \textsf{SMDV} are motor neurons
+% that initiate ventral turns. As state 4 is in close correspondence with
+% the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
+% sense that these neurons would be activated.
+
+\clearpage
 
@@
 \centering
-\includegraphics[width=6in]{figures/figure4} 
+\includegraphics[width=6in]{figures/v2/figure4v2} 
 \caption{ \textit{Discrete state transitions show a non-Markovian,
@@
   lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
-  transitions occur near the origin.  The recurrent SLDS incorporates
+  transitions occur near the origin.  With the exception of state 7,
+  which is unstable, the recurrent SLDS incorporates
   this dependency and generates state durations (solid black lines)
-  that are a much better fit to those inferred from the data (colored
+  that are a better fit to those inferred from the data (colored
   bars).  }
@@
 \centering
-\includegraphics[width=6in]{figures/figure5} 
+\includegraphics[width=6in]{figures/v1/figure5} 
 \caption{\textit{Incorporating exogenous factors like oxygen level

### 2018-01-03 experiments/fit_arhmm.py

@@ -55,7 +55,7 @@ from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states
     plot_3d_dynamics, plot_2d_dynamics, plot_state_overlap, plot_state_usage_by_worm, plot_all_transition_matrices, \
     plot_simulated_trajectories, make_state_predictions_3d_movie, plot_simulated_trajectories2, plot_simulated_trajectories3,\
     plot_recurrent_transitions, plot_x_at_changepoints, plot_latent_trajectories_vs_time, \
-    plot_state_usage_by_worm_matrix, plot_duration_histogram
+    plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_duration_cdfs
 
 # LDS Results
 lds_dir = os.path.join("results", "2017-11-03-hlds", "run002")
@@ -538,11 +538,18 @@ def plot_best_model_results(do_plot_expected_states=True,
     if do_plot_duration_histogram:
         durss = [np.array([x_sim.shape[0] for x_sim in x_sims]) for x_sims in x_simss]
 
-        plot_duration_histogram(best_model.trans_distn,
+        # plot_duration_histogram(best_model.trans_distn,
+        #                         z_finals,
+        #                         durss,
+        #                         perm=perm,
+        #                         results_dir=fig_dir)
+
+        plot_duration_cdfs(best_model.trans_distn,
                                 z_finals,
                                 durss,
                                 perm=perm,
                                 results_dir=fig_dir)
+
         plt.close("all")
 
     if do_plot_eigenspectrum:
@@ -642,7 +649,7 @@ if __name__ == "__main__":
     d = lds_results['best_model'].D[:,0]
     N_clusters = lds_results['N_clusters']
     neuron_clusters = lds_results['neuron_clusters']
-    C_norm = C[:, :-1] / np.linalg.norm(C[:, :-1], axis=1)[:, None]
+    C_norm = C / np.linalg.norm(C, axis=1)[:, None]
     C_clusters = np.array([C[neuron_clusters == c].mean(0) for c in range(N_clusters)])
     d_clusters = np.array([d[neuron_clusters == c].mean(0) for c in range(N_clusters)])
 
@@ -705,11 +712,11 @@ if __name__ == "__main__":
         do_plot_state_overlap=False,
         do_plot_state_usage=False,
         do_plot_transition_matrices=False,
-        do_plot_simulated_trajs=True,
+        do_plot_simulated_trajs=False,
[…]

### 2018-01-03 experiments/fit_arhmm_nichols.py

@@ -11,7 +11,7 @@ from tqdm import tqdm
 
 # Plotting stuff
 import matplotlib
-matplotlib.use("Agg")
+# matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from hips.plotting.colormaps import gradient_cmap
 import seaborn as sns
@@ -61,6 +61,9 @@ from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states
     plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_driven_transition_matrices, \
     plot_driven_transition_mod
 
+from hips.plotting.layout import create_axis_at_location
+from hips.plotting.colormaps import white_to_color_cmap
+
 # LDS Results
 lds_dir = os.path.join("results", "nichols", "2017-11-13-hlds", "run001")
 assert os.path.exists(lds_dir)
@@ -71,13 +74,17 @@ assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
 # Datasets
-worms_groups_conditions = [(i, 0, "n2_1_prelet") for i in range(11)] + \
-                          [(i, 1, "n2_2_let") for i in range(12)] + \
-                          [(i, 2, "npr1_1_prelet") for i in range(10)] + \
-                          [(i, 3, "npr1_2_let") for i in range(11)]
+from zimmer.io import WormData
+condition_names = ["n2_1_prelet", "n2_2_let", "npr1_1_prelet", "npr1_2_let"]
+short_condition_names = ["N2 pre-leth.", "N2 leth.", "npr1 pre-leth.", "npr1 leth."]
+worms_groups_conditions = [(i, 0, condition_names[0]) for i in range(11)] + \
+                          [(i, 1, condition_names[1]) for i in range(12)] + \
+                          [(i, 2, condition_names[2]) for i in range(10)] + \
+                          [(i, 3, condition_names[3]) for i in range(11)]
 worm_names = ["{} worm {}".format(condition, i)
               for (i, group, condition) in worms_groups_conditions]
 N_worms = len(worms_groups_conditions)
+N_groups = 4
 
 # Fitting parameters
 N_lags = 1
@@ -105,6 +112,54 @@ def cached(results_name):
     return _cache
 
 
+def load_data(include_unnamed=True):
+    # Load the data
+    worm_datas = [WormData(i,
+                           name=worm_names[j],
+                           version="nichols",
+                           condition=condition)
[…]

### 2018-01-03 experiments/fit_hlds_nichols.py

@@ -711,6 +711,13 @@ if __name__ == "__main__":
     ys, ms, us, z_trues, z_true_key, neuron_names = load_data(include_unnamed=False)
     D_obs = ys[0].shape[1]
 
+    # Encode both the absolute o2 level and the time since o2 change
+    window = 20 * 3
+    tau = 1 * 3
+    delta_o2s = [np.concatenate(([0], np.diff(u))) for u in us]
+    delta_o2s_filt = [np.convolve(d_o2, np.exp(-np.arange(window) / tau), "full")[:-window] for d_o2 in delta_o2s]
+    us = [np.column_stack(u, d_o2) for u, d_o2 in zip(us, delta_o2s_filt)]
+
     # Split test train
     ytrains, ytests, train_inds = list(zip(*[_split_test_train(y, train_frac=0.8) for y in ys]))
     mtrains, mtests, _ = list(zip(*[_split_test_train(m, train=train) for m, train in zip(ms, train_inds)]))
@@ -748,7 +755,6 @@ if __name__ == "__main__":
     neuron_perm, neuron_clusters = cluster_neruons(best_model)
 
     # plot_likelihoods(final_lls, hlls, best_index)
-    #
     plot_best_model_results(best_model,
                             do_plot_x_3d=False,
                             do_plot_x_2d=False,
@@ -758,28 +764,29 @@ if __name__ == "__main__":
                             N_plot=N_worms)
 
     # heldout_neuron_identification()
-    #
-    # # Save out the results
-    # results = dict(
-    #     xtrains=xtrains,
-    #     xtests=xtests,
-    #     ytrains=ytrains,
-    #     ytests=ytests,
-    #     mtrains=mtrains,
-    #     mtests=mtests,
-    #     utrains=utrains,
-    #     utests=utests,
-    #     z_true_trains=z_true_trains,
-    #     z_true_tests=z_true_tests,
-    #     z_key=z_true_key,
-    #     best_model=best_model,
-    #     D_latent=best_model.D_latent,
-    #     C=C,
-    #     perm=dim_perm,
-    #     N_clusters=N_clusters,
-    #     neuron_clusters=neuron_clusters,
-    #     neuron_perm=neuron_perm
-    # )
-    #
-    # with open(os.path.join(results_dir, "lds_data.pkl"), "wb") as f:
-    #     pickle.dump(results, f)
+
+    # Save out the results
+    results = dict(
+        xtrains=xtrains,
+        xtests=xtests,
[…]

### 2018-01-03 zimmer/plotting.py

@@ -705,20 +705,19 @@ def plot_state_overlap(z_finals, z_trues,
                        z_key=None,
                        colors=None,
                        z_colors=None,
-                       filename=None,
                        titles=None,
                        results_dir=None):
     colors = default_colors if colors is None else colors
     z_colors = default_colors if z_colors is None else z_colors
     Kmax = np.max(np.concatenate(z_finals)) + 1
     K_zimmer = np.concatenate(z_trues).max() + 1
-    titles = ["State Overlap (Worm {})".format(i+1) for i in range(len(z_trues))] \
+    titles = ["state overlap (worm {})".format(i+1) for i in range(len(z_trues))] \
       if titles is None else titles
 
     def _plot_overlap(overlap, perm, filename, title):
 
-        fig = plt.figure(figsize=(2., 2))
-        ax1 = create_axis_at_location(fig, .5, .5, 1., 1)
+        fig = plt.figure(figsize=(1.5, 1.25))
+        ax1 = create_axis_at_location(fig, .4, .3, .75, .75)
         im = ax1.imshow(overlap[:, perm], vmin=0, vmax=1.0, cmap="Greys", interpolation="nearest", aspect="auto")
         ax1.set_xticks([])
         if z_key is None:
@@ -727,25 +726,27 @@ def plot_state_overlap(z_finals, z_trues,
             ax1.set_yticks(np.arange(K_zimmer))
             ax1.set_yticklabels(z_key, fontdict=dict(size=6))
             ax1.tick_params(axis='y', which='major', pad=11)
-        ax1.set_title(title)
+        ax1.set_title(title, fontsize=8)
 
-        lax = create_axis_at_location(fig, .4, .5, .06, 1)
+        lax = create_axis_at_location(fig, .3, .3, .06, .75)
         lax.imshow(np.arange(K_zimmer)[:, None], cmap=gradient_cmap(z_colors[:K_zimmer]), interpolation="nearest",
                    aspect="auto")
         lax.set_xticks([])
         lax.set_yticks([])
 
         if z_key is None:
-            lax.set_ylabel("Zimmer State", fontsize=8)
+            lax.set_ylabel("manual state", fontsize=6)
 
-        bax = create_axis_at_location(fig, .5, .4, 1., .06)
[…]

### 2018-01-04 experiments/fit_arhmm_nichols.py
updating for two o2 features

@@ -11,7 +11,7 @@ from tqdm import tqdm
 
 # Plotting stuff
 import matplotlib
-# matplotlib.use("Agg")
+matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from hips.plotting.colormaps import gradient_cmap
 import seaborn as sns
@@ -65,11 +65,11 @@ from hips.plotting.layout import create_axis_at_location
 from hips.plotting.colormaps import white_to_color_cmap
 
 # LDS Results
-lds_dir = os.path.join("results", "nichols", "2017-11-13-hlds", "run001")
+lds_dir = os.path.join("results", "nichols", "2018-01-03-hlds", "run001")
 assert os.path.exists(lds_dir)
 
 # AR-HMM RESULTS
-results_dir = os.path.join("results", "nichols", "2017-11-14-arhmm", "run001")
+results_dir = os.path.join("results", "nichols", "2018-01-03-arhmm", "run001")
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
@@ -560,21 +560,19 @@ def plot_best_model_results(do_plot_expected_states=True,
         plt.close("all")
 
     if do_plot_state_overlap:
-        # Combine states from each condition
-        # for condition, title in zip(condition_names, short_condition_names):
-        #     plot_state_overlap([np.concatenate([z_finals[i] for i in range(len(z_finals)) if worms_groups_conditions[i][2] == condition])],
-        #                        [np.concatenate([z_trues[i][N_lags:] for i in range(len(z_finals)) if worms_groups_conditions[i][2] == condition])],
-        #                        z_key=z_key,
-        #                        z_colors=zimmer_colors,
-        #                        titles=[title],
-        #                        results_dir=fig_dir)
-
-        plot_state_overlap([np.concatenate([z_finals[i] for i in range(len(z_finals))])],
-                           [np.concatenate([z_trues[i][N_lags:] for i in range(len(z_finals))])],
-                           z_key=z_key,
-                           z_colors=zimmer_colors,
-                           titles=["state overlap"],
-                           results_dir=fig_dir)
+        # Combine states from each condition        
+        plot_state_overlap(
+            [np.concatenate([z_finals[i] for i in range(len(z_finals))
+                             if worms_groups_conditions[i][2] == condition])
+             for condition in condition_names],
[…]

### 2018-01-04 zimmer/plotting.py

@@ -706,6 +706,7 @@ def plot_state_overlap(z_finals, z_trues,
                        colors=None,
                        z_colors=None,
                        titles=None,
+                       permute=True,
                        results_dir=None):
     colors = default_colors if colors is None else colors
     z_colors = default_colors if z_colors is None else z_colors
@@ -751,21 +752,26 @@ def plot_state_overlap(z_finals, z_trues,
         if results_dir is not None:
             plt.savefig(os.path.join(results_dir, filename))
 
-    # Use the Hungarian algorithm to find a permutation of states that
-    # yields the highest overlap
+    # Compute total overlap from all worms
     overlap = np.zeros((K_zimmer, Kmax), dtype=float)
     for z_true, z_inf in zip(z_trues, z_finals):
         for k1 in range(K_zimmer):
             for k2 in range(Kmax):
                 overlap[k1, k2] = np.sum((z_true == k1) & (z_inf == k2))
+                
+    if permute:
+        # Use the Hungarian algorithm to find a permutation of states that
+        # yields the highest overlap                    
+        from scipy.optimize import linear_sum_assignment
+        _, perm = linear_sum_assignment(-overlap)
 
-    from scipy.optimize import linear_sum_assignment
-    _, perm = linear_sum_assignment(-overlap)
+        # Add any unused inferred states
+        if Kmax > K_zimmer:
+            unused = np.array(list(set(np.arange(Kmax)) - set(perm)))
+            perm = np.concatenate((perm, unused))
 
-    # Add any unused inferred states
-    if Kmax > K_zimmer:
-        unused = np.array(list(set(np.arange(Kmax)) - set(perm)))
-        perm = np.concatenate((perm, unused))
+    else:
+        perm = np.arange(Kmax)
 
     # Normalize the overlap from all worms and plot
     overlap /= overlap.sum(1)[:, None]
@@ -1575,7 +1581,8 @@ def plot_x_at_changepoints(zs, xs, window=9, colors=None,
             plt.savefig(os.path.join(results_dir, basename + "_{}.pdf".format(k)))
 
 
-def plot_driven_transition_matrices(u_values,
+def plot_driven_transition_matrices(u_index,
+                                    u_values,
                                     trans_distns,
                                     z_finals,
                                     colors=None,
[…]

### 2018-01-04 experiments/fit_arhmm_nichols.py
minor edits

@@ -883,9 +883,6 @@ def rolling_predictions(T_pred, worm, N_sim=50):
 
 
 if __name__ == "__main__":
-    # Load the continuous states found with the LDS
-    ys, ms, us, z_trues, z_true_key, neuron_names = load_data(include_unnamed=False)
-
     with open(os.path.join(lds_dir, "lds_data.pkl"), "rb") as f:
         lds_results = pickle.load(f)
 

### 2018-01-04 experiments/fit_arhmm_nichols.py
finalizing figure 5

@@ -11,7 +11,7 @@ from tqdm import tqdm
 
 # Plotting stuff
 import matplotlib
-matplotlib.use("Agg")
+# matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from hips.plotting.colormaps import gradient_cmap
 import seaborn as sns
@@ -650,17 +650,19 @@ def plot_best_model_results(do_plot_expected_states=True,
         #                                 condition_names=["N2 pre-leth.", "N2 leth.", "npr1 pre-leth.", "npr1 leth."],
         #                                 results_dir=fig_dir)
 
-        plot_driven_transition_mod(0, [-0.7, 1.4],
+        plot_driven_transition_mod(2, [[-0.7, 1.4], [0, 28]],
                                    best_model.trans_distns,
+                                   ["log effect\n(O$_2$)", "log effect\n($\\Delta$O$_2$)"],
                                    perm=perm,
                                    condition_names=short_condition_names,
                                    results_dir=fig_dir)
 
-        plot_driven_transition_mod(1, [0, 28],
-                                   best_model.trans_distns,
-                                   perm=perm,
-                                   condition_names=short_condition_names,
-                                   results_dir=fig_dir)
+        # plot_driven_transition_mod(1, [0, 28],
+        #                            best_model.trans_distns,
+        #                            perm=perm,
+        #                            condition_names=short_condition_names,
+        #                            ylabel="log effect of $\\Delta$O$_2$",
+        #                            results_dir=fig_dir)
 
         plt.close("all")
 
@@ -798,6 +800,7 @@ def plot_best_model_results(do_plot_expected_states=True,
         # find all the times where we enter state k
         k = 7
         Tpre, Tpost = 15 * 3, 30 * 3
+        tt = np.arange(-Tpre, Tpost) / 3.0
         xks = np.nan * np.zeros((1000, Tpre + Tpost, D_latent))
         index = 0
 
@@ -823,26 +826,25 @@ def plot_best_model_results(do_plot_expected_states=True,
         nstd = 1
 
         # Plot the state triggered average
[…]

### 2018-01-04 experiments/fit_hlds_nichols.py

@@ -390,23 +390,40 @@ def plot_best_model_results(best_model,
         sigma_obs_mask = np.array([np.any(~m, axis=0) for m in ms])
         sigma_obs[sigma_obs_mask] = np.nan
 
+        # Plot observed neurons instead
+        sigma_obs = np.array([m[0] for m in ms], dtype=np.float)
+        sigma_obs[sigma_obs_mask] = np.nan
+
         cmap = gradient_cmap([np.ones(3), colors[0]])
         cmap.set_bad(0.7 * np.ones(3))
 
-        fig = plt.figure(figsize=(5.5, 2.))
+        fig = plt.figure(figsize=(1.5, 5.))
         ax = fig.add_subplot(111)
-        im = ax.imshow(np.sqrt(sigma_obs), vmin=0, aspect="auto", cmap=cmap)
-        ax.set_xticks(np.arange(D_obs))
-        ax.set_xticklabels(neuron_names, rotation="90", fontsize=5)
-        ax.set_yticks(np.arange(0, N_worms, 5))
-        ax.set_yticklabels(np.arange(0, N_worms, 5) + 1)
-        ax.set_ylabel("Worm")
+        im = ax.imshow(np.sqrt(sigma_obs).T, vmin=0, aspect="auto", cmap=cmap)
 
-        ax.set_title("$\sigma_{\mathsf{obs}}$")
+        group_sizes = np.array([11, 12, 10, 11])
+        for w in np.cumsum(group_sizes)[:-1]:
+            plt.plot([w, w], [-0.5, D_obs-0.5], ':k')
+        plt.xlim(-0.5, N_worms-0.5)
+        plt.ylim(D_obs-0.5, -0.5)
 
-        plt.colorbar(im)
+
+        ax.set_yticks(np.arange(D_obs))
+        ax.set_yticklabels(neuron_names, rotation="0", fontsize=5)
+        ax.set_xticks([])
+        # ax.set_xticks(np.arange(0, N_worms, 5))
+        # ax.set_xticklabels(np.arange(0, N_worms, 5) + 1)
+        ax.set_xlabel("worm", fontsize=8, labelpad=20)
+
+        # ax.set_title("standard deviation of\ndifferenced Ca++", fontsize=8)
+        ax.set_title("observed neurons", fontsize=8)
+
+        # divider = make_axes_locatable(ax)
+        # cax = divider.append_axes("bottom", size="1%", pad=0.5)
[…]

### 2018-01-04 zimmer/plotting.py

@@ -1670,9 +1670,10 @@ def plot_driven_transition_matrices(u_index,
         plt.savefig(os.path.join(results_dir, "driven_trans_matrices.pdf"))
 
 
-def plot_driven_transition_mod(u_index,
+def plot_driven_transition_mod(D_input,
                                u_values,
                                trans_distns,
+                               ylabels,
                                colors=None,
                                perm=None,
                                results_dir=None,
@@ -1681,46 +1682,52 @@ def plot_driven_transition_mod(u_index,
     N_trans = len(trans_distns)
     K = trans_distns[0].num_states
     perm = np.arange(K) if perm is None else perm
-    condition_names = ["Cond {} ({} - {})".format(i+1, u_values[1] - u_values[0]) for i in range(N_trans)] \
-      if condition_names is None else condition_names
 
-    # Get the differential effect of O2 = 21% - O2 = 10%
-    effects = []
-    for i in range(N_trans):
-        w = trans_distns[i].W[-2+u_index][perm]
-        effects.append(w * (u_values[1] - u_values[0]))
-    lim = abs(np.array(effects)).max()
+    pad = 0.25
+    left = 0.5
+    fwidth = 2.75
+    width = (fwidth - left - 3 * pad - .1) / 4
+    height = .5
+    fig = plt.figure(figsize=(fwidth, height * D_input + .25))
 
-    pad = 0.2
-    left = 0.3
-    fwidth = 2.5
-    width = (fwidth - left - 3 * pad) / 4
-    fig = plt.figure(figsize=(fwidth, 1))
+    for j in range(D_input):
+        # Get the differential effect of the input
+        condition_names = ["Cond {} ({} - {})".
+                               format(i + 1, u_values[j][1] - u_values[j][0]) for i in range(N_trans)] \
+            if condition_names is None else condition_names
 
-    for i in range(N_trans):
-        ax = create_axis_at_location(fig, left + i * (width + pad), 0.05, width, 0.75)
+        effects = []
+        for i in range(N_trans):
+            w = trans_distns[i].W[-D_input + j][perm]
+            effects.append(w * (u_values[j][1] - u_values[j][0]))
+        lim = abs(np.array(effects)).max()
 
-        eff = effects[i]
-        for k in range(K):
-            ax.bar(k, eff[k], width=0.8, color=colors[k])
[…]

### 2018-01-05 drafts/elife/draft.tex
adding updated figures.

@@
 \centering%
-\includegraphics[width=6in]{figures/v2/figure1} 
+\includegraphics[width=6in]{figures/v3/figure1} 
 \caption{
@@
 \centering
-\includegraphics[width=6in]{figures/v2/figure2part1} 
-\caption{ \textit{Hierarchical state space models smooth noisy recordings and
-  make predictions about activity of unseen neurons.}
-  \textbf{a.}~The observed neural activity for the
-  five worms of the \citet{kato2015global} dataset along with the
-  smoothed activity under the hierarchical model.  Only a subset of neurons
-  are labeled in each worm; the rest are treated as missing data.
-  The model uses correlations learned across the population to predict the activity
-  of unobserved neurons.
-  The neurons have been clustered based on similarity, as shown in Fig.~\ref{fig:clustering}.
-  \textbf{b.} The observed activity (green lines) and smoothed activity (black lines)
-  for 3 minutes of activity from worm 5, as denoted by the dotted black box in panel a.
-  Neurons are in the same order as panels a and c.
-  \textbf{c.} Zooming in on 30 seconds of activity denoted by the box in panel b.
-}
+\includegraphics[width=6in]{figures/v3/figure2} 
+\caption{ \textit{Hierarchical state space models smooth noisy
+    recordings and predict the activity of unseen neurons.}  We show
+  the neural activity (differenced Ca++ traces) for two of the
+  five worms from the \citet{kato2015global} dataset.  Of these five
+  worms, worm 1 (left) has the fewest labeled neurons and worm 5
+  (right) has the most.  Green lines are the measured activity of the
+  labeled neurons; black lines are the smoothed activities predicted
+  by the model for the 61 neurons that were labeled in at least one
+  worm.  The model uses correlations learned across all recordings to
+  predict the activity of unlabeled neurons, like \textsf{SMBDL} in
+  worm 1.  The neurons have been clustered based on similarity, as
+  discussed in Fig.~\ref{fig:clustering}.  }
 \label{fig:smoothing}
@@
 structure of neural activity in the head ganglia that can only be
-obtained by pooling information across many partial recordings.  With
-this aggregated information, we can smooth noisy observations and
-predict the activity of unobserved neurons. \citet{kato2015global}
-used a smoothing procedure to estimate temporal principal components.
-By contrast, we fit our model directly to the first-order differences
-in calcium fluorescence; our only preprocessing step is to correct for
-photo-bleaching. Fig~\ref{fig:smoothing}a show our results for the 
-five worms in this dataset. We see that the model smooths out
-noise in the observed activity and, moreover, makes predictions about
-the activity of neurons that could not be identified in this worm.
+obtained by pooling information across many partial recordings.  We
+fit our model directly to the first-order differences in calcium
+fluorescence; our only preprocessing step is to correct for
+photo-bleaching. The green lines in Fig.~\ref{fig:smoothing} show a
+three minute window of bleaching-corrected, differenced calcium traces
+for two of the five worms in the~\citet{kato2015global} dataset. Of
+these five worms, Worm 1 has the fewest labeled neurons (24 neurons)
+and worm 5 has the most (48 neurons). Across the five worms, 61 unique
+neuron labels were identified in at least one worm.
 
-Fig.~\ref{fig:smoothing}b and shows the segment of neural activity
-denoted by the dotted black box in a. The neurons are clustered based
-on similarity, as discussed in the next section.  Here, the green lines
-show the observed activity and the black lines show the smoothed activity.
-Neurons that were not identified in this worm do not have observed activity,
-but the model still makes predictions about their activity. Fig.~\ref{fig:smoothing}c
-zooms in further to show a 30 sec window of activity. This shows the
-degree of noise in the raw data.
+By aggregating information across worms, the model learns correlation
+structure that enables it to smooth noisy traces and predict the
+activity of unobserved neurons.  The smoothed and predicted activity
+is shown as black lines in Fig.~\ref{fig:smoothing}. For example,
+even though \textsf{SMDVR} is not labeled in worm 1, the model
+learns how its activity covaries with that of other neurons
+by leveraging information from other worms, like worm 5, in which
+both~\textsf{SMDVR} and~\textsf{SMDVL} are labeled.  Since these
+two neurons are highly correlated, the model can make strong predictions
+about the activity of~\textsf{SMDVR} in worm 1 on the basis of the
+activity of~\textsf{SMDVL}.
 
-Even though each individual recording contains only a subset of neurons,
-the hierarchical model can still estimate the dimensionality of
-neural activity in the entire head ganglia.  We reserve the last 20\% of
-the data (3.6 min) for testing and estimate the dimensionality
-of neural activity based on the marginal log likelihood the model assigns
-to this held-out data. We find that the test likelihood peaks with a
-ten-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).  Moreover, the likelihood is substantially
-higher in the model that includes per-worm observation variances
-compared to the model with shared variances for all worms. Fig.~\ref{fig:supp_emissions}d
-shows the inferred variance for each worm in the \citet{kato2015global}
+The degree to which the smoothed activity matches the observed
+activity is governed by the dimensionality of the latent state space.
+We estimate this dimensionality with cross validation, reserving the
+last 20\% of the data (3.6 min) for testing and measuring the marginal
+log likelihood of the held-out data as we increase the latent
+dimensionality. We find that the test likelihood peaks with a
+10-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).
+
+Moreover, the likelihood is substantially higher in the model that
+includes per-worm observation variances compared to the model with
+shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
+the inferred variance for each worm in the \citet{kato2015global}
 dataset.  Neurons that were not identified are marked with a gray box.
-This shows the extent of the missing data and the need for hierarchical
-modeling.\todo{make supplement figure}
-
-\subsection*{Predictions aid in neural identification}
-
-\todo[inline]{Make a figure for this result}
+This shows the extent of the missing data and the need for
+hierarchical modeling.\todo{make supplement figure}
 
@@
 \centering
-\includegraphics[width=6in]{figures/v2/figure2part2} 
-\caption{ \textit{The mapping from continuous latent states reveals clusters of
-    functionally similar neurons.}
-  We can view the linear mapping from latent states to observations
-  as either a matrix (\textbf{a.}) or as an embedding of neurons (\textbf{b.}).
-  \textbf{c.}~Viewed as an embedding, we see that neurons cluster into
-  correlated groups with similar tuning to the continuous latent state.
-  \textbf{d.}~The embedding similarity reflects correlations between
-  neurons, but since it is learned as part of a hierarchical model, it
-  predicts similarities for all pairs of neurons, even those that are
-  never simultaneously recorded.  These pairs are denoted by gray boxes
-  in the empirical correlation matrix.
-  \textbf{e.}~Neurons in the same cluster are not necessarily close
-  to one another in space, though we do see that, on average, neurons in cluster 2
-  are more anterior whereas neurons in clusters 5 and 6 are more posterior.}
+\includegraphics[width=6in]{figures/v3/figure3B} 
+\caption{ \textit{The mapping from continuous latent states to neural
+    activity reveals clusters of functionally similar neurons.}
+  \textbf{a.} The emission matrix is a linear map from latent states;
+  the rows can be seen as ``tuning vectors'' that specify how each
+  neuron responds to changes in the latent state.  These tuning
+  vectors cluster into discrete groups (separated by black lines),
+  indicating that there are subpopulations of similarly tuned, and
+  hence highly correlated, neurons. Neurons are ordered as in
+  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
+  tuning vectors supports this clustering and shows differences
+  between subpopulations (clusters have been sorted to make this as
+  close to diagonal as possible).  \textbf{c.}~The tuning similarity
+  reflects correlations between neurons, but since it is learned as
+  part of a hierarchical model, it predicts similarities for all pairs
+  of neurons, even pairs that are never simultaneously recorded (gray,
+  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
+  neurons from each cluster, as specified by WormAtlas.  Neurons in
+  the same cluster are not necessarily close to one another in space,
+  though some clusters are spatially localized. Black bars denote
+  approximate locations of the anterior (AG), dorsal (DG), ventral
+  (VG) and retrovesicular ganglia.}
 \label{fig:clustering}
@@
 
-% --------- neural identification ----------------
-% The predicted activity is based on the correlations between neurons
-% learned from many recordings.  These predictions are a useful aid for
-% identifying unlabeled neurons.  For example, we see that
-% \textsf{SMBDL} was not found in the first worm
-% (Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
-% neurons closely matches the model's predictions for \textsf{SMBDL},
-% then this provides some evidence in favor of assigning it this
-% label. We tested this idea by withholding the labels of some
-% confidently-identified neurons and comparing their activity to the
-% model's predicted activity for all unlabeled neurons.  We measured the
-% similarity between the actual and predicted activity and found that
-% correct label, or ``candidate,'' tends to be much more similar than
-% other incorrect candidates, as summarized in
-% Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
-% the location of the neuron, this provides a powerful aid for identifying
-% neurons.  The more neurons that we can identify, the better we will be
-% able to learn about the dynamics of neural activity.
+The linear mapping from latent states to observations reveals clusters
+of similarly tuned neurons.  The mapping is given by a matrix with
+rows corresponding to neurons and columns corresponding to latent
+dimensions, shown in Fig.~\ref{fig:clustering}a.  Each row can be seen
+as a \emph{tuning vector} that specifies how the corresponding neuron
+responds to each dimension of the latent state.  This matrix reveals
+groups of similarly tuned neurons. To capture this, we used k-means to
+cluster the neurons based on their tuning vectors and found that 12
+clusters adequately captures the diversity of the population.  Using
+fewer clusters leads to qualitatively similar groups but omits some of
+the finer within-group differences; allowing more clusters eventually
+leads to groups of lateral pairs (like \textsf{ASKL} and \textsf{ASKR}
+in a.).  The neurons are ordered as in Fig.~\ref{fig:smoothing}, and
+clusters are separated by black lines.
 
-The linear mapping from latent states to observations reveals clusters
-of similarly tuned neurons.  The mapping maybe viewed as either an $N$
-(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
-vectors in $D$-dimensional space, i.e. as an embedding of neurons.
-These two views are shown in Fig.~\ref{fig:emissions}d. We note that
-many vectors point in similar directions, which implies that there are
-groups of correlated neurons.  We clustered the neurons based on the
-cosine similarity between their embedding vectors and found five clusters
-of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
-as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
-looked at their spatial distribution along the head-to-tail axis of the
-worm's body.  We found that clusters 2-4 correspond to activity closer to
-the head and working backward toward the tail.  The last and largest cluster
-contains many neurons in the nerve ring (location 0.15) \todo{check} as well
-as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}
+Fig.~\ref{fig:clustering}b shows the cosine similarity of tuning
+vectors for each pair of neurons, illustrating the differences
+and similarities between clusters.  We notice two prominent
+features.  First, the block structure of this
+matrix lends support to the notion of discrete clusters, indicating
+strong similarity between neurons within the same cluster. Second,
+we have sorted the clusters in order to make this matrix as
+diagonal as possible.  Upon doing so, we see a banded structure
+indicating that in addition to the discrete clustering, there is
+also a continuum along which these clusters vary, with the first
+clusters more similar to each other than to the latter clusters.
+Since the diagonal values equal one by definition, these
+have been grayed out.
+
+In our model, the emission matrix determines the correlations between
+neurons.  As such, we expect the similarity between rows of the
+emission matrix to mimic the empirical correlations observed in the
+data.  Fig.~\ref{fig:clustering}c shows that this is indeed the case,
+as neurons with similar tuning vectors also have higher empirical
+correlation coefficients.  However, without a hierarchical model, we
+cannot empirically estimate the correlation between pairs of neurons
+that were not simultaneously observed.  For this dataset, this leaves
+many entries in the empirical correlation matrix missing, as denoted
+by the gray off-diagonal entries. A key advantage of our hierarchical
+model is its ability to draw inferences in the face of significant
+amounts of missing data.
+
+What can explain this clustering?  We first look at the spatial
+locations of neurons in each cluster, under the hypothesis that
+clusters could be trivially explained by an inability to resolve
+differences in activity between nearby neurons.
+Fig.~\ref{fig:clustering}d shows the locations of all 61 labeled
+neurons, using the positions from WormAtlas\todo{cite} along the
+head-tail axis of the worm's body.  The size of the dot indicates the
+number of neurons at that location; the bars below denote the
+approximate location of the anterior (AG), dorsal (DG), ventral (VG)
+and retrovesicular (RVG) ganglia.  Some cluster (e.g. cluster 6) are
+spatially localized, but many clusters contain neurons from multiple
+head ganglia.  This suggests that location alone cannot explain the
+similarity between tuning vectors.
+
+\todo[inline]{Break down clustes by type (sensory/inter/motor
+  neuron).}
+
+\todo[inline]{Ask Manuel/Annika to label functional role of neurons.}
+
+\subsection*{Model predictions aid in neural identification}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=6in]{figures/v3/figure_id} 
+\caption{ \textit{The predicted neural activity is a useful aid for
+    neural identification.}  \textbf{a.} The model uses observed and
+  labeled activity (green traces) to predict the activity of unlabeled
+  neurons, as in Fig.~\ref{fig:smoothing}. \textbf{b.} Alongside the
+  labeled activity, we also have traces from many neurons that could
+  not be unambiguously labeled. To test whether the model's
+  predictions could aid in labeling neurons, we conducted a synthetic
+  experiment, withholding the activity of some labeled neurons. Here,
+  the red trace is the withheld activity of VB01 and the blue traces
+  are the activity of other unlabeled neurons.  Together, these make
+  up a pool of candidates for the VB01 label.  \textbf{c.} For each
+  candidate, we measured the correlation between its activity and the
+  predicted activity of VB01.  \textbf{d.} Repeating this experiment
+  50 times, we found that the correct candidates tend to be much more
+  correlated with the model's predicted activity than the other,
+  incorrect candidates.  Thus, the model's predictions serve as a
+  useful aid for labeling neurons.  }
+\label{fig:id}
+\end{figure}
+
+The predicted activity is based on the correlations between neurons
+learned from many recordings.  These predictions should be a useful
+aid for laneling neurons whose activity could be measured but whose
+identity could not be unambiguously resolved.  To test this
+hypothesis, we conducted a synthetic experiment in which we withheld
+the labels of ten neurons per worm.  We fit the model to the remaining
+neurons and then compared the model's predicted activity to the
+activity of the heldout neurons, as well as to the activity of the
+unlabeled neurons identified by~\citet{kato2015global}.
+
+Fig.~\ref{fig:id}a shows a schematic of the experiment: though VB01
+was labeled in the dataset, we withheld its activity and used the
+model to predict it (black trace) based on the observations of the
+other, labeled neurons (green traces).  We then compared the model
+predictions to the true activity of VB01 (red trace) and to the
+activity of other neurons that were identified but not labeled (blue
+traces), as shown in Fig.~\ref{fig:id}b.  We computed the correlation
+coefficient between these traces and the predicted activity of VB01,
+as shown in Fig.~\ref{fig:id}c.
+We find that in this case, the model's predictions and the true activity
+are highly correlated, even though the true activity was not used
+when training the model.  We repeated this experiment for 50 heldout
+neurons and found that this pattern holds: the predicted activity is
+typically much more correlated with the activity of the correct
+heldout trace (red histogram) than with the activity of other unlabeled
+neurons (blue histogram).  This suggests that our hierarchical model
+may serve an additional role in aiding neural identification in
+ambiguous, noisy data.
+
+\todo[inline]{Also discuss simultaneous matching experiments? Possible
+  improvements from including more labeled neurons.  Work with Gonzalo
+  on learning permutations.  Importance of capturing uncertainty in matching.}
 
@@
 \centering
-\includegraphics[width=6in]{figures/v2/figure3part1} 
+\includegraphics[width=6in]{figures/v3/figure4} 
 \caption{ \textit{The hierarchical SLDS automatically segments neural
-  activity based on its dynamics.}
-  \textbf{a.}~Three dimensions of the inferred continuous latent states
-  for each of the five worms, color coded by the corresponding discrete latent
-  state.
-  \textbf{b.}~We select the number of discrete states and justify the
-  hierarchcial, robust, and recurrent extensions by measuring the marginal likelihood
-  of held-out test data. We find that the hierarchical, robust, recurrent model
-  with~$K=8$ states performs best.
-  \textbf{c.}~The inferred segmentations align with the manual segmentations
-  of~\citet{kato2015global}.
-  \textbf{d.}~Moreover, the inferred states are in close correspondence with
-  the manually labeled states, which were shown to indicate different behaviors.
-}
+    activity based on its dynamics.}  \textbf{a.}~The top three
+  dimensions of the inferred continuous latent states for each of the
+  five worms, color coded by the corresponding discrete latent state.
+  \textbf{b.}~The inferred segmentations (top) align with the manual
+  segmentations (bottom) of~\citet{kato2015global}, though some
+  manually labeled states are split into multiple inferred states
+  (e.g. forest green state is often split into green and brown) and
+  others are combined (e.g. light purple, light blue, and white are
+  often combined into pink).  \textbf{c.}~We quantified the overlap
+  between the manual and inferred states and found a close
+  correspondence.  As~\citet{kato2015global} showed, the manually
+  labeled states map onto different behaviors (\textsf{REVSUS}:
+  sustained reversal; \textsf{SLOW}: slow forward crawl; \textsf{VT}:
+  ventral turn; \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn;
+  \textsf{REV1, REV2}: reversals; \textsf{NOSTATE}:
+  undetermined). The correspondence between inferred and manually
+  labeled states implies that hierarchical state space models can
+  infer behaviorally meaningful states directly from neural activity.}
 \label{fig:syllables}
@@
 states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
-worm's neural activity in the first three dimensions of the continuous
+worm's neural activity in the top three dimensions of the continuous
 state space, color coded by the discrete regime inferred under the
-SLDS.  We see that different regimes correspond to distinct loops
-through latent space.  Compare this with Fig. 4b
+SLDS.  These are the three dimensions with the highest percentage of
+expained variance. We see that different regimes correspond to
+distinct loops through latent space.  Compare this with Fig. 4b
 of~\citet{kato2015global}---they achieved a similar result by manually
@@
 
-Moreover, we can use the likelihood of held-out test data to determine
+We use the likelihood of held-out test data to determine
 the number of discrete regimes and assess the value of the robust and
-recurrent model extensions, as shown in Fig.~\ref{fig:syllables}b.
+recurrent model extensions, as shown in Supplemental Fig.~\ref{fig:modelselection}.
 While the non-robust models with Gaussian dynamics noise achieve
@@
 close correspondence between the manually labeled and inferred states.
-Fig.~\ref{fig:syllables}c shows ten-dimensional latent states and the
+Fig.~\ref{fig:syllables}b shows ten-dimensional latent states and the
 manual and inferred segmentations for a 5.5 minute section of data.
@@
 see instances where there appears to be a change in continuous state
-dynamics that are identified by the SLDS but missed in the manual
-segmentation (e.g. around $t=700$ seconds).  Fig.~\ref{fig:syllables}d
+dynamics that are identified by the SLDS before it is detected in the
+manual segmentation (e.g. around 11.8 min).  Fig.~\ref{fig:syllables}c
 quantifies the overlap in states for the first worm and for all worms.
 Many inferred states are in one-to-one correspondence with manually
-labeled states, and the remainder tend to be split between two manual
-states.  This confirms that the SLDS is a useful tool for automatically
-segmenting brain data.
+labeled states. Some manual states are split into two inferred states
+(e.g. ventral turns (\textsf{VT}) are split into green and brown) and
+others are combined (e.g. reversals and undetermined states
+(\textsf{REV1}, \textsf{REV2}, \textsf{NOSTATE}) map onto pink).
+The correspondence between manual and inferred states shows that
+the hierarchical SLDS is a useful tool for automatically segmenting
+brain data into behaviorally meaningful states.
 
 \subsection*{Quantifying the dynamics of neural activity in each discrete state}
+The hierarchical SLDS not only automates the laborious manual
+segmentation process, it justifies its segmentation with a
+quantitative definition of each inferred state. Each state is
+associated with a set of canonical linear dynamics, as shown in
+Fig.~\ref{fig:dynamics}a. The arrows in these vector fields show where
+the continuous state is mapped to under the corresponding discrete
+state (for the top three latent dimensions).  Arrows are only drawn
+at those locations where the discrete state was actually employed.
+
+States 1 (blue), 3 (yellow), and 5 (purple) are all characterized
+by convergence toward the origin, which corresponds, essentially, to
+no change in neural activity. These three states differ in terms of how
+quickly they converge to the origin and the amount of rotation as
+they do so.  By contrast, states 2 (red), 4 (green), 6 (orange), 7(brown)
+and 8 (pink) form parts of rotational loops in latent state space.
+As the latent state traces out a loop, it transiently activates
+the clusters of neurons that are tuned to the loop direction.
+These states are associated with dorsal turns, ventral turns, and
+reversals.
+
+We use these dynamics to simulate how brain activity evolves
+under the different states. Fig.~\ref{fig:dynamics}b shows how four
+simulated trajectories of each state from various starting locations map 
+into activations in the space of neural activity. For each discrete state,
+we chose the starting locations randomly from the set of locations where
+that state was entered in the real data.  We then simulated forward in
+time according to that state's dynamics until the simulation predicted
+a state change.  We then mapped these four simulations for each state
+through the emission matrix and measured the average activation induced
+on each of the 12 neural clusters.
+
+These activations lend further evidence in support of the behavioral
+labels attributed to each state. For eaxample, state 4 (green) coincides
+with the manually labeled ``ventral turn'' state.  Its dynamics produce
+a pronounced rise in the activity of cluster 6, which (referring back
+to Fig.~\ref{fig:clustering}), consists of the \textsf{RIVL/R} and
+\textsf{SMDVL/R} neurons.  The \textsf{RIV} pair are polymodal inter-
+and motor-neurons that specify the ventral bias of turns, and the
+\textsf{SMDV} are motor neurons that initiate ventral turns.  State 4
+dynamics also tend to suppress the activation of cluster 8, which
+consists mostly of the \textsf{SMB*} neurons.  As these neurons are
+implicated in dorsal turns, this again makes sense. Finally, we see
+that states 6 (orange) and 5 (purple) often produce a rise and fall, respectively,
+of activation on cluster 8, which together produce a dorsal turn.
+These two states do indeed coincide with the manually labeled dorsal
+turn state, again confirming the behavioral interpretation of these
+dynamical states.
+
+Finally, Fig.~\ref{fig:dynamics}c shows the eigenspectra of the linear
+dynamics for each worm and each state.  Imaginary eigenvalues give
+rise to rotational dynamics, and the magnitude of the eigenvalues
+determines whether the state is stable or not.  If the magnitude is
+greater than one (outside the unit disk denoted by the black arc),
+the dynamics produce exponential growth in the latent state space; if
+the eigenvalues are all less than unit magnitude, the dynamics converge
+to a fixed point.  As expected, the rotational states do indeed have
+more imaginary eigenvalues, and most states have eigenvalues with
+magnitude near one.  Some states, like the brown state, have eigenvalues
+larger than one, which produce unstable dynamics, though this is
+somewhat balanced by the recurrent extensions discussed in the next
+section.  Comparing across rows, we see only minor differences from worm
+to worm, substantiating our hierarchical modeling assumptions.
+
+\begin{figure} [b!]
+  \caption{ \textit{(Next page.) Discrete states are characterized by
+      simple linear dynamics that give rise to interpretable patterns
+      of neural activity.}  \textbf{a.}~Each discrete state is
+    associated with a canonical set of linear dynamics, which can be
+    viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show
+    the mappings for the first three dimensions here, with vectors
+    placed at locations where each state was used.  \textbf{b.}~We
+    simulated four latent state trajectories with these linear
+    dynamics from different starting locations, and then we mapped
+    these latent state trajectories through the emission matrix to
+    compute a time course of resulting activations for each cluster of
+    neurons.  \textbf{c.}~The eigenspectra of the dynamics matrices
+    provide insight into the nature of the states: imaginary eigenvalues
+    give rise to rotational dynamics and the magnitude determines
+    whether the stability of the dynamics. The arc denotes
+    the unit magnitude line; eigenvalues to the left are contractive
+    and to the right are expansive. Looking across rows, we
+    see how the spectra vary from worm to worm. }
+\end{figure}
+\addtocounter{figure}{-1}
+
 \begin{figure}[h!]
   \centering
-\includegraphics[width=6in]{figures/v2/figure3part2} 
-\caption{ \textit{Each state is characterized by simple linear dynamics.}
-    \textbf{a.}~Each discrete state is associated with a canonical set of linear
-  dynamics, which can be viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.
-  We show the mappings for the first three dimensions here, with vectors placed
-  at locations where each state was used.
-  \textbf{b.}~We simulate these linear dynamics from a variety of starting
-  locations and compute the time course of resulting activations for each
-  cluster of neurons.
-  \textbf{c.}~The eigenspectra of the dynamics matrices highlight the rotational
-  vs. contractive nature of the states and illustrate worm-to-worm variability. 
-}
+  \vspace{-.5in}
+  \includegraphics[width=6in]{figures/v3/figure5B}
+  \caption{Caption on preceding page.}
   \label{fig:dynamics}
@@
 
-Beyond simply automating a manual process, the SLDS also provides a
-quantitative description of the different states.  In particular, each
-state is associated with a set of linear dynamics.
-Fig.~\ref{fig:dynamics}a shows the canonical dynamics associated with
-each of the eight discrete states. The arrows in these vector fields
-show where the continuous state is mapped to under the corresponding
-discrete state (for the first three latent dimensions).  Arrows are
-only drawn at those locations where the discrete state was actually
-employed.  We see that these dynamics capture the rotational patterns
-of the various loops.
-
-We use these dynamics to simulate how brain activity would evolve
-under the different states. Fig.~\ref{fig:dynamics}b shows how four
-simulations of each state from various starting locations  map 
-into activations in the space of neural activity. In particular, we show how they would
-activate the six clusters shown in Fig.~\ref{fig:clustering}.  We see
-that state 4 corresponds to a pronounced dip followed by a rise in the
-activity of cluster 3, alongside a slow ramp in the activity of
-clusters 1 and 2.
-% Referring back to Fig~\ref{fig:emissions}a, we see
-% that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
-% neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
-% pair are sensory neurons responsible for initiating local search behavior,
-% the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
-% the ventral bias of turns, and the \textsf{SMDV} are motor neurons
-% that initiate ventral turns. As state 4 is in close correspondence with
-% the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
-% sense that these neurons would be activated.
-
 \clearpage
 
-\subsection*{Recurrent models learn spatially localized states and transition boundaries}
+\subsection*{Recurrent models learn spatially localized states and non-Markovian dynamics}
 
@@
 \centering
-\includegraphics[width=6in]{figures/v2/figure4v2} 
-\caption{ \textit{Discrete state transitions show a non-Markovian,
-    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
+\includegraphics[width=6in]{figures/v3/figure6} 
+\caption{ \textit{Discrete state transitions show a chain-like structure
+    with non-Markovian durations.}  \textbf{a.}~Counts of inferred transitions
   from one state to another. The states are permuted to emphasize the
@@
   imply a geometric distribution of state durations (dotted black
-  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
+  lines), but Fig.~\ref{fig:syllables} suggests that state
   transitions occur near the origin.  With the exception of state 7,
@@
 
-The final and most critical component of the dynamical model for
-neural activity in the head ganglia is the discrete transition
-distribution.  In contrast to standard SLDS, we allow these transition
-probabilities to depend on the location in continuous state space (the
-\emph{recurrent} dependency) as well as the preceding discrete state
-(the \emph{Markovian} dependency).  Fig.~\ref{fig:recurrent}a shows
-the contribution of the Markovian term.  Each cell in this matrix
-denotes the number of transitions from one state to another. We are
-only counting transitions in which the state changes, hence the
-diagonal is zero.  We have permuted the states to emphasize the
-largely feed-forward structure.
+The final component of our hierarchical state space model is the
+discrete transition distribution.  In contrast to standard SLDS, we
+allow these transition probabilities to depend on the location in
+continuous state space (via the \emph{recurrent} dependency) as well as
+the preceding discrete state (the \emph{Markovian} dependency).
+Fig.~\ref{fig:recurrent}a shows the contribution of the Markovian
+term.  Each cell in this matrix denotes the number of transitions from
+one state to another. We are only counting transitions in which the
+state changes, hence the diagonal is zero.  We have permuted the
+states to emphasize the largely feed-forward structure.
 
@@
 \centering
-\includegraphics[width=6in]{figures/v1/figure5} 
+\includegraphics[width=6in]{figures/v3/figure7} 
 \caption{\textit{Incorporating exogenous factors like oxygen level
@@
     different genetic strains and developmental stages.}
-  \textbf{a.}~The observed neurons and their inferred observation variances
-  for each of the 41 worms in the~\citet{nichols2017global} dataset.
-  \textbf{b.}~Worms of different genetic strains (N2, npr1) and developmental
-  stages (pre-lethargus, lethargus) exhibit different levels of overall
-  state usage, with blue and yellow much more likely in the lethargus stage.
-  \textbf{c.}~As above, the inferred states closely match the manually
-  labeled states of~\citet{nichols2017global}, with blue and yellow mapping
-  onto the forward (\textsf{Fwd}) and quiescent (\textsf{Q}) states.
-  \textbf{d.}~We extended the SLDS to allow exogenous factors like oxygen
-  level influence discrete transition probabilities.
-  \textbf{e., f.}~Upon fitting the model, we find that oxygen has a limited
-  effect on N2 worms and a much stronger effect on npr1 worms. In npr1 worms,
-  increasing the oxygen level decreases the probability of quiescent and
-  forward states and elicits reversals and turns.
-}
+  \textbf{a.}~The observed neurons variances for each of the 41 worms
+  in the~\citet{nichols2017global} dataset.  \textbf{b.}~Worms of
+  different genetic strains (N2, npr1) and developmental stages
+  (pre-lethargus, lethargus) exhibit different average state usages as
+  a function of the oxygen level.  Blue and red states are much more
+  likely in the lethargus stage for both genetic strains, but they are
+  suppressed in npr1 worms at high oxygen concentrations. Moreover, we
+  see a pronounced increase in the pink state usage immediately after
+  the oxygen level increases to 21\%.  \textbf{c.}~As above, the
+  inferred states closely match the manually labeled states
+  of~\citet{nichols2017global}, with blue, red, and to a lesser extent
+  green mapping onto the forward (\textsf{Fwd}) and quiescent
+  (\textsf{Q}) states.  \textbf{d.}~We extended the model to
+  incorporate oxygen level (O$_2$) and change in oxygen level
+  ($\Delta$O$_2$) as exogenous factors that drive discrete transition
+  probabilities.  We find that oxygen level has a limited effect on N2
+  worms but strongly suppresses the blue (\textsf{FWD/Q}) state in
+  npr1 worms and promotes the purple and orange
+  (\textsf{Rev/DT/VT/Other}) states. The largest effect arises from
+  the change in oxygen: an increase in $O_2$ strongly elicits the pink
+  state.  \textbf{e.} The state-triggered average latent state
+  trajectory following entry into the pink state shows a dip in
+  activity along the last dimension 7 seconds after entry (black
+  arrows).  \textbf{f.} This dip gives rise to strong opposing
+  responses of sensory neurons \textsf{URXL/R} and \textsf{BAGL}
+  (among others), which are known to respond to increases and
+  decreases, respectively, in oxygen concentration.  }
 \label{fig:o2}
@@
 
-Upon fitting our model, we immediately see a difference in the overall
-usage of the eight states (Fig.~\ref{fig:o2}b). The worms in the
-lethargus stage predominantly use the blue and yellow states.  Again,
+Upon fitting our model, we immediately see a difference in the average
+usage of the eight states  as a function of oxygen level(Fig.~\ref{fig:o2}b). The worms in the
+lethargus stage predominantly use the blue and red states.  Again,
 we found a close correspondence between our inferred states and the
 manually labeled states of~\citet{nichols2017global}
-(Fig.~\ref{fig:o2}c). Here, blue and yellow both map strongly onto the
+(Fig.~\ref{fig:o2}c). Here, blue and red both map strongly onto the
 quiescent (\textsf{Q}) state. This is consistent with the
@@
 lethargus N2 worms. We can quantify this effect with our models by
-including oxygen level as an exogenous input.  Fig.~\ref{fig:o2}d
-shows how the graphical model of the recurrent SLDS changes: now the
-discrete state at time~$t$ (red circle) depends on the preceding discrete
-state, the preceding continuous state, and the current oxygen level.
-\todo[inline]{consider alternative models in which oxygen level
-  influences continuous states and observations?}
+including oxygen level (O$_2$) and change in oxygen level ($\Delta$O$_2$)
+as exogenous inputs.
+% Fig.~\ref{fig:o2}d
+% shows how the graphical model of the recurrent SLDS changes: now the
+% discrete state at time~$t$ (red circle) depends on the preceding discrete
+% state, the preceding continuous state, and the current oxygen level.
+% \todo[inline]{consider alternative models in which oxygen level
+%   influences continuous states and observations?}
+Fig.~\ref{fig:o2}d (top) shows the inferred effect of oxygen level for each
+of the four strain/developmental stage pairs.  As predicted,
+higher oxygen levels increase the probability of reversals and
+turning related states (orange, purple) and decrease the probability
+of forward and quiescent (blue, red, green) states in npr1 worms, but
+has limited effect on N2 worms. We do see an increase in turning related
+states (orange) in the N2 lethargus worms, but inspection of panel b
+indicates this state is still used very sparingly.  Somewhat
+surprisingly, the magnitude of the effect is comparable for both
+pre-lethargus and lethargus npr1 worms, even though the baseline level
+of reversals and turns is much higher in the pre-lethargus worms.
 
-Fig.~\ref{fig:o2}e and~\ref{fig:o2}f show the inferred effect of
-oxygen level for each of the four strain/developmental stage pairs.
-As predicted, only the npr1 worms show a much larger response to
-oxygen. Increasing the oxygen level increases the probability
-of reversals and turning related states (red, orange, purple) and
-decreases the probability of forward and quiescent (blue, yellow)
-states. Somewhat surprisingly, the magnitude of the effect is
-comparable for both pre-lethargus and lethargus npr1 worms, even
-though the baseline level of reversals and turns is much higher in
-the pre-lethargus worms.
+The largest effect is in response to the change in oxygen level, as
+shown in Fig.~\ref{fig:o2}d (bottom).  Across all worms, the change
+from 10\% to 21\% oxygen concentration sharply increases the probability
+of transitioning into the pink state. Moreover, this state is seldom
+used except for when the oxygen level increases.  This suggests that
+sharp increases in oxygen level induce a transient and characteristic
+neural response that is not seen in steady-state oxygen levels.
+
+Fig.~\ref{fig:o2}e shows the average latent state trajectory triggered
+upon entry into the pink state (for this dataset, the optimal performance
+was achieved by a 6-dimensional state space model).  We see a dip and
+rebound along the last dimension approximately 7 seconds after entry
+(denoted by the black arrows).  We identified the neurons that showed
+the largest response at this point in time and plotted the top ten
+in Fig.~\ref{fig:o2}f.  Unsurprisingly, these neurons are largely sensory
+and interneurons.  In particular, \textsf{URXL/R} and \textsf{BAGL/R} are
+known to jointly encode changes in oxygen level with opposing signs: \textsf{URXL/R}
+respond to increases in oxygen, \textsf{BAGL/R} respond to dcreases.  Our
+model recapitulates these results and suggests that many other sensory
+neurons respond similarly to this stimulus.
 

### 2018-01-05 experiments/fit_arhmm.py

@@ -58,12 +58,12 @@ from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states
     plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_duration_cdfs
 
 # LDS Results
-lds_dir = os.path.join("results", "2017-11-03-hlds", "run002")
+lds_dir = os.path.join("results", "2017-11-03-hlds", "run003")
 # lds_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
 assert os.path.exists(lds_dir)
 
 # AR-HMM RESULTS
-results_dir = os.path.join("results", "2017-11-04-arhmm", "run003")
+results_dir = os.path.join("results", "2017-11-04-arhmm", "run004")
 # results_dir = os.path.join("results", "2017-11-04-arhmm", "run004_dff_bc")
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
@@ -552,18 +552,67 @@ def plot_best_model_results(do_plot_expected_states=True,
 
         plt.close("all")
 
-    if do_plot_eigenspectrum:
-        markers = ['o', '^', 's', 'p', 'h']
-        for i, hdd in enumerate(hier_dynamics_distns):
-            width = 1.0 if i == 0 else 0.7
-            left = 0.3 if i == 0 else 0.05
-            fig = plt.figure(figsize=(width, 1.0))
+    # if do_plot_eigenspectrum:
+    #     markers = ['o', '^', 's', 'p', 'h']
+    #     for i, hdd in enumerate(hier_dynamics_distns):
+    #         width = 1.0 if i == 0 else 0.7
+    #         left = 0.3 if i == 0 else 0.05
+    #         fig = plt.figure(figsize=(width, 1.0))
+    #
+    #         # ax = fig.add_subplot(111, aspect="equal")
+    #         from hips.plotting.layout import create_axis_at_location
+    #         ax = create_axis_at_location(fig, left, 0.2, 0.6, 0.6)
+    #         for w, dd in enumerate(hdd.regressions):
+    #             evs = np.linalg.eigvals(dd.A[:,:-1])
+    #             assert np.all(evs.real >= 0.45)
+    #             assert np.all(evs.real <= 1.2)
+    #             assert np.all(evs.imag >= -0.3)
+    #             assert np.all(evs.imag <= 0.3)
+    #
+    #             ax.plot(np.real(evs), np.imag(evs),
+    #                     ls='',
+    #                     marker=markers[w],
[…]

### 2018-01-05 experiments/fit_hlds.py

@@ -48,17 +48,15 @@ from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
 
 # IO
-run_num = 2
+run_num = 3
 results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
 signal = "dff_diff"
-# results_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
-# signal = "dff_bc"
 
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
 N_worms = 5
-N_clusters = 6
+N_clusters = 12
 
 
 def cached(results_name):
@@ -258,10 +256,10 @@ def plot_likelihoods(D_latents, final_lls, hlls, best_index,
     return ax1, ax2
 
 
-def order_latent_dims(xs, C, ytrains, mtrains):
+def order_latent_dims(xtrains, C, ytrains, mtrains):
 
     # Sort latent dimensions by how much variance they account for
-    D_latent = xs[0].shape[1]
+    D_latent = xtrains[0].shape[1]
     corrcoeffs = np.zeros(D_latent)
     for d in range(D_latent):
         yobss = []
@@ -269,37 +267,43 @@ def order_latent_dims(xs, C, ytrains, mtrains):
         for i in range(N_worms):
             mask = mtrains[i][0]
             yobss.append(ytrains[i][:, mask].ravel())
-            yhats.append(np.outer(xs[i][:,d], C[mask,d]).ravel())
+            yhats.append(np.outer(xtrains[i][:, d], C[mask, d]).ravel())
         yobss = np.concatenate(yobss)
         yhats = np.concatenate(yhats)
-        corrcoeffs[d] = np.corrcoef(yobss.ravel(), yhats.ravel())[0,1]
-    return np.argsort(corrcoeffs)[::-1]
+
+        corrcoeffs[d] = np.corrcoef(yobss.ravel(), yhats.ravel())[0, 1]
+
+    perm = np.argsort(corrcoeffs)[::-1]
+    return perm
 
 
 def cluster_neruons(best_model, seed=0):
     from pyhsmm.util.general import relabel_by_permutation
-    from sklearn.cluster import SpectralClustering, KMeans
-    C_true = best_model.emission_distn.A[:, :-1].copy()
-    C_true /= np.linalg.norm(C_true, axis=1)[:, None]
[…]

### 2018-01-05 zimmer/plotting.py

@@ -294,7 +294,7 @@ def plot_vector_field_2d(ii, z, x, perm_dynamics_distns, colors,
     ax.set_ylim(lims)
 
 
-def plot_vector_field_3d(ii, z, x, perm_dynamics_distns, colors,
+def plot_vector_field_3d(ii, z, x, dynamics_distns, colors,
                          ax=None, lims=(-3,3), N_plot=500,
                          **kwargs):
 
@@ -314,9 +314,9 @@ def plot_vector_field_3d(ii, z, x, perm_dynamics_distns, colors,
         ini_inds = np.random.choice(ini.size, replace=False, size=N_plot)
         ini = ini[ini_inds]
 
-    Ai_full = perm_dynamics_distns[ii].A \
-        if hasattr(perm_dynamics_distns[ii], 'A') \
-        else perm_dynamics_distns[ii].A_0
+    Ai_full = dynamics_distns[ii].A \
+        if hasattr(dynamics_distns[ii], 'A') \
+        else dynamics_distns[ii].A_0
 
     Ai = Ai_full[:, :D]
     bi = Ai_full[:, D] if Ai_full.shape[1] == D+1 else 0
@@ -597,7 +597,7 @@ from pyhsmm.util.general import rle
 def plot_latent_trajectories_vs_time(xs, zs,
                                      colors=None,
                                      plot_slice=(0, 500),
-                                     alpha=0.5,
+                                     alpha=.75,
                                      title=None,
                                      show_xticks=True,
                                      basename="x_segmentation",
@@ -610,8 +610,8 @@ def plot_latent_trajectories_vs_time(xs, zs,
         x = x / (2 * lim)
         D_latent  = x.shape[1]
 
-        fig = plt.figure(figsize=(2.4, 1.2))
-        ax = create_axis_at_location(fig, 0.3, 0.3, 2.0, .7)
+        fig = plt.figure(figsize=(4.25, 1.35))
+        ax = create_axis_at_location(fig, 0.3, 0.3, 3.85, .85)
 
         # Plot z in background
         offset = 0
@@ -717,19 +717,19 @@ def plot_state_overlap(z_finals, z_trues,
 
     def _plot_overlap(overlap, perm, filename, title):
 
-        fig = plt.figure(figsize=(1.5, 1.25))
-        ax1 = create_axis_at_location(fig, .4, .3, .75, .75)
+        fig = plt.figure(figsize=(1.75, 1.25))
[…]

### 2018-01-10 drafts/elife/draft.tex
syncing

@@
   states of population neural activity, along with the discrete
-  behavioral modes that drive dynamics in this latent state space.  We
-  find stochastic transition patterns between these discrete
-  behavioral modes, and we see that transition probabilities are
-  determined by a combination of current brain state and environmental
-  cues.  In addition to quantifying neural dynamics, this
-  probabilistic framework aids in neural identification---currently a
-  laborious, manual task---and reveals clusters of neurons that are
-  similarly tuned in latent state space.  Finally, we find a
-  significant overlap between our inferred states and the
-  manually-labeled states of~\citet{kato2015global}
-  and~\citet{nichols2017global}, which were shown to correspond to
-  different behaviors, like forward crawling, reversals, and
-  turns. Our methods automatically discover and quantify these
-  behaviorally-meaningful states directly from neural activity,
-  yielding powerful new tools for neuro-behavioral analysis.
+  behavioral ``syllables'' that drive dynamics in this latent state
+  space.  We find stochastic transition patterns between these
+  syllables, and we see that transition probabilities are determined
+  by a combination of current brain state and environmental cues.  In
+  addition to quantifying neural dynamics, this probabilistic
+  framework aids in neural identification---currently a laborious,
+  manual task---and reveals clusters of neurons that are similarly
+  tuned in latent state space.  Finally, we find a significant overlap
+  between our inferred syllables and the manually-labeled states
+  of~\citet{kato2015global} and~\citet{nichols2017global}, which were
+  shown to correspond to different behaviors, like forward crawling,
+  reversals, and turns. Our methods automatically discover and
+  quantify these behaviorally-meaningful states directly from neural
+  activity, yielding powerful new tools for neuro-behavioral analysis.
 \end{abstract}
@@
 \section*{Introduction}
+
+\textbf{Context:}
+\begin{itemize}
+\item Novel recording technologies enable simultaneous recording of
+  O(100) of neurons simultaneously.  In \celegans, this constitutes
+  the bulk of the head ganglia, and in other organisms this may constitue
+  the majority of neurons in a specific circuit.
+\item The promise is that these new recordings will lead to new
+  insights into brain function. For example, with a complete view of a
+  circuit, we should be able to study how it responds to external
+  inputs, how it represents its internal state, and how it transforms
+  internal states into circuit outputs.
+\item \celegans~is the ideal model organism for realizing
+  this goal.  It has a stereotyped nervous system with 302 identifiable
+  neurons.  Its head ganglia is responsible for
+  processing sensory inputs and generating outputs that initiate
+  motor commands, like forward crawling, reversals, and turns.
+\item With new recording capabilities, we can study how the
+  coordinated activity of the \celegans~head ganglia encodes internal
+  state, updates it in response to sensory information (like oxygen
+  concentration), and transmits it via output (motor) neurons.
+\item Moreover, we can study how the encoding, dynamics, and outputs
+  differ across individual worms, genetic strains, and developmental stages.
+\end{itemize}
+
+\textbf{Three Problems: How to conceptualize the latent states of
+  brain activity, their dynamics, and their relation to observed
+  activity?  How to learn these dynamics from partial recordings of
+  many individual worms? How to quantify differences between worms of
+  different genetic strains and developmental stages?}
+\begin{itemize}
+\item We seek a model that is flexible enough to capture the complexity
+  of neural dynamics and the relationship between latent states and
+  observed activity, but we also require that the model be composed
+  of interpretable building blocks. 
+\item We have a spectrum of options --- simple factor analyzers and
+  clustering algorithms, to Kalman filters and hidden Markov models,
+  all the way up to highly nonlinear neural network models.
+\item The challenge is to strike a balance: flexible yet interpretable.
+\item While, in aggregate, we have a great deal of data to fit these models
+  with, the recordings are spread over many individuals. In
+  each individual we can only label a handful of neurons and we only
+  record them for tens of minutes (photobleaching).
+\item Moreover, each individual has its own idiosyncrasies due to
+  factors like different levels of GCaMP expression from one worm
+  to another, as well as subtle differences in the underlying brain
+  dynamics. 
+\item We need methods that combine these limited, partial recordings
+  in order to learn shared structure shared across the population.
+\item Finally, in order to understand the differences between
+  genetic strains and developmental stages, we need models that allow
+  us to flexibly test hypotheses about how they differ. 
+\end{itemize}
+
+\textbf{Main results, in broad brushstrokes:}
+\begin{itemize}
+\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms. The framework is:
+  \begin{itemize}
+  \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
+  \item \emph{robust}, ameliorating some concerns of model misspecification;
+  \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
+  \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
+  \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
+  \end{itemize}
+\item We use this framework to extract low-dimensional state
+  trajectories for each worm, using shared correlations across the
+  population to align the trajectories and make predictions about
+  unseen activity.
+\item We discover clusters of similarity-tuned and highly correlated
+  neurons, and we find that these neurons are functionally grouped
+  according the different behaviors they are implicated in.
+\item We show how our model's predictions can aid in identifying neurons (which will then drive down posterior uncertainty in the underlying dynamics of neural activity).
+\item We show that brain dynamics are well-approximated as switching between discrete regimes, or ``syllables,'' each associated with simple linear dynamics.
+These syllables map onto the manual segmentations of experimental experts, which were based on known behavioral implications of certain patterns of brain activity.
+\item We quantify these linear dynamical regimes, show how they drive activity in clusters of neurons, and evaluate their differences from worm to worm.
+\item We find that transitions between syllables are not simply Markovian; rather, the transitions depend on a combination of syllable history and latent brain state.
+\item We measure how oxygen concentration (and change in oxygen concentration) differentially impact state transitions in worms of different genetic strains and developmental stages, and we find that sudden increases in oxygen elicit a characteristic, transient response regardless of genetic make-up or developmental stage, but that only certain genetic strains show a sustained response to increased oxygen levels.
+\end{itemize}
 
@@
 can be learned from noisy, partial recordings with trial-to-trial
-variability.
-
-We study the data presented in~\citet{kato2015global} and
+variability.  We call this model \textbf{TBD}.\todo{name it!}
+
+We analyze the data presented in~\citet{kato2015global} and
 \citet{nichols2017global}, which consist of a collection of optical
@@
   information across multiple worms in order to learn a shared
-  dynamical system for \celegans.}
+  dynamics model for \celegans.}
   \textbf{a.}~The data consists of observed neural activity collected
@@
   the vector field.
+  \textbf{e.}~Though location in continuous state space helps inform
+  transitions, the current discrete state does as well. This is captured
+  by the Markov transition
   \textbf{e.}~To increase robustness to model misspecification, we
@@
 \centering
-\includegraphics[width=6in]{figures/v3/figure2} 
+\includegraphics[width=6in]{figures/v4/figure2} 
 \caption{ \textit{Hierarchical state space models smooth noisy
@@
 log likelihood of the held-out data as we increase the latent
-dimensionality. We find that the test likelihood peaks with a
-10-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).
-
-Moreover, the likelihood is substantially higher in the model that
-includes per-worm observation variances compared to the model with
-shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
-the inferred variance for each worm in the \citet{kato2015global}
-dataset.  Neurons that were not identified are marked with a gray box.
-This shows the extent of the missing data and the need for
-hierarchical modeling.\todo{make supplement figure}
+dimensionality. We find that the test likelihood is maximized with a
+10-dimensional latent state, and that variances differ substantially
+between worms.
+
+% Moreover, the likelihood is substantially higher in the model that
+% includes per-worm observation variances compared to the model with
+% shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
+% the inferred variance for each worm in the \citet{kato2015global}
+% dataset.  Neurons that were not identified are marked with a gray box.
+% This shows the extent of the missing data and the need for
+% hierarchical modeling.\todo{make supplement figure}
 
@@
 \centering
-\includegraphics[width=6in]{figures/v3/figure3B} 
+\includegraphics[width=6in]{figures/v4/figure3} 
 \caption{ \textit{The mapping from continuous latent states to neural
@@
 \centering
-\includegraphics[width=6in]{figures/v3/figure_id} 
+\includegraphics[width=6in]{figures/v4/figure_4} 
 \caption{ \textit{The predicted neural activity is a useful aid for
@@
 \centering
-\includegraphics[width=6in]{figures/v3/figure4} 
+\includegraphics[width=6in]{figures/v4/figure5} 
 \caption{ \textit{The hierarchical SLDS automatically segments neural
     activity based on its dynamics.}  \textbf{a.}~The top three
-  dimensions of the inferred continuous latent states for each of the
-  five worms, color coded by the corresponding discrete latent state.
-  \textbf{b.}~The inferred segmentations (top) align with the manual
-  segmentations (bottom) of~\citet{kato2015global}, though some
-  manually labeled states are split into multiple inferred states
-  (e.g. forest green state is often split into green and brown) and
-  others are combined (e.g. light purple, light blue, and white are
-  often combined into pink).  \textbf{c.}~We quantified the overlap
-  between the manual and inferred states and found a close
-  correspondence.  As~\citet{kato2015global} showed, the manually
-  labeled states map onto different behaviors (\textsf{REVSUS}:
-  sustained reversal; \textsf{SLOW}: slow forward crawl; \textsf{VT}:
-  ventral turn; \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn;
-  \textsf{REV1, REV2}: reversals; \textsf{NOSTATE}:
-  undetermined). The correspondence between inferred and manually
-  labeled states implies that hierarchical state space models can
-  infer behaviorally meaningful states directly from neural activity.}
+  dimensions, according to variance explained, of the inferred
+  continuous latent states for each of the five worms, color coded by
+  the corresponding discrete latent state.  \textbf{b.}~The same
+  continuous latent states but now color coded according to the manual
+  segmentation of~\citet{kato2015global}.  \textbf{c.}~The inferred
+  segmentations (top) and the manual segmentations (bottom) over time
+  for worm 5. The segmentations largely align, though some manually
+  labeled states are split into multiple inferred states (e.g. forest
+  green state is often split into green and brown) and others are
+  combined (e.g. light purple, light blue, and white are often
+  combined into pink).  \textbf{d.}~We quantified the overlap between
+  the manual and inferred states and found a close correspondence.
+  As~\citet{kato2015global} showed, the manually labeled states map
+  onto different behaviors (\textsf{REVSUS}: sustained reversal;
+  \textsf{SLOW}: slow forward crawl; \textsf{VT}: ventral turn;
+  \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn; \textsf{REV1,
+    REV2}: reversals; \textsf{NOSTATE}: undetermined). The
+  correspondence between inferred and manually labeled states implies
+  that hierarchical state space models can infer behaviorally
+  meaningful states directly from neural activity.}
 \label{fig:syllables}
@@
 
-Finally, Fig.~\ref{fig:dynamics}c shows the eigenspectra of the linear
-dynamics for each worm and each state.  Imaginary eigenvalues give
-rise to rotational dynamics, and the magnitude of the eigenvalues
-determines whether the state is stable or not.  If the magnitude is
-greater than one (outside the unit disk denoted by the black arc),
-the dynamics produce exponential growth in the latent state space; if
-the eigenvalues are all less than unit magnitude, the dynamics converge
-to a fixed point.  As expected, the rotational states do indeed have
-more imaginary eigenvalues, and most states have eigenvalues with
-magnitude near one.  Some states, like the brown state, have eigenvalues
-larger than one, which produce unstable dynamics, though this is
-somewhat balanced by the recurrent extensions discussed in the next
-section.  Comparing across rows, we see only minor differences from worm
-to worm, substantiating our hierarchical modeling assumptions.
-
-\begin{figure} [b!]
-  \caption{ \textit{(Next page.) Discrete states are characterized by
+\begin{figure}[h!]
+  \centering
+  \vspace{-.5in}
+  \includegraphics[width=6in]{figures/v4/figure6}
+  \caption{\textit{Discrete states are characterized by
       simple linear dynamics that give rise to interpretable patterns
@@
     compute a time course of resulting activations for each cluster of
-    neurons.  \textbf{c.}~The eigenspectra of the dynamics matrices
-    provide insight into the nature of the states: imaginary eigenvalues
-    give rise to rotational dynamics and the magnitude determines
-    whether the stability of the dynamics. The arc denotes
-    the unit magnitude line; eigenvalues to the left are contractive
-    and to the right are expansive. Looking across rows, we
-    see how the spectra vary from worm to worm. }
-\end{figure}
-\addtocounter{figure}{-1}
-
-\begin{figure}[h!]
+    neurons.}
+  \label{fig:dynamics}
+\end{figure}
+
+\clearpage
+
+\subsection*{Recurrent models learn spatially localized states and non-Markovian dynamics}
+
+\begin{figure}[t!]
 \centering
-  \vspace{-.5in}
-  \includegraphics[width=6in]{figures/v3/figure5B}
-  \caption{Caption on preceding page.}
-  \label{fig:dynamics}
-\end{figure}
-
-\clearpage
-
-\subsection*{Recurrent models learn spatially localized states and non-Markovian dynamics}
-
-\begin{figure}[t!]
-\centering
-\includegraphics[width=6in]{figures/v3/figure6} 
+\includegraphics[width=6in]{figures/v4/figure7} 
 \caption{ \textit{Discrete state transitions show a chain-like structure
@@
 \centering
-\includegraphics[width=6in]{figures/v3/figure7} 
+\includegraphics[width=6in]{figures/v4/figure8} 
 \caption{\textit{Incorporating exogenous factors like oxygen level
@@
 
+\section{Supplementary Figures}
+
+\begin{itemize}
+\item Graphical model (with arrows $y_t \to y_{t+1}$)
+\item Neural activity for worms 2, 3, 4?
+\item Model selection based on held-out log likelihood
+\item Table of matching results?
+\item Eigenspectra of inferred states
+\item Model fitting 
+\end{itemize}
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=5in]{figures/v3/figure1_supp} 
+\caption{\textit{Graphical depiction of a hierarchical, robust, recurrent,
+    and input-driven switching linear dynamical system (SLDS).}
+  The blue arrow denotes the \textbf{hierarchical} component: each worm~$w$ has its
+  own set of parameters~$\theta^{(w)}$, which contain the dynamics libraries, emission
+  matrix, etc. These are slight perturbations of the same global mean parameters~$\theta$.
+  In some cases, the worm-specific parameters are required to equal their global
+  counterparts (as with the emission matrix). 
+  Red arrows highlight the \textbf{robust} noise model that links~$x_t$ to~$x_{t+1}$.
+  Green arrows indicate the \textbf{recurrent} extension: the next discrete syllable depends
+  on both the preceding syllable as well as the preceding continuous latent state.
+  Purple arrows carry the \textbf{input-dependence}: the next discrete syllable
+  can also be governed by an exogenous input, like the oxygen level or the change
+  in oxygen level. The remaining black arrows denote standard dependencies
+  of the SLDS: discrete syllables and continuous latent states co-evolve, with
+  the syllables specifying the linear dynamics to use when updating the continuous
+  latent states.  The observations are a function of these underlying latent states.
+  The only non-standard addition is an autoregressive dependency between~$y_t$ and
+  $y_{t+1}$; here, the SLDS models only the differences in observed fluoresence
+  value between two consecutive frames.
+}
+\label{fig:graphical_model}
+\end{figure}
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure5_supp1} 
+\caption{\textit{Model selection on the basis of marginal log
+    likelihood of test data.}  We withheld the last 20\% of the
+  continuous latent states from each worm in order to compare
+  different models.  For each combination of hierarchical
+  vs. standard, robust vs standard, and recurrent vs. standard, we
+  swept from~$K=2$ syllables to~$K=20$ (from white to purple)
+  syllables in steps of size two. We also included the special case
+  of~$K=1$ (grey bars), which reduces to standard autoregressive
+  models. For each combination, we fit the model and evaluated the
+  marginal likelihood of the held-out test data. The best-performing
+  model was the hierarchical, robust, recurrent SLDS with $K=8$
+  syllables.}
+\label{fig:model_selection}
+\end{figure}
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure5_supp2} 
+\caption{\textit{Discrete states carry substantial information about
+    continuous latent states.} How much information is carried by the
+  discrete syllables alone? In blue we show the mean and two standard deviations of
+  the predictive distribution~$p(x \given z)$.  The esimated values of~$x$
+  are shown in black. Below we show the inferred syllables~$z$. We see that
+  the discrete syllable carry a surprising amount of information about the continuous
+  latent states.}
+\label{fig:z_predictions}
+\end{figure}
+
+\subsection{Eigenspectra of the inferred dynamics}
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure6_supp} 
+\caption{The eigenspectra of the dynamics matrices
+    provide insight into the nature of the states: imaginary eigenvalues
+    give rise to rotational dynamics and the magnitude determines
+    whether the stability of the dynamics. The arc denotes
+    the unit magnitude line; eigenvalues to the left are contractive
+    and to the right are expansive. Looking across rows, we
+    see how the spectra vary from worm to worm. 
+}
+\label{fig:eigen}
+\end{figure}
+
+Fig.~\ref{fig:eigen} shows the eigenspectra of the linear
+dynamics for each worm and each state.  Imaginary eigenvalues give
+rise to rotational dynamics, and the magnitude of the eigenvalues
+determines whether the state is stable or not.  If the magnitude is
+greater than one (outside the unit disk denoted by the black arc),
+the dynamics produce exponential growth in the latent state space; if
+the eigenvalues are all less than unit magnitude, the dynamics converge
+to a fixed point.  As expected, the rotational states do indeed have
+more imaginary eigenvalues, and most states have eigenvalues with
+magnitude near one.  Some states, like the brown state, have eigenvalues
+larger than one, which produce unstable dynamics, though this is
+somewhat balanced by the recurrent extensions discussed in the next
+section.  Comparing across rows, we see only minor differences from worm
+to worm, substantiating our hierarchical modeling assumptions.
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure8_supp1} 
+\caption{\textit{Smoothing and prediction of neural responses
+    in \citet{nichols2017global} dataset.} Like Fig.~\ref{fig:smoothing},
+  this shows a three minute section of data.  We show one instance
+  from each combination of N2 vs. npr1 and pre-lethargic vs. lethargic.
+  The three minute window has been centered on the 6 minute mark, at
+  which point the oxygen level was increased from 10\% to 21\%.  }
+\label{fig:nichols_supp1}
+\end{figure}
+
+% This elicits a strong response in some clusters of neurons, regardless
+% of genetic background and developmental stage.  In npr1 worms, the
+% elevated oxygen level will elicit a sustained period of activity, even
+% in the lethargic stage, whereas N2 worms remain largely quiescent
+% despite the increased oxygen concentration.
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure8_supp2} 
+\caption{\textit{Inferred continuous latent states for the four worms
+    shown in Fig.~\ref{fig:nichols_supp1}.} As in
+  Fig.~\ref{fig:syllables}, we show the continuous trajectories
+  color-coded by both the inferred syllable sequence and by the manual
+  segmentation. We show this for dimensions (1, 2, 3) and (4, 5, 6),
+  sorted by percentage of variance explained. Below, we show the total
+  overlap between manual and inferred states, averaging over all worms
+  of the given genetic strain and developmental stage.  In all cases,
+  there is a reasonable mapping between manual and inferred states,
+  though it is not as clear-cut as in the~\citet{kato2015global}
+  dataset. In particular, the distinction between forward (Fwd) and
+  quiescent (Q) is less clear, with both being mapped to the blue
+  state.}
+\label{fig:nichols_supp2}
+\end{figure}
+
 \end{document}

### 2018-01-10 drafts/elife/figures/figure1/make_figure_1.py

@@ -332,6 +332,39 @@ def plot_noise_density(lims=(-3, 3), n_pts=50):
     plt.savefig("fig1_noise.pdf")
     plt.show()
 
+def plot_trans_matrix():
+    T = np.eye(3, 3) + np.random.gamma(shape=2, scale=.5, size=(3,3))
+    T /= T.sum(axis=1, keepdims=True)
+    print(T)
+
+    fig = plt.figure(figsize=(2.5, 1.5))
+    ax = fig.add_subplot(111)
+
+    im = ax.imshow(T, vmin=0, vmax=1, aspect="equal", cmap="Greys")
+    ax.set_xticks([])
+    ax.set_yticks([])
+
+    from mpl_toolkits.axes_grid1 import make_axes_locatable
+    from hips.plotting.colormaps import gradient_cmap
+    cmap = gradient_cmap(colors[:3])
+    divider = make_axes_locatable(ax)
+    lax = divider.append_axes("left", size="5%", pad=0.05)
+    lax.imshow(np.arange(K)[:, None], cmap=cmap, vmin=0, vmax=2, aspect="auto", interpolation="nearest")
+    lax.set_xticks([])
+    lax.set_yticks([])
+
+    bax = divider.append_axes("bottom", size="5%", pad=0.05)
+    bax.imshow(np.arange(K)[None, :], cmap=cmap, vmin=0, vmax=2, aspect="auto", interpolation="nearest")
+    bax.set_xticks([])
+    bax.set_yticks([])
+
+    cax = divider.append_axes("right", size="5%", pad=0.05)
+    plt.colorbar(im, cax=cax)
+
+    plt.tight_layout(pad=0.1)
+    plt.savefig("fig1_transmatrix.pdf")
+    plt.show()
+
 if __name__ == "__main__":
     # Simulate data
     As, bs = make_dynamics_library_2D()
@@ -373,4 +406,5 @@ if __name__ == "__main__":
     # plot_continuous_latent_states()
     # plot_neural_activity()
     # plot_observation_variance()
-    plot_noise_density()
+    # plot_noise_density()
+    plot_trans_matrix()

### 2018-01-10 experiments/fit_arhmm.py

@@ -41,7 +41,7 @@ zimmer_colors = [cm(i) for i in np.linspace(0.05, 0.95, 8)]
 # Modeling stuff
 from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
 
-from autoregressive.models import ARWeakLimitStickyHDPHMM
+from autoregressive.models import ARWeakLimitStickyHDPHMM, ARHMM
 from pyslds.util import get_empirical_ar_params
 from pybasicbayes.distributions import AutoRegression, RobustAutoRegression
 from rslds.models import SoftmaxRecurrentARHMM
@@ -94,6 +94,28 @@ def cached(results_name):
             return results
         return func_wrapper
     return _cache
+
+
+def compute_baseline_likelihood(xtrains, xtests, alpha=3.):
+    model = ARHMM(init_state_distn="uniform",
+                  obs_distns=[AutoRegression(**ar_params)],
+                  alpha=alpha)
+
+    for xtr in xtrains:
+        model.add_data(xtr)
+
+    # Fit the model with Gibbs
+    lls = []
+    for itr in tqdm(range(N_samples)):
+        model.resample_model()
+        lls.append(model.log_likelihood())
+
+    # Compute heldout likelihood
+    hll = 0
+    for xte in xtests:
+        hll += model.log_likelihood(xte)
+
+    return hll
 
 
 def _fit_model_wrapper(K, alpha=3, gamma=100., kappa=100.,
@@ -177,42 +199,48 @@ def _fit_model_wrapper(K, alpha=3, gamma=100., kappa=100.,
     return model, np.array(lls), hll, None
 
 
-def plot_likelihoods(group, Ks, final_lls, hlls, best_index,
-                     color, name, axs=None):
+def plot_likelihoods(group, Ks, hlls, best_index,
+                     name,
+                     baseline=None,
+                     axs=None):
 
     # Plot results of searching over models
     if axs is None:
-        plt.figure(figsize=(2.4,1.8))
+        plt.figure(figsize=(6, 3))
         # ax1 = plt.subplot(211)
         # ax2 = plt.subplot(212)
         ax = plt.subplot(111)
     else:
         ax = axs
+    cmap = gradient_cmap([np.ones(3), colors[4]])
 
     M = len(Ks)
     T_te = T_tests[:N_worms].sum()
 
-    cmap = gradient_cmap([np.ones(3), colors[4]])
[…]

### 2018-01-10 experiments/fit_arhmm_nichols.py

@@ -65,11 +65,11 @@ from hips.plotting.layout import create_axis_at_location
 from hips.plotting.colormaps import white_to_color_cmap
 
 # LDS Results
-lds_dir = os.path.join("results", "nichols", "2018-01-03-hlds", "run001")
+lds_dir = os.path.join("results", "nichols", "2018-01-08-hlds", "run001")
 assert os.path.exists(lds_dir)
 
 # AR-HMM RESULTS
-results_dir = os.path.join("results", "nichols", "2018-01-03-arhmm", "run001")
+results_dir = os.path.join("results", "nichols", "2018-01-09-arhmm", "run001")
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
 
@@ -314,6 +314,19 @@ def plot_likelihoods(Ks, final_lls, hlls, best_index,
     return ax1, ax2
 
 
+def compute_baseline_likelihood(xs):
+    from scipy.stats import norm
+
+    ll = 0
+    for x in xs:
+        mus = np.mean(xs, axis=0)
+        sigmas = np.std(xs, axis=0)
+
+        for d in range(D_latent):
+            ll += norm(mus[d], sigmas[d]).logpdf(xs[:,d]).sum()
+    return ll
+
+
 def fit_all_models(Ks=np.arange(4, 21, 2)):
     axs = None
     is_hierarchical = True
@@ -321,6 +334,12 @@ def fit_all_models(Ks=np.arange(4, 21, 2)):
     is_recurrent = True
     # for index, (is_hierarchical, is_robust, is_recurrent, is_driven) in \
     #         enumerate(it.product(*([(True, False)] * 4))):
+
+    # Compute the baseline probability
+    baseline = compute_baseline_likelihood(xtrains)
+    baseline /= T_trains[:N_worms].sum()
+    print("baseline ll: ", baseline)
+    assert False
 
     all_models = []
     all_lls = []
@@ -509,14 +528,40 @@ def plot_best_model_results(do_plot_expected_states=True,
     if do_plot_x_3d:
         for i in range(N_worms):
             plot_3d_continuous_states(xtrains[i], z_finals[i], colors,
-                                      figsize=(2.7, 2.7),
-                                      # title="LDS Worm {} States (ARHMM Labels)".format(i + 1),
+                                      figsize=(1.2, 1.2),
                                       title="{}".format(worm_names[i]),
                                       results_dir=fig_dir,
[…]

### 2018-01-10 experiments/fit_hlds.py

@@ -1151,9 +1151,9 @@ def heldout_neuron_identification_corr(N_heldout=10, D_latent=10, is_hierarchica
     lim = 1
     bins = np.linspace(-lim, lim, 26)
     p_other, _ = np.histogram(np.concatenate(s_others), bins)
-    p_other = p_other.astype(float) / np.concatenate(s_others).size
+    p_other = p_other.astype(np.float) / np.concatenate(s_others).size
     p_heldouts, _ = np.histogram(np.concatenate(s_heldouts), bins)
-    p_heldouts = p_heldouts.astype(float) /  np.concatenate(s_heldouts).size
+    p_heldouts = p_heldouts.astype(np.float) / np.concatenate(s_heldouts).size
 
     plt.bar(bins[:-1], p_other, width=bins[1]-bins[0],
             color=colors[0], alpha=0.75, label="incorrect")
@@ -1237,7 +1237,7 @@ def heldout_neuron_identification_corr(N_heldout=10, D_latent=10, is_hierarchica
             break
 
     plt.yticks(-np.arange(N_plot), yticklabels)
-    plt.gca().yaxis.tick_right()
+    # plt.gca().yaxis.tick_right()
     plt.xlabel("time (sec)", fontsize=6)
     plt.tick_params(labelsize=6)
     for sp in ["top", "left", "right"]:
@@ -1262,6 +1262,7 @@ if __name__ == "__main__":
     n_tests = np.array([mte.sum() for mte in mtests])
 
     D_latents = np.arange(2, 21, 2)
+    # D_latents = np.arange(2, 21, 2)
     best_models = fit_all_models(D_latents)
     best_model = best_models[0]
 

### 2018-01-10 experiments/fit_hlds_nichols.py

@@ -9,7 +9,7 @@ from functools import partial
 
 # Plotting stuff
 import matplotlib
-matplotlib.use("Agg")
+# matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 from hips.plotting.colormaps import gradient_cmap
@@ -50,7 +50,7 @@ from pylds.models import MissingDataLDS
 
 # IO
 run_num = 1
-results_dir = os.path.join("results", "nichols", "2018-01-03-hlds", "run{:03d}".format(run_num))
+results_dir = os.path.join("results", "nichols", "2018-01-08-hlds", "run{:03d}".format(run_num))
 signal = "dff_diff"
 
 
@@ -63,7 +63,7 @@ worms_and_conditions = [(i, "n2_1_prelet") for i in range(11)] + \
                        [(i, "npr1_2_let") for i in range(11)]
 worm_names = ["{} worm {}".format(condition, i) for (i, condition) in worms_and_conditions]
 N_worms = len(worms_and_conditions)
-N_clusters = 5
+N_clusters = 12
 
 
 def cached(results_name):
@@ -143,7 +143,7 @@ def _split_test_train(y, train=None, train_frac=0.8):
     return y[train], y[~train], train
 
 
-def _fit_lds(D_latent, alpha_0=1.0, beta_0=1.0,
+def _fit_lds(D_latent, alpha_0=3.0, beta_0=2.0,
              is_hierarchical=True,
              use_covariates=False,
              datas=None,
@@ -176,6 +176,7 @@ def _fit_lds(D_latent, alpha_0=1.0, beta_0=1.0,
             HierarchicalDiagonalRegression(
                 D_obs, D_latent + D_in, N_groups=N_worms,
                 alpha_0=alpha_0, beta_0=beta_0)
+        emission_distn.sigmasq_flat = 0.01 * np.ones((N_worms, D_obs))
 
         model = HierarchicalLDS(dynamics_distn, emission_distn)
     else:
@@ -310,9 +311,10 @@ def order_latent_dims(xs, C, ytrains, mtrains):
             mask = mtrains[i][0]
             yobss.append(ytrains[i][:, mask].ravel())
             yhats.append(np.outer(xs[i][:,d], C[mask,d]).ravel())
-        yobss = np.concatenate(yobss)
-        yhats = np.concatenate(yhats)
[…]

### 2018-01-10 zimmer/io.py

@@ -132,7 +132,7 @@ def _load_nichols(index, name, condition, sample_rate):
         npr1_2_let="npr1_2_Let"
     )
     dir1 = os.path.join(nichols_dir, condition_to_dir[condition.lower()])
-    dir2 = glob.glob(os.path.join(dir1, "AN*"))[index]
+    dir2 = sorted(glob.glob(os.path.join(dir1, "AN*")))[index]
     zimmer_data = loadmat(os.path.join(dir2, "wbdataset.mat"))
 
     # Get the neuron names

### 2018-01-10 zimmer/plotting.py

@@ -1382,6 +1382,8 @@ def plot_duration_histogram(trans_distn, zs, sim_durs,
                             colors=None,
                             results_dir=None):
 
+    import ipdb; ipdb.set_trace()
+
     from pyhsmm.util.general import rle
     from scipy.stats import geom
     colors = default_colors if colors is None else colors
@@ -1391,8 +1393,6 @@ def plot_duration_histogram(trans_distn, zs, sim_durs,
     durs = np.concatenate(durs)
     K = np.max(states) + 1
     perm = np.arange(K) if perm is None else perm
-
-    import ipdb; ipdb.set_trace()
 
     logpi = trans_distn.logpi[np.ix_(perm, perm)]
     P = np.exp(logpi)
@@ -1407,25 +1407,26 @@ def plot_duration_histogram(trans_distn, zs, sim_durs,
         bins = np.linspace(0, dmax+1, 15)
         width = (bins[1] - bins[0]) / 2.0
 
-        print(len(dk))
-
         plt.figure(figsize=(1.8, 1.8))
 
         # Plot the histogram of inferred transitions
         tmp, _ = np.histogram(dk, bins, density=True)
-        # plt.bar(bins[:-1] / 3.0, tmp * 3.0, width=width / 3.0,
-        #         color=colors[k], edgecolor='k', label="Empirical")
         plt.bar(bins[:-1] / 3.0, tmp * 3.0, width=width * 2.0 / 3.0,
                 color=colors[k], alpha=0.75, edgecolor='k', label="Emp")
 
 
-        # Plot the histogram of simulatedtransitions
+        # Plot the histogram of simulated transitions
+        if dmax < 30:
+            bins = np.arange(1, dmax+1, step=3)
+        else:
+            bins = np.linspace(1, dmax, 10).round()
+
         tmp, _ = np.histogram(sim_durs[k], bins, density=True)
-        # plt.bar((bins[:-1] + width) / 3.0, tmp * 3.0, width=width / 3.0,
-        #          color=colors[k], alpha=0.5, edgecolor='k', label="Simulated")
         plt.plot(bins[:-1] / 3.0, tmp * 3.0, '-k', label="Rec")
 
-        plt.plot(np.arange(1, dmax) / 3.0, g.pmf(np.arange(1, dmax)) * 3.0, ':k', label="Mkv")
+        # Plot the geometric distribution under Markov
[…]

### 2018-02-26 drafts/elife/draft.tex
updates leading up to job talk

@@
 % \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
-\title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
+% \title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
+\title{Hierarchical state space models reveal dynamics of neural activity in \textit{C. elegans}}
 \author{Scott Linderman$^{\text{1}}$,
@@
 \begin{abstract}
-  Recent advances in neural recording technologies have enabled
+  Recent advances in neural recording technologies enable
   simultaneous measurements of the majority of head ganglia neurons in
@@
   space models that decompose complex time-series into segments with
-  simple, linear dynamics. We incorporate these models into a robust,
+  simple, linear dynamics. We incorporate these models into a
   hierarchical framework for combining information across whole-brain
@@
   syllables, and we see that transition probabilities are determined
-  by a combination of current brain state and environmental cues.  In
+  by both current brain state and environmental cues.  In
   addition to quantifying neural dynamics, this probabilistic
@@
   reversals, and turns. Our methods automatically discover and
-  quantify these behaviorally-meaningful states directly from neural
+  quantify these states directly from neural
   activity, yielding powerful new tools for neuro-behavioral analysis.
@@
 \begin{itemize}
-\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms. The framework is:
-  \begin{itemize}
-  \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
-  \item \emph{robust}, ameliorating some concerns of model misspecification;
-  \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
-  \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
-  \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
-  \end{itemize}
+\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms.
+  % The framework is:
+  % \begin{itemize}
+  % \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
+  % \item \emph{robust}, ameliorating some concerns of model misspecification;
+  % \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
+  % \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
+  % \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
+  % \end{itemize}
 \item We use this framework to extract low-dimensional state
@@
   trials to learn canonical dynamics}
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=6in]{figures/v4/figure1} 
+\caption{
+  \textit{A probabilistic framework for combining
+  information across multiple worms in order to learn a shared
+  dynamics model for \celegans.}
+  \textbf{a.}~The data consists of observed neural activity collected
+  from many different worms.  In each recording we observe a different
+  subset of labeled neurons in the head ganglia.
+  \textbf{b.}~We model the observed activity as a linear function of
+  an underlying, continuous, and typically low-dimensional latent state,
+  here shown as a two-dimensional signal for each recording.
+  \textbf{c.}~The dynamics of these continuous states are governed by
+  a co-evolving discrete state, or \emph{syllable}.
+  \textbf{d.}~These syllables index into a library of linear
+  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
+  Each dynamics function is selectively used in different regions of
+  continuous latent space, as shown by the different intensities of
+  the vector field.
+  \textbf{e.}~Though location in continuous state space helps inform
+  transitions, the current syllable does as well. This is captured
+  by a Markov transition matrix, which is modulated by location in
+  continuous state space.
+  \textbf{f.}~To increase robustness to model misspecification, we
+  use a multivariate-t distribution for the noise in the continuous
+  state updates.
+  \textbf{g.}~Each worm has its own dynamics library, but they are
+  constrained to be close to the shared, canonical library.  This
+  combines information to learn shared structure while also allowing
+  for worm-to-worm variability.
+  \textbf{h.} Likewise, we allow each worm to have its own observation
+  variances to account for individual differences that may arise from,
+  for example, variable levels of fluorescent protein expression.
+  Missing neurons are denoted by gray boxes. 
+}
+\label{fig:model}
+\end{figure}
 
@@
 recording. \celegans~is genetically stereotyped and each of its 302
-neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
-units could be unambiguously labeled in each
-recording. Fig.~\ref{fig:model}a illustrates this type of data. The
-three panels represent recordings from three separate worms, and the
-rows correspond to different neuron labels.  Neural activity is
-measured for only a subset of neurons in each worm.
-
-\begin{figure}[t!]
-\centering%
-\includegraphics[width=6in]{figures/v3/figure1} 
-\caption{
-  \textit{A probabilistic framework for combining
-  information across multiple worms in order to learn a shared
-  dynamics model for \celegans.}
-  \textbf{a.}~The data consists of observed neural activity collected
-  from many different worms.  In each recording we observe a different
-  subset of labeled neurons in the head ganglia.
-  \textbf{b.}~We model the observed activity as a linear function of
-  an underlying, continuous, and typically low-dimensional latent state,
-  here shown as a two-dimensional signal for each recording.
-  \textbf{c.}~The dynamics of these continuous states are governed by
-  a co-evolving discrete state.
-  \textbf{d.}~These discrete states index into a library of linear
-  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
-  Each dynamics function is selectively used in different regions of
-  continuous latent space, as shown by the different intensities of
-  the vector field.
-  \textbf{e.}~Though location in continuous state space helps inform
-  transitions, the current discrete state does as well. This is captured
-  by the Markov transition
-  \textbf{e.}~To increase robustness to model misspecification, we
-  use a multivariate-t distribution for the noise in the continuous
-  state updates.
-  \textbf{f.}~Each worm has its own dynamics library, but they are
-  constrained to be close to the shared, canonical library.  This
-  combines information to learn shared structure while also allowing
-  for worm-to-worm variability.
-  \textbf{g.} Likewise, we allow each worm to have its own observation
-  variances to account for individual differences that may arise from,
-  for example, variable levels of fluorescent protein expression.
-  Missing neurons are denoted by gray boxes. 
-}
-\label{fig:model}
-\end{figure}
+neurons has a unique label (e.g. \textsf{AVAL}). In
+the~\citet{kato2015global} dataset, between 24 and 48 neurons could be
+unambiguously labeled in each recording, and 61 unique labels were
+found in at least one of the five worms.  Of these, 13 labels were
+found in only worm, 14 were found in two, 6 were found in three, 13 in
+four, and 15 were found in all five
+worms. The~\citet{nichols2017global} dataset is similar, but it
+contains 44 worms from two different genetic strains and two different
+developmental stages.  Across these 44 worms, 73 neuron labels were
+identified at least once. Fig.~\ref{fig:model}a illustrates this type
+of data. The three panels represent recordings from three separate
+worms, and the rows correspond to different neuron labels. Since
+calcium fluorescence is highly autocorrelated from one time bin to the
+next, we model the first-order differences in fluorescence instead.
+Most importantly, this neural activity is measured for only a subset
+of labeled neurons in each worm.
 
@@
 % evolve.
-To model this data, we build on switching linear dynamical systems
-(SLDS) \citep{chang1978state, ackerson1970state, hamilton1990analysis,
-  ghahramani1996switching, murphy1998switching}, a type of state space
-model that views data as a projection of a low-dimensional latent
-state that switches between linear dynamical regimes.  The expected
-activity is a linear function of an underlying, continuous latent
-state, illustrated as two-dimensional time series for each worm in
-Fig.~\ref{fig:model}b.  We typically assume this latent state is lower
-dimensional than the number of observed neurons. SLDS simplify the
-dynamics of the continuous latent states by introducing discrete
-latent states that indicate the current dynamical regime
-(Fig.~\ref{fig:model}c).  Each discrete regime is associated with a
-linear map (Fig.~\ref{fig:model}d), and the discrete state specifies
-which map to use in order to propagate the continuous latent state to
-the next time step.  By switching between different linear dynamical
+We model this data with an extended version of a classical switching
+linear dynamical system (SLDS) \citep{chang1978state,
+  ackerson1970state, hamilton1990analysis, ghahramani1996switching,
+  murphy1998switching}, a type of state space model that treats data
+as a projection of a low-dimensional latent state that switches
+between linear dynamical regimes.  The expected activity is a linear
+function of an underlying, continuous latent state, illustrated as
+two-dimensional time series for each worm in Fig.~\ref{fig:model}b.
+We typically assume this latent state is lower dimensional than the
+number of observed neurons. SLDS simplify the dynamics of the
+continuous latent states by introducing discrete latent states, or
+\emph{syllables}, that indicate the current dynamical regime
+(Fig.~\ref{fig:model}c).  Each syllable is associated with a linear
+map (Fig.~\ref{fig:model}d), and the discrete syllable specifies which
+map to use in order to propagate the continuous latent state to the
+next time step.  By switching between different linear dynamical
 regimes, the SLDS captures highly nonlinear dynamics.  We seek to
-infer the continuous and discrete latent states and simultaneously
-learn the linear maps, the distribution of the noise in the
-observations and the continuous states, and the dynamics that govern
-how the discrete latent states evolve.
+infer both the continuous latent states and discrete syllables and
+simultaneously learn the linear dynamics libraries, the distribution
+of the noise in the observations and the continuous states, and the
+transition structure that governs how the discrete syllables evolve
+over time.
 
 % Recurrent model
-Standard SLDS treat the discrete latent states as a simple Markov
-process: the next state depends only on the current state. However,
+Standard SLDS treat the discrete syllables as a simple Markov
+process: the next syllable depends only on the current syllable according
+to a transition matrix, as shown in~Fig.~\ref{fig:model}e. However,
 this neglects the more nuanced ways in which discrete and continuous
-states may co-evolve. For example, the next discrete state may depend
-on both the current discrete state and the current location in
+states may co-evolve. For example, the next discrete syllable may depend
+on both the current syllable and the current location in
 continuous latent space; intuitively, different dynamical regimes may
@@
 are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
-ameliorate this concern by lessening the penalty on observations that
-do not quite match the dynamics of the current regime.  We achieve
-this robustness by replacing the standard Gaussian noise model with a
-multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
-of this distribution Fig.~\ref{fig:model}e compares a multivariate-t
-distribution to a Gaussian with the same mean; though they have
-the same covariance shape, the tails of the robust model fall off
-more slowly than those of the Gaussian, allowing for outliers and a
-degree of model misspecification.
+ameliorate this concern by reducing the penalty placed on observations
+that do not quite match the dynamics of the current regime.  We
+achieve this robustness by replacing the standard Gaussian noise model
+with a multivariate-t distribution~\citep{lange1989robust}.  The heavy
+tails of this distribution Fig.~\ref{fig:model}f show a multivariate-t
+distribution. Its tails decay more slowly than those of the
+Gaussian, allowing for outliers and a degree of model
+misspecification.
 
@@
 having a separate variance for each worm and neuron, as shown in
-Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
+Fig.~\ref{fig:model}h.  More importantly, the dynamics of neural
 activity may be slightly different from worm to worm.  We account for
 this by introducing a canonical library of dynamical regimes
-(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
+(Fig.~\ref{fig:model}g) that are shared by all worms as well as a
 unique library for each individual worm (Fig.~\ref{fig:model}d);
@@
 discrete and continuous latent states with approximate Bayesian
-inference as described in the Supplementary Material. Briefly, we
-separate the problem into two steps: learning the continuous latent
-states and the mapping from continuous latent states to observations,
-then learning the dynamics libraries that govern the continuous latent
-states and the corresponding discrete state sequences.  Both steps are
-performed with a block Gibbs sampler, leveraging the structure of the
-probabilistic model to perform efficient conditional updates of the
-various parameters.
+inference as described in Section~\ref{sec:slds}. Briefly, we separate
+the problem into two steps: learning the continuous latent states and
+the mapping from continuous latent states to observations, then
+learning the dynamics libraries that govern the continuous latent
+states and the corresponding discrete syllable sequences.  This
+separation is justified when the observed neural activity, combined
+with a weak prior distribution to prefer smooth latent states,
+provides sufficient information to constrain the continous latent
+state sequence.  Having obtained these latent states, inferring the
+discrete syllables is equivalent to performing inference in an
+autoregressive hidden Markov model. Both stages are performed with a
+block Gibbs sampler, leveraging the conditionally conjugate structure
+of the probabilistic model to perform efficient updates of the states
+and parameters.
+
+Section~\ref{sec:slds} provides the technical details of our
+probabilistic model and our algorithm for inferring the latent states
+and syllables and learning the model parameters.
+Fig.~\ref{fig:graphical_model} summarizes the probabilistic framework
+in the form of a probabilistic graphical model.
 
@@
 
-The parameters of the learned model offer novel insights into the
-structure of neural activity in the head ganglia that can only be
-obtained by pooling information across many partial recordings.  We
-fit our model directly to the first-order differences in calcium
-fluorescence; our only preprocessing step is to correct for
-photo-bleaching. The green lines in Fig.~\ref{fig:smoothing} show a
-three minute window of bleaching-corrected, differenced calcium traces
-for two of the five worms in the~\citet{kato2015global} dataset. Of
-these five worms, Worm 1 has the fewest labeled neurons (24 neurons)
-and worm 5 has the most (48 neurons). Across the five worms, 61 unique
+The parameters of the learned model offer insights into the structure
+of neural activity in the head ganglia that can only be obtained by
+pooling information across many partial recordings.  We fit our model
+to the first-order differences in calcium fluorescence with minimal
+preprocessing; we simply correct for photo-bleaching, z-score the
+resulting traces, and compute the differences between consecutive
+frames. The green lines in Fig.~\ref{fig:smoothing} show a three
+minute window of bleaching-corrected, differenced calcium traces for
+two of the five worms in the~\citet{kato2015global} dataset. Of these
+five worms, worm 1 has the fewest labeled neurons (24 neurons) and
+worm 5 has the most (48 neurons). Across the five worms, 61 unique
 neuron labels were identified in at least one worm.
@@
 By aggregating information across worms, the model learns correlation
-structure that enables it to smooth noisy traces and predict the
-activity of unobserved neurons.  The smoothed and predicted activity
-is shown as black lines in Fig.~\ref{fig:smoothing}. For example,
+structure that enables it to smooth the noisy traces and predict the
+activity of unlabeled neurons.  The smoothed and predicted activity
+are shown as black lines in Fig.~\ref{fig:smoothing}. For example,
 even though \textsf{SMDVR} is not labeled in worm 1, the model
@@
 both~\textsf{SMDVR} and~\textsf{SMDVL} are labeled.  Since these
-two neurons are highly correlated, the model can make strong predictions
+two neurons are highly correlated, the model makes strong predictions
 about the activity of~\textsf{SMDVR} in worm 1 on the basis of the
-activity of~\textsf{SMDVL}.
+activity of~\textsf{SMDVL} and other the other observed neurons.
 
@@
 We estimate this dimensionality with cross validation, reserving the
-last 20\% of the data (3.6 min) for testing and measuring the marginal
-log likelihood of the held-out data as we increase the latent
+last 20\% of the data ($\approx$3.6 min) for testing and measuring the
+marginal log likelihood of the held-out data as we increase the latent
 dimensionality. We find that the test likelihood is maximized with a
-10-dimensional latent state, and that variances differ substantially
-between worms.
+10-dimensional latent state, and that performance improves
+substantially after allowing neurons different observation variances
+in each worm. This may account for differences in fluorescence level,
+for example.
 
@@
   hence highly correlated, neurons. Neurons are ordered as in
-  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
-  tuning vectors supports this clustering and shows differences
-  between subpopulations (clusters have been sorted to make this as
-  close to diagonal as possible).  \textbf{c.}~The tuning similarity
-  reflects correlations between neurons, but since it is learned as
-  part of a hierarchical model, it predicts similarities for all pairs
-  of neurons, even pairs that are never simultaneously recorded (gray,
-  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
-  neurons from each cluster, as specified by WormAtlas.  Neurons in
+  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity
+  between tuning vectors supports this clustering and shows
+  differences between subpopulations (clusters have been sorted to
+  make this as close to diagonal as possible).  \textbf{c.}~The
+  empirical correlation matrix between neurons mimics the similarity
+  between tuning vectors, but has a substantial number of missing
+  entries (gray, off-diagonal boxes) for pairs that were never
+  simultaneously recorded .  \textbf{d.}~We show the spatial locations
+  of neurons from each cluster, as specified by WormAtlas.  Neurons in
   the same cluster are not necessarily close to one another in space,
@@
   approximate locations of the anterior (AG), dorsal (DG), ventral
-  (VG) and retrovesicular ganglia.}
+  (VG) and retrovesicular ganglia.  \textbf{e.}~Clusters 2 and 7
+  contain only sensory (S) neurons, but the other clusters contain
+  neurons with a mix of interneurons (I), motor (M) and polymodal (P) neurons, and
+  neurons of unknown (U) type. \textbf{f.}~However, these clusters of
+  interneurons and motor neurons are functionally grouped, according
+  to expert labels. }
 \label{fig:clustering}
@@
 responds to each dimension of the latent state.  This matrix reveals
-groups of similarly tuned neurons. To capture this, we used k-means to
-cluster the neurons based on their tuning vectors and found that 12
-clusters adequately captures the diversity of the population.  Using
-fewer clusters leads to qualitatively similar groups but omits some of
-the finer within-group differences; allowing more clusters eventually
-leads to groups of lateral pairs (like \textsf{ASKL} and \textsf{ASKR}
-in a.).  The neurons are ordered as in Fig.~\ref{fig:smoothing}, and
-clusters are separated by black lines.
+groups of similarly tuned neurons. To capture these groups, we used
+k-means to cluster the neurons based on their tuning vectors and found
+that 12 clusters adequately captures the diversity of the population.
+Using fewer clusters leads to qualitatively similar groups but omits
+some of the finer within-group differences; allowing more clusters
+eventually leads to groups of lateral pairs (like \textsf{ASKL} and
+\textsf{ASKR} in Fig.~\ref{fig:clustering}a).  The neurons are ordered
+as in Fig.~\ref{fig:smoothing}, and clusters are separated by black
+lines.
 
@@
 Fig.~\ref{fig:clustering}d shows the locations of all 61 labeled
-neurons, using the positions from WormAtlas\todo{cite} along the
+neurons, using the positions from WormAtlas \citep{wormatlas} along the
 head-tail axis of the worm's body.  The size of the dot indicates the
@@
 
-\todo[inline]{Break down clustes by type (sensory/inter/motor
-  neuron).}
-
-\todo[inline]{Ask Manuel/Annika to label functional role of neurons.}
+Next, we considered the types of cells contained in each cluster.
+Fig.~\ref{fig:clustering}e shows the composition of each cluster in
+terms of sensory (S), inter (I), motor (M), polymodal (P) and unknown
+(U) neuron types.  Clusters 2 and 7 consist of only sensory neurons,
+cluster 3 contains mostly interneurons, and clusters 8 and 11 are
+primarily motor neurons, but the other clusters are composed of a mix
+of inter and motor neurons. Thus, cell type alone appears insufficient to
+explain these clusters as well.
 
@@
 \centering
-\includegraphics[width=6in]{figures/v4/figure_4} 
+\includegraphics[width=6in]{figures/v4/figure4} 
 \caption{ \textit{The predicted neural activity is a useful aid for
@@
   labeled activity, we also have traces from many neurons that could
-  not be unambiguously labeled. To test whether the model's
+  not be labeled with certainty. To test whether the model's
   predictions could aid in labeling neurons, we conducted a synthetic
@@
 predictions to the true activity of VB01 (red trace) and to the
-activity of other neurons that were identified but not labeled (blue
-traces), as shown in Fig.~\ref{fig:id}b.  We computed the correlation
-coefficient between these traces and the predicted activity of VB01,
-as shown in Fig.~\ref{fig:id}c.
-We find that in this case, the model's predictions and the true activity
-are highly correlated, even though the true activity was not used
-when training the model.  We repeated this experiment for 50 heldout
-neurons and found that this pattern holds: the predicted activity is
-typically much more correlated with the activity of the correct
-heldout trace (red histogram) than with the activity of other unlabeled
-neurons (blue histogram).  This suggests that our hierarchical model
-may serve an additional role in aiding neural identification in
-ambiguous, noisy data.
-
-\todo[inline]{Also discuss simultaneous matching experiments? Possible
-  improvements from including more labeled neurons.  Work with Gonzalo
-  on learning permutations.  Importance of capturing uncertainty in matching.}
+activity of other neurons that were recorded but could not be labeled
+(blue traces), as shown in Fig.~\ref{fig:id}b.  We computed the
+correlation coefficient between these traces and the predicted
+activity of VB01, as shown in Fig.~\ref{fig:id}c.  In this case, the
+model's predictions and the true activity are highly correlated, even
+though the true activity was not used when training the model.  We
+repeated this experiment for 50 heldout neurons and found that this
+pattern holds: the predicted activity is typically much more
+correlated with the activity of the correct heldout trace (red
+histogram) than with the activity of other unlabeled neurons (blue
+histogram).  This suggests that our hierarchical model may serve an
+additional role in aiding neural identification in ambiguous, noisy
+data.
+
+Table~\ref{tab:neuron_id} shows the results of a more challenging, yet
+more realistic experiment.  Here we withheld the identities of ten
+neurons at a time in each worm.  We combined the activity of these
+neurons with those of the other measured but unlabeled neurons to form
+a candidate pool that ranged from 76 neurons to 111 neurons in size.
+Then we solved a minimum weight bipartite matching problem to
+simultaneously match the candidates to the set of all possible labels
+(which consists of 23 to 47 possible labels, not just the ten that
+were artificially withheld).  We find that in all five worms, 6 to 8
+of the 10 withheld labels were accurately matched.  While this is
+impressive performance for this challenging task, we expect that
+further improvements could be obtained with recent work on variational
+Bayesian inference for permutation and matching inference
+problems~\citep{linderman2018reparameterizing}.
 
@@
   continuous latent states for each of the five worms, color coded by
-  the corresponding discrete latent state.  \textbf{b.}~The same
+  the corresponding discrete syllable.  \textbf{b.}~The same
   continuous latent states but now color coded according to the manual
@@
     REV2}: reversals; \textsf{NOSTATE}: undetermined). The
-  correspondence between inferred and manually labeled states implies
+  correspondence between inferred and manually labeled states shows
   that hierarchical state space models can infer behaviorally
@@
 Having learned a mapping from latent state space to observed activity,
-we can now reason about the dynamics of these continuous latent
+we can now study the dynamics of these continuous latent
 states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
@@
 three principal components.  In contrast, our probabilistic framework
-is automatic, operates on ten-dimensional latent states, and requires
-no additional smoothing.
-
-We use the likelihood of held-out test data to determine
-the number of discrete regimes and assess the value of the robust and
-recurrent model extensions, as shown in Supplemental Fig.~\ref{fig:modelselection}.
-While the non-robust models with Gaussian dynamics noise achieve
-higher training likelihood, the robust models generalize better to
-test data.  Moreover, the recurrent models provide further improvement
-over the standard Markovian models.  We find that test likelihood is
-maximized with eight latent states, which matches the number manually
-chosen by~\citet{kato2015global}.
+is automatic, operates on ten-dimensional latent states, combines
+recordings across worms, and requires no additional preprocessing of
+the data.
+
+We use the likelihood of held-out test data to determine the number of
+discrete syllables and assess the value of the robust and recurrent
+model extensions, as shown in Fig.~\ref{fig:model_selection}.  Across
+the board, we see a significant improvement when moving from a single
+syllable, which is equivalent to an autoregressive model, to multiple
+syllables.  Likewise, we see substantial improvements when adding the
+robust and hierarchical extensions.  Moreover, in the hierarchical and
+robust models, we see a further improvement by introducing the
+recurrent extensions.  The test likelihood is maximized with
+eight latent states, which matches the number manually chosen
+by~\citet{kato2015global}.
 
@@
 Fig.~\ref{fig:syllables}b shows ten-dimensional latent states and the
-manual and inferred segmentations for a 5.5 minute section of data.
+manual and inferred segmentations for a three minute section of data.
 The manual and inferred states often change at similar times. We also
@@
 
-\subsection*{Quantifying the dynamics of neural activity in each discrete state}
+\subsection*{Syllable dynamics differentially activate clusters of neurons}
+
+\begin{figure}[t!]
+  \centering
+  \vspace{-.5in}
+  \includegraphics[width=6in]{figures/v4/figure6B}
+  \caption{\textit{Discrete syllables are characterized by simple
+      linear dynamics that give rise to interpretable patterns of
+      neural activity.}  \textbf{a.}~Each syllable is associated with
+    a canonical set of linear dynamics, which can be viewed as
+    mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show the
+    mappings for the first three dimensions here, with colored arrows
+    placed at locations where each syllable was used.  The black
+    arrows show a simulated trajectory with the syllable's
+    dynamics. The syllables are ordered by their overall usage, with
+    syllable 1 being used most and syllable 8 least. \textbf{b.}~We
+    mapped the simulated latent state trajectories (black arrows in
+    a.) through the emission matrix to compute a time course of
+    resulting activations for each cluster of neurons.  \textbf{c.}
+    Using 1000 simulations, we computed the average activation each
+    syllable gives rise to on each cluster, as well as the average
+    change in activation.  }
+  \label{fig:dynamics}
+\end{figure}
+
 The hierarchical SLDS not only automates the laborious manual
-segmentation process, it justifies its segmentation with a
-quantitative definition of each inferred state. Each state is
+segmentation process, it justifies the segmentation with a
+quantitative definition of each inferred syllable. Each syllable is
 associated with a set of canonical linear dynamics, as shown in
-Fig.~\ref{fig:dynamics}a. The arrows in these vector fields show where
-the continuous state is mapped to under the corresponding discrete
-state (for the top three latent dimensions).  Arrows are only drawn
-at those locations where the discrete state was actually employed.
-
-States 1 (blue), 3 (yellow), and 5 (purple) are all characterized
+Fig.~\ref{fig:dynamics}a. The colored arrows in these vector fields
+show how the continuous state in one time bin is mapped to a new
+location in the next time bin according to the corresponding discrete
+syllable (for the top three latent dimensions).  Arrows are only drawn
+at those locations where the syllable was actually employed.
+
+Syllables 1 (blue), 3 (yellow), and 5 (purple) are all characterized
 by convergence toward the origin, which corresponds, essentially, to
-no change in neural activity. These three states differ in terms of how
+no change in neural activity. These three syllables differ in terms of how
 quickly they converge to the origin and the amount of rotation as
-they do so.  By contrast, states 2 (red), 4 (green), 6 (orange), 7(brown)
+they do so.  By contrast, syllables 2 (red), 4 (green), 6 (orange), 7(brown)
 and 8 (pink) form parts of rotational loops in latent state space.
@@
 
-We use these dynamics to simulate how brain activity evolves
-under the different states. Fig.~\ref{fig:dynamics}b shows how four
-simulated trajectories of each state from various starting locations map 
-into activations in the space of neural activity. For each discrete state,
-we chose the starting locations randomly from the set of locations where
-that state was entered in the real data.  We then simulated forward in
-time according to that state's dynamics until the simulation predicted
-a state change.  We then mapped these four simulations for each state
-through the emission matrix and measured the average activation induced
-on each of the 12 neural clusters.
+We use these dynamics to simulate how brain activity evolves under the
+different syllables. The black arrows in Fig.~\ref{fig:dynamics}a show
+one simulated trajectory for each syllable. Fig.~\ref{fig:dynamics}b
+shows how these simulated trajectories map into activations in the
+space of neural activity. For each discrete
+syllable, we chose the starting locations randomly from the set of
+locations where that syllable was entered in the real data.  We then
+simulated forward in time according to that syllable's dynamics until
+the simulation predicted a transition to another syllable.  We then
+mapped these simulations for each state through the emission
+matrix and measured the average activation induced on each of the 12
+neural clusters.
 
 These activations lend further evidence in support of the behavioral
-labels attributed to each state. For eaxample, state 4 (green) coincides
-with the manually labeled ``ventral turn'' state.  Its dynamics produce
-a pronounced rise in the activity of cluster 6, which (referring back
-to Fig.~\ref{fig:clustering}), consists of the \textsf{RIVL/R} and
-\textsf{SMDVL/R} neurons.  The \textsf{RIV} pair are polymodal inter-
-and motor-neurons that specify the ventral bias of turns, and the
-\textsf{SMDV} are motor neurons that initiate ventral turns.  State 4
-dynamics also tend to suppress the activation of cluster 8, which
-consists mostly of the \textsf{SMB*} neurons.  As these neurons are
-implicated in dorsal turns, this again makes sense. Finally, we see
-that states 6 (orange) and 5 (purple) often produce a rise and fall, respectively,
-of activation on cluster 8, which together produce a dorsal turn.
-These two states do indeed coincide with the manually labeled dorsal
-turn state, again confirming the behavioral interpretation of these
-dynamical states.
-
-\begin{figure}[h!]
-  \centering
-  \vspace{-.5in}
-  \includegraphics[width=6in]{figures/v4/figure6}
-  \caption{\textit{Discrete states are characterized by
-      simple linear dynamics that give rise to interpretable patterns
-      of neural activity.}  \textbf{a.}~Each discrete state is
-    associated with a canonical set of linear dynamics, which can be
-    viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show
-    the mappings for the first three dimensions here, with vectors
-    placed at locations where each state was used.  \textbf{b.}~We
-    simulated four latent state trajectories with these linear
-    dynamics from different starting locations, and then we mapped
-    these latent state trajectories through the emission matrix to
-    compute a time course of resulting activations for each cluster of
-    neurons.}
-  \label{fig:dynamics}
-\end{figure}
-
-\clearpage
-
-\subsection*{Recurrent models learn spatially localized states and non-Markovian dynamics}
+labels attributed to each state. For eaxample, state 4 (green)
+coincides with the manually labeled ``ventral turn'' state.  Its
+dynamics produce a pronounced rise and fall in the activity of cluster 6, which
+(referring back to Fig.~\ref{fig:clustering}), consists of the
+\textsf{RIVL/R} and \textsf{SMDVL/R} neurons.  The \textsf{RIV} pair
+are polymodal inter- and motor-neurons that specify the ventral bias
+of turns, and the \textsf{SMDV} are motor neurons that initiate
+ventral turns.  We also see that states 6 (orange) and 5 (purple) often
+produce a dip followed by a rise, respectively, of activation on cluster 8,
+which together produce a dorsal turn.
+\todo{Need confirmation of these results.  Hard for me to tell difference between
+dorsal turn activation (orange to purple) and ventral turn (green to brown).}
+
+To understand the overall effect of these different syllables, we
+simulated 1000 continuous latent state trajectories for each, with
+the same procedure as above.  We again mapped these into neural space
+and computed the activation of each cluster.  We averaged over all
+the simulations to compute the average cluster activation shown in
+Fig.~\ref{fig:dynamics}c (left).  We did the same with the difference
+between starting and ending activation to compute the average change
+in cluster activation shown in Fig.~\ref{fig:dynamics}c (right).
+We notice that different subsets of clusters tend
+to be activated by different syllables.  In other words, syllables
+engage unique subsets of neurons.
+
+\subsection*{Recurrent models learn spatially localized syllables and non-Markovian transitions}
 
@@
 \includegraphics[width=6in]{figures/v4/figure7} 
-\caption{ \textit{Discrete state transitions show a chain-like structure
-    with non-Markovian durations.}  \textbf{a.}~Counts of inferred transitions
-  from one state to another. The states are permuted to emphasize the
-  feed-forward structure.  \textbf{b.}~An alternative view of the
-  transitions; each state is a node in a graph and edges denote
-  high-probability transitions.  This shows transitions between
-  forward and reverse crawling linked by dorsal and ventral turns.
-  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
-  imply a geometric distribution of state durations (dotted black
-  lines), but Fig.~\ref{fig:syllables} suggests that state
-  transitions occur near the origin.  With the exception of state 7,
-  which is unstable, the recurrent SLDS incorporates
-  this dependency and generates state durations (solid black lines)
-  that are a better fit to those inferred from the data (colored
-  bars).  }
+\caption{ \textit{Discrete syllable transitions show a chain-like
+    structure with non-Markovian durations.}  \textbf{a.}~Counts of
+  inferred transitions from one syllable to another. The syllables are
+  permuted to emphasize the feed-forward structure.  \textbf{b.}~An
+  alternative view of the transitions; each syllable is a node in a graph
+  and edges denote high-probability transitions.  This shows
+  transitions between forward and reverse crawling linked by dorsal
+  and ventral turns.  \textbf{c.}~The standard SLDS assumes Markovian
+  transitions, which imply a geometric distribution of syllable durations
+  (dotted black lines), but Fig.~\ref{fig:syllables} suggests that
+  transitions occur near the origin.  With the exception of
+  syllable 7, which is unstable, the recurrent SLDS incorporates this
+  dependency and generates syllable durations (solid black lines) that
+  are a better fit to those inferred from the data (colored bars).  }
 \label{fig:recurrent}
@@
 continuous state space (via the \emph{recurrent} dependency) as well as
-the preceding discrete state (the \emph{Markovian} dependency).
+the preceding discrete syllable (the \emph{Markovian} dependency).
 Fig.~\ref{fig:recurrent}a shows the contribution of the Markovian
 term.  Each cell in this matrix denotes the number of transitions from
-one state to another. We are only counting transitions in which the
-state changes, hence the diagonal is zero.  We have permuted the
-states to emphasize the largely feed-forward structure.
+one syllable to another. We are only counting transitions in which the
+syllable changes, hence the diagonal is zero.  We have permuted the
+syllables to emphasize the largely feed-forward structure.
 
@@
 
-To accurately model these dynamics, the changes must occur at precise
-locations in continuous latent space.  Namely,
+Though these syllables were inferred from neural activity alone, the
+transition structure---forward and reverse crawling linked by dorsal
+and ventral turns---mimics the natural behavior of freely crawling
+worms.  This point was made by~\citet{kato2015global} based on their
+manual segmentation of neural activity; here we see that this finding
+holds given our automatic segmentation as well.  Moreover, we obtain
+further insights by studying the precise transition structure that
+links these syllables together according to the current syllable
+as well as the current continuous latent brain state.
+
+To accurately model the transitions between syllables, the changes
+must occur at specific boundaries in continuous latent space.  Namely,
 Fig.~\ref{fig:syllables}a shows that the loops tend to start and end
@@
 distribution with monotonically decreasing probability. By contrast,
-the recurrent model learns to transition near the origin, which
-gives rise to characteristic duration distributions that closely
-match the empirical distributions.
+the recurrent model learns to transition near the origin, which gives
+rise to characteristic duration distributions that closely match the
+empirical distributions.
 
@@
 \caption{\textit{Incorporating exogenous factors like oxygen level
-    into the dynamics model yield differential responses in worms from
+    into the dynamics model reveals how external covariates
+    differentially modulate transition probabilities in worms of
     different genetic strains and developmental stages.}
@@
   different genetic strains (N2, npr1) and developmental stages
-  (pre-lethargus, lethargus) exhibit different average state usages as
-  a function of the oxygen level.  Blue and red states are much more
-  likely in the lethargus stage for both genetic strains, but they are
-  suppressed in npr1 worms at high oxygen concentrations. Moreover, we
-  see a pronounced increase in the pink state usage immediately after
-  the oxygen level increases to 21\%.  \textbf{c.}~As above, the
-  inferred states closely match the manually labeled states
-  of~\citet{nichols2017global}, with blue, red, and to a lesser extent
-  green mapping onto the forward (\textsf{Fwd}) and quiescent
-  (\textsf{Q}) states.  \textbf{d.}~We extended the model to
+  (pre-lethargus, lethargus) exhibit different average syllable usages
+  as a function of the oxygen level.  Blue and red syllables are much
+  more likely in the lethargus stage for both genetic strains, but
+  they are suppressed in npr1 worms at high (21\%) oxygen
+  concentrations. Moreover, we see a pronounced increase in the pink
+  syllable usage immediately after the oxygen level increases to 21\%.
+  \textbf{c.}~As above, the inferred states closely match the manually
+  labeled states of~\citet{nichols2017global}, with blue, red, and to
+  a lesser extent green mapping onto the forward (\textsf{Fwd}) and
+  quiescent (\textsf{Q}) states.  \textbf{d.}~We extended the model to
   incorporate oxygen level (O$_2$) and change in oxygen level
@@
   the change in oxygen: an increase in $O_2$ strongly elicits the pink
-  state.  \textbf{e.} The state-triggered average latent state
-  trajectory following entry into the pink state shows a dip in
+  syllable.  \textbf{e.} The average continuous latent state
+  trajectory following entry into the pink syllable shows a dip in
   activity along the last dimension 7 seconds after entry (black
-  arrows).  \textbf{f.} This dip gives rise to strong opposing
-  responses of sensory neurons \textsf{URXL/R} and \textsf{BAGL}
-  (among others), which are known to respond to increases and
-  decreases, respectively, in oxygen concentration.  }
+  arrows).  \textbf{f.} This dip gives rise to strong excitatory
+  responses of sensory neurons \textsf{URXL/R} and \textsf{AUAL/R}
+  (among others), which are known to respond to increases in oxygen
+  concentration. Likewise, we see strong inhibition of other neruons,
+  like the chemosensory neurons \textsf{ASGL/R} and \textsf{BAGL/R}.}
 \label{fig:o2}
@@
 of~\citet{kato2015global}.  We now turn to the dataset studied
-in~\citet{nichols2017global}, which contains many more worms with
-roughly the same number of neurons per recording but is complicated by
-additional factors.  First, there are two strains of worms (N2 and
-npr-1) at two different stages of development (pre-lethargus and
-lethargus); second, the oxygen level is manipulated over the course of
-the recordings to elicit different responses.  Here we show how our
-framework can be extended to study the effects of these manipulations.
-
-Fig.~\ref{fig:o2}a shows which neurons were labeled in each of the
-recordings along with the inferred observation variance. There are 41
-worms, roughly ten from each strain and developmental stage
-combination (N2 + pre-lethargus, N2 + lethargus, npr1 + pre-lethargus,
-and npr1 + lethargus). Some neurons are consistently identified in
-almost all worms; others are identified in only a handful.  As before,
-our hierarchical framework combines all this information.
+in~\citet{nichols2017global}. It contains many more worms, but it has
+roughly the same number of neurons per recording.  However, the
+experimental design is more complicated.  First, there are two strains
+of worms (N2 and npr-1) at two different stages of development
+(pre-lethargus and lethargus); second, the oxygen concentration is
+manipulated over the course of the recordings to elicit different
+responses.  Here we show how our framework can be extended to study
+the effects of these manipulations.
+
+Fig.~\ref{fig:o2}a shows the neurons that were labeled in each of the
+recordings. There are 41 worms, roughly ten from each strain and
+developmental stage combination (N2 + pre-lethargus, N2 + lethargus,
+npr1 + pre-lethargus, and npr1 + lethargus). Some neurons are
+consistently identified in almost all worms; others are identified in
+only a handful.  As before, our hierarchical framework combines all
+this information to learn a canonical dynamical system.
 
 Upon fitting our model, we immediately see a difference in the average
-usage of the eight states  as a function of oxygen level(Fig.~\ref{fig:o2}b). The worms in the
-lethargus stage predominantly use the blue and red states.  Again,
-we found a close correspondence between our inferred states and the
-manually labeled states of~\citet{nichols2017global}
-(Fig.~\ref{fig:o2}c). Here, blue and red both map strongly onto the
-quiescent (\textsf{Q}) state. This is consistent with the
-characterization of the lethargus stage---at this stage, worms exhibit
-much less crawling behavior. The pre-lethargus worms execute a broader
-variety of behaviors, including reversals (\textsf{Rev.}) and dorsal
-and ventral turns  (\textsf{DT} and \textsf{VT}, respectively).
+usage of the eight syllables as a function of oxygen
+level(Fig.~\ref{fig:o2}b). The worms in the lethargus stage
+predominantly use the blue and red syllables.  Again, we found a close
+correspondence between our inferred syllables and the manually labeled
+states of~\citet{nichols2017global} (Fig.~\ref{fig:o2}c). Here, blue
+and red both map strongly onto the quiescent (\textsf{Q}) state. This
+is consistent with the characterization of the lethargus stage---at
+this stage, worms exhibit much less crawling behavior. The
+pre-lethargus worms execute a broader variety of behaviors, including
+reversals (\textsf{Rev.}) and dorsal and ventral turns (\textsf{DT}
+and \textsf{VT}, respectively).
 
 A key finding of~\citet{nichols2017global} was that increasing the
-oxygen level can elicit activity in lethargus npr1 worms but not in
-lethargus N2 worms. We can quantify this effect with our models by
+oxygen level elicits activity in lethargus npr1 worms but not in
+lethargus N2 worms. We quantify this effect with our models by
 including oxygen level (O$_2$) and change in oxygen level ($\Delta$O$_2$)
@@
 %   influences continuous states and observations?}
-Fig.~\ref{fig:o2}d (top) shows the inferred effect of oxygen level for each
-of the four strain/developmental stage pairs.  As predicted,
-higher oxygen levels increase the probability of reversals and
-turning related states (orange, purple) and decrease the probability
-of forward and quiescent (blue, red, green) states in npr1 worms, but
-has limited effect on N2 worms. We do see an increase in turning related
-states (orange) in the N2 lethargus worms, but inspection of panel b
-indicates this state is still used very sparingly.  Somewhat
-surprisingly, the magnitude of the effect is comparable for both
-pre-lethargus and lethargus npr1 worms, even though the baseline level
-of reversals and turns is much higher in the pre-lethargus worms.
+Fig.~\ref{fig:o2}d (top) shows the inferred effect of oxygen level for
+each of the four strain/developmental stage pairs.  As predicted,
+higher oxygen levels increase the probability of reversals and turning
+related states (orange, purple) and decrease the probability of
+forward and quiescent (blue, red, green) states in npr1 worms, but has
+limited effect on N2 worms.  Somewhat surprisingly, the magnitude of
+the effect is comparable for both pre-lethargus and lethargus npr1
+worms, even though the baseline level of reversals and turns is much
+higher in the pre-lethargus worms.
 
@@
 from 10\% to 21\% oxygen concentration sharply increases the probability
-of transitioning into the pink state. Moreover, this state is seldom
+of transitioning into the pink syllable. Moreover, this syllable is seldom
 used except for when the oxygen level increases.  This suggests that
 sharp increases in oxygen level induce a transient and characteristic
-neural response that is not seen in steady-state oxygen levels.
+neural response in worms of both genetic strains and developmental stages,
+however the sustained upregulation of active states in response to elevated
+oxygen concentration is unique to npr1 worms.
 
 Fig.~\ref{fig:o2}e shows the average latent state trajectory triggered
-upon entry into the pink state (for this dataset, the optimal performance
-was achieved by a 6-dimensional state space model).  We see a dip and
-rebound along the last dimension approximately 7 seconds after entry
-(denoted by the black arrows).  We identified the neurons that showed
-the largest response at this point in time and plotted the top ten
-in Fig.~\ref{fig:o2}f.  Unsurprisingly, these neurons are largely sensory
-and interneurons.  In particular, \textsf{URXL/R} and \textsf{BAGL/R} are
-known to jointly encode changes in oxygen level with opposing signs: \textsf{URXL/R}
-respond to increases in oxygen, \textsf{BAGL/R} respond to dcreases.  Our
-model recapitulates these results and suggests that many other sensory
-neurons respond similarly to this stimulus.
+upon entry into the pink syllable (for this dataset, the optimal
+performance was achieved by a 6-dimensional continuous late state
+space).  We see a dip and rebound along the last dimension
+approximately 7 seconds after entry (denoted by the black arrows).  We
+identified the neurons that showed the largest excitatory and
+inhibitory responses at this point in time and plotted the top ten in
+Fig.~\ref{fig:o2}f and g, respectively.  Unsurprisingly, these neurons
+are largely sensory and interneurons.  In particular, \textsf{URXL/R}
+and \textsf{BAGL/R} are known to jointly encode changes in oxygen
+level with opposing signs: \textsf{URXL/R} respond to increases in
+oxygen, \textsf{BAGL/R} respond to dcreases.  Our model recapitulates
+these results and suggests that many other sensory neurons respond
+similarly to this stimulus.
 
@@
 \appendix
-
-\section{Data}
+\counterwithin{figure}{section}
+\counterwithin{table}{section}
+
+\section{Data representation and preprocessing}
 \label{sec:data}
@@
 
-\section{Hierarchical Generative Model}
-
-We model the neural activity with a switching linear
-dynamical system (SLDS) with three new extensions: (i)~a \emph{hierarchical} model
-to share parameters across worms while also allowing for worm-to-worm
-variability; (ii)~a \emph{robust} model for dynamics noise to help address
-model misspecification; and (iii)~a \emph{recurrent} model to capture
-how continuous latent states influence discrete state transition probabilities.
-We will introduce the SLDS first and then present each of these extensions
-in turn.
-
-\subsection{Switching linear dynamical systems}
+\section{Hierarchical state space model for neural data}
 \label{sec:slds}
+
+\begin{figure}[t!]
+\centering%
+\includegraphics[width=5in]{figures/v4/figure1_supp} 
+\caption{\textit{Graphical depiction of a hierarchical, robust,
+    recurrent, and input-driven switching linear dynamical system
+    (SLDS).}  The blue arrow denotes the \textbf{hierarchical}
+  component: each worm~$w$ has its own set of
+  parameters~$\theta^{(w)}$, which contain the dynamics libraries,
+  emission matrix, etc. These are slight perturbations of the same
+  global mean parameters~$\theta$.  In some cases, the worm-specific
+  parameters are required to equal their global counterparts (as with
+  the emission matrix).  Red arrows highlight the \textbf{robust}
+  noise model that links~$x_t$ to~$x_{t+1}$.  Green arrows indicate
+  the \textbf{recurrent} extension: the next discrete syllable depends
+  on both the preceding syllable as well as the preceding continuous
+  latent state.  Purple arrows carry the \textbf{input-dependence}:
+  the next discrete syllable can also be governed by an exogenous
+  input, like the oxygen level or the change in oxygen level. The
+  remaining black arrows denote standard dependencies of the SLDS:
+  discrete syllables and continuous latent states co-evolve, with the
+  syllables specifying the linear dynamics to use when updating the
+  continuous latent states.  The observations are a function of these
+  underlying latent states.  The only non-standard addition is an
+  autoregressive dependency between~$y_t$ and $y_{t+1}$; here, the
+  SLDS models only the differences in observed fluoresence value
+  between two consecutive frames.  }
+\label{fig:graphical_model}
+\end{figure}
+
+We model the neural activity with a switching linear dynamical system
+(SLDS) with three new extensions: (i)~a \emph{hierarchical} model to
+share parameters across worms while also allowing for worm-to-worm
+variability; (ii)~a \emph{robust} model for dynamics noise to help
+address model misspecification; and (iii)~a \emph{recurrent} model to
+capture how continuous latent states influence discrete state
+transition probabilities.  We will introduce the SLDS first and then
+present each of these extensions in turn.
+
+\subsection{The standard SLDS}
 Switching linear dynamical system models (SLDS) break down complex, nonlinear
@@
 
-\section{Supplementary Figures}
-
-\begin{itemize}
-\item Graphical model (with arrows $y_t \to y_{t+1}$)
-\item Neural activity for worms 2, 3, 4?
-\item Model selection based on held-out log likelihood
-\item Table of matching results?
-\item Eigenspectra of inferred states
-\item Model fitting 
-\end{itemize}
-
-\begin{figure}[t!]
-\centering%
-\includegraphics[width=5in]{figures/v3/figure1_supp} 
-\caption{\textit{Graphical depiction of a hierarchical, robust, recurrent,
-    and input-driven switching linear dynamical system (SLDS).}
-  The blue arrow denotes the \textbf{hierarchical} component: each worm~$w$ has its
-  own set of parameters~$\theta^{(w)}$, which contain the dynamics libraries, emission
-  matrix, etc. These are slight perturbations of the same global mean parameters~$\theta$.
-  In some cases, the worm-specific parameters are required to equal their global
-  counterparts (as with the emission matrix). 
-  Red arrows highlight the \textbf{robust} noise model that links~$x_t$ to~$x_{t+1}$.
-  Green arrows indicate the \textbf{recurrent} extension: the next discrete syllable depends
-  on both the preceding syllable as well as the preceding continuous latent state.
-  Purple arrows carry the \textbf{input-dependence}: the next discrete syllable
-  can also be governed by an exogenous input, like the oxygen level or the change
-  in oxygen level. The remaining black arrows denote standard dependencies
-  of the SLDS: discrete syllables and continuous latent states co-evolve, with
-  the syllables specifying the linear dynamics to use when updating the continuous
-  latent states.  The observations are a function of these underlying latent states.
-  The only non-standard addition is an autoregressive dependency between~$y_t$ and
-  $y_{t+1}$; here, the SLDS models only the differences in observed fluoresence
-  value between two consecutive frames.
-}
-\label{fig:graphical_model}
-\end{figure}
+\subsection{Extensions of the standard SLDS}
+\label{sec:extensions}
+
+To better model the \celegans~data, we introduce three extensions
+to the standard SLDS.
+
+\paragraph{Hierarchical model to allow limited variability across worms.}
+The first extension captures shared patterns across worms while
+still allowing for worm-to-worm variability.  We do this in
+two ways. First, we allow the dynamics parameters to vary
+slightly from worm to worm with the following model,
+\begin{align*}
+  \mathrm{vec}([A_k, b_k]) &\sim \distNormal(\lambda), \\
+  \mathrm{vec}([A_k^{(w)}, b_k^{(w)}]) &\sim \distNormal(\mathrm{vec}([A_k, b_k]), \sigma^2 I), \\
+  Q_k^{(w)} &\sim \mathrm{IW}(\lambda).
+\end{align*}
+This allows worm-specific dynamics parameters~$(A_k^{(w)}, b_k^{(w)})$ but
+requires that they not deviate too far from the global mean~$(A_k, b_k)$.
+It also allows for worm-specific dynamics noise.
+
+Second, we allow each worm to have neuron-specific emission variances,
+\begin{align*}
+  s_{n}^{(w)} & \iid\sim \mathrm{IG}(\eta).
+\end{align*}
+We may expect, for example, that calcium indicator expression
+varies from worm to worm, which in turn could give rise to
+different levels of observed activity.  While this could be
+modeled as a scaling of the emission matrix, a simpler approach
+is to let the noise term capture these different fluctuations.
+We could share the global mean of the worm-specific variances,
+but we have little reason to expect the~$s_{n}^{(w)}$ and~$s_{n}^{(w')}$
+to be correlated under the prior.
+
+Figure~\ref{fig:graphical_model} illustrates the complete
+probabilistic graphical model.  Each worm has a set of discrete
+states~$z_{1:T}$, continuous states~$x_{1:T}$, and observations for a
+subset of neurons.  The worms share a set of global dynamics
+parameters and emission matrices, but they also have worm-specific
+perturbations of the global dynamics, as well as worm-specific
+observation variances.
+
+\paragraph{Robust dynamics model.}  While the SLDS can capture nonlinear
+dynamics by composing simple linear pieces, it is still an approximation
+to the true data-generating process.  To account for some of this
+model misspecification, we introduce a \emph{robust} extension of the
+SLDS by allowing for heavy-tailed noise in the dynamics.  Specifically,
+we replace~\eqref{eq:slds_start} with,
+\begin{align}
+  x_{t+1}^{(w)} &= A_{z_{t+1}^{(w)}} x_{t}^{(w)} + b_{z_{t+1}^{(w)}} +  u_t^{(w)},
+  &
+  u_t^{(w)} &\iid\sim \mathrm{t}(0, Q_{z_{t+1}^{(w)}}, \nu_{z_{t+1}^{(w)}}),
+\end{align}
+where~$\mathrm{t}(\cdot, \cdot, \cdot)$ denotes the multivariate-t
+distribution.  This distribution has heavier tails than its Gaussian
+counterpart; indeed, it can be derived by a mixture of multivariate
+Gaussians with a~$\chi^2$-distributed scaling of the covariance matrix.
+
+\paragraph{Recurrent model of discrete state transition probabilities.}
+Finally, we consider a third extension that allows for more complex
+transition probabilities.  We replace~\eqref{eq:markov_z} with,
+\begin{align}
+  z_{t+1}^{(w)} \given z_t^{(w)}, x_t^{(w)}
+  &\sim \mathrm{softmax}(r_{z_t^{(w)}} + R x_t^{(w)}),
+    \label{eq:recurrent}
+\end{align}
+where~${r_k \in \reals^K}$ and~${R \in \reals^{K \times D}}$
+parameterize a map from previous discrete and continuous states to a
+distribution over next discrete states.  The link function is the
+softmax function, which exponentiates and normalizes
+the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.
+
+\paragraph{Input-driven transition probabilities}
+todo
+\todo[inline]{Detail the model for how transition probabilities depend
+  on oxygen and change in oxygen concentration.  Straightforward extension
+  of the recurrent model above.}
+
+\section{Model fitting}
+
+We have proposed a variety of Bayesian algorithms for learning the
+model parameters~$\theta$ and performing joint inference of the
+latent states~$z$ and~$x$~\citep{linderman2017recurrent, linderman2017structure},
+but for simplicity, here we treat inference of~$x$ and~$z$ separately.
+That is, first we find low-dimensional continuous states to model the
+observations~$y$, then we find a set of discrete states~$z$ that best
+explain the dynamics in~$x$. We learn the corresponding parameters
+at both stages.
+
+To be more specific, first we fit linear dynamical systems to infer
+the continuous latent states~$x$ given the observed values of~$y$.
+Since many entries of~$y$ are masked off, this requires inference
+with missing data.  Under the assumption of diagonal observation
+covariance, this inference is straightforward.  We will compare the
+standard LDS with the hierarchical LDS, which has worm- and neuron-specific
+variances.
+
+Once we have inferred the continuous latent states, we will fit
+an autoregressive hidden Markov model (AR-HMM) to~$x$ in order
+to infer~$z$ and learn the dynamics parameters~$(A_k, b_k, Q_k)$.
+Here, again, we will consider all possible variants of hierarchical,
+robust, and recurrent models.
+
+Both fitting procedures use Markov chain Monte Carlo, specifically
+block Gibbs sampling, to sample from the conditional distribution
+of latent states given parameters and of parameters given latent
+states.  We run these algorithms for 1000 iterations, at which
+point we find them to have converged on the basis of held-out
+log likelihood.
 
@@
 
+We choose the dimensionality of the continuous latent state space
+and the number of discrete syllables using the log likelihood of
+held-out test data, marginalizing over the corresponding latent
+states.  Since we have split inference into two steps, this amounts
+to computing the marginal likelihood in a linear dynamical system (LDS)
+and an autoregressive hidden Markov model (AR-HMM), respectively.
+Both model classes admit efficient marginal likelihood computation
+with simple message passing algorithms. Fig.~\ref{fig:model_selection}
+shows the held-out log likelihoods used to select the number of
+syllables and to justify the hierarchical, robust, and recurrent
+extensions.
+
 \begin{figure}[t!]
@@
 
-\subsection{Eigenspectra of the inferred dynamics}
+Fig.~\ref{fig:z_predictions} suggests that the switching linear model
+for the dynamics of the continuous latent states is indeed a good
+model.  In blue we show the predictive distribution~$p(x \given z^*)$
+where~$z^*$ is a posterior sample of the discrete syllable sequence.
+That this predictive distribution (blue) generally places high
+probability on the true values of~$x$ (black lines) indicates that the
+switching linear model is a good approximation for the continuous
+latent brain states.
+
+\clearpage
+
+\section{Simultaneous identification with an entire population of unlabeled neurons}
+
+\begin{table}[h!]
+  \caption{\textit{Artificial neuron identification task results.}  In
+    each worm we have~$N_{\mathsf{labeled}}$ labeled neurons
+    and~$N_{\mathsf{unlabeled}}$ unlabeled neurons. Of the unlabeled
+    neurons, 10 had known labels that were held out for this task.  We
+    compare the cosine similarity of the unlabeled neuron embeddings
+    to the embeddings of the labeled neurons, sort the similarities,
+    and compute the rank of each held-out neuron to its true label
+    (most similar~=~1; least similar~=~$N_{\mathsf{unlabeled}}$).  We
+    use the Hungarian algorithm to find the most likely matching of
+    unlabeled neurons to possible labels, and we report the fraction
+    of the 10 held-out neurons that were correctly matched~(``Acc.'').
+  }
+  \vspace{1em}
+  \label{tab:neuron_id}
+  \centering
+  \begin{tabular}{cccccccccccccc}
+    \toprule
+    & $N_{\mathsf{labeled}}$ & $N_{\mathsf{unlabeled}}$ & \multicolumn{10}{c}{Matching Rank (10 Held-out Neurons)} & Acc. \\
+    \cmidrule(r){2-2}
+    \cmidrule(rl){3-3}
+    \cmidrule(rl){4-13}
+    \cmidrule(l){14-14}
+    % \midrule
+    Worm 1 & 14 & 95 & 7 & 16 & 1 & 1 & 77 & 3 & 2 & 6 & 4 & 2 & 0.60 \\
+    Worm 2 & 31 & 76 & 1 & 5 & 1 & 2 & 2 & 4 & 3 & 2 & 1 & 12 & 0.60 \\
+    Worm 3 & 20 & 111 & 7 & 66 & 10 & 9 & 1 & 1 & 1 & 1 & 4 & 1 & 0.70 \\
+    Worm 4 & 33 & 93 & 1 & 1 & 3 & 1 & 1 & 48 & 1 & 3 & 4 & 2 & 0.60 \\
+    Worm 5 & 38 & 91 & 2 & 1 & 1 & 1 & 1 & 2 & 2 & 3 & 4 & 66 & 0.80 \\
+    % Worm 1 & 14 & 95 & 10 & 25 & 1 & 1 & 66 & 2 & 1 & 1 & 2 & 1 & 0.30 \\
+    % Worm 2 & 31 & 76 & 3 & 1 & 1 & 1 & 1 & 39 & 1 & 1 & 1 & 4 & 0.50 \\
+    % Worm 3 & 20 & 111 & 5 & 53 & 3 & 22 & 1 & 2 & 1 & 1 & 1 & 1 & 0.50 \\
+    % Worm 4 &33 & 93 & 2 & 1 & 26 & 2 & 1 & 56 & 1 & 1 & 1 & 1 & 0.30 \\
+    % Worm 5 & 38 & 91 & 1 & 1 & 1 & 1 & 1 & 2 & 1 & 26 & 1 & 67 & 0.70 \\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+\todo[inline]{Discuss this experiment in detail because it is a bit complicated but the results are impressive.}
+
+\clearpage
+
+\section{Eigenspectra of the inferred dynamics}
 \begin{figure}[t!]
@@
 
+\clearpage
+
+\section{Further analysis of~\citet{nichols2017global} data}
+
 \begin{figure}[t!]
@@
 
-% This elicits a strong response in some clusters of neurons, regardless
-% of genetic background and developmental stage.  In npr1 worms, the
-% elevated oxygen level will elicit a sustained period of activity, even
-% in the lethargic stage, whereas N2 worms remain largely quiescent
-% despite the increased oxygen concentration.
+Fig.~\ref{fig:nichols_supp1} shows 3 minute windows of data from
+four worms in the~\citet{nichols2017global} dataset, one worm
+from each combination of genetic background (N2 vs npr1) and
+developmental stage (pre-lethargus vs lethargus). The window is
+centered on the 6 minute mark, at which point the oxygen level
+was increased from 10\% to 21\%. At this point in time, we see
+a transient response in some clusters of neurons, as discussed
+in the main text.  We also note that the noise level varies
+substantially from one worm to another, motivating the use of
+our hierarchical modeling framework.
 
@@
 
+Fig.~\ref{fig:nichols_supp2} shows the 6-dimensional latent state
+trajectories for the four worms whose activity was shown in
+Fig.~\ref{fig:nichols_supp1}.  As in Fig.~\ref{fig:syllables} in the
+main text, we have color coded the trajectories by the inferred
+discrete syllables in panels a. and c., and by the manual segmentation
+of~\citet{nichols2017global} in panels b. and d.  Panel e. shows the
+average state overlap from all worms of the corresponding genetic
+background and developmental stage.  Thought the overlap is not as
+one-to-one as we saw in the \citet{kato2015global} dataset, we still
+see a strong overlap between manually-labeled quiescent (Q) and
+forward (FWD) states and inferred blue and red syllables.  Likewise,
+reversals (Rev), dorsal turns (DT), and ventral turns (VT) are often
+mapped onto the latter syllables.  We note that the pre-lethargus
+worms tend to execute many more turns, making these states much easier
+to identify in these worms.  As in the main text, the syllbles
+are ordered (blue, red, ...)  according to overall usage.
+
 \end{document}

### 2018-02-26 experiments/fit_arhmm.py

@@ -6,6 +6,7 @@ import itertools as it
 
 import numpy as np
 np.random.seed(0)
+from scipy.ndimage import gaussian_filter1d
 
 from tqdm import tqdm
 
@@ -45,33 +46,39 @@ from autoregressive.models import ARWeakLimitStickyHDPHMM, ARHMM
 from pyslds.util import get_empirical_ar_params
 from pybasicbayes.distributions import AutoRegression, RobustAutoRegression
 from rslds.models import SoftmaxRecurrentARHMM
-from zimmer.models import HierarchicalARWeakLimitStickyHDPHMM, HierarchicalRecurrentARHMM
+from zimmer.models import HierarchicalARWeakLimitStickyHDPHMM, HierarchicalRecurrentARHMM, HierarchicalRecurrentARHMMWithNN
 from zimmer.dynamics import HierarchicalAutoRegression, HierarchicalRobustAutoRegression
 
 import importlib
+from zimmer.io import WormData, load_kato_key
 import zimmer.plotting
 importlib.reload(zimmer.plotting)
 from zimmer.plotting import plot_3d_continuous_states, plot_2d_continuous_states, plot_expected_states, \
     plot_3d_dynamics, plot_2d_dynamics, plot_state_overlap, plot_state_usage_by_worm, plot_all_transition_matrices, \
-    plot_simulated_trajectories, make_state_predictions_3d_movie, plot_simulated_trajectories2, plot_simulated_trajectories3,\
+    plot_simulated_trajectories, make_state_predictions_3d_movie, plot_simulated_trajectories2, plot_simulated_trajectories4,\
     plot_recurrent_transitions, plot_x_at_changepoints, plot_latent_trajectories_vs_time, \
-    plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_duration_cdfs
+    plot_state_usage_by_worm_matrix, plot_duration_histogram, plot_duration_cdfs, plot_simulated_cluster_activation, \
+    make_states_3d_movie, make_states_dynamics_movie, plot_changepoint_prs
 
 # LDS Results
+# lds_dir = os.path.join("results", "2018-01-17-hlds", "run001_D10")
 lds_dir = os.path.join("results", "2017-11-03-hlds", "run003")
+signal = "dff_diff"
 # lds_dir = os.path.join("results", "2017-11-03-hlds", "run003_dff_bc")
 assert os.path.exists(lds_dir)
 
 # AR-HMM RESULTS
[…]

### 2018-02-26 experiments/fit_hlds.py

@@ -44,13 +44,19 @@ from zimmer.io import WormData, load_kato_key
 from zimmer.models import HierarchicalLDS
 from zimmer.emissions import HierarchicalDiagonalRegression
 from zimmer.plotting import plot_3d_continuous_states
+
+from zimmer.plotting import make_states_3d_movie
 from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
 
 # IO
-run_num = 3
-results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
-signal = "dff_diff"
+# run_num = 3
+# results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
+# signal = "dff_diff"
+run_num = 1
+results_dir = os.path.join("results", "2018-01-17-hlds", "run{:03d}".format(run_num))
+signal = "dff_deriv"
+
 
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
@@ -189,7 +195,7 @@ def _fit_lds(D_latent, D_in=1, alpha_0=1.0, beta_0=1.0,
 
     datas = ytrains if datas is None else datas
     masks = mtrains if masks is None else masks
-    for i in range(N_worms):
+    for i in range(len(datas)):
         data_kwargs = dict(group=i) if is_hierarchical else dict()
         model.add_data(datas[i], mask=masks[i],
                        inputs=np.ones((datas[i].shape[0], 1)),
@@ -315,6 +321,7 @@ def cluster_neruons(best_model, seed=0):
 
 
 def plot_best_model_results(best_model,
+                            do_plot_pca=True,
                             do_plot_x_3d=True,
                             do_plot_x_2d=True,
                             do_plot_sigmasq=True,
@@ -326,29 +333,82 @@ def plot_best_model_results(best_model,
                             do_plot_data_wide=True,
                             do_plot_data_zoom=True,
                             do_plot_data_as_matrix=True,
-                            do_plot_smoothed_data_as_matrix=True):
+                            do_plot_smoothed_data_as_matrix=True,
+                            do_plot_cluster_types=True):
+
+    if do_plot_pca:
+        from sklearn.decomposition import PCA
+        for i in range(N_worms):
+            # plot_3d_continuous_states(xtrains[i], np.zeros(xtrains[i].shape[0], dtype=int), colors,
+            #                           figsize=(1.2, 1.2),
[…]

### 2018-02-26 experiments/fit_hlds_nichols.py

@@ -832,6 +832,8 @@ if __name__ == "__main__":
     n_trains = np.array([mtr.sum() for mtr in mtrains])
     n_tests = np.array([mte.sum() for mte in mtests])
 
+    assert False
+
     # D_latents = np.arange(2, 31, 2)
     D_latents = [6]
     best_models = fit_all_models(D_latents)
@@ -874,27 +876,27 @@ if __name__ == "__main__":
     # heldout_neuron_identification()
 
     # Save out the results
-    results = dict(
-        xtrains=xtrains,
-        xtests=xtests,
-        ytrains=ytrains,
-        ytests=ytests,
-        mtrains=mtrains,
-        mtests=mtests,
-        utrains=utrains,
-        utests=utests,
-        z_true_trains=z_true_trains,
-        z_true_tests=z_true_tests,
-        z_key=z_true_key,
-        best_model=best_model,
-        D_latent=best_model.D_latent,
-        C=C,
-        perm=dim_perm,
-        N_clusters=N_clusters,
-        neuron_clusters=neuron_clusters,
-        neuron_perm=neuron_perm,
-        neuron_names=neuron_names,
-    )
-
-    with open(os.path.join(results_dir, "lds_data.pkl"), "wb") as f:
-        pickle.dump(results, f)
+    # results = dict(
+    #     xtrains=xtrains,
+    #     xtests=xtests,
+    #     ytrains=ytrains,
+    #     ytests=ytests,
+    #     mtrains=mtrains,
+    #     mtests=mtests,
+    #     utrains=utrains,
+    #     utests=utests,
+    #     z_true_trains=z_true_trains,
+    #     z_true_tests=z_true_tests,
+    #     z_key=z_true_key,
+    #     best_model=best_model,
+    #     D_latent=best_model.D_latent,
+    #     C=C,
+    #     perm=dim_perm,
+    #     N_clusters=N_clusters,
+    #     neuron_clusters=neuron_clusters,
+    #     neuron_perm=neuron_perm,
+    #     neuron_names=neuron_names,
+    # )
+    #
+    # with open(os.path.join(results_dir, "lds_data.pkl"), "wb") as f:
+    #     pickle.dump(results, f)

### 2018-02-26 experiments/make_talk_figures.py

@@ -1,0 +1,646 @@
+import os
+import pickle
+
+import numpy as np
+np.random.seed(1234)
+from scipy.ndimage import gaussian_filter1d
+
+
+# Plotting stuff
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+from hips.plotting.colormaps import gradient_cmap
+from hips.plotting.layout import create_axis_at_location
+import seaborn as sns
+sns.set_style("white")
+sns.set_context("talk")
+color_names = ["windows blue",
+               "red",
+               "amber",
+               "faded green",
+               "dusty purple",
+               "orange",
+               "clay",
+               "pink",
+               "greyish",
+               "mint",
+               "light cyan",
+               "steel blue",
+               "forest green",
+               "pastel purple",
+               "salmon",
+               "dark brown"]
+colors = sns.xkcd_palette(color_names)
+cmap = gradient_cmap(colors)
+
+
+# Modeling stuff
+from pyhsmm.util.general import relabel_by_usage, rle
+
+# Load worm modeling specific stuff
+from zimmer.io import WormData, load_kato_key
+import zimmer.plotting as zplt
+
+# LDS Results
+lds_dir = os.path.join("results", "2017-11-03-hlds", "run003")
+assert os.path.exists(lds_dir)
+
+results_dir = os.path.join("results", "2018-01-19-arhmm", "run001")
+
+# Figure dir
+fig_dir = os.path.join("results", "talk")
+
+N_worms = 5
+N_clusters = 12
+
+
+def load_data(include_unnamed=True):
+    # Load the data
+    worm_datas = [WormData(i, name="worm{}".format(i), version="kato") for i in range(N_worms)]
+
+    # Get the "true" discrete states as labeled by Zimmer
+    z_trues = [wd.zimmer_states for wd in worm_datas]
+    z_trues, newlabels = relabel_by_usage(z_trues, return_mapping=True)
+
+    # Get the key
+    z_key = load_kato_key()
+    z_key = [z_key[i] for i in np.argsort(newlabels)]
+
+    # Get the names of the neurons
[…]

### 2018-02-26 zimmer/models.py

@@ -89,6 +89,60 @@ class _HierarchicalARHMMMixin(object):
     # todo: implement generate function
 
 
+    def generate(self, T=100, keep=True, init_x=None, init_z=None, covariates=None, with_noise=True, group=None,
+                 tau=0, noise_scale=1):
+        from pybasicbayes.util.stats import sample_discrete, sample_gaussian
+        # Generate from the prior and raise exception if unstable
+        K, n = self.num_states, self.D
+
+        # Initialize discrete state sequence
+        zs = np.empty(T, dtype=np.int32)
+        if init_z is None:
+            zs[0] = sample_discrete(self.init_state_distn.pi_0.ravel())
+        else:
+            zs[0] = init_z
+
+        xs = np.empty((T, n), dtype='double')
+        if init_x is None:
+            xs[0] = np.random.randn(n)
+        else:
+            xs[0] = init_x
+
+        for t in range(1, T):
+            # Sample discrete state given previous continuous state
+            A = self.trans_distn.trans_matrix
+            zs[t] = sample_discrete(A[zs[t-1], :])
+
+            # Sample continuous state given current discrete state
+            mu = self.obs_distns[zs[t]].predict(xs[t - 1][None, :], group=group)[0]
+            sigma = self.obs_distns[zs[t]].sigmas[group] * noise_scale
+            J_dyn = np.linalg.inv(sigma)
+            h = J_dyn.dot(mu)
+
+            # Combine predicted mean with stability penalty from
+            #   p(1^T x | 0, \tau)
+            # where \tau = stable_prec. Thisintroduces potential of the form
+            #   exp( -\tau/2 (1^T x)^2 )
+            # = exp( -\tau/2 x^T 1 1^T x)
+            #
+            # In information form, this is N(x | J = \tau 11^T, h = 0)
+            J_pen = tau * np.eye(self.D)
+
+            if with_noise:
+                # xs[t] = self.obs_distns[zs[t]].rvs(xs[t-1][None, :], return_xy=False, group=group)
+                xs[t] = sample_gaussian(J=J_dyn+J_pen, h=h)
+            else:
+                xs[t] = np.linalg.solve(J_dyn + J_pen, h)
+
+            # assert np.all(abs(xs[t]) < 5), "RARHMM appears to be unstable!"
+
+        # TODO:
+        # if keep:
+        #     ...
+
[…]

### 2018-02-26 zimmer/plotting.py

@@ -1,11 +1,3 @@
-# Plot the vector field showing inferred (discrete-state-dependent)
-# dynamics at each point in continuous latent space.
-# Sum over time bins and smoothed continuous latent states.
-# so that each point in continuous latent space is
-#    E_{z} E_x_t [A x_{t} + b]
-#
-# We would ideally sample over A as well, but to start, just take
-# a single sample of A.
 import os
 import numpy as np
 from scipy.stats import multivariate_normal
@@ -14,7 +6,9 @@ import matplotlib.pyplot as plt
 import matplotlib.gridspec as gridspec
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 from mpl_toolkits.mplot3d import Axes3D
+from mpl_toolkits.mplot3d import proj3d
 import matplotlib.animation as manimation
+from matplotlib.patches import FancyArrowPatch
 
 from hips.plotting.colormaps import gradient_cmap
 from hips.plotting.sausage import sausage_plot
@@ -47,6 +41,20 @@ default_cmap = gradient_cmap(default_colors)
 
 from zimmer.util import states_to_changepoints
 from tqdm import tqdm
+
+
+
+class Arrow3D(FancyArrowPatch):
+    # From https://stackoverflow.com/questions/22867620/putting-arrowheads-on-vectors-in-matplotlibs-3d-plot
+    def __init__(self, xs, ys, zs, *args, **kwargs):
+        FancyArrowPatch.__init__(self, (0,0), (0,0), *args, **kwargs)
+        self._verts3d = xs, ys, zs
+
+    def draw(self, renderer):
+        xs3d, ys3d, zs3d = self._verts3d
+        xs, ys, zs = proj3d.proj_transform(xs3d, ys3d, zs3d, renderer.M)
+        self.set_positions((xs[0],ys[0]),(xs[1],ys[1]))
+        FancyArrowPatch.draw(self, renderer)
 
 
 def plot_1d_continuous_states(x_inf, z_inf, z_zimmer, colors,
@@ -123,7 +131,9 @@ def plot_3d_continuous_states(x, z, colors,
 
     if ax is None:
         fig = plt.figure(figsize=figsize)
+        fig.patch.set_alpha(0)
         ax = fig.add_subplot(111, projection="3d")
+        ax.patch.set_alpha(0)
 
     cps = states_to_changepoints(z)
 
[…]

### 2018-02-26 zimmer/io.py
fixing SMB neuron names per Manuel's instructions

@@ -78,6 +78,16 @@ def _load_kato(index, name, sample_rate):
                                else str(x[0, 1][0]),
                 neuron_ids)))
 
+        # Fix labels of SMB neurons per Manuel's instructions on Jan 16, 2018
+        mapping = dict(SMBDL="SMDDL", SMBDR="SMDDR", SMBVL="SMDDL", SMBVR="SMDDR")
+        def translate(ids):
+            for j, id in enumerate(ids):
+                if id in mapping:
+                    ids[j] = mapping[id]
+                    
+        translate(neuron_ids_1)
+        translate(neuron_ids_2)
+
         neuron_names = _get_neuron_names(neuron_ids_1, neuron_ids_2, name)
 
         # Get the calcium trace (corrected for bleaching)

### 2018-02-26 examples/hierarchical_factor_analysis.py
fit hlds

@@ -1,131 +1,0 @@
-import numpy as np
-np.random.seed(1)
-
-from pybasicbayes.util.text import progprint_xrange
-import matplotlib.pyplot as plt
-from matplotlib.cm import get_cmap
-from pybasicbayes.util.stats import sample_mniw
-
-from zimmer.models import HierarchicalFactorAnalysis
-
-N = 2000
-D_obs = 20
-D_latent = 2
-
-def principal_angle(A,B):
-    """
-    Find the principal angle between two subspaces
-    spanned by columns of A and B
-    """
-    from numpy.linalg import qr, svd
-    qA, _ = qr(A)
-    qB, _ = qr(B)
-    U,S,V = svd(qA.T.dot(qB))
-    return np.arccos(min(S.min(), 1.0))
-
-
-def generate_synth_data(N_groups=5):
-
-    # Create a true model and sample from it
-    mask = np.random.rand(N,D_obs) < 0.9
-    true_model = HierarchicalFactorAnalysis(D_obs, D_latent, N_groups)
-    true_datas = []
-    for group in range(N_groups):
-        true_datas.append(true_model.generate(N=N, mask=mask, group=group, keep=True))
-    return true_model, true_datas
-
-
-def plot_results(lls, angles, Ztrue, Zinf, sigmasq_true, sigmasq_inf):
-    # Plot log probabilities
-    plt.figure()
-    plt.plot(lls)
-    plt.ylabel("Log Likelihood")
-    plt.xlabel("Iteration")
-
-    plt.figure()
-    plt.plot(np.array(angles) / np.pi * 180.)
-    plt.ylabel("Principal Angle")
-    plt.xlabel("Iteration")
-
-    # Plot locations, color by angle
-    N = Ztrue.shape[0]
-    inds_to_plot = np.random.randint(0, N, min(N, 500))
-    th = np.arctan2(Ztrue[:,1], Ztrue[:,0])
-    nperm = np.argsort(np.argsort(th))
-    cmap = get_cmap("jet")
-
-    plt.figure()
-    plt.subplot(121)
-    for n in inds_to_plot:
-        plt.plot(Ztrue[n,0], Ztrue[n,1], 'o', markerfacecolor=cmap(nperm[n] / float(N)), markeredgecolor="none")
-    plt.title("True Embedding")
-    plt.xlim(-4,4)
-    plt.ylim(-4,4)
-
-    plt.subplot(122)
-    for n in inds_to_plot:
[…]

### 2018-02-26 experiments/demo.py

@@ -1,171 +1,0 @@
-from __future__ import division
-import numpy as np
-import numpy.random as npr
-npr.seed(0)
-
-import matplotlib.pyplot as plt
-import matplotlib.gridspec as gridspec
-import seaborn as sns
-from hips.plotting.colormaps import gradient_cmap
-
-sns.set_style("white")
-sns.set_context("paper")
-
-color_names = ["red",
-               "windows blue",
-               "medium green",
-               "dusty purple",
-               "orange",
-               "amber",
-               "clay",
-               "pink",
-               "greyish",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "mint",
-               "salmon",
-               "dark brown"]
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-from pybasicbayes.distributions import Regression, Gaussian
-from pyhsmm.util.general import rle
-from pylds.util import random_rotation
-
-from zimmer.models import HierarchicalHMMSLDS
-from zimmer.emissions import HierarchicalDiagonalRegression
-
-
-###################
-#  generate data  #
-###################
-N_groups = 4
-K = 5                               # number of latent discrete states
-D_latent = 2                        # latent linear dynamics' dimension
-D_obs = 2                           # data dimension
-D_input = 1                         # input dimension
-T = 1000                            # duration of data
-
-true_mu_inits = [np.ones(D_latent) for _ in range(K)]
-true_sigma_inits = [np.eye(D_latent) for _ in range(K)]
-true_As = [.99 * random_rotation(D_latent, theta=np.pi/((k+1) * 4)) for k in range(K)]
-true_Bs = [3 * npr.randn(D_latent, D_input) for k in range(K)]
-true_Qs = [np.eye(D_latent) for _ in range(K)]
-true_C = npr.randn(D_obs, D_latent + D_input)
-true_Rs = np.tile(np.logspace(-1, -1 + N_groups, N_groups, endpoint=False)[:,None], (1, D_obs))
-
-init_dynamics_distns = [Gaussian(mu=mu, sigma=sigma)
[…]

### 2018-02-26 experiments/fit_hlds.py

@@ -53,10 +53,14 @@ from pylds.models import MissingDataLDS
 # run_num = 3
 # results_dir = os.path.join("results", "2017-11-03-hlds", "run{:03d}".format(run_num))
 # signal = "dff_diff"
-run_num = 1
-results_dir = os.path.join("results", "2018-01-17-hlds", "run{:03d}".format(run_num))
-signal = "dff_deriv"
 
+# run_num = 1
+# results_dir = os.path.join("results", "2018-01-17-hlds", "run{:03d}".format(run_num))
+# signal = "dff_deriv"
+
+run_num = 1
+results_dir = os.path.join("results", "kato", "2018-02-26-hlds", "run{:03d}".format(run_num))
+signal = "dff_diff"
 
 assert os.path.exists(results_dir)
 fig_dir = os.path.join(results_dir, "figures")
@@ -220,7 +224,45 @@ def _fit_lds(D_latent, D_in=1, alpha_0=1.0, beta_0=1.0,
     return model, np.array(lls), hll
 
 
-def plot_likelihoods(D_latents, final_lls, hlls, best_index,
+def fit_all_models(D_latents=np.arange(2, 21, 2)):
+
+    results = {}
+
+    for index, is_hierarchical in enumerate([True, False]):
+
+        models = []
+        llss = []
+        hlls = []
+
+        for D_latent in D_latents:
+            name = "{}_{}".format(
+                "hlds" if is_hierarchical else "lds",
+                D_latent
+            )
+            print("Fitting model: {}".format(name))
+
+            fit = cached(name)(
+                partial(_fit_lds,
+                        is_hierarchical=is_hierarchical))
+            model, lls, hll = fit(D_latent)
+
+            # Append results
+            models.append(model)
+            llss.append(lls)
+            hlls.append(hll)
+
+        final_lls = np.array([lls[-1] for lls in llss])
+        hlls = np.array(hlls)
+        best_index = np.argmax(hlls)
+        print("Best dimension: {}".format(D_latents[best_index]))
+
+        results["hier" if is_hierarchical else "no_hier"] = \
+            models[best_index], best_index, models, llss, final_lls, hlls
+
+    return results
+
+
+def _plot_likelihoods(D_latents, final_lls, hlls, best_index,
                      color, name, axs=None):
 
     # Plot results of searching over models
@@ -260,6 +302,17 @@ def plot_likelihoods(D_latents, final_lls, hlls, best_index,
[…]

### 2018-02-26 experiments/fit_hslds.py

@@ -1,868 +1,0 @@
-import os, pickle, copy
-
-import numpy as np
-np.random.seed(0)
-
-
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.gridspec as gridspec
-import matplotlib.animation as manimation
-
-from hips.plotting.colormaps import gradient_cmap
-from hips.plotting.sausage import sausage_plot
-from hips.plotting.layout import create_axis_at_location
-
-# Come up with a set of colors
-import seaborn as sns
-sns.set_style("white")
-# sns.set_context("paper")
-
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-import importlib
-from sklearn.decomposition import PCA
-
-from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
-from pybasicbayes.util.text import progprint_xrange
-from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
-from pyslds.util import get_empirical_ar_params
-
-
-
-import zimmer.io
-importlib.reload(zimmer.io)
-from zimmer.io import WormData, N_worms, find_shared_neurons, load_kato_key
-
-import zimmer.states
-importlib.reload(zimmer.states)
-
-import zimmer.models
-importlib.reload(zimmer.models)
-from zimmer.models import HierarchicalWeakLimitStickyHDPHMMSLDS
-
-from zimmer.emissions import HierarchicalDiagonalRegressionFixedScale
-from zimmer.emissions import HierarchicalDiagonalRegressionTruncatedScale
-
-import zimmer.plotting
-importlib.reload(zimmer.plotting)
-from zimmer.plotting import plot_1d_continuous_states, plot_3d_continuous_states, plot_vector_field_3d
-
-
-from zimmer.util import states_to_changepoints
-
-# IO
-run_num = 1
[…]

### 2018-02-26 experiments/fit_nonconj_rslds.py

@@ -1,871 +1,0 @@
-import os
-import pickle
-
-import numpy as np
-import numpy.random as npr
-npr.seed(1)
-
-import matplotlib.pyplot as plt
-from mpl_toolkits.axes_grid1 import make_axes_locatable
-import matplotlib.gridspec as gridspec
-from matplotlib.font_manager import FontProperties
-
-from mpl_toolkits.mplot3d import Axes3D
-from mpl_toolkits.axes_grid1 import make_axes_locatable
-
-
-import seaborn as sns
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-
-colors = sns.xkcd_palette(color_names)
-sns.set_style("white")
-sns.set_context("paper")
-
-from hips.plotting.colormaps import gradient_cmap
-cmap = gradient_cmap(colors)
-from hips.plotting.layout import create_axis_at_location
-
-from pybasicbayes.util.text import progprint_xrange
-from pybasicbayes.models import FactorAnalysis
-from pybasicbayes.distributions import \
-    Regression, Gaussian, DiagonalRegression, AutoRegression
-
-from pybasicbayes.util.text import progprint_xrange
-
-from pyhsmm.util.general import relabel_by_permutation, relabel_by_usage
-from autoregressive.models import ARWeakLimitStickyHDPHMM
-from pyslds.util import get_empirical_ar_params
-from pylds.util import random_rotation
-
-from pinkybrain.models import MixedEmissionHMMSLDS
-from pyslds.models import WeakLimitStickyHDPHMMSLDS
-from rslds.rslds import RecurrentSLDS
-from rslds.nonconj_rslds import SoftmaxRecurrentOnlySLDS
-from rslds.util import compute_psi_cmoments
-
-from zimmer.io import WormData, N_worms, find_shared_neurons, load_kato_key
-from zimmer.util import states_to_changepoints
-
-### Global parameters
-K = 5
-D_latent = 3
-
[…]

### 2018-02-26 experiments/fit_rhslds.py

@@ -1,1135 +1,0 @@
-import os, pickle, copy
-
-import numpy as np
-np.random.seed(0)
-
-
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.gridspec as gridspec
-import matplotlib.animation as manimation
-
-from hips.plotting.colormaps import gradient_cmap
-from hips.plotting.sausage import sausage_plot
-from hips.plotting.layout import create_axis_at_location
-
-# Come up with a set of colors
-import seaborn as sns
-sns.set_style("white")
-# sns.set_context("paper")
-
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-import importlib
-from sklearn.decomposition import PCA
-
-from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
-from pybasicbayes.util.text import progprint_xrange
-from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
-from pyslds.util import get_empirical_ar_params
-
-from pypolyagamma.distributions import MultinomialRegression
-
-from rslds.util import one_hot
-
-import zimmer.io
-importlib.reload(zimmer.io)
-from zimmer.io import WormData, N_worms, find_shared_neurons, load_kato_key
-
-import zimmer.states
-importlib.reload(zimmer.states)
-
-import zimmer.models
-importlib.reload(zimmer.models)
-from zimmer.models import HierarchicalWeakLimitStickyHDPHMMSLDS, \
-    HierarchicalRecurrentOnlySLDS, \
-    HierarchicalRecurrentSLDS,  \
-    HierarchicalStickyRecurrentOnlySLDS, \
-    HierarchicalStickyRecurrentSLDS
-
-import zimmer.emissions
-importlib.reload(zimmer.emissions)
-from zimmer.emissions import HierarchicalDiagonalRegression
-from zimmer.emissions import HierarchicalDiagonalRegressionFixedScale
[…]

### 2018-02-26 experiments/fit_rhslds2.py

@@ -1,1044 +1,0 @@
-import os, pickle, copy
-
-import numpy as np
-np.random.seed(0)
-
-
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.gridspec as gridspec
-import matplotlib.animation as manimation
-
-from hips.plotting.colormaps import gradient_cmap
-from hips.plotting.sausage import sausage_plot
-from hips.plotting.layout import create_axis_at_location
-
-# Come up with a set of colors
-import seaborn as sns
-sns.set_style("white")
-# sns.set_context("paper")
-
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-import importlib
-from sklearn.decomposition import PCA
-
-from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
-from pybasicbayes.util.text import progprint_xrange
-from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
-from pyslds.util import get_empirical_ar_params
-
-
-
-import zimmer.io
-importlib.reload(zimmer.io)
-from zimmer.io import WormData, N_worms, find_shared_neurons, load_kato_key
-
-import zimmer.states
-importlib.reload(zimmer.states)
-
-import zimmer.models
-importlib.reload(zimmer.models)
-from zimmer.models import HierarchicalWeakLimitStickyHDPHMMSLDS
-
-from zimmer.emissions import HierarchicalDiagonalRegressionFixedScale
-from zimmer.emissions import HierarchicalDiagonalRegressionTruncatedScale
-
-import zimmer.plotting
-importlib.reload(zimmer.plotting)
-from zimmer.plotting import plot_1d_continuous_states, plot_3d_continuous_states, plot_vector_field_3d
-
-
-from zimmer.util import states_to_changepoints
-
-# IO
-run_num = 1
[…]

### 2018-02-26 experiments/fit_rhslds_final.py

@@ -1,1065 +1,0 @@
-import os, pickle, copy
-
-import numpy as np
-np.random.seed(0)
-
-
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.gridspec as gridspec
-import matplotlib.animation as manimation
-
-from hips.plotting.colormaps import gradient_cmap
-from hips.plotting.sausage import sausage_plot
-from hips.plotting.layout import create_axis_at_location
-
-# Come up with a set of colors
-import seaborn as sns
-sns.set_style("white")
-# sns.set_context("paper")
-
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-import importlib
-from sklearn.decomposition import PCA
-
-from pybasicbayes.distributions import Gaussian, Regression, DiagonalRegression
-from pybasicbayes.util.text import progprint_xrange
-from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
-from pyslds.util import get_empirical_ar_params
-
-import zimmer.io
-importlib.reload(zimmer.io)
-from zimmer.io import WormData, N_worms, find_shared_neurons, load_kato_key
-
-import zimmer.states
-importlib.reload(zimmer.states)
-
-import zimmer.models
-importlib.reload(zimmer.models)
-from zimmer.models import HierarchicalWeakLimitStickyHDPHMMSLDS, HierarchicalRecurrentSLDS
-from zimmer.emissions import HierarchicalDiagonalRegression
-
-import zimmer.plotting
-importlib.reload(zimmer.plotting)
-from zimmer.plotting import plot_1d_continuous_states, plot_3d_continuous_states, plot_vector_field_3d
-
-
-from zimmer.util import states_to_changepoints
-
-# IO
-run_num = 1
-results_dir = os.path.join("results", "08_16_17", "run{:03d}".format(run_num))
-assert os.path.exists(results_dir)
-
-# Hyperparameters
[…]

### 2018-02-26 experiments/nonconj_nascar.py

@@ -1,532 +1,0 @@
-import os
-import pickle
-
-import numpy as np
-import numpy.random as npr
-npr.seed(0)
-
-from pybasicbayes.util.text import progprint_xrange
-from pybasicbayes.models import FactorAnalysis
-from pybasicbayes.distributions import \
-    Regression, Gaussian, AutoRegression
-
-from autoregressive.models import ARWeakLimitStickyHDPHMM
-from pyslds.util import get_empirical_ar_params
-from pylds.util import random_rotation
-
-from pyslds.models import HMMSLDS
-from zimmer.emissions import HierarchicalDiagonalRegression
-from zimmer.models import HierarchicalRecurrentOnlySLDS, HierarchicalHMMSLDS
-
-import matplotlib.pyplot as plt
-import matplotlib.gridspec as gridspec
-from matplotlib.font_manager import FontProperties
-
-from rslds.plotting import plot_most_likely_dynamics, plot_trajectory, plot_z_samples, \
-    plot_data, plot_trajectory_and_probs, plot_all_dynamics
-
-from rslds.util import cached
-
-### Global parameters
-T, K, K_true, D_obs, D_latent = 1000, 4, 4, 10, 2
-N_groups = 3
-N_trials_per_group = 1
-mask_start, mask_stop = 0, 0
-N_iters = 1000
-
-# Save / cache the outputs
-RUN_NUMBER = 1
-RESULTS_DIR = os.path.join("results", "nascar", "run{:03d}".format(RUN_NUMBER))
-
-
-### Plotting code
-def make_figure(true_model, z_true, x_true, y,
-                rslds, zs_rslds, x_rslds,
-                z_rslds_gen, x_rslds_gen, y_rslds_gen,
-                slds, zs_slds, x_slds,
-                z_slds_gen, x_slds_gen, y_slds_gen):
-    """
-    Show the following:
-     - True latent dynamics (for most likely state)
-     - Segment of trajectory in latent space
-     - A few examples of observations in 10D space
-     - ARHMM segmentation of factors
-     - rSLDS segmentation of factors
-     - ARHMM synthesis
-     - rSLDS synthesis
-    """
-    # fig = plt.figure(figsize=(6.5,3.5))
-    fig = plt.figure(figsize=(13,7))
-    gs = gridspec.GridSpec(2,3)
-
-    fp = FontProperties()
[…]

### 2018-02-26 experiments/fit_hlds.py
comment out plotting

@@ -8,31 +8,31 @@ from tqdm import tqdm
 from functools import partial
 
 # Plotting stuff
-import matplotlib.pyplot as plt
-from mpl_toolkits.axes_grid1 import make_axes_locatable
-from hips.plotting.colormaps import gradient_cmap
-from hips.plotting.layout import create_axis_at_location
-import seaborn as sns
-sns.set_style("white")
-sns.set_context("paper")
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
+# import matplotlib.pyplot as plt
+# from mpl_toolkits.axes_grid1 import make_axes_locatable
+# from hips.plotting.colormaps import gradient_cmap
+# from hips.plotting.layout import create_axis_at_location
+# import seaborn as sns
+# sns.set_style("white")
+# sns.set_context("paper")
+# color_names = ["windows blue",
+#                "red",
+#                "amber",
+#                "faded green",
+#                "dusty purple",
+#                "orange",
+#                "clay",
+#                "pink",
+#                "greyish",
+#                "mint",
+#                "light cyan",
+#                "steel blue",
+#                "forest green",
+#                "pastel purple",
+#                "salmon",
+#                "dark brown"]
+# colors = sns.xkcd_palette(color_names)
+# cmap = gradient_cmap(colors)
 
 
 # Modeling stuff
@@ -43,9 +43,10 @@ from zimmer.io import WormData, load_kato_key
 
 from zimmer.models import HierarchicalLDS
 from zimmer.emissions import HierarchicalDiagonalRegression
-from zimmer.plotting import plot_3d_continuous_states
 
-from zimmer.plotting import make_states_3d_movie
+# from zimmer.plotting import plot_3d_continuous_states
+# from zimmer.plotting import make_states_3d_movie
+
 from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
 
@@ -261,58 +262,58 @@ def fit_all_models(D_latents=np.arange(2, 21, 2)):
 
     return results
[…]

### 2018-02-26 zimmer/io.py
move load data functions to io

@@ -5,6 +5,8 @@ import numpy as np
 from scipy.io import loadmat
 from scipy.interpolate import interp1d
 
+from pyhsmm.util.general import relabel_by_usage
+
 # Load the data
 kato_dir = os.path.join("data", "kato2015")
 kato_files = ["TS20140715e_lite-1_punc-31_NLS3_2eggs_56um_1mMTet_basal_1080s.mat",
@@ -275,6 +277,97 @@ def find_shared_neurons(worm_datas):
     print(shared_neurons)
 
     return shared_neurons
+
+
+def load_kato_data(include_unnamed=True, N_worms=5, signal="dff_diff"):
+    # Load the data
+    worm_datas = [WormData(i, name="worm{}".format(i), version="kato") for i in range(N_worms)]
+
+    # Get the "true" discrete states as labeled by Zimmer
+    z_trues = [wd.zimmer_states for wd in worm_datas]
+    z_trues, newlabels = relabel_by_usage(z_trues, return_mapping=True)
+
+    # Get the key
+    z_key = load_kato_key()
+    z_key = [z_key[i] for i in np.argsort(newlabels)]
+
+    # Get the names of the neurons
+    neuron_names = np.unique(np.concatenate([wd.neuron_names for wd in worm_datas]))
+    if not include_unnamed:
+        print("Only including named neurons.")
+        neuron_names = neuron_names[:61]
+    else:
+        print("Including all neurons, regardless of whether they were identified.")
+
+    N_neurons = neuron_names.size
+    print("{} neurons across all {} worms".format(N_neurons, N_worms))
+
+    # Construct a big dataset with all neurons for each worm
+    ys = []
+    masks = []
+    for wd in worm_datas:
+        y_indiv = getattr(wd, signal)
+        y = np.zeros((wd.T, N_neurons))
+        mask = np.zeros((wd.T, N_neurons), dtype=bool)
+        indices = wd.find_neuron_indices(neuron_names)
+        for n, index in enumerate(indices):
+            if index is not None:
+                y[:, n] = y_indiv[:, index]
+                mask[:, n] = True
+
+        ys.append(y)
+        masks.append(mask)
+
+    return ys, masks, z_trues, z_key, neuron_names
+
+
+def load_nichols_data(worms_and_conditions, worm_names, signal="dff_diff", include_unnamed=True):
+    # Load the data
+    worm_datas = [WormData(i,
+                           name=worm_names[j],
+                           version="nichols",
[…]

### 2018-02-26 experiments/fit_hlds.py
minor tweaks

@@ -1383,7 +1383,7 @@ if __name__ == "__main__":
     D_latents = np.arange(2, 21, 2)
     fit_results = fit_all_models(D_latents)
     best_model = fit_results["hier"][0]
-
+    
     # Do an E step to smooth the latent states
     C = best_model.emission_distn.A
     xtrains = []
@@ -1427,12 +1427,15 @@ if __name__ == "__main__":
         perm=dim_perm,
         N_clusters=N_clusters,
         neuron_clusters=neuron_clusters,
-        neuron_perm=neuron_perm
+        neuron_perm=neuron_perm,
+        fit_results=fit_results
     )
 
     with open(os.path.join(results_dir, "lds_data.pkl"), "wb") as f:
         pickle.dump(results, f)
 
+    print("Finished.")
+    
     # Plotting
     # plot_likelihoods(fit_results)
     # plot_best_model_results(best_model,

### 2018-02-26 zimmer/io.py

@@ -295,6 +295,7 @@ def load_kato_data(include_unnamed=True, N_worms=5, signal="dff_diff"):
     neuron_names = np.unique(np.concatenate([wd.neuron_names for wd in worm_datas]))
     if not include_unnamed:
         print("Only including named neurons.")
+        import ipdb; ipdb.set_trace()
         neuron_names = neuron_names[:61]
     else:
         print("Including all neurons, regardless of whether they were identified.")

### 2018-02-26 experiments/fit_hlds.py
moving plotting code to notebooks

@@ -7,45 +7,11 @@ np.random.seed(1234)
 from tqdm import tqdm
 from functools import partial
 
-# Plotting stuff
-# import matplotlib.pyplot as plt
-# from mpl_toolkits.axes_grid1 import make_axes_locatable
-# from hips.plotting.colormaps import gradient_cmap
-# from hips.plotting.layout import create_axis_at_location
-# import seaborn as sns
-# sns.set_style("white")
-# sns.set_context("paper")
-# color_names = ["windows blue",
-#                "red",
-#                "amber",
-#                "faded green",
-#                "dusty purple",
-#                "orange",
-#                "clay",
-#                "pink",
-#                "greyish",
-#                "mint",
-#                "light cyan",
-#                "steel blue",
-#                "forest green",
-#                "pastel purple",
-#                "salmon",
-#                "dark brown"]
-# colors = sns.xkcd_palette(color_names)
-# cmap = gradient_cmap(colors)
-
-
-# Modeling stuff
-from pyhsmm.util.general import relabel_by_usage
-
 # Load worm modeling specific stuff
-from zimmer.io import WormData, load_kato_key
-
+from zimmer.io import load_kato_data
 from zimmer.models import HierarchicalLDS
 from zimmer.emissions import HierarchicalDiagonalRegression
 
-# from zimmer.plotting import plot_3d_continuous_states
-# from zimmer.plotting import make_states_3d_movie
 
 from pybasicbayes.distributions import Regression, DiagonalRegression
 from pylds.models import MissingDataLDS
@@ -90,82 +56,11 @@ def cached(results_name):
         return func_wrapper
     return _cache
 
-
-def load_data(include_unnamed=True):
-    # Load the data
-    worm_datas = [WormData(i, name="worm{}".format(i), version="kato") for i in range(N_worms)]
-
-    # Get the "true" discrete states as labeled by Zimmer
-    z_trues = [wd.zimmer_states for wd in worm_datas]
-    z_trues, newlabels = relabel_by_usage(z_trues, return_mapping=True)
-
-    # Get the key
-    z_key = load_kato_key()
-    z_key = [z_key[i] for i in np.argsort(newlabels)]
-
-    # Get the names of the neurons
-    neuron_names = np.unique(np.concatenate([wd.neuron_names for wd in worm_datas]))
-    if not include_unnamed:
-        print("Only including named neurons.")
-        neuron_names = neuron_names[:61]
-    else:
[…]

### 2018-02-26 experiments/fit_arhmm.py
removing plotting from arhmm fitting

@@ -11,33 +11,33 @@ from scipy.ndimage import gaussian_filter1d
 from tqdm import tqdm
 
 # Plotting stuff
-import matplotlib.pyplot as plt
-from hips.plotting.colormaps import gradient_cmap
-import seaborn as sns
-sns.set_style("white")
-sns.set_context("paper")
-color_names = ["windows blue",
-               "red",
-               "amber",
-               "faded green",
-               "dusty purple",
-               "orange",
-               "clay",
-               "pink",
-               "greyish",
-               "mint",
-               "light cyan",
-               "steel blue",
-               "forest green",
-               "pastel purple",
-               "salmon",
-               "dark brown"]
-colors = sns.xkcd_palette(color_names)
-cmap = gradient_cmap(colors)
-
-from matplotlib.cm import get_cmap
-cm = get_cmap("cubehelix")
-zimmer_colors = [cm(i) for i in np.linspace(0.05, 0.95, 8)]
+# import matplotlib.pyplot as plt
+# from hips.plotting.colormaps import gradient_cmap
+# import seaborn as sns
+# sns.set_style("white")
+# sns.set_context("paper")
+# color_names = ["windows blue",
+#                "red",
+#                "amber",
+#                "faded green",
+#                "dusty purple",
+#                "orange",
+#                "clay",
+#                "pink",
+#                "greyish",
+#                "mint",
+#                "light cyan",
+#                "steel blue",
+#                "forest green",
+#                "pastel purple",
+#                "salmon",
+#                "dark brown"]
+# colors = sns.xkcd_palette(color_names)
+# cmap = gradient_cmap(colors)
+#
+# from matplotlib.cm import get_cmap
+# cm = get_cmap("cubehelix")
+# zimmer_colors = [cm(i) for i in np.linspace(0.05, 0.95, 8)]
 
 # Modeling stuff
 from pyhsmm.util.general import relabel_by_usage, relabel_by_permutation
@@ -49,28 +49,32 @@ from rslds.models import SoftmaxRecurrentARHMM
 from zimmer.models import HierarchicalARWeakLimitStickyHDPHMM, HierarchicalRecurrentARHMM, HierarchicalRecurrentARHMMWithNN
 from zimmer.dynamics import HierarchicalAutoRegression, HierarchicalRobustAutoRegression
 
-import importlib
[…]

### 2018-02-26 experiments/fit_arhmm.py
fix data loading

@@ -49,7 +49,7 @@ from rslds.models import SoftmaxRecurrentARHMM
 from zimmer.models import HierarchicalARWeakLimitStickyHDPHMM, HierarchicalRecurrentARHMM, HierarchicalRecurrentARHMMWithNN
 from zimmer.dynamics import HierarchicalAutoRegression, HierarchicalRobustAutoRegression
 
-from zimmer.io import WormData, load_kato_key
+from zimmer.io import WormData, load_kato_key, load_kato_data
 
 # import importlib
 # import zimmer.plotting
@@ -105,49 +105,6 @@ def cached(results_name):
             return results
         return func_wrapper
     return _cache
-
-
-
-def load_data(include_unnamed=True):
-    # Load the data
-    worm_datas = [WormData(i, name="worm{}".format(i), version="kato") for i in range(N_worms)]
-
-    # Get the "true" discrete states as labeled by Zimmer
-    z_trues = [wd.zimmer_states for wd in worm_datas]
-    z_trues, newlabels = relabel_by_usage(z_trues, return_mapping=True)
-
-    # Get the key
-    z_key = load_kato_key()
-    z_key = [z_key[i] for i in np.argsort(newlabels)]
-
-    # Get the names of the neurons
-    neuron_names = np.unique(np.concatenate([wd.neuron_names for wd in worm_datas]))
-    if not include_unnamed:
-        print("Only including named neurons.")
-        neuron_names = neuron_names[:61]
-    else:
-        print("Including all neurons, regardless of whether they were identified.")
-
-    N_neurons = neuron_names.size
-    print("{} neurons across all {} worms".format(N_neurons, N_worms))
-
-    # Construct a big dataset with all neurons for each worm
-    ys = []
-    masks = []
-    for wd in worm_datas:
-        y_indiv = getattr(wd, signal)
-        y = np.zeros((wd.T, N_neurons))
-        mask = np.zeros((wd.T, N_neurons), dtype=bool)
-        indices = wd.find_neuron_indices(neuron_names)
-        for n, index in enumerate(indices):
-            if index is not None:
-                y[:, n] = y_indiv[:, index]
-                mask[:, n] = True
-
-        ys.append(y)
-        masks.append(mask)
-
-    return ys, masks, z_trues, z_key, neuron_names
 
 
 def compute_baseline_likelihood(xtrains, xtests, alpha=3.):
[…]

### 2018-02-26 experiments/fit_arhmm.py
scan over nn too

@@ -82,7 +82,7 @@ mov_dir = os.path.join(results_dir, "movies")
 
 # Fitting parameters
 N_lags = 1
-N_samples = 5000
+N_samples = 1000
 N_worms = 5
 
 
@@ -223,24 +223,26 @@ def fit_all_models(Ks=np.arange(4, 21, 2)):
 
     results = dict(baseline_hll=baseline_hll)
 
-    for index, (is_hierarchical, is_robust, is_recurrent) in \
-            enumerate(it.product(*([(False, True)] * 3))):
+    for index, (is_hierarchical, is_robust, is_recurrent, is_nn) in \
+            enumerate(it.product(*([(False, True)] * 4))):
 
         models = []
         llss = []
         hlls = []
         z_smplss = []
 
-        group_name = "{}\n{}\n{}".format(
-            "hierarchical" if is_hierarchical else "standard",
-            "robust" if is_robust else "standard",
-            "recurrent" if is_recurrent else "standard"
+        group_name = "{}_{}_{}_{}".format(
+            "hier" if is_hierarchical else "nohier",
+            "rob" if is_robust else "norob",
+            "rec" if is_recurrent else "norec",
+            "nn" if is_nn else "linear"
         )
         for K in Ks:
-            name = "{}_{}_{}_{}".format(
+            name = "{}_{}_{}_{}_{}".format(
                 "hier" if is_hierarchical else "nohier",
                 "rob" if is_robust else "norob",
                 "rec" if is_recurrent else "norec",
+                "nn" if is_nn else "linear",
                 K
             )
             print("Fitting model: {}".format(name))
@@ -249,7 +251,8 @@ def fit_all_models(Ks=np.arange(4, 21, 2)):
                 partial(_fit_model_wrapper,
                         is_hierarchical=is_hierarchical,
                         is_robust=is_robust,
-                        is_recurrent=is_recurrent))
+                        is_recurrent=is_recurrent,
+                        is_nn=is_nn))
             mod, lls, hll, z_smpls = fit(K)
 
             # Append results

### 2018-08-29 drafts/elife/draft.tex
adding hierarchical ar observations

@@
 % \title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
-\title{Hierarchical state space models reveal dynamics of neural activity in \textit{C. elegans}}
+\title{Recurrent state space models capture discrete and continuous dynamics of neural activity in \textit{C. elegans}}
 \author{Scott Linderman$^{\text{1}}$,
@@
 can be learned from noisy, partial recordings with trial-to-trial
-variability.  We call this model \textbf{TBD}.\todo{name it!}
+variability.  We call this model \textbf{TBD}.
+%\todo{name it!}
 
@@
 which together produce a dorsal turn.
-\todo{Need confirmation of these results.  Hard for me to tell difference between
-dorsal turn activation (orange to purple) and ventral turn (green to brown).}
+%\todo{Need confirmation of these results.  Hard for me to tell difference between dorsal turn activation (orange to purple) and ventral turn (green to brown).}
 
@@
 \paragraph{Input-driven transition probabilities}
-todo
+
 \todo[inline]{Detail the model for how transition probabilities depend

### 2018-08-29 zimmer/observations.py

@@ -236,3 +236,233 @@ class HierarchicalIndependentAutoRegressiveObservations(_Observations):
         mask = np.ones((T, self.D), dtype=bool) 
         mus = self._compute_mus(data, input, mask, tag)
         return (expectations[:, :, None] * mus).sum(1)
+
+
+class HierarchicalAutoRegressiveObservations(_Observations):
+    def __init__(self, K, D, G=1, M=0, lags=1, eta=0.1):
+        super(HierarchicalAutoRegressiveObservations, self).__init__(K, D, M)
+        
+        # Distribution over initial point
+        self.mu_init = np.zeros(D)
+        self.inv_sigma_init = np.zeros(D)
+        
+        # Global AR parameters
+        assert lags > 0 
+        self.lags = lags
+        self.As = .95 * np.array([
+            np.column_stack([random_rotation(D), np.zeros((D, (lags-1) * D))]) 
+            for _ in range(K)])
+        self.shared_bs = npr.randn(K, D)
+        self.shared_Vs = npr.randn(K, D, M)
+        
+        # Per-group AR parameters
+        self.G = G
+        self.eta = eta
+        self.As = .95 * np.array([
+                [np.column_stack([random_rotation(D), np.zeros((D, (lags-1) * D))]) for _ in range(K)]
+                for _ in range(G)]
+            )
+        self.bs = npr.randn(G, K, D)
+        self.Vs = npr.randn(G, K, D, M)
+        self.inv_sigmas = -4 + npr.randn(G, K, D)
+
+    @property
+    def params(self):
+        return self.shared_As, self.shared_bs, self.shared_Vs, \
+               self.As, self.bs, self.Vs, self.inv_sigmas
+        
+    @params.setter
+    def params(self, value):
+        self.shared_As, self.shared_bs, self.shared_Vs, \
+        self.As, self.bs, self.Vs, self.inv_sigmas = value
+        
+    def permute(self, perm):
+        self.shared_As = self.shared_As[perm]
+        self.shared_bs = self.shared_bs[perm]
+        self.shared_Vs = self.shared_Vs[perm]
+
+        for g in range(self.G):
+            self.As[g] = self.As[g, perm]
+            self.bs[g] = self.bs[g, perm]
+            self.Vs[g] = self.Vs[g, perm]
+            self.inv_sigmas[g] = self.inv_sigmas[g, perm]
+
+    def initialize(self, datas, inputs=None, masks=None, tags=None):
[…]

### 2018-08-29 zimmer/observations.py
updating observations and transitions with initialization code and correct prior. updating notebook to sweep over model size

@@ -88,12 +88,24 @@ class HierarchicalIndependentAutoRegressiveObservations(_Observations):
                 for g in range(self.G):
                     self.inv_sigmas[g, k, d] = np.log(sigmas + 1e-8)
 
+    def initialize_from_standard(self, ar):
+        # Copy the observation parameters
+        self.shared_As = ar.As.copy()
+        self.shared_Vs = ar.Vs.copy()
+        self.shared_bs = ar.bs.copy()
+
+        for g in range(self.G):
+            self.As[g] = ar.As.copy()
+            self.Vs[g] = ar.Vs.copy()
+            self.bs[g] = ar.bs.copy()
+            self.inv_sigmas[g] = ar.inv_sigmas.copy()
+                    
     def log_prior(self):
         lp = 0
         for g in range(self.G):
-            lp += np.sum(norm.logpdf(self.As[g], self.shared_As[g], np.sqrt(self.eta)))
-            lp += np.sum(norm.logpdf(self.bs[g], self.shared_bs[g], np.sqrt(self.eta)))
-            lp += np.sum(norm.logpdf(self.Vs[g], self.shared_Vs[g], np.sqrt(self.eta)))
+            lp += np.sum(norm.logpdf(self.As[g], self.shared_As, np.sqrt(self.eta)))
+            lp += np.sum(norm.logpdf(self.bs[g], self.shared_bs, np.sqrt(self.eta)))
+            lp += np.sum(norm.logpdf(self.Vs[g], self.shared_Vs, np.sqrt(self.eta)))
         return lp
                     
     def _compute_mus(self, data, input, mask, tag):
@@ -319,12 +331,24 @@ class HierarchicalAutoRegressiveObservations(_Observations):
             for g in range(self.G):
                 self.inv_sigmas[g, k] = np.log(sigmas + 1e-8)
 
+    def initialize_from_standard(self, ar):
+        # Copy the observation parameters
+        self.shared_As = ar.As.copy()
+        self.shared_Vs = ar.Vs.copy()
+        self.shared_bs = ar.bs.copy()
+
+        for g in range(self.G):
+            self.As[g] = ar.As.copy()
+            self.Vs[g] = ar.Vs.copy()
+            self.bs[g] = ar.bs.copy()
+            self.inv_sigmas[g] = ar.inv_sigmas.copy()
+
     def log_prior(self):
         lp = 0
         for g in range(self.G):
-            lp += np.sum(norm.logpdf(self.As[g], self.shared_As[g], np.sqrt(self.eta)))
[…]

### 2018-08-29 zimmer/transitions.py

@@ -60,6 +60,17 @@ class HierarchicalRecurrentTransitions(_Transitions):
             self.Ws[g] = self.Ws[g, perm]
             self.Rs[g] = self.Rs[g, perm]
 
+    def initialize_from_standard(self, tr):
+        # Copy the transition parameters
+        self.shared_log_Ps = tr.log_Ps.copy()
+        self.shared_Ws = tr.Ws.copy()
+        self.shared_Rs = tr.Rs.copy()
+
+        for g in range(self.G):
+            self.log_Ps[g] = tr.log_Ps.copy()
+            self.Ws[g] = tr.Ws.copy()
+            self.Rs[g] = tr.Rs.copy()
+                        
     def log_prior(self):
         lp = 0
         for g in range(self.G):

### 2018-08-29 zimmer/observations.py
minor bug fixes

@@ -203,7 +203,7 @@ class HierarchicalIndependentAutoRegressiveObservations(_Observations):
                 # Update the variances
                 yhats = xs.dot(muk)
                 sqerr = (ys - yhats)**2
-                self.inv_sigmas[g, k, d] = np.log(np.average(sqerr, weights=weights[:,k], axis=0))
+                self.inv_sigmas[g, k, d] = np.log(np.average(sqerr, weights=weights[:,k], axis=0) + 1e-16)
 
     def _m_step_shared(self, expectations, datas, inputs, masks, tags):
         G, K, D, M = self.G, self.K, self.D, self.M
@@ -261,7 +261,7 @@ class HierarchicalAutoRegressiveObservations(_Observations):
         # Global AR parameters
         assert lags > 0 
         self.lags = lags
-        self.As = .95 * np.array([
+        self.shared_As = .95 * np.array([
             np.column_stack([random_rotation(D), np.zeros((D, (lags-1) * D))]) 
             for _ in range(K)])
         self.shared_bs = npr.randn(K, D)
@@ -421,7 +421,7 @@ class HierarchicalAutoRegressiveObservations(_Observations):
         # Otherwise, fit a weighted linear regression for each discrete state
         for k in range(K):
             # Check for zero weights (singular matrix)
-            if np.sum(weights[:, k]) < lags + M + 1:
+            if np.sum(weights[:, k]) < D * lags + M + 1:
                 self.As[g, k] = self.shared_As[k]
                 self.Vs[g, k] = self.shared_Vs[k]
                 self.bs[g, k] = self.shared_bs[k]
@@ -445,7 +445,7 @@ class HierarchicalAutoRegressiveObservations(_Observations):
                 # Update the variances
                 yhats = xs.dot(muk)
                 sqerr = (ys[:, d] - yhats)**2
-                self.inv_sigmas[g, k, d] = np.log(np.average(sqerr, weights=weights[:,k], axis=0))
+                self.inv_sigmas[g, k, d] = np.log(np.average(sqerr, weights=weights[:,k], axis=0) + 1e-16)
 
     def _m_step_shared(self, expectations, datas, inputs, masks, tags):
         G, K, D, M = self.G, self.K, self.D, self.M

### 2018-08-30 zimmer/observations.py
adding hierarchical robust observations

@@ -490,3 +490,173 @@ class HierarchicalAutoRegressiveObservations(_Observations):
         mask = np.ones((T, self.D), dtype=bool) 
         mus = self._compute_mus(data, input, mask, tag)
         return (expectations[:, :, None] * mus).sum(1)
+
+
+class HierarchicalRobustAutoRegressiveObservations(_Observations):
+    def __init__(self, K, D, G=1, M=0, lags=1, eta=0.1):
+        super(HierarchicalRobustAutoRegressiveObservations, self).__init__(K, D, M)
+        
+        # Distribution over initial point
+        self.mu_init = np.zeros(D)
+        self.inv_sigma_init = np.zeros(D)
+        
+        # Global AR parameters
+        assert lags > 0 
+        self.lags = lags
+        self.shared_As = .95 * np.array([
+            np.column_stack([random_rotation(D), np.zeros((D, (lags-1) * D))]) 
+            for _ in range(K)])
+        self.shared_bs = npr.randn(K, D)
+        self.shared_Vs = npr.randn(K, D, M)
+        
+        # Per-group AR parameters
+        self.G = G
+        self.eta = eta
+        self.As = .95 * np.array([
+                [np.column_stack([random_rotation(D), np.zeros((D, (lags-1) * D))]) for _ in range(K)]
+                for _ in range(G)]
+            )
+        self.bs = npr.randn(G, K, D)
+        self.Vs = npr.randn(G, K, D, M)
+        self.inv_sigmas = -4 + npr.randn(G, K, D)
+        self.inv_nus = np.log(4) * np.ones(G, K)
+
+    @property
+    def params(self):
+        return self.shared_As, self.shared_bs, self.shared_Vs, \
+               self.As, self.bs, self.Vs, self.inv_sigmas, self.inv_nus
+        
+    @params.setter
+    def params(self, value):
+        self.shared_As, self.shared_bs, self.shared_Vs, \
+        self.As, self.bs, self.Vs, self.inv_sigmas, self.inv_nus = value
+        
+    def permute(self, perm):
+        self.shared_As = self.shared_As[perm]
+        self.shared_bs = self.shared_bs[perm]
+        self.shared_Vs = self.shared_Vs[perm]
+
+        for g in range(self.G):
+            self.As[g] = self.As[g, perm]
+            self.bs[g] = self.bs[g, perm]
+            self.Vs[g] = self.Vs[g, perm]
[…]

### 2018-08-30 zimmer/observations.py
minor fix

@@ -586,6 +586,7 @@ class HierarchicalRobustAutoRegressiveObservations(_Observations):
             self.Vs[g] = ar.Vs.copy()
             self.bs[g] = ar.bs.copy()
             self.inv_sigmas[g] = ar.inv_sigmas.copy()
+            self.inv_nus[g] = ar.inv_nus.copy()
 
     def log_prior(self):
         lp = 0

### 2018-08-30 zimmer/observations.py
fitting regular and robust models

@@ -519,7 +519,7 @@ class HierarchicalRobustAutoRegressiveObservations(_Observations):
         self.bs = npr.randn(G, K, D)
         self.Vs = npr.randn(G, K, D, M)
         self.inv_sigmas = -4 + npr.randn(G, K, D)
-        self.inv_nus = np.log(4) * np.ones(G, K)
+        self.inv_nus = np.log(4) * np.ones((G, K))
 
     @property
     def params(self):
@@ -631,7 +631,7 @@ class HierarchicalRobustAutoRegressiveObservations(_Observations):
         D = self.D
         mus = self._compute_mus(data, input, mask, tag)
         sigmas = self._compute_sigmas(data, input, mask, tag)
-        nus = np.exp(self.inv_nus)
+        nus = np.exp(self.inv_nus)[tag]
 
         resid = data[:, None, :] - mus
         z = resid / sigmas
